### 2023-01-30 paper.tex
initial commit

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%    INSTITUTE OF PHYSICS PUBLISHING                                   %
%                                                                      %
%   `Preparing an article for publication in an Institute of Physics   %
%    Publishing journal using LaTeX'                                   %
%                                                                      %
%    LaTeX source code `ioplau2e.tex' used to generate `author         %
%    guidelines', the documentation explaining and demonstrating use   %
%    of the Institute of Physics Publishing LaTeX preprint files       %
%    `iopart.cls, iopart12.clo and iopart10.clo'.                      %
%                                                                      %
%    `ioplau2e.tex' itself uses LaTeX with `iopart.cls'                %
%                                                                      %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%
% First we have a character check
%
% ! exclamation mark    " double quote  
% # hash                ` opening quote (grave)
% & ampersand           ' closing quote (acute)
% $ dollar              % percent       
% ( open parenthesis    ) close paren.  
% - hyphen              = equals sign
% | vertical bar        ~ tilde         
% @ at sign             _ underscore
% { open curly brace    } close curly   
% [ open square         ] close square bracket
% + plus sign           ; semi-colon    
% * asterisk            : colon
% < open angle bracket  > close angle   
% , comma               . full stop
% ? question mark       / forward slash 
% \ backslash           ^ circumflex
%
% ABCDEFGHIJKLMNOPQRSTUVWXYZ 
% abcdefghijklmnopqrstuvwxyz 
% 1234567890
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%

%\documentclass[12pt]{iopart}
\documentclass[12pt]{article}
%\usepackage{amsmath}
\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xcolor}
\usepackage{graphicx}
%Uncomment next line if AMS fonts required
%\usepackage{iopams}  
\begin{document}

\title{Optogenetic Maxwell Demon to Exploit Intrinsic Noise and Control Cell Differentiation Despite Time Delays and Extrinsic Variability}
\maketitle

\author{M P May$^1$, B Munsky$^{1,2}$}

%\address{$^1$ School of Bioengineering, Colorado State University, Fort Collins, Co}
%\address{$^2$ Dept. of Chemical and Biological Engineering, Colorado State University, Fort Collins, Co}

%\vspace{10pt}
%\begin{indented}
%\item[]August 2017
%\end{indented}

\begin{abstract}
The field of synthetic biology focuses on creating modular components which can generate complex and controllable synthetic biological systems, but the intrinsic noise of gene regulation can create undesired effects in these systems. This noise is often treated as a nuisance which should be removed, but theoretical noise-exploiting controllers have been shown to use noise to their advantage to control difficult control problems which would be difficult otherwise. In such work, a difficult control problem similar to Maxwells Demon was proposed and showed that noise exploiting controllers could send two identical systems to different stable points reliably by exploiting the natural noise of gene regulation. Those controllers relied upon systems with no time delay well known parameters but, in this work the performance of noise-exploiting controllers are measured under time delay, parameter perturbation and system scale.

\end{abstract}
Keywords: Stochastic control, optogenetic, synthetic biology

%
% Uncomment for keywords
%\vspace{2pc}
%\noindent{\it Keywords}: Stochastic control, optogenetic, synthetic biology
%
% Uncomment for Submitted to journal title message
%\submitto{\JPA}
%
% Uncomment if a separate title page is required
%\maketitle
% 
% For two-column output uncomment the next line and choose [10pt] rather than [12pt] in the \documentclass declaration
%\ioptwocol

\section{Introduction}
\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{CartoonAndControler.png}
\caption{(A) Schematic of the T7 optogenetic control system (adapted from \cite{May2021}). A microscope observer is partially (PAC) or fully (FAC) aware  of cells; the resulting controller modulates a single spatially homogeneous light signal that perturbs all cells equally seeks to drive each genetically identical cells to a different, specified states. (B) Schematic of the optogenetically controllable T7 polymerase model, which is identical in all cells: Light activates the T7 polymerase-based production of protein, and the protein self-actuates its own production. (C) The FAC (top) and PAC (bottom) control laws for two cells ($\mathbf{u}^{XAC}(x_1,x_2)$ shown on left) break the symmetry between Cell 1 and Cell 2 and result in distinct marginal distributions for the cells' responses (right).}
\label{cartoons}
\end{center}
\vspace{-0.3in}
\end{figure}

Synthetic biology seeks to create modular\cite{Ng2019} and orthogonal\cite{Liu2018} components to sense and actuate\cite{Sheets2020} complex logical systems\cite{Groseclose2020}, and which are capable of performing a variety of advanced biological behaviors\cite{Shin2020}. New optogenetic tools have increased our ability to actuate embedded systems within cells reliably and with strong control performance\cite{Sheets2020,Baumschlager2017,Chen2020,Lillacci2018}. Great advances in these two fields have enabled computer programmable control of cellular protein production through the use of external optogenetic inputs and smart microscope techniques\cite{Fox2021,Baumschlager2021,Lugagne2017}. These types of digital-synthetic actuators allows for fine-tuned, computer-modulated control of cellular systems that were previously impossible\cite{Rullan2018, Baumschlager2017} and with fast response times in comparison to chemical diffusion.

Classical and modern control methods like PID control and model predictive control have been implemented in such systems\cite{} which can control synthetic systems to different stable points. Theoretical works have shown that stochastic systems can be controlled to unstable points by oscillating around those points \cite{Guarino2020} but new control techniques which exploit the full probability distribution information of the system can be used in certain scenarios to control simple systems to unstable points by exploiting the natural noise of protein production to its advantage. Previous control effort focus on deterministic ODE formulations of the control problem, but only a few studies seek to analyze the full chemical master equation to exploit noise to {\em improve} the control of synthetic biology processes \cite{Szymanska2015,May2021}.

In \cite{May2021}, it was suprisingly found that, when noise, auto-regulation, and noise-exploiting feedback control were all present at once, the system could be controlled to bring two similar cells to two different arbitrary points simultaneously (figure \ref{cartoons}A), but when any one of these components were missing the system could not break its symmetry. More specifically, this required that the symmetry between the protein probability distributions of each cell would not be not the same (or asymmetric). Three effective controllers were found which could do this, a Fully Aware Feedback Controller (FAC) that observed the protein counts of all observed cells, a Partially Aware Feedback Controller (PAC) that observed the protein counts of only one cell, and a probabilistic Model Predictive Feedback Controller (pMPC) that observed only the primary cell and integrated a probabilistic model to represent the non-stationary probability distribution of the secondary cells in the same environment.

The solutions of the controllers was carried out by solving the chemical master equation (i.e., the forward Komogorov equation) to estimate the probability of possible cellular states. For a two cell, single-input-single output (SIMO) control problem, the FAC performed the best (figure\ \ref{cartoons}C(top)), and the PAC (figure\ \ref{cartoons}C(bottom)) 2nd, and pMPC controllers 3rd. In all cases, the controller's knowledge and explicit treatment of intrinsic noise was critical to improve the control performance, break symmetry, and enable the SIMO control of all cells.

In this paper, the control problem is extended to examine how the FAC and PAC controllers are affected by time delay, the system granularity, and parameter uncertainty. In the `Methods' section, we briefly set up a stochastic chemical master equation model for multiple individual cells in the same spatially homogeneous environment, and we formulate the effects of FAC and PAC controllers on the dynamics of cells within that environment. Next, in the `Results' section, we examine the effects of volume changes, time delays, and parametric uncertainties on the FAC and PAC control performance. Finally, in `Conclusions' we summarize our findings and discuss potential implications that these controllers may have on future systems and synthetic biology investigations.

\section{Methods}
We begin with a five species model for the production of an arbitratry protein through the activation of light activatible T7 polymerase (figure 1B) motivated by Baumschlager, et al.
This model for the production of protein uses a split T7 domain ($T7_N$,and $T7_C$) which reversibly dimerizes in the presence of light to create a T7 dimer. The T7 monomer is created at a constant rate and degrades linearly with the number of monomer. The split monomers of T7 can bind to create a T7 dimer (equation ...) . The T7 dimer can then bind to a gene which enables the production of protein with a leak term, and the protein can bind to a secondary self-activated promoter described by a Hill function for the creation of more protein. This auto-regulatory motif in protein production is responsible for creating non-linearity in the system that enables noise-exploiting control. The final step was the degredation of protein

These model dynamics were then simplified to a single ODE with auto-regulation describing the protein accumulation as: 
\begin{equation}
\frac{dp}{dt}=u(x_1,x_2)  + \kappa\frac{p^\eta}{p^\eta+\beta^\eta}+ k_0 -\gamma p,
\label{rateeq}
\end{equation}

where $p$ is the protein count; $\kappa$ is the maximum auto-regulation promoter strength; $\beta$ is the concentration at which auto-regulation reaches half strength; $\eta$ is the auto-regulation promoter cooperativity; $\gamma$ is the degradation rate; $k_0$ is the promoter leakage rate; and $u$ is the T7 promoter strength.  The protein production rate response to the light level $u$ can be modulated in response to the state of the system $u(x_1,x_2)$ and introduces the possibility of state-dependent optogenetic feedback to the system.  We note that $u(x_1,x_2)$ refers to the effect of light on the optogenetic system instead of light activity itself. A schematic of the model is shown in figure \ \ref{cartoons}B, and parameters were chosen to reproduce dynamics measured in \cite{Baumschlager2017} and are shown in are defined in Table 1\cite{May2021}. Two optimized controls laws $u=u^{\rm (FAC)}$ and $u=u^{\rm (PAC)}$ are depicted in figure \ \ref{cartoons}C (top left and top right, respectively).

\begin{table}[]
\caption{Model Parameters}
\begin{center}
\begin{tabular}{|c|c|c|}
\hline
Parameter & Value   & units                     \\ \hline
$\kappa$         & 0.406   & Molecules Per Minute \\
$\beta$          & 20.0    & Molecules            \\
$\eta $           & 8.00    & unit less          \\
$k_0$           & 0.0001  & Molecules per Minute \\
$\gamma$         & 0.0203  & per Minute         \\
$u$              & dynamic & Molecules per Minute \\ \hline
\end{tabular}
\label{table}
\end{center}
\vspace{-0.2in}
\end{table}

\subsection{Stochastic Simulation Algorithm and the Finite State Projection}
Solving discrete stochastic representations of chemical dynamics is critical to the development of noise-exploiting controllers, and can be analyzed using the Stochastic Simulation Algorithm (SSA) , or the Finite State Projection (FSP).
To recast the above ODE formulation into the discrete stochastic representation, we define the enumerated $i = 1,2,\ldots$ states of the system as the tuple of the numbers of proteins in each cell as $\mathbf{X}_i = [x_{i1},x_{i2},\ldots,x_{iN_c}]$, and we define the stoichiometry vectors as the change in state following one reaction event (e.g., $\mathbf{X}_i \rightarrow \mathbf{X}_i + \mathbf{s}_\nu$), where the $2N_c$ possible reactions are:

\begin{align}
\mathbf{s}_1 &= \mathbf{e}_1,\ \mathbf{s}_2 = -\mathbf{e}_1,\nonumber\\ 
\mathbf{s}_3 &= \mathbf{e}_2,\ \mathbf{s}_4 = -\mathbf{e}_2,\nonumber\\
&\vdots \nonumber\\ 
\mathbf{s}_{\rm 2N_c-1} &= \mathbf{e}_{\rm N_c},\ \mathbf{s}_{\rm 2N_c} = -\mathbf{e}_{\rm N_c}, \label{Stochs}
 \end{align}

where each $\mathbf{e}_{i_c}$ corresponds to the Euclidean basis vector for the $i_c^{\rm th}$ cell. The corresponding propensity functions are:
\begin{align}
w_1 &= u(\mathbf{X},t)  + \kappa \frac{x_1^\eta}{x_1^{\eta}+\beta^{\eta}} + k_0,\ w_2 = \gamma x_1,\nonumber\\
w_3 &= u(\mathbf{X},t)  + \kappa \frac{x_2^\eta}{x_2^{\eta}+\beta^{\eta}} + k_0,\ w_4 = \gamma x_2,\nonumber\\
&\vdots \nonumber\\
w_{\rm 2N_c-1} &= u(\mathbf{X},t)  + \kappa \frac{x_N^\eta}{x_N^{\eta}+\beta^{\eta}} + k_0,\ w_{\rm 2N_c} = \gamma x_N,\label{Props}
 \end{align}
With these stoichiometry and propensities, we can formulate and run the Gillespie stochastic simulation algorithm\cite{Gillespie1992,Gillespie1977} to generate representative trajectories of the stochastic process.

For the same specifications of the stoichiometry and propensity functions, the chemical master equation can also be formulated in matrix/vector form as:
\begin{equation}
\frac{d}{dt}\mathbf{P}=(\mathbf{A}_0+\textbf{Bu}^{\mathcal{C}})\mathbf{P},\label{CME}
\end{equation}
where $\mathbf{P} = [P(\mathbf{X}_1), P(\mathbf{X}_2), \ldots ]^T$ is the enumerated probability mass vector for all possible states of the system; $\mathbf{A}_0$ is the infinitesimal generator of the stochastic process due to the autoregulation promoter and degradation events;  $\textbf{u}^{\mathcal{C}} =[u^{\mathcal{C}}(\mathbf{X}_1), u^{\mathcal{C}}(\mathbf{X}_2), \ldots ]^T$ is the collection of state-dependent control inputs associated with each state; and $\textbf{Bu}^{\mathcal{C}}$ is the infinitesimal generator for the T7 promoter expression by which the effects of controller $\mathcal{C}$ are introduced into the process.

The generator $\mathbf{A}_0$ can be constructed according to:
\begin{equation}
[\mathbf{A}_0]_{ij} = \left\{
\begin{array}{rl}
-\sum_{\nu =1}^{2N_c} w_{\nu}(\mathbf{X}_j), &\text{for }i=j,\\
w_{\nu}(\mathbf{X}_j), &\text{for }\mathbf{X}_i = \mathbf{X}_j+\mathbf{s}_\nu\\
0, & \text{otherwise,}
\end{array}\right. 
\label{InfA}
\end{equation}
and the infinitesimal generator $\mathbf{Bu}^{\mathcal{C}}$ of the controller is given by
\begin{equation}
[\mathbf{Bu^{\mathcal{C}}]}_{ij} = \left\{
\begin{array}{rl}
- N_c{u}^{\mathcal{C}}(\mathbf{X}_j), &\text{for }i=j,\\
{u}^{\mathcal{C}}(\mathbf{X}_j), &\text{for }\begin{array}{ll}\mathbf{X}_i =\mathbf{X}_j + \mathbf{e}_{i_c}, \\ \text{and } i_c = 1,\ldots,N_c\end{array}\\
0, & \text{otherwise,}
\end{array}\right.
\label{InfB}
\end{equation}
where ${u}^{\mathcal{C}}(\mathbf{X}_j)$ is the specification of the $\mathcal{C}$ controller in terms of the (partially observed) state.

For a given controller, the steady state of the system ($\mathbf{P}^*$) found by solving Eq.\ \ref{CME} is given by
\begin{equation}
\mathbf{P}^*=null(\mathbf{A}+\textbf{Bu}^{\mathcal{C}}).\label{SSDist}
\end{equation}

and the infinitesimal generator $\mathbf{Bu}^{\mathcal{C}}$ of the controller is given by

\subsection{Control Performance, Optimization, \& Design}
Score metrics of steady state distributions were made in order to enable optimization of the system by adjusting controllers. In the control of a SIMO stochastic system, a control input which tries to perfectly control one system will create increased noise in the other system because their inputs are shared. Therefore, a controller which strikes a balance between the two cells should be important when it is being optimized.

 The control performance ($J$), is defined as the expected steady state squared Euclidean distance of the process from a specified target ($\mathbf{T}$) state according to $E\{(\mathbf{X}-\mathbf{T})^2\}$. This score can be rewritten in terms of a precomputed set of linear weights ($\mathbf{C}$) and the steady state probability distribution ($\mathbf{P}^*$ from Eq.\ (\ref{SSDist})) given by
{\small \begin{align}
J&= \lim_{t\rightarrow \infty}\mathbb{E}\{|\mathbf{X}(t)-\mathbf{T}|_2^2\} \nonumber \\ 
&=\sum_{i,j,\ldots} P^*(x_1=i,x_2=j,\ldots) ((i-\mathbf T_1)^2 + (j-\mathbf T_2)^2 +\ldots)\nonumber  \\
&=\sum_{i,j,\ldots} P^*_{ij\ldots}(t)C_{ij\ldots} =\mathbf{C}\mathbf{P}^*
\label{Euclid}
\end{align}}
Where $J$ is a scalar, $\mathbf{C}$ is a vector of weights given by the Euclidian distance, and $\mathbf{P}^*$ the vectorized steady state distribution of the FSP.

Two different controller designs were chosen at two different levels of information for the controller to use. A fully aware controller ($\mathbf{u}^{FAC}$) used direct observation of the protein count of both cells simultaneously while making its control input decision, while the partially aware controller $\mathbf{u}^{PAC}$) used only the direct observation of a single cell.

\begin{equation}
\mathbf{u}^{FAC}=u(x_0,x_1)
\end{equation}
\begin{equation}
\mathbf{u}^{PAC}=u(x_0)
\end{equation}
where $x_0$ and $x_1$ are discrete integers greater than or equal to 0 which represent the number of protein in cell one and cell two.  The $\mathbf{u}^{FAC}$ and $\mathbf{u}^{PAC}$ controllers were optimized to minimize $J$ in order to maximize the amount of probability near the target point, and both $\mathbf{u}^{FAC}$ and $\mathbf{u}^{PAC}$ were saved as lookup tables after being optimized.

\subsection{System Granularity Scalings}
A granularity parameter ($\alpha$) was developed that linearly re-streches the dynamics of the system to occur over at a different range of protein while maintaining the general behavior of the model.  Any arbitrary propensity function ($w$) can be rescaled to a different level of discreteness by substituting all rate equations with $w'$ given by
\begin{equation}
w'(x)=w(x/\alpha)
\end{equation}
where $\alpha=M'/M$, $M$ the size of the original system, and $M'$ the size of the re-scaled system. When $\alpha <1 $ the system dynamics occur over a smaller range of protein counts, and when $\alpha >1$ the system dynamics occurs over a larger range of protein counts.

The performance metric and controller must also be scaled by $1/\alpha$ before computing the appropriate functions of that state, e.g., ${u}^{\mathcal{C}} = {u}^{\mathcal{C}}(\mathbf{X}/\alpha)$ and 
{\small \begin{align}
J &= \sum_{i=0}^{M'}  \sum_{j=0}^{M'}P^*(x_1=i,x_2=j) ((i/\alpha - \mathbf T_1)^2 + (j/\alpha -\mathbf T_2)^2),\nonumber \\
& =\mathbf{C}'\mathbf{P}^*
\label{EuclidV}
\end{align}}
Since $\mathbf{u}^{FAC}$ and $\mathbf{u}^{PAC}$ are both matrices, fractional number inputs after rescaling are handled by 2D cubic interpolation of the nearest values.

\begin{equation}
u'=u(x_0 / \alpha, x_1 / \alpha) 
\end{equation}
\begin{equation}
u'=u(\alpha x_0)
\end{equation}

\subsection{Introduction of Observation/Actuation Time Delays}
A time-delay SSA was developed by recording the piecewise constant history of the system and inputing the current control activation at time $t$ in the following delayed manner
\begin{equation}
u_{\tau}(t)=\left\{
\begin{array}{rl}
      0 ,&\text{ for }  t \leq \tau, \\
      {u}^{\mathcal{C}}(x_1(t-\tau), x_2(t-\tau)) , &\text{ for }   t > \tau,\\
\end{array}\right. 
\label{scoreeq}
\end{equation}
where $\tau$ is the time delay between observation and actuation, and ${u}^{\mathcal{C}}$ is the previously optimized control law for $\mathbf{u}^{FAC}$ or $\mathbf{u}^{PAC}$. We note that the time delay stochastic process was only simulated using the SSA because an appropriate direct FSP/CME integration procedure is not known.

\subsection{Parameter Scalings and Sensitivity}
Parameter sensitivities of the system can be performed by calculating the control performance score ($J$), over a range of parameters which are scaled linearly with a scaling parameter ($s$) while holding all other parameters constant. Any new arbitrary parameter $\lambda\in \{ \kappa,\beta,k_0,\eta,\gamma\}$ can be rescaled from an old one ($\lambda$) according to
\begin{equation}
\lambda' = s \lambda
\end{equation}
It should be noted that each cell receives its own set of parameters, and therefore the number of parameters double during sensitivity analysis of a two cell control process. Two types of sensitivity analysis can be performed, one which performs the analysis on one cell at a time while keeping the other constant, and another which scales similar parameters in both cells at the same time. Steady state distributions of the perrtubed models were solved using the FSP.
\section{Results}
\subsection{Effects of Changes to Granularity}
\begin{figure*}[t!]
\begin{center}
\includegraphics[width=1\textwidth]{GranularityPerturbation.png}
\vspace{-0.1in}
\caption{Effects of Scale Changes. (Ai-Fi) Joint (Ai,Ci,Di) and marginal (Bi,Di,Fi) probability distributions for the FAC control method at three different $\alpha$. $\alpha=0.2$ (top), $\alpha=1$ (middle) and $\alpha=2.0$ (bottom) relative to the target point (blue circles). (Aii-Fii) Same as (Ai-Fi), but for PAC controller. (G)  FAC (cyan) and PAC (magenta) controller performance as function of system volume. Dashed lines correspond to SSA analyses at $\alpha=100$. In all analyses, the system model and PAC and FAC controllers are unchanged from \cite{May2021} other than their adaptation to the new system volume size.}
\label{Volume}
\end{center}
\vspace{-0.2in}
\end{figure*}
Despite noise being a critical component of noise-exploiting control, the removal of noise through the addition of system granularity is likely beneficial to control performance. This two-edged characteristic that noise is critically important yet bad was explored by observing how control performance changes with system granularity, a parameter that is known to make systems behave more deterministically.

Control performance scores for the $\mathbf{u}^{FAC}$ and $\mathbf{u}^{PAC}$ controllers were calculated at different levels of granularity ($\alpha$) between 0.1 and 2.0. Figures.\ \ref{Volume} (Ai - Fi) show the joint (left plots) and marginal (right plots) distributions of the two cells relative to the specified target position (circles in joint distributions and vertical dashed lines in marginal distributions)
for the FAC controller and for $\mathbf{\alpha}$=.2 (A,B), 1 (C,D) and 2 (E,F). As the granularity changes from $\alpha$=0.2 , to $\alpha$=2, the distributions become focused at a point near the target, and the performance score improves from $J=$333 to 85 to 27, respectively. The PAC controller performance also improves considerably with alpha as shown in figures. \ref{Volume} (Aii - Fii).

Figure \ \ref{Volume}G shows the trend of the performance score versus alpha for both the FAC (solid cyan line) and the PAC (solid magenta line) controller. This improvement in performance appears to approach a small value as the granularity goes to infinity, but since the size of the FSP increases with the square of the system size, systems much larger than $\alpha$=2 (where $\mathbf{A}_0\in \mathbb{R}^{10^4\times10^4}$) are difficult to calculate. To bypass this limit in the FSP, sixteen SSA simulations were used to sample the CME of a system with a much larger volume of $\alpha$=100. Each SSA was ran for 5e7 minutes and only the last 4e7 minutes were sampled from. The performance score estimates of this high granularity SSA using the FAC and PAC were 4.13 and 27.5 respectively, which are plotted as dashed lines in figure \ref{Volume}G. We note however, that systems with larger granularity take longer to reach a steady state distribution, and it is unclear if performance improvements can be obtained with further increases to the system volume.  However, for all volumes considered, we found that both controllers monotonically improved with increased volume and that $\mathbf{u}^{FAC}$ always outperforms the $\mathbf{u}^{PAC}$ at any system volume.

These data suggest that increasing granularity increases steady state control performance even if the controller itself depends on noise for its implementation. Also, designing a controller to work for a system of one volume size can result in a controller that works even for another much larger volume.  This has practical significance, because it is relatively easy to search over a control law defined on a small $M\times M$ grid of states (e.g., when $M$=50 as in \cite{May2021}), but to implement the controller for a larger system which may be computationally prohibitive. This result implies that it may be possible to optimize controller using coarse-grained FSP analyses for small granularity problems and adapt these via interpolation for use with larger, more realistic systems, that exceed the computational limit of standard FSP computations.

\begin{figure*}
\begin{center}
\includegraphics[width=1\textwidth]{TimeDelayPerturbation.png}
\vspace{-0.1in}
\caption{Effects of time delays. (Ai-Fi) Joint (Ai,Ci,Di) and marginal (Bi,Di,Fi) probability distributions for the FAC control method at three different time delays of $\tau/\gamma$=0.01 (top), 0.1 (middle) and 1 (bottom) relative to the target point (blue circles). (Aii-Fii) Same as (Ai-Fi), but for PAC controller. (G) FAC PAC performance as function of time delay show that PAC outperforms the FAC when time delay is greater than $0.07 /\gamma$. In all analyses, the system model and PAC and FAC controllers are unchanged from \cite{May2021}.}
\label{Time}
\end{center}
\vspace{-0.2in}
\end{figure*}

\subsection{Effects of Time Delays on Controller Performance}

Time delay was hypothesized to adversely affect control performance of the $\mathbf{u}^{FAC}$ and $\mathbf{u}^{PAC}$ controllers. To test the effect of time delay on control performance, groups of time delayed SSA were simulated while changing controller and the level of time delay between each group. Each SSA was sampled for 1000 times over 10000 minutes after reaching a steady state by simulating the system for 10000 minutes.

Figure\ \ref{Time} shows the joint and marginal of each controller and time-delay pair and the resulting score, with panels Ai - Fi showing results for the FAC controller and panels Aii-Fii showing results for the PAC controller, and panel G showing the score of both controllers versus the time delay. From the figures, it is clear that performance is rapidly degraded as the delay approaches and then exceeds the characteristic time scale of the process. At low time delays (below $\tau = 0.07/\gamma$), the FAC outperforms the PAC ($J=$87 versus 146 at $\tau = 0.01/\gamma$) but at moderate time delays (above $\tau = 0.07/\gamma$) the PAC outperforms the FAC ($J=$170 versus 219 at $\tau = 0.1/\gamma$)).

This data taken together show that time delays much larger that $0.07 / \gamma$ is detrimental to the FAC controller; delays beyond $0.2 / \gamma$ are detrimental for the PAC controller; and the best choice in controller is dependent the level of time delay in the system.  For extreme levels of time delay, both systems lose their asymmetry, and their scores become much worse ($J=$1041 and 931).

\begin{figure*}
\begin{center}
\includegraphics[width=1\textwidth]{ParameterPerturbation.png}
\caption{Effects of parametric variations to Cell 1 (top row), Cell 2 (middle row) or to both cells simultaneously (bottom row). Each column show the effect of changing a single model parameter (e.g., $\kappa$ in the leftmost row). Trends for the $FAC$ (solid lines) and PAC (dashed lines), but the optimal controller can change under parameter variations. In all analyses, the system model and PAC and FAC controllers are unchanged from \cite{May2021}.}
\label{Parameter}
\end{center}
\end{figure*}

\subsection{Effects of Parameter Errors or Extrinsic Uncertainties}

It is rare that any parameters are exactly known for biological processes, and real systems often exhibit slightly different parameters due to extrinsic variations. This parameter variation between cells is factor that could influence control performance. To test this, parameter sensitivity analysis was performed on the cell one while holding the parameters of cell two constant figure \ref{Parameter}(Row I). For each perturbation analysis, both the $\mathbf{u}^{FAC}$ and $\mathbf{u}^{FAC}$ controller was used over a range of $s$ between 0.5 and 2. The data for such perturbations are shown in figure \ref{Parameter}. The analysis was then repeated, this time holding the parameters of Cell 1 constant and perturbing only the parameters of Cell 2  (shown in figure\ \ref{Parameter}(Row II)). Finally, both cell one and cell two had their parameters jointly perturbed by the same amount (figure \ref{Parameter}(Row III)).

Parameter changes were found to have a broad range of effects, including increasing performance, decreasing performance, or having little effect. Figure \ref{Parameter}(Bi) shows that increasing $\beta$ in Cell 1 worsens performance while increasing $\beta$ in Cell 2 improves performance. In some cases, these effects were not monotonic; for example increasing $\kappa$ in Cell 1 is highly advantageous up to a limit after which the control performance degrades rapidly. In other cases (such as for $k_0$), the effect of parameter perturbations on performance is insignificant even for relatively large ($s$=2) perturbations. When parameters of both Cell 1 and Cell 2 are jointly changed (Fig.\ \ref{Parameter} bottom row) some unique trends were observed. It was found that the $\mathbf{u}^{FAC}$ often outperformed the $\mathbf{u}^{PAC}$, but for large $s$ in a few parameters the inverse was found (figure \ref{parameter} (Row III, E and At).

When examining the effects of changing parameters when using the $\mathbf{u}^{FAC}$ (solid lines) or the $\mathbf{u}^{PAC}$ (dashed lines), general trends typically remained the same in that changes to a parameter which caused a decrease in the control performance of the FAC also tended to decrease the performance of the PAC. Some parameters (e.g., $\kappa$, $\gamma$) reached minima that performed better than the original, which suggests that the physical system itself has room for optimization beyond the original design that could lead to better control performance, {\em even without changing the control law}.  One explanation for why changes to $\eta$ had no effect on control performance is that the origonal parameter was so small that it may require a larger scaling on a different magnitude before changes are observed. For situations where the $\mathbf{u}^{PAC}$ outperformed the $\mathbf{u}^{FAC}$ at large $s$, this may be an indication that re-optimizing the controller for this parameter combination could yield better control results, since the controller is not re-optimized at each parameter combination.

\section{Conclusion}
Key results from \cite{May2021} showed that noise was a key requirement for the development of a noise-exploiting controller, and that deterministic systems could not be controlled to two different stable points if they both started at the same initial condition. As a stochastic system becomes increasingly more granular, it also becomes less noisy and more similar to its ODE solution which is known to be impossible to control to two different fates, but here, we have shown that the removal of noise through system granularity led to better control performance.  Therefore, stochastic controllers may perform work best on models which are nearly deterministic, but not quite. Unfortunately, such systems were found to take longer time to achieve steady state.

The ability to analyze the model at one system granularity apply it elsewhere might help analyze models for which the computational effort may be too large. Since the computation time of the FSP solution to the CME grows with the square of the number of states, this can cause an explosion in computational requirements for large systems. One alternative is to learn a controller at a computationally feasible number of states and apply them to large systems which cannot be solves for using the FSP.

Parameter perturbation analysis showed that there is room to improve control performance by adjusting system parameters, and that joint optimization of the controller with the system parameters may lead to better control performance. It also shows that populations of cells where each cell mightt have its own parameter combination could still be reasonably controlled for small changes in parameter scale, but large changes could be detrimental.

Time delay analysis showed that increasing time delay decreased control performance. It was also found that a controller with less control information outperformed a controller with more information at higher levels of time delay. We believe this is happening because a controller with more information can afford to be more aggressive to implement its control, and time delay can cause this aggression to backfire. A controller was optimized which required no information in May et al which had a score of 402. This controller should be immune to time delay since no feedback is required for this controller. Here we saw both controllers did performed worse than 402 at very large time delay.

\begin{thebibliography}{10}
\providecommand{\url}[1]{#1}
\csname url@samestyle\endcsname
\providecommand{\newblock}{\relax}
\providecommand{\bibinfo}[2]{#2}
\providecommand{\BIBentrySTDinterwordspacing}{\spaceskip=0pt\relax}
\providecommand{\BIBentryALTinterwordstretchfactor}{4}
\providecommand{\BIBentryALTinterwordspacing}{\spaceskip=\fontdimen2\font plus
\BIBentryALTinterwordstretchfactor\fontdimen3\font minus
  \fontdimen4\font\relax}
\providecommand{\BIBforeignlanguage}[2]{{%
\expandafter\ifx\csname l@#1\endcsname\relax
\typeout{** WARNING: IEEEtran.bst: No hyphenation pattern has been}%
\typeout{** loaded for the language `#1'. Using the pattern for}%
\typeout{** the default language instead.}%
\else
\language=\csname l@#1\endcsname
\fi
#2}}
\providecommand{\BIBdecl}{\relax}
\BIBdecl

\bibitem{May2021}
M.~May and B.~Munsky, ``{Exploiting noise, nonlinearity, and feedback to
  differentially control multiple synthetic cells with a single optogenetic
  input},'' pp. 1--28, 2021.
eps
\bibitem{Ng2019}
A.~H. Ng, T.~H. Nguyen, M.~G{\'{o}}mez-Schiavon, G.~Dods, R.~A. Langan, S.~E.
  Boyken, J.~A. Samson, L.~M. Waldburger, J.~E. Dueber, D.~Baker, and
  H.~El-Samad, ``{Modular and tunable biological feedback control using a de
  novo protein switch},'' pp. 265--269, 2019.

\bibitem{Liu2018}
C.~C. Liu, M.~C. Jewett, J.~W. Chin, and C.~A. Voigt, ``{Toward an orthogonal
  central dogma},'' \emph{Nature Chemical Biology}, vol.~14, no.~2, pp.
  103--106, 2018.

\bibitem{Sheets2020}
M.~B. Sheets, W.~W. Wong, and M.~J. Dunlop, ``{Light-Inducible Recombinases for
  Bacterial Optogenetics},'' \emph{ACS Synthetic Biology}, vol.~9, no.~2, pp.
  227--235, 2020.

\bibitem{Groseclose2020}
\BIBentryALTinterwordspacing
T.~M. Groseclose, R.~E. Rondon, Z.~D. Herde, C.~A. Aldrete, and C.~J. Wilson,
  ``{Engineered systems of inducible anti-repressors for the next generation of
  biological programming},'' \emph{Nature Communications}, vol.~11, no.~1, pp.
  1--15, 2020. [Online]. Available:
  \url{http://dx.doi.org/10.1038/s41467-020-18302-1}
\BIBentrySTDinterwordspacing

\bibitem{Shin2020}
J.~Shin, S.~Zhang, B.~S. Der, A.~A. Nielsen, and C.~A. Voigt, ``{ Programming
  Escherichia coli to function as a digital display },'' \emph{Molecular
  Systems Biology}, vol.~16, no.~3, pp. 1--12, 2020.

\bibitem{Baumschlager2017}
A.~Baumschlager, S.~K. Aoki, and M.~Khammash, ``{Dynamic blue light-inducible
  T7 RNA polymerases (Opto-T7RNAPs) for precise spatiotemporal gene expression
  control},'' \emph{ACS synthetic biology}, vol.~6, no.~11, pp. 2157--2167,
  2017.

\bibitem{Chen2020}
\BIBentryALTinterwordspacing
S.~Y. Chen, L.~C. Osimiri, M.~Chevalier, L.~J. Bugaj, T.~H. Nguyen, R.~A.
  Greenstein, A.~H. Ng, J.~Stewart-Ornstein, L.~T. Neves, and H.~El-Samad,
  ``{Optogenetic Control Reveals Differential Promoter Interpretation of
  Transcription Factor Nuclear Translocation Dynamics},'' \emph{Cell Systems},
  vol.~11, no.~4, pp. 336----353.e24, 2020. [Online]. Available:
  \url{http://dx.doi.org/10.1016/j.cels.2020.08.009}
\BIBentrySTDinterwordspacing

\bibitem{Lillacci2018}
G.~Lillacci, Y.~Benenson, and M.~Khammash, ``{Synthetic control systems for
  high performance gene expression in mammalian cells},'' \emph{Nucleic acids
  research}, vol.~46, no.~18, pp. 9855--9863, 2018.

\bibitem{Fox2021}
\BIBentryALTinterwordspacing
Z.~R. Fox, S.~Fletcher, A.~Fraisse, C.~Aditya, and S.~Sosa, ``{MicroMator: Open
  and Flexible Software for Reactive Microscopy},'' \emph{bioRxiv}, pp. 1--9,
  2021. [Online]. Available:
  \url{http://biorxiv.org/cgi/content/short/2021.03.12.435206v1}
\BIBentrySTDinterwordspacing

\bibitem{Baumschlager2021}
\BIBentryALTinterwordspacing
A.~Baumschlager and M.~Khammash, ``{Synthetic Biological Approaches for
  Optogenetics and Tools for Transcriptional Light-Control in Bacteria},''
  \emph{Advanced Biology}, vol.~5, no.~5, p. 2000256, 2021. [Online].
  Available:
  \url{https://onlinelibrary.wiley.com/doi/abs/10.1002/adbi.202000256}
\BIBentrySTDinterwordspacing

\bibitem{Lugagne2017}
\BIBentryALTinterwordspacing
J.~B. Lugagne, S.~{Sosa Carrillo}, M.~Kirch, A.~K{\"{o}}hler, G.~Batt, and
  P.~Hersen, ``{Balancing a genetic toggle switch by real-time feedback control
  and periodic forcing},'' \emph{Nature Communications}, vol.~8, no.~1, pp.
  1--7, 2017. [Online]. Available:
  \url{http://dx.doi.org/10.1038/s41467-017-01498-0}
\BIBentrySTDinterwordspacing

\bibitem{Rullan2018}
\BIBentryALTinterwordspacing
M.~Rullan, D.~Benzinger, G.~W. Schmidt, A.~Milias-Argeitis, and M.~Khammash,
  ``{An Optogenetic Platform for Real-Time, Single-Cell Interrogation of
  Stochastic Transcriptional Regulation},'' \emph{Molecular Cell}, vol.~70,
  no.~4, pp. 745--756.e6, 2018. [Online]. Available:
  \url{https://doi.org/10.1016/j.molcel.2018.04.012}
\BIBentrySTDinterwordspacing

\bibitem{Guarino2020}
A.~Guarino, D.~Fiore, D.~Salzano, and M.~{Di Bernardo}, ``{Balancing Cell
  Populations Endowed with a Synthetic Toggle Switch via Adaptive Pulsatile
  Feedback Control},'' \emph{ACS Synthetic Biology}, vol.~9, no.~4, pp.
  793--803, 2020.

\bibitem{Munsky2012}
B.~Munsky, G.~Neuert, and A.~Van~Oudenaarden, ``{Using gene expression noise to
  understand gene regulation},'' \emph{Science}, vol. 336, no. 6078, pp.
  183--187, 2012.

\bibitem{Szymanska2015}
P.~Szyma{\'{n}}ska, N.~Gritti, J.~M. Keegstra, M.~Soltani, and B.~Munsky,
  ``{Using noise to control heterogeneity of isogenic populations in homogenous
  environments},'' \emph{Physical Biology}, vol.~12, no.~4, 2015.

\bibitem{Gillespie1992}
D.~T. Gillespie, ``{A rigorous derivation of the chemical master equation},''
  \emph{Physica A: Statistical Mechanics and its Applications}, vol. 188, no.
  1-3, pp. 404--425, 1992.

\bibitem{Gillespie1977}
D.~T. Gillespie, ``{Exact stochastic simulation of coupled chemical reactions},''
  \emph{The journal of physical chemistry}, vol.~81, no.~25, pp. 2340--2361,
  1977.

\end{thebibliography}

\end{document}

### 2023-01-31 paper.tex
Update abstract.

@@
 \begin{abstract}
-The field of synthetic biology focuses on creating modular components which can generate complex and controllable synthetic biological systems, but the intrinsic noise of gene regulation can create undesired effects in these systems.
-This noise is often treated as a nuisance which should be removed, but theoretical noise-exploiting controllers have been shown to use noise to their advantage to control difficult control problems which would be difficult otherwise.
-In such work, a difficult control problem similar to Maxwells Demon was proposed and showed that noise exploiting controllers could send two identical systems to different stable points reliably by exploiting the natural noise of gene regulation.
-Those controllers relied upon systems with no time delay well known parameters but, in this work the performance of noise-exploiting controllers are measured under time delay, parameter perturbation and system scale.
+Synthetic biology seeks to create modular components that generate complex and controllable biological behaviors, often at microscopic scales.
+In single-cell applications, where important regulatory RNA or proteins are present in small numbers, intrinsic fluctuations cause biochemical `noise' that frequently results in undesired effects, and this noise is usually treated as a nuisance that must be removed.
+However, recent theoretical analyses have shown that noise can sometimes be used to one's advantage when seeking to manipulate nonlinear regulatory systems that would be difficult or impossible to control in a fully deterministic setting.
+Specifically, it was shown that a noise-exploiting optogenetic controller reminiscent of Maxwell's Demon could systematically force multiple identical systems to reach different independently-assigned stable points, even for a few special cases where the model is inexact or if observations were incomplete.
+This work expands upon those analyses to explore in greater detail how control performance is affected by errors or approximations in the model or by temporal delays in the observations.
+We find that noise-exploiting controllers can remain highly effective despite coarse approximations to the model's scale or incorrect estimations of key model parameters, and these controllers can even retain performance for significant time delays.
+Together, these findings suggest that noise-exploiting control should be possible in real experiments where models are always approximate, where parameters are always uncertain, and where observations are corrupted by errors.
 
 \end{abstract}
-Keywords: Stochastic control, optogenetic, synthetic biology
+Keywords: Optogenetic control, Stochastic gene regulation, Synthetic biology
 
@@
 % Uncomment if a separate title page is required
-%\maketitle
+\maketitle
 % 
@@
 Great advances in these two fields have enabled computer programmable control of cellular protein production through the use of external optogenetic inputs and smart microscope techniques\cite{Fox2021,Baumschlager2021,Lugagne2017}.
-These types of digital-synthetic actuators allows for fine-tuned, computer-modulated control of cellular systems that were previously impossible\cite{Rullan2018, Baumschlager2017} and with fast response times in comparison to chemical diffusion.
+These types of digital-synthetic actuators allow for fine-tuned, computer-modulated control of cellular systems that were previously impossible\cite{Rullan2018, Baumschlager2017} and with fast response times in comparison to chemical diffusion.
 
@@
 
-In \cite{May2021}, it was suprisingly found that, when noise, auto-regulation, and noise-exploiting feedback control were all present at once, the system could be controlled to bring two similar cells to two different arbitrary points simultaneously (figure \ref{cartoons}A), but when any one of these components were missing the system could not break its symmetry.
+In \cite{May2021}, it was surprisingly found that, when noise, auto-regulation, and noise-exploiting feedback control were all present at once, the system could be controlled to bring two similar cells to two different arbitrary points simultaneously (Figure \ref{cartoons}A), but when any one of these components were missing the system could not break its symmetry.
 More specifically, this required that the symmetry between the protein probability distributions of each cell would not be not the same (or asymmetric).
@@
 \section{Methods}
-We begin with a five species model for the production of an arbitratry protein through the activation of light activatible T7 polymerase (figure 1B) motivated by Baumschlager, et al.
+We begin with a five species model for the production of an arbitrary protein through the activation of light activatible T7 polymerase (figure 1B) motivated by Baumschlager, et al.
 This model for the production of protein uses a split T7 domain ($T7_N$,and $T7_C$) which reversibly dimerizes in the presence of light to create a T7 dimer.

### 2023-01-31 paper.tex
Update introduction - needs some citations but is getting close.

@@
 
-In \cite{May2021}, it was surprisingly found that, when noise, auto-regulation, and noise-exploiting feedback control were all present at once, the system could be controlled to bring two similar cells to two different arbitrary points simultaneously (Figure \ref{cartoons}A), but when any one of these components were missing the system could not break its symmetry.
-More specifically, this required that the symmetry between the protein probability distributions of each cell would not be not the same (or asymmetric).
-Three effective controllers were found which could do this, a Fully Aware Feedback Controller (FAC) that observed the protein counts of all observed cells, a Partially Aware Feedback Controller (PAC) that observed the protein counts of only one cell, and a probabilistic Model Predictive Feedback Controller (pMPC) that observed only the primary cell and integrated a probabilistic model to represent the non-stationary probability distribution of the secondary cells in the same environment.
-
-The solutions of the controllers was carried out by solving the chemical master equation (i.e., the forward Komogorov equation) to estimate the probability of possible cellular states.
+Szymanska et al.~\cite{Szymanska2015} examined a theoretical system inspired by the genetic toggle switch from Kobayashi et al.\cite{Kobayashi:2004}, where two gene products repress one another, but where degradation of one protein was under the control of UV radiation via the XXX pathway.
+Their analysis showed that control was possible despite certain parameter errors and time delays due to maturation of fluorescent proteins and limited observation of the regulatory proteins.
+To explore this concept in the context of more realistic parameters, May et al.~\cite{May2021} fit a simplified stochastic model to reproduce data measured in Baumschlager, et al.\cite{} for the expression from the XXX promoter under the optogenetic control of a UV-activated T7 polymerase.
+An extended model showed that addition of a positive auto-regulation module (see Figure XXX) could help to maintain an elevated expression phenotype in the presence of UV excitation.
+Using discrete stochastic models based on direct solutions to the chemical master equation, it was shown that   the combination of biochemical noise, nonlinear auto-regulation, and a {\em single} optogenetic feedback could allow two genetically identical cells with arbitrary initial conditions to be controlled simultaneously to reach different specified phenotypes (Figure \ref{cartoons}A).
+Moreover, effective control was possible not only for a Fully Aware Feedback Controller (FAC) that observed the protein counts of all observed cells, but control could be achieved almost as well for a Partially Aware Feedback Controller (PAC) that observed the protein counts of only one cell and used a probabilistic model to estimate the expression for unobserved cells.
+Moreover, it was again seen that control remained effective even in the circumstance where the model used to design the control law was an inexact simplification of the process dynamics to which the controller was later applied.
+%The solutions of the controllers was carried out by solving the chemical master equation (i.e., the forward Komogorov equation) to estimate the probability of possible cellular states.
 For a two cell, single-input-single output (SIMO) control problem, the FAC performed the best (figure\ \ref{cartoons}C(top)), and the PAC (figure\ \ref{cartoons}C(bottom)) 2nd, and pMPC controllers 3rd.
@@
 
-In this paper, the control problem is extended to examine how the FAC and PAC controllers are affected by time delay, the system granularity, and parameter uncertainty.
-In the `Methods' section, we briefly set up a stochastic chemical master equation model for multiple individual cells in the same spatially homogeneous environment, and we formulate the effects of FAC and PAC controllers on the dynamics of cells within that environment.
-Next, in the `Results' section, we examine the effects of volume changes, time delays, and parametric uncertainties on the FAC and PAC control performance.
-Finally, in `Conclusions' we summarize our findings and discuss potential implications that these controllers may have on future systems and synthetic biology investigations.
+In this paper, the control problem is extended to examine how the fully- and partially-aware controllers (FAC and PAC, respectively) are affected by additional model uncertainties, including coarse grained approximations to the system dynamics, uncertainties or errors in system parameters, and delays in the biochemical sensors or optogenetic controllers through which the external feedback is implemented.
+First, in the next section, we briefly describe our methods to set up a stochastic chemical master equation model for multiple individual cells in the same spatially homogeneous environment; we formulate the FAC and PAC controllers on the dynamics of cells within that environment; and we explain how the control law is optimized to achieve desired performance objectives.
+Next, in the `Results' section, we examine how the granularity of the model formulation, the inaccuracy of parametric assumptions, and the inclusion of time delays affect the FAC and PAC control performance.
+Finally, in `Conclusions' we summarize our findings and discuss the implications that these controllers, and their robustness to modeling errors, may have on future systems in synthetic biology.
 

### 2023-02-01 paper.tex
More updates to introduction.

@@
 Classical and modern control methods like PID control and model predictive control have been implemented in such systems\cite{} which can control synthetic systems to different stable points. 
+% XXX Missing Citation above.
 Theoretical works have shown that stochastic systems can be controlled to unstable points by oscillating around those points \cite{Guarino2020} but new control techniques which exploit the full probability distribution information of the system can be used in certain scenarios to control simple systems to unstable points by exploiting the natural noise of protein production to its advantage.
@@
 Szymanska et al.~\cite{Szymanska2015} examined a theoretical system inspired by the genetic toggle switch from Kobayashi et al.\cite{Kobayashi:2004}, where two gene products repress one another, but where degradation of one protein was under the control of UV radiation via the XXX pathway.
+% XXX - add citation to Kobayashi paper, and update name of the pathway.  I think it was the SOS pathway.
 Their analysis showed that control was possible despite certain parameter errors and time delays due to maturation of fluorescent proteins and limited observation of the regulatory proteins. 
-To explore this concept in the context of more realistic parameters, May et al.~\cite{May2021} fit a simplified stochastic model to reproduce data measured in Baumschlager, et al.\cite{} for the expression from the XXX promoter under the optogenetic control of a UV-activated T7 polymerase.
-An extended model showed that addition of a positive auto-regulation module (see Figure XXX) could help to maintain an elevated expression phenotype in the presence of UV excitation.
-Using discrete stochastic models based on direct solutions to the chemical master equation, it was shown that   the combination of biochemical noise, nonlinear auto-regulation, and a {\em single} optogenetic feedback could allow two genetically identical cells with arbitrary initial conditions to be controlled simultaneously to reach different specified phenotypes (Figure \ref{cartoons}A).
-Moreover, effective control was possible not only for a Fully Aware Feedback Controller (FAC) that observed the protein counts of all observed cells, but control could be achieved almost as well for a Partially Aware Feedback Controller (PAC) that observed the protein counts of only one cell and used a probabilistic model to estimate the expression for unobserved cells.
-Moreover, it was again seen that control remained effective even in the circumstance where the model used to design the control law was an inexact simplification of the process dynamics to which the controller was later applied.
+To explore this concept in the context of more realistic models and parameters, May et al.~\cite{May2021} fit a simplified stochastic model to reproduce data measured in Baumschlager, et al.\cite{} for the expression from the XXX promoter under the optogenetic control of a UV-activated T7 polymerase (see model in Figure\ \ref{cartoons}A, top promoter).
+% XXX - add details about the promoter and citation.
+% XXX - let's change the order of panels A and B for this figure so that we can refer to them in the proper order.  I adjusted the text below to assume this change.
+An extended model showed that addition of a positive auto-regulation module (Figure\ \ref{cartoons}A, bottom promoter) could help to maintain an elevated expression phenotype in the presence of UV excitation.
+% XXX - Missing figure reference
+Using discrete stochastic models based on direct solutions to the chemical master equation, it was shown that the combination of biochemical noise, nonlinear auto-regulation, and a {\em single} optogenetic feedback could allow two genetically identical cells with arbitrary initial conditions to be controlled simultaneously to reach different specified phenotypes (Figure \ref{cartoons}B).
+For example, Figure\ \ref{cartoons}C(left) reproduces the heat maps for two control laws, where the level of a single external UV excitation signal was modulated depending on the observed cellular activity, and Figure\ \ref{cartoons}C(right) shows the corresponding marginal stationary distributions that could be achieved under these control laws.
+Effective control was possible not only for a Fully Aware Feedback Controller (FAC, Figure\ \ref{cartoons}C(top)) that observed the protein counts of all observed cells, but control could be achieved nearly as well for a Partially Aware Feedback Controller (PAC, Figure\ \ref{cartoons}C(bottom)) that observed the protein counts of only one cell.
+Moreover, it was again seen that these control laws remained effective even in the circumstance where the model used to design the control law was an inexact simplification of the process dynamics to which the controller was later applied.
 %The solutions of the controllers was carried out by solving the chemical master equation (i.e., the forward Komogorov equation) to estimate the probability of possible cellular states.
@@
 
-In this paper, the control problem is extended to examine how the fully- and partially-aware controllers (FAC and PAC, respectively) are affected by additional model uncertainties, including coarse grained approximations to the system dynamics, uncertainties or errors in system parameters, and delays in the biochemical sensors or optogenetic controllers through which the external feedback is implemented.
-First, in the next section, we briefly describe our methods to set up a stochastic chemical master equation model for multiple individual cells in the same spatially homogeneous environment; we formulate the FAC and PAC controllers on the dynamics of cells within that environment; and we explain how the control law is optimized to achieve desired performance objectives.
-Next, in the `Results' section, we examine how the granularity of the model formulation, the inaccuracy of parametric assumptions, and the inclusion of time delays affect the FAC and PAC control performance.
+In this paper, the analysis of this Single-Input-Multiple-Output (SIMO) multicellular control problem is extended to examine how the FAC and PAC controllers are affected by additional model uncertainties, including coarse grained approximations to the system dynamics, uncertainties or errors in system parameters, and delays in the biochemical sensors or optogenetic controllers through which the external feedback is implemented.
+In the next section, we briefly describe our methods to formulate the chemical master equation to analyze multiple individual cells in the same spatially homogeneous environment; we define the FAC and PAC controllers and their effects the dynamics of cells within that environment; and we explain how the control law is optimized to achieve desired performance objectives.
+Next, in the `Results' section, we explore how the granularity of the model formulation, the inaccuracy of parametric assumptions, and the inclusion of time delays would affect the FAC and PAC control performance.
 Finally, in `Conclusions' we summarize our findings and discuss the implications that these controllers, and their robustness to modeling errors, may have on future systems in synthetic biology.

### 2023-02-01 paper.tex
Change figure 1 caption -- request changes to figure layout.

@@
 \includegraphics[width=\columnwidth]{CartoonAndControler.png}
-\caption{(A) Schematic of the T7 optogenetic control system (adapted from \cite{May2021}).
-A microscope observer is partially (PAC) or fully (FAC) aware  of cells; the resulting controller modulates a single spatially homogeneous light signal that perturbs all cells equally seeks to drive each genetically identical cells to a different, specified states.
-(B) Schematic of the optogenetically controllable T7 polymerase model, which is identical in all cells: Light activates the T7 polymerase-based production of protein, and the protein self-actuates its own production.
-(C) The FAC (top) and PAC (bottom) control laws for two cells ($\mathbf{u}^{XAC}(x_1,x_2)$ shown on left) break the symmetry between Cell 1 and Cell 2 and result in distinct marginal distributions for the cells' responses (right).}
+\caption{{\bf Optogenetic control of multiple cells using a single light course}
+({\bf A}) Schematic of the gene expression system.
+UV light activates T7 polymerase based transcription and protein expression from the first promoter-gene construct, and the protein actuates its own production via a second promoter-gene construct.
+({\bf B}) Schematic of the optogenetic control system.
+An observer is partially (PAC) or fully (FAC) aware of cell expression.
+The controller modulates a single spatially homogeneous light signal in attempt to drive each genetically identical cell to its own specified state.
+({\bf C}) The FAC (top) and PAC (bottom) control laws for two cells ($\mathbf{u}^{XAC}(x_1,x_2)$ shown on left) break the symmetry between Cell 1 and Cell 2 and result in distinct marginal distributions for the cells' responses (right).
+Figure adapted from \cite{May2021}.}
 \label{cartoons}
@@
 \end{figure}
+% XXX - Please modify figure as follows: (1) switch panels A and B to match caption.
+(2) Change title of panel B (previously panel A) to "External Observer, controller and actuator".
+(3) Change title of panel A (previously panel B) to "Internal gene regulation pathway".
+(4) Change title of panel C to "Feedback control law and resulting performance"
 

### 2023-02-01 paper.tex
Update methods up to section 2.2.

@@
 \section{Methods}
-We begin with a five species model for the production of an arbitrary protein through the activation of light activatible T7 polymerase (figure 1B) motivated by Baumschlager, et al.
-This model for the production of protein uses a split T7 domain ($T7_N$,and $T7_C$) which reversibly dimerizes in the presence of light to create a T7 dimer.
-The T7 monomer is created at a constant rate and degrades linearly with the number of monomer.
-The split monomers of T7 can bind to create a T7 dimer (equation ...) .
-The T7 dimer can then bind to a gene which enables the production of protein with a leak term, and the protein can bind to a secondary self-activated promoter described by a Hill function for the creation of more protein.
-This auto-regulatory motif in protein production is responsible for creating non-linearity in the system that enables noise-exploiting control.
-The final step was the degredation of protein
+In previous work~\cite{May2021}, a six-species gene regulation model was postulated for the light-activated association of two T7 split domains (species 1 and 2) to form active T7 polymerases (species 3) which could then in turn associate with inactive T7 promoters (species 4) to create an active gene construct (species 5) that would produce its eventual protein product (species 6).
+A second, far simpler single-species model was also postulated by assuming quasi-steady equilibrium for the first five species.
+Both models were independently parametrized by fitting to the same experimental data from Baumschlager, et al.~\cite{XXX}, and then extended to include a secondary self-activated promoter-gene construct for which the expression rate was defined by a Hill function (see Eqn.\ \ref{prodRate} below).
+In \cite{May2021}, we showed that a control law identified using the simple model would work to achieve differential control for the more complex model, thus motivating our current study's efforts to explore further how control performance changes upon additional simplifying assumptions or errors in model specification.
 
-These model dynamics were then simplified to a single ODE with auto-regulation describing the protein accumulation as: 
+\subsection{Model}
+To examine how model approximations would affect the application of noise-enhanced control strategies, we start with the one species model from \cite{May2021} (see Fig.\ \ref{cartoons}A), which consists of two reactions: protein production and degradation.
+The nonlinear light-activated protein production rate is given by
  \begin{equation}
-\frac{dp}{dt}=u(x_1,x_2)  + \kappa\frac{p^\eta}{p^\eta+\beta^\eta}+ k_0 -\gamma p,
-\label{rateeq}
+\nu_1(x,t)= \kappa\frac{x^\eta}{x^\eta+\beta^\eta}+ k_0 + u(UV(t)),
+\label{prodRate}
 \end{equation}
-
-where $p$ is the protein count; $\kappa$ is the maximum auto-regulation promoter strength; $\beta$ is the concentration at which auto-regulation reaches half strength; $\eta$ is the auto-regulation promoter cooperativity; $\gamma$ is the degradation rate; $k_0$ is the promoter leakage rate; and $u$ is the T7 promoter strength.
-The protein production rate response to the light level $u$ can be modulated in response to the state of the system $u(x_1,x_2)$ and introduces the possibility of state-dependent optogenetic feedback to the system.
-We note that $u(x_1,x_2)$ refers to the effect of light on the optogenetic system instead of light activity itself.
-A schematic of the model is shown in figure \ \ref{cartoons}B, and parameters were chosen to reproduce dynamics measured in \cite{Baumschlager2017} and are shown in are defined in Table 1\cite{May2021}.
-Two optimized controls laws $u=u^{\rm (FAC)}$ and $u=u^{\rm (PAC)}$ are depicted in figure \ \ref{cartoons}C (top left and top right, respectively).
+%
+%These model dynamics were then simplified to a single ODE with auto-regulation describing the protein accumulation as: 
+%\begin{equation}
+%\frac{dp}{dt}=u(x_1,x_2)  + \kappa\frac{p^\eta}{p^\eta+\beta^\eta}+ k_0 -\gamma p,
+%\label{rateeq}
+%\end{equation}
+%
+where $x$ is the instantaneous protein level; $\kappa$ is the maximum strength of the auto-regulation promoter; $\beta$ is the concentration at which auto-regulation promoter reaches half its strength; $\eta$ is the cooperativity in the auto-regulation promoter; $k_0$ is the promoter leakage rate; and $u(UV(t))$ is the T7 promoter strength.
+Through external feedback, the level of light excitation, and therefore the T7 promoter strength, $u(UV(t))$, can be modulated in response to the state of the system, thus removing explicit dependence on time.
+For example, in a system with the observation of two cells, one could define $u(UV(t)) = u(x_1,x_2)$ as shown in Fig.\ \ref{cartoons}C(top left) for the optimized control law $u=u^{\rm (FAC)}$, and for a system with the observation of only one cell, one could define $u(UV(t)) = u(x_1)$ as shown in Fig.\ \ref{cartoons}C(bottom left) for the optimized control law $u=u^{\rm (PAC)}$.
+Protein degradation is assumed to be a first order process with rate $\gamma$:
+\begin{equation}
+\nu_2(x) = \gamma x.
+\end{equation}
+Parameters describing the auto-regulation promoter, $\kappa$, $\eta$, $\beta$, and $k_0$, and the degradation rate $\gamma$
+%were chosen to reproduce dynamics measured in \cite{Baumschlager2017} and 
+% XXX - I don't think that this is right.  Only the parameters u(UV) and gamma were fit to data.
+are presented in Table 1~\cite{May2021}.
 
@@
 
-\subsection{Stochastic Simulation Algorithm and the Finite State Projection}
-Solving discrete stochastic representations of chemical dynamics is critical to the development of noise-exploiting controllers, and can be analyzed using the Stochastic Simulation Algorithm (SSA) , or the Finite State Projection (FSP).
-To recast the above ODE formulation into the discrete stochastic representation, we define the enumerated $i = 1,2,\ldots$ states of the system as the tuple of the numbers of proteins in each cell as $\mathbf{X}_i = [x_{i1},x_{i2},\ldots,x_{iN_c}]$, and we define the stoichiometry vectors as the change in state following one reaction event (e.g., $\mathbf{X}_i \rightarrow \mathbf{X}_i + \mathbf{s}_\nu$), where the $2N_c$ possible reactions are:
-
+\subsection{Stochastic analyses of the model}
+%Solving discrete stochastic representations of chemical dynamics is critical to the development of noise-exploiting controllers, and can be analyzed using the Stochastic Simulation Algorithm (SSA) , or the Finite State Projection (FSP).
+%To recast the above ODE formulation into the discrete stochastic representation, 
+To describe the discrete behavior of the above model for a population system with $N_{\rm c}$ cells, we define the current state of the system as the tuple of the non-negative numbers of proteins in each cell: $\mathbf{X}_i = [x_{1},x_{2},\ldots,x_{N_c}]_i\in \mathbb{Z}_{\ge 0}$, where the index $i$ denotes the enumeration of the state within the countably infinite set of all possible states, $\mathbf{X}_i\in\mathcal{X}=\{\mathbf{X}_1,\mathbf{X}_2,\ldots\}$.
+The stoichiometry vector for a reaction in any given cell is then defined as the change in state following that reaction event (e.g., $\mathbf{X}_i \rightarrow \mathbf{X}_i + \mathbf{s}_\mu$), where the $2N_c$ possible reactions are defined in pairs corresponding to production (odd indices) and degradation (even indices) as:
 \begin{align}
@@
  \end{align}
-
-where each $\mathbf{e}_{i_c}$ corresponds to the Euclidean basis vector for the $i_c^{\rm th}$ cell. The corresponding propensity functions are:
+where each $\mathbf{e}_{i}\in\mathbb{Z}^{{\rm N_ c}}$ is is a Euclidean vector (i.e., unity for the $i^{\rm th}$ entry and otherwise zero). The corresponding propensity functions are:
 \begin{align}
-w_1 &= u(\mathbf{X},t)  + \kappa \frac{x_1^\eta}{x_1^{\eta}+\beta^{\eta}} + k_0,\ w_2 = \gamma x_1,\nonumber\\
-w_3 &= u(\mathbf{X},t)  + \kappa \frac{x_2^\eta}{x_2^{\eta}+\beta^{\eta}} + k_0,\ w_4 = \gamma x_2,\nonumber\\
+w_1(\mathbf{X}) &= u(\mathbf{X},t)  + \kappa \frac{x_1^\eta}{x_1^{\eta}+\beta^{\eta}} + k_0,\ w_2(\mathbf{X}) = \gamma x_1,\nonumber\\
+w_3(\mathbf{X}) &= u(\mathbf{X},t)  + \kappa \frac{x_2^\eta}{x_2^{\eta}+\beta^{\eta}} + k_0,\ w_4(\mathbf{X}) = \gamma x_2,\nonumber\\
 &\vdots \nonumber\\
-w_{\rm 2N_c-1} &= u(\mathbf{X},t)  + \kappa \frac{x_N^\eta}{x_N^{\eta}+\beta^{\eta}} + k_0,\ w_{\rm 2N_c} = \gamma x_N,\label{Props}
+w_{\rm 2N_c-1}(\mathbf{X}) &= u(\mathbf{X},t)  + \kappa \frac{x_N^\eta}{x_N^{\eta}+\beta^{\eta}} + k_0,\ w_{\rm 2N_c}(\mathbf{X}) = \gamma x_N.\label{Props}
  \end{align}
@@
 
-For the same specifications of the stoichiometry and propensity functions, the chemical master equation can also be formulated in matrix/vector form as:
+For the same specifications of the stoichiometry and propensity functions, the chemical master equation can be formulated in matrix/vector form as:
 \begin{equation}
@@
 \end{equation}
-where $\mathbf{P} = [P(\mathbf{X}_1), P(\mathbf{X}_2), \ldots ]^T$ is the enumerated probability mass vector for all possible states of the system; $\mathbf{A}_0$ is the infinitesimal generator of the stochastic process due to the autoregulation promoter and degradation events;  $\textbf{u}^{\mathcal{C}} =[u^{\mathcal{C}}(\mathbf{X}_1), u^{\mathcal{C}}(\mathbf{X}_2), \ldots ]^T$ is the collection of state-dependent control inputs associated with each state; and $\textbf{Bu}^{\mathcal{C}}$ is the infinitesimal generator for the T7 promoter expression by which the effects of controller $\mathcal{C}$ are introduced into the process.
+where $\mathbf{P} = [P(\mathbf{X}_1), P(\mathbf{X}_2), \ldots ]^T$ is the enumerated probability mass vector for all possible states of the system; $\mathbf{A}_0$ is the infinitesimal generator of the stochastic process due to the autoregulation promoter and degradation events;  $\textbf{u}^{\mathcal{C}} =[u^{\mathcal{C}}(\mathbf{X}_1), u^{\mathcal{C}}(\mathbf{X}_2), \ldots ]^T$ is the collection of control inputs associated with each state; and $\textbf{Bu}^{\mathcal{C}}$ is the contribution that these control inputs make to the infinitesimal generator when included into the feedback process.
 
-The generator $\mathbf{A}_0$ can be constructed according to:
+More specifically, the open look infinitesimal generator, $\mathbf{A}_0$, is constructed according to:
 \begin{equation}
@@
 \begin{array}{rl}
--\sum_{\nu =1}^{2N_c} w_{\nu}(\mathbf{X}_j), &\text{for }i=j,\\
-w_{\nu}(\mathbf{X}_j), &\text{for }\mathbf{X}_i = \mathbf{X}_j+\mathbf{s}_\nu\\
+-\sum_{\mu =1}^{2N_{\rm c}} w_{\mu}(\mathbf{X}_j), &\text{for }i=j,\\
+w_{\mu}(\mathbf{X}_j), &\text{for }\mathbf{X}_i = \mathbf{X}_j+\mathbf{s}_\mu\\
 0, & \text{otherwise,}
@@
 \end{equation}
-and the infinitesimal generator $\mathbf{Bu}^{\mathcal{C}}$ of the controller is given by
+and the feedback infinitesimal generator, $\mathbf{Bu}^{\mathcal{C}}$, of the controller is constructed according to
 \begin{equation}
@@
 \begin{array}{rl}
-- N_c{u}^{\mathcal{C}}(\mathbf{X}_j), &\text{for }i=j,\\
-{u}^{\mathcal{C}}(\mathbf{X}_j), &\text{for }\begin{array}{ll}\mathbf{X}_i =\mathbf{X}_j + \mathbf{e}_{i_c}, \\ \text{and } i_c = 1,\ldots,N_c\end{array}\\
+- N_{\rm c}{u}^{\mathcal{C}}(\mathbf{X}_j), &\text{for }i=j\\
+{u}^{\mathcal{C}}(\mathbf{X}_j), &\text{for }\begin{array}{ll}\mathbf{X}_i =\mathbf{X}_j + \mathbf{e}_{i_{\rm c}}, \\ \text{and } i_{\rm c} = 1,\ldots,N_{\rm c}\end{array},\\
 0, & \text{otherwise,}
@@
 \end{equation}
-where ${u}^{\mathcal{C}}(\mathbf{X}_j)$ is the specification of the $\mathcal{C}$ controller in terms of the (partially observed) state.
+where ${u}^{\mathcal{C}}(\mathbf{X}_j)$ is the specification of the $\mathcal{C}$ controller in terms of the state, or its partial observations.
 
-For a given controller, the steady state of the system ($\mathbf{P}^*$) found by solving Eq.\ \ref{CME} is given by
+For a given controller, the equilibrium distribution of the system ($\mathbf{P}^*$) found by solving Eq.\ \ref{CME} is given by:
 \begin{equation}
-\mathbf{P}^*=null(\mathbf{A}+\textbf{Bu}^{\mathcal{C}}).\label{SSDist}
+\mathbf{P}^*={\rm null}(\mathbf{A}+\textbf{Bu}^{\mathcal{C}}).\label{SSDist}
 \end{equation}
 
-and the infinitesimal generator $\mathbf{Bu}^{\mathcal{C}}$ of the controller is given by
-
-\subsection{Control Performance, Optimization, \& Design}
+\subsection{Quantification and optimization of control performance}
 Score metrics of steady state distributions were made in order to enable optimization of the system by adjusting controllers.

### 2023-02-01 paper.tex
Couple more edits to methods section.

@@
 Both models were independently parametrized by fitting to the same experimental data from Baumschlager, et al.~\cite{XXX}, and then extended to include a secondary self-activated promoter-gene construct for which the expression rate was defined by a Hill function (see Eqn.\ \ref{prodRate} below).
-In \cite{May2021}, we showed that a control law identified using the simple model would work to achieve differential control for the more complex model, thus motivating our current study's efforts to explore further how control performance changes upon additional simplifying assumptions or errors in model specification.
+In \cite{May2021}, we showed that a control law identified using the simple model would work to achieve differential control for the more complex model, thus motivating our current study's efforts to explore further how control performance would change upon additional simplifying assumptions or errors in model specification.
+In the following subsections, we discuss our methods to formulate this model, show how it is analyzed numerically, and discuss the specification of the control objective and the determination of an optimal controller.
 
@@
 Through external feedback, the level of light excitation, and therefore the T7 promoter strength, $u(UV(t))$, can be modulated in response to the state of the system, thus removing explicit dependence on time.
-For example, in a system with the observation of two cells, one could define $u(UV(t)) = u(x_1,x_2)$ as shown in Fig.\ \ref{cartoons}C(top left) for the optimized control law $u=u^{\rm (FAC)}$, and for a system with the observation of only one cell, one could define $u(UV(t)) = u(x_1)$ as shown in Fig.\ \ref{cartoons}C(bottom left) for the optimized control law $u=u^{\rm (PAC)}$.
+For example, in a system with the observation of two cells, one could define $u(UV(t)) = u^{\rm (FAC)}(x_1,x_2)$ as shown in Fig.\ \ref{cartoons}C(top left) for the fully aware control law, and for a system with the observation of only one cell, one could define $u^{\rm (PAC)}(UV(t)) = u(x_1)$ as shown in Fig.\ \ref{cartoons}C(bottom left) for the partially aware control law.
 Protein degradation is assumed to be a first order process with rate $\gamma$:
@@
  \end{align}
+
 With these stoichiometry and propensities, we can formulate and run the Gillespie stochastic simulation algorithm\cite{Gillespie1992,Gillespie1977} to generate representative trajectories of the stochastic process.
-
-For the same specifications of the stoichiometry and propensity functions, the chemical master equation can be formulated in matrix/vector form as:
+However, quantifying performance and designing an optimal controller requires a more direct analysis of the chemical master equation (CME).
+For the same specifications of the stoichiometry and propensity functions, the CME is a linear ODE that describes how probability mass changes in time due to the specified reaction propensities and stoichiometries, which can be written:
+\begin{equation}
+\frac{d}{dt}P(\mathbf{X}_i)=\sum_{\mu=1}^{2{\rm N_c}}
+\left(
+-w_{\mu}(\mathbf{X}_i)P(\mathbf{X}_i)
++w_{\mu}(\mathbf{X}_i-\mathbf{s}_{\mu})P(\mathbf{X}_i-\mathbf{s}_{\mu})
+\right).\label{CMEindex}
+\end{equation}
+For convenience, the CME can also be formulated more compactly in matrix/vector form as:
 \begin{equation}
@@
 
-More specifically, the open look infinitesimal generator, $\mathbf{A}_0$, is constructed according to:
+More specifically, the openloop infinitesimal generator, $\mathbf{A}_0$, is constructed according to:
 \begin{equation}
@@
 
-For a given controller, the equilibrium distribution of the system ($\mathbf{P}^*$) found by solving Eq.\ \ref{CME} is given by:
+For a given controller, the equilibrium distribution of the system ($\mathbf{P}^*$) can be found by solving Eq.\ \ref{CME} and is given by:
 \begin{equation}

### 2023-02-01 paper.tex
Edit up to section 2.3.

@@
 \subsection{Quantification and optimization of control performance}
-Score metrics of steady state distributions were made in order to enable optimization of the system by adjusting controllers.
-In the control of a SIMO stochastic system, a control input which tries to perfectly control one system will create increased noise in the other system because their inputs are shared.
-Therefore, a controller which strikes a balance between the two cells should be important when it is being optimized.
+%To quantify the performance of the controlled system, Score metrics of steady state distributions were made in order to enable optimization of the system by adjusting controllers.
 
- The control performance ($J$), is defined as the expected steady state squared Euclidean distance of the process from a specified target ($\mathbf{T}$) state according to $E\{(\mathbf{X}-\mathbf{T})^2\}$.
-This score can be rewritten in terms of a precomputed set of linear weights ($\mathbf{C}$) and the steady state probability distribution ($\mathbf{P}^*$ from Eq.\ (\ref{SSDist})) given by
-{\small \begin{align}
-J&= \lim_{t\rightarrow \infty}\mathbb{E}\{|\mathbf{X}(t)-\mathbf{T}|_2^2\} \nonumber \\ 
-&=\sum_{i,j,\ldots} P^*(x_1=i,x_2=j,\ldots) ((i-\mathbf T_1)^2 + (j-\mathbf T_2)^2 +\ldots)\nonumber  \\
-&=\sum_{i,j,\ldots} P^*_{ij\ldots}(t)C_{ij\ldots} =\mathbf{C}\mathbf{P}^*
+For the single-input-multiple-output control of the multiple cell system, any input signal that is applied to one cell in the population will be felt by all cells in the population.
+As a result, it is not possible to independently control any individual cell without introducing changes to the behavior of the others.
+Therefore, an effective controller must be able to strike a balance among the desired behaviors of all cells in the system.
+
+To quantify performance success, we define the performance error score, $J$, as the expected steady state squared Euclidean distance of the process from the specified target state, $\mathbf{T}$:
+ \begin{equation}
+ J = E\{(\mathbf{X}-\mathbf{T})^2\}.
+ \end{equation}
+This score is easily calculated by applying a linear operation on steady state probability distribution $\mathbf{P}^*$ from Eq.\ \ref{SSDist}) as follows:
+{ \begin{align}
+J&= \lim_{t\rightarrow \infty}\mathbb{E}\{|\mathbf{X}(t)-\mathbf{T}|_2^2\}, \nonumber \\ 
+&=\sum_{i_1,i_2,\ldots} P^*(x_1=i_1,x_2=i_2,\ldots) \left[(i-\mathbf T_1)^2 + (j-\mathbf T_2)^2 +\ldots\right],\nonumber  \\
+&=\sum_{i,j,\ldots} P^*_{ij\ldots}(t)C_{ij\ldots} =\mathbf{C}\mathbf{P}^*,
 \label{Euclid}
-\end{align}}
-Where $J$ is a scalar, $\mathbf{C}$ is a vector of weights given by the Euclidian distance, and $\mathbf{P}^*$ the vectorized steady state distribution of the FSP.
+\end{align}} where $\mathbf{C}$ is simply a vector that contains the squared Euclidian distance of each state from the specified target T, i.e., $C_i = |\mathbf{X}_i-\mathbf{T}|_2^2$.
+As a result of this calculation, $J$ is a non-negative scalar that is zero if any only if $\mathbf{P}^*$ is a delta distribution located exactly at the target vector $\mathbf{T}$.
 
-Two different controller designs were chosen at two different levels of information for the controller to use.
-A fully aware controller ($\mathbf{u}^{FAC}$) used direct observation of the protein count of both cells simultaneously while making its control input decision, while the partially aware controller $\mathbf{u}^{PAC}$) used only the direct observation of a single cell.
-
+In this work, we explore two different controller designs that allow for the controller to use two different types of information.
+A fully aware controller ($\mathbf{u}^{FAC}$) uses direct observation of the protein count of two cells simultaneously while making its control input decision, while the partially aware controller $\mathbf{u}^{PAC}$) uses only the direct observation of a single cell:
 \begin{equation}
-\mathbf{u}^{FAC}=u(x_0,x_1)
+\mathbf{u}^{FAC}=u(x_1,x_2)
 \end{equation}
 \begin{equation}
-\mathbf{u}^{PAC}=u(x_0)
+\mathbf{u}^{PAC}=u(x_1)
 \end{equation}
-where $x_0$ and $x_1$ are discrete integers greater than or equal to 0 which represent the number of protein in cell one and cell two.
-The $\mathbf{u}^{FAC}$ and $\mathbf{u}^{PAC}$ controllers were optimized to minimize $J$ in order to maximize the amount of probability near the target point, and both $\mathbf{u}^{FAC}$ and $\mathbf{u}^{PAC}$ were saved as lookup tables after being optimized.
+% XXX - Do we need to use 0 and 1.
+Why not use 1 and 2 so that this can match better to the formulaton above?
+I changed them to x1 and x2 here, but please chekc to make sure that this is sonsistent throughout text and figures.
+where $x_1$ and $x_2$ are discrete integers greater than or equal to zero that represent the number of proteins in cell one and cell two.
+In \cite{May2021}, the controllers $\mathbf{u}^{FAC}$ and $\mathbf{u}^{PAC}$ controllers were optimized to minimize $J$, and both $\mathbf{u}^{FAC}$ and $\mathbf{u}^{PAC}$ were saved as lookup tables after being optimized.
+In what follows, we will use these controllers exactly and without modification to explore how performance changes when the controllers are applied to systems that are approximations of the model under they were originally designed.
 

### 2023-02-01 paper.tex
Edits to section 2.4.

@@
 \subsection{System Granularity Scalings}
-A granularity parameter ($\alpha$) was developed that linearly re-streches the dynamics of the system to occur over at a different range of protein while maintaining the general behavior of the model.
-Any arbitrary propensity function ($w$) can be rescaled to a different level of discreteness by substituting all rate equations with $w'$ given by
+One of the major computational obstacles in using the CME or FSP formulations to analyze discrete stochastic chemical kinetics is that the dimension of the FSP scales linearly with each chemical species, and therefore exponentially with the number of independent species.
+To circumvent this issue, several previous studies have projected the FSP onto lower dimensional spaces based using various techniques such as time scale separations\cite{Peles2006}, Krylov subspaces\cite{Macnemara&Sidje?}, coarse meshes\cite{Munsky2008IEEE,Tapia2012CDC}, or principle orthogonal decompositions\cite{Vo2019}.
+% XXX - would be good to look for a few other examples outside of our own papers where the CME dimension is reduced by linear projections.
+On the other hand, an experimental obstacle to using CME analyses is that gene expression quantification based on fluorescence intensities (e.g., using fluorescent protein reporters) usually does not provide molecular resolution, but rather measures only relative changes in expression over time.
+
+To address how variations in model or observation scales might impact the effectiveness of noise-based control, we use the controller optimized at one resolution and ask how effective it would be when applied to systems with the same dynamics but with different resolution.
+To accomplish this rescaling, let $M$ denote some metric for the size of the original system (e.g., the average number of proteins at steady state and maximal UV) and let $M'$ denote the same metric for the system at a different resolution.
+Toward this goal, we define a granularity parameter ($\alpha=M'/M$) that linearly scales the species populations to increase ($\alpha>1$) or decrease ($\alpha<1$) the maximum populations size, while maintaining the dynamics and general behavior of the model.
+To apply the granularity parameter, we assume that each propensity function, $w_{\mu}$, from Eqns.\ \ref{Props} is rescaled to a different level of discreteness by substituting
 \begin{equation}
-w'(x)=w(x/\alpha)
+w'(\mathbf{X})=w(\mathbf{X}/\alpha).
 \end{equation}
-where $\alpha=M'/M$, $M$ the size of the original system, and $M'$ the size of the re-scaled system.
-When $\alpha <1 $ the system dynamics occur over a smaller range of protein counts, and when $\alpha >1$ the system dynamics occurs over a larger range of protein counts.
+For example, the production and degradation of protein in cell one would become:
+\begin{equation}
+w'_1(\mathbf{X}) = u(\mathbf{X}/\alpha,t)  + \kappa \frac{(x_1/\alpha)^\eta}{(x_1/\alpha)^{\eta}+\beta^{\eta}} + k_0,\ w'_2(\mathbf{X}) = \gamma x_1/\alpha.
+\end{equation}
+% XXX - Please check this against what you did computationally. 
+After this rescaling, if $\alpha<1 $ the system dynamics will now occur over a smaller range of protein counts, and if $\alpha >1$ the system dynamics occurs over a larger range of protein counts.
 
-The performance metric and controller must also be scaled by $1/\alpha$ before computing the appropriate functions of that state, e.g., ${u}^{\mathcal{C}} = {u}^{\mathcal{C}}(\mathbf{X}/\alpha)$ and 
-{\small \begin{align}
-J &= \sum_{i=0}^{M'}  \sum_{j=0}^{M'}P^*(x_1=i,x_2=j) ((i/\alpha - \mathbf T_1)^2 + (j/\alpha -\mathbf T_2)^2),\nonumber \\
+We note that in transforming the propensity functions, the inputs to the controller have also be scaled by $1/\alpha$ before computing the appropriate functions of that state, e.g., ${u}^{\mathcal{C}} = {u}^{\mathcal{C}}(\mathbf{X}/\alpha)$.
+Because identification of the original control formulation, $\mathbf{u}^{FAC}(x_1,x_2)$ and $\mathbf{u}^{PAC}(x_1)$ only considered integer values for $(x_1,x_2)$, fractional number inputs for after rescaling $(x_1/\alpha,x_2/\alpha)$ are handled by 2D cubic interpolation of the nearest values.
+%\begin{equation}
+%u'=u(x_0 / \alpha, x_1 / \alpha) 
+%\end{equation}
+%\begin{equation}
+%u'=u(\alpha x_0)
+%\end{equation}
+Finally, to provide a consistency relative scoring, the definition of the performance score is also adjusted according to sale magnitudes, i.e.,:
+{\begin{align}
+J' &= \sum_{i=0}^{M'}  \sum_{j=0}^{M'}P^*(x_1=i,x_2=j) ((i/\alpha - \mathbf T_1)^2 + (j/\alpha -\mathbf T_2)^2),\nonumber \\
 & =\mathbf{C}'\mathbf{P}^*
@@
 \end{align}}
-Since $\mathbf{u}^{FAC}$ and $\mathbf{u}^{PAC}$ are both matrices, fractional number inputs after rescaling are handled by 2D cubic interpolation of the nearest values.
-
-\begin{equation}
-u'=u(x_0 / \alpha, x_1 / \alpha) 
-\end{equation}
-\begin{equation}
-u'=u(\alpha x_0)
-\end{equation}
 

### 2023-02-02 paper.tex
Edits up to section 3.1

@@
 In the next section, we briefly describe our methods to formulate the chemical master equation to analyze multiple individual cells in the same spatially homogeneous environment; we define the FAC and PAC controllers and their effects the dynamics of cells within that environment; and we explain how the control law is optimized to achieve desired performance objectives.
-Next, in the `Results' section, we explore how the granularity of the model formulation, the inaccuracy of parametric assumptions, and the inclusion of time delays would affect the FAC and PAC control performance.
+Next, in the `Results' section, we explore how granularity of the model formulation, inaccuracy of parametric assumptions, or introduction of time delays would affect the FAC and PAC control performance.
 Finally, in `Conclusions' we summarize our findings and discuss the implications that these controllers, and their robustness to modeling errors, may have on future systems in synthetic biology.
@@
 
-More specifically, the openloop infinitesimal generator, $\mathbf{A}_0$, is constructed according to:
+More specifically, the open-loop infinitesimal generator, $\mathbf{A}_0$, is constructed according to:
 \begin{equation}
@@
 
-\subsection{System Granularity Scalings}
+\subsection{Scaling for system granularity}
 One of the major computational obstacles in using the CME or FSP formulations to analyze discrete stochastic chemical kinetics is that the dimension of the FSP scales linearly with each chemical species, and therefore exponentially with the number of independent species.
@@
 
-\subsection{Introduction of Observation/Actuation Time Delays}
-A time-delay SSA was developed by recording the piecewise constant history of the system and inputing the current control activation at time $t$ in the following delayed manner
+\subsection{Observation and actuation time delays}
+No real optogenetic control system can provide instantaneous feedback -- there will always be delays in biochemical reactions needed for observations (e.g., transcription and translation dynamics as well as folding and maturation of fluorescent proteins), delays in data analysis and decision making (e.g., image processing and calculation of the corresponding control signals), and delays in the actuation dynamics (e.g., activation of optogenetically excitable molecules).
+To explore the impacts of delays between the time a cellular fluctuation occurs and the time at which the corresponding control action can take effect, we developed a simple time-delay stochastic simulation algorithm.
+In this analysis, the state history is recorded following every stochastic event, allowing for reconstruction of the piecewise constant history of the species' populations, e.g., $x_1(s)$ and $x_2(s)$ for all $s\in [0,t]$.
+Using this information, the time-delayed control signal at time $t$ can be specified as:
 \begin{equation}
@@
 \end{array}\right. 
-\label{scoreeq}
+\label{timeDelaySSA}
 \end{equation}
-where $\tau$ is the time delay between observation and actuation, and ${u}^{\mathcal{C}}$ is the previously optimized control law for $\mathbf{u}^{FAC}$ or $\mathbf{u}^{PAC}$.
-We note that the time delay stochastic process was only simulated using the SSA because an appropriate direct FSP/CME integration procedure is not known.
+where $\tau$ is the time delay between observation and actuation, and ${u}^{\mathcal{C}}$ is the previously optimized control law (e.g., $\mathbf{u}^{\rm FAC}$ or $\mathbf{u}^{\rm PAC}$).
+We note that the time delay stochastic process was only simulated using the SSA because to our knowledge an appropriate direct FSP/CME integration procedure has not yet been developed.
 
-\subsection{Parameter Scalings and Sensitivity}
-Parameter sensitivities of the system can be performed by calculating the control performance score ($J$), over a range of parameters which are scaled linearly with a scaling parameter ($s$) while holding all other parameters constant.
-Any new arbitrary parameter $\lambda\in \{ \kappa,\beta,k_0,\eta,\gamma\}$ can be rescaled from an old one ($\lambda$) according to
-\begin{equation}
-\lambda' = s \lambda
-\end{equation}
-It should be noted that each cell receives its own set of parameters, and therefore the number of parameters double during sensitivity analysis of a two cell control process.
-Two types of sensitivity analysis can be performed, one which performs the analysis on one cell at a time while keeping the other constant, and another which scales similar parameters in both cells at the same time.
-Steady state distributions of the perrtubed models were solved using the FSP.
+%\subsection{Parameter Scalings and Sensitivity}
+%Finally, to explore the sensitivity of the control performance to inaccuracies in model parameters for one or more cells, we calculate the control performance score ($J$), over a range for each parameter scaled linearly with a scaling parameter ($\zeta$) while holding all other parameters constant.
+%%Any new arbitrary parameter $\lambda\in \{ \kappa,\beta,k_0,\eta,\gamma\}$ can be rescaled from an old one ($\lambda$) according to
+%%\begin{equation}
+%%\lambda' = s \lambda
+%%\end{equation}
+%We explore two types of sensitivity analysis: one which performs the analysis on one cell at a time while keeping the other constant, and another which scales similar parameters in both cells at the same time.
+%%In the former case, it should be noted that each cell receives its own set of parameters, so that the number of parameters double during sensitivity analysis of a two cell control process. 
+%In either case, steady state distributions of the perturbed models are solved using the FSP.
+% XXX - I don't think that we need this section in Methods -- but perhaps some of the text should be moved to the results if needed.
+
 \section{Results}
-\subsection{Effects of Changes to Granularity}
+
+To assess the potential robustness of noise-enhanced control of the proposed synthetic auto-repression toggle switch, we adopt the model and  parameters from \cite{May2021} as well as the fully aware and partially aware controllers, $u^{\rm FAC}(x_1,x_2)$ and  $u^{\rm FAC}(x_1)$, respectively as described in the methods section.
+In the following subsections, we explore how well these controllers would perform in the more realistic setting where the approximate model used to define the controller no longer matches exactly to the dynamics of the process that the controller is being used to manipulate.
+
+We first explore how the model defined for a given assumed system size (i.e., in which protein numbers have a range of $M$) when applied to a system with larger or smaller systems sizes (i.e., where protein numbers now have a range of $M' = \alpha M$.
+This modification could arise from two situations related to numerical convenience or inexact resolution in the measurement of concentrations.
+First, from the perspective of numerical convenience when optimizing a controller, it can be advantageous to reduce the assumed range of protein levels in order to reduce the dimension of the CME and FSP analyses, thereby achieving a more tractable optimization problem.
+For example, the current model (see Methods) takes XXX seconds to solve the FSP analysis when $\alpha=1$ but the same analysis takes XXX seconds to solve when $\alpha=10$.
+Since numerical optimization of the controller can take many millions of calculation for different control laws, learning a controller using $\alpha=1$ that works for $\alpha=10$ would yield substantial savings in computational effort.
+The second situation under which sensitivity to system scaling could become important is when measurements are relative and do not provide single-molecule resolution.
+For example, in measurements like those on \cite{Baumschlager} which are based on fluorescence protein intensities, it is not clear how many proteins correspond to what level of fluorescence intensity, and the scalar $\alpha$ is an unknown quantity that must be estimated (e.g., through calibration with single-molecule measurements or dilution experiment\cite{XXX - look for elowitz paper or something similar that tried to calibrate for single-molecule counts}).
+
+It is well established that the relative size of stochastic fluctuations (i.e., noise) in a chemical process decreases with the inverse square root of the process scale \cite{XXX van kampen?
+Gillespei look for this}.
+As the system size increases (with all species being scaled by the same factor), the dynamics approach those of a deterministic process with probability one (with zero-measure exceptions only for initial conditions lying on manifolds that separate attractors for different steady state behaviors)\cite{XXX}.
+For the purpose of noise-induced control, this introduces a tradeoff where noise is critical to break symmetry and enable differential control, yet noise is also deleterious to maintaining desired states once they have been achieved.
+Considering these competing objectives, it is not immediately clear how the reduction of noise through the addition of system granularity will affect control performance.
+
+To shed light on this tradeoff, control performance scores for the $\mathbf{u}^{\rm FAC}$ and $\mathbf{u}^{\rm PAC}$ controllers were calculated at different levels of granularity ($\alpha$) between 0.1 and 2.0.
+% XXX - Why only go up to 2?  Did it become intractable?
+Figure \ref{Volume} (Ai - Fi) shows the joint (left plots) and marginal (right plots) distributions of the two cells relative to the specified target position (circles in joint distributions and vertical dashed lines in marginal distributions) for the FAC controller and for $\mathbf{\alpha}$=.2 (A,B), 1 (C,D) and 2 (E,F).
+As the granularity changes from $\alpha$=0.2 , to $\alpha$=2, the distributions become focused at a point near the target, and the performance score improves from $J=$333 to 85 to 27, respectively.
+The PAC controller performance also improves considerably with alpha as shown in Fig.\ \ref{Volume} (G-L).
+% XXX - Please change the panel labels to A-F, G-L, and M.
+
+\subsection{Effects of changes to system granularity}
 \begin{figure*}[t!]
@@
 \caption{Effects of Scale Changes.
-(Ai-Fi) Joint (Ai,Ci,Di) and marginal (Bi,Di,Fi) probability distributions for the FAC control method at three different $\alpha$.
-$\alpha=0.2$ (top), $\alpha=1$ (middle) and $\alpha=2.0$ (bottom) relative to the target point (blue circles).
-(Aii-Fii) Same as (Ai-Fi), but for PAC controller.
-(G)  FAC (cyan) and PAC (magenta) controller performance as function of system volume.
-Dashed lines correspond to SSA analyses at $\alpha=100$.
-In all analyses, the system model and PAC and FAC controllers are unchanged from \cite{May2021} other than their adaptation to the new system volume size.}
+(A-F) Joint (A,C,D,) and marginal (B,D,F) probability distributions for the FAC control method at $\alpha=0.2$ (A,B), $\alpha=1$ (C,D) and $\alpha=2.0$ (E,F).
+Target point shown as blue circle.
+(G-L) Same as (A-F), but for PAC controller.
+(M)  FAC (cyan) and PAC (magenta) controller performance as function of system volume.
+Horizontal dashed lines correspond to SSA analyses at $\alpha=100$.
+In all analyses, the system model and PAC and FAC controllers are unchanged from \cite{May2021} other than their adaptation to the new system volume size (see Methods).}
 \label{Volume}
@@
 \end{figure*}
-Despite noise being a critical component of noise-exploiting control, the removal of noise through the addition of system granularity is likely beneficial to control performance.
-This two-edged characteristic that noise is critically important yet bad was explored by observing how control performance changes with system granularity, a parameter that is known to make systems behave more deterministically.
 
-Control performance scores for the $\mathbf{u}^{FAC}$ and $\mathbf{u}^{PAC}$ controllers were calculated at different levels of granularity ($\alpha$) between 0.1 and 2.0.
-Figures.\ \ref{Volume} (Ai - Fi) show the joint (left plots) and marginal (right plots) distributions of the two cells relative to the specified target position (circles in joint distributions and vertical dashed lines in marginal distributions)
-for the FAC controller and for $\mathbf{\alpha}$=.2 (A,B), 1 (C,D) and 2 (E,F).
-As the granularity changes from $\alpha$=0.2 , to $\alpha$=2, the distributions become focused at a point near the target, and the performance score improves from $J=$333 to 85 to 27, respectively.
-The PAC controller performance also improves considerably with alpha as shown in figures.
-\ref{Volume} (Aii - Fii).
+Figure \ref{Volume}M shows the trend of the performance score versus alpha for both the FAC (solid cyan line) and the PAC (solid magenta line) controller.
+This improvement in performance appears to approach a small value as the granularity goes to infinity, but since the size of the FSP increases with the square of the system size, systems much larger than $\alpha$=2 (where $\mathbf{A}_0\in \mathbb{R}^{10^4\times10^4}$) become more difficult to calculate.
+% XXX - I would think you should be able to compute larger values of 10^6.  They would be slow, but should be doable in a few minutes per calculation.
+To bypass this limit in the FSP, sixteen SSA simulations were used to sample the CME of a system with a much larger volume of $\alpha$=100.
+Each SSA was ran for $5\times10^7$ minutes and only the last $4\times10^7$ minutes were sampled to estimate the stationary distribution and the calculate the performance score.
+The performance score estimates of this high granularity SSA using the FAC and PAC were 4.13 and 27.5 respectively, which are plotted as dashed lines in Fig.\ \ref{Volume}G.
+Although it is unclear if further performance improvements could be obtained with further increases to the system scale, for all cases considered so far, we found that both controllers monotonically improved with increased $\alpha$ and that $\mathbf{u}^{\rm FAC}$ always outperforms the $\mathbf{u}^{\rm PAC}$.
 
-Figure \ \ref{Volume}G shows the trend of the performance score versus alpha for both the FAC (solid cyan line) and the PAC (solid magenta line) controller.
-This improvement in performance appears to approach a small value as the granularity goes to infinity, but since the size of the FSP increases with the square of the system size, systems much larger than $\alpha$=2 (where $\mathbf{A}_0\in \mathbb{R}^{10^4\times10^4}$) are difficult to calculate.
-To bypass this limit in the FSP, sixteen SSA simulations were used to sample the CME of a system with a much larger volume of $\alpha$=100.
-Each SSA was ran for 5e7 minutes and only the last 4e7 minutes were sampled from.
-The performance score estimates of this high granularity SSA using the FAC and PAC were 4.13 and 27.5 respectively, which are plotted as dashed lines in figure \ref{Volume}G.
-We note however, that systems with larger granularity take longer to reach a steady state distribution, and it is unclear if performance improvements can be obtained with further increases to the system volume.
-However, for all volumes considered, we found that both controllers monotonically improved with increased volume and that $\mathbf{u}^{FAC}$ always outperforms the $\mathbf{u}^{PAC}$ at any system volume.
+We note however, that the improved steady state performance for systems with larger scales, comes at a cost that they also take longer to reach a steady state distribution.
+This is illustrated in Fig.\ \ref{XXX} which shows the time needed to adapt to a change in the target point $\mathbf{T} = [XXX,XXX]$ to its mirror point $\mathbf{T}_m = [XXX,XXX]$.
+% XXX - Please add a new figure to show this point.
+Start with an initial condition that is the steady state with target T and then flip it to T' = T(::-1).
+Then solve for P(t) and calculated J(t) from this.
+Finally, make a plot of J(t) versus t for the three cases \alpha = 0.2, 1, 2.
+ From the figure, it is clear that the smaller systems XXX.
+ % XXX - update this to explain what you see in this new figure.
 
-These data suggest that increasing granularity increases steady state control performance even if the controller itself depends on noise for its implementation.
+%These data suggest that increasing granularity increases steady state control performance even if the controller itself depends on noise for its implementation.
 Also, designing a controller to work for a system of one volume size can result in a controller that works even for another much larger volume.
@@
 
-Time delay was hypothesized to adversely affect control performance of the $\mathbf{u}^{FAC}$ and $\mathbf{u}^{PAC}$ controllers.
+Time delay was hypothesized to adversely affect control performance of the $\mathbf{u}^{\rm FAC}$ and $\mathbf{u}^{\rm PAC}$ controllers.
 To test the effect of time delay on control performance, groups of time delayed SSA were simulated while changing controller and the level of time delay between each group.
@@
 Each column show the effect of changing a single model parameter (e.g., $\kappa$ in the leftmost row).
-Trends for the $FAC$ (solid lines) and PAC (dashed lines), but the optimal controller can change under parameter variations.
+Trends for the FAC (solid lines) and PAC (dashed lines), but the optimal controller can change under parameter variations.
 In all analyses, the system model and PAC and FAC controllers are unchanged from \cite{May2021}.}
@@
 To test this, parameter sensitivity analysis was performed on the cell one while holding the parameters of cell two constant figure \ref{Parameter}(Row I).
-For each perturbation analysis, both the $\mathbf{u}^{FAC}$ and $\mathbf{u}^{FAC}$ controller was used over a range of $s$ between 0.5 and 2.
+For each perturbation analysis, both the $\mathbf{u}^{\rm FAC}$ and $\mathbf{u}^{\rm PAC}$ controller was used over a range of $s$ between 0.5 and 2.
 The data for such perturbations are shown in figure \ref{Parameter}.
@@
 When parameters of both Cell 1 and Cell 2 are jointly changed (Fig.\ \ref{Parameter} bottom row) some unique trends were observed.
-It was found that the $\mathbf{u}^{FAC}$ often outperformed the $\mathbf{u}^{PAC}$, but for large $s$ in a few parameters the inverse was found (figure \ref{parameter} (Row III, E and At).
+It was found that the $\mathbf{u}^{\rm FAC}$ often outperformed the $\mathbf{u}^{\rm PAC}$, but for large $s$ in a few parameters the inverse was found (figure \ref{parameter} (Row III, E and At).
 
-When examining the effects of changing parameters when using the $\mathbf{u}^{FAC}$ (solid lines) or the $\mathbf{u}^{PAC}$ (dashed lines), general trends typically remained the same in that changes to a parameter which caused a decrease in the control performance of the FAC also tended to decrease the performance of the PAC.
+When examining the effects of changing parameters when using the $\mathbf{u}^{\rm FAC}$ (solid lines) or the $\mathbf{u}^{\rm PAC}$ (dashed lines), general trends typically remained the same in that changes to a parameter which caused a decrease in the control performance of the FAC also tended to decrease the performance of the PAC.
 Some parameters (e.g., $\kappa$, $\gamma$) reached minima that performed better than the original, which suggests that the physical system itself has room for optimization beyond the original design that could lead to better control performance, {\em even without changing the control law}.
 One explanation for why changes to $\eta$ had no effect on control performance is that the origonal parameter was so small that it may require a larger scaling on a different magnitude before changes are observed.
-For situations where the $\mathbf{u}^{PAC}$ outperformed the $\mathbf{u}^{FAC}$ at large $s$, this may be an indication that re-optimizing the controller for this parameter combination could yield better control results, since the controller is not re-optimized at each parameter combination.
+For situations where the $\mathbf{u}^{\rm PAC}$ outperformed the $\mathbf{u}^{\rm FAC}$ at large $s$, this may be an indication that re-optimizing the controller for this parameter combination could yield better control results, since the controller is not re-optimized at each parameter combination.
 

### 2023-02-02 paper.tex
Minore change to Section 3.1.

@@
 We note however, that the improved steady state performance for systems with larger scales, comes at a cost that they also take longer to reach a steady state distribution.
-This is illustrated in Fig.\ \ref{XXX} which shows the time needed to adapt to a change in the target point $\mathbf{T} = [XXX,XXX]$ to its mirror point $\mathbf{T}_m = [XXX,XXX]$.
+This tradeoff is illustrated in Fig.\ \ref{XXX} which shows the performance score $J(t)$ as the system adapts to a change from the original target point $\mathbf{T} = [XXX,XXX]$ to its mirror point $\mathbf{T}_m = [XXX,XXX]$.
 % XXX - Please add a new figure to show this point.

### 2023-02-02 paper.tex
Edits to results section.

@@
 \end{figure*}
+% XXX - This figure needs larger fonts for all the labels and axes. 
+% Update panel letters as written in caption
+% Add tic marks to left side of the A and C heat maps
 
@@
 \begin{center}
+\includegraphics[width=1\textwidth]{ParameterPerturbation.png}
+\caption{Effects of parametric variations on control performance.
+({\bf A}) Sensitivity of control performance score, $J$, to modifications in parameters applied to Cell 1 only, one at a time for parameters $\kappa$, XXX, XXX, XXX, and XXX (in order left to right).
+({\bf B}) Same as {\bf A}, but where modifications are made to the corresponding parameter in Cell 2.
+({\bf C}) Same as ({\bf A}), but where the modification is made simultaneously to both cells.
+In all panels, trends lines are shown for the FAC (solid) and PAC (dashed) controllers.
+Controllers are unchanged from \cite{May2021}.}
+\label{Parameter}
+\end{center}
+\end{figure*}
+% XXX - This figure needs much larger fonts for all the labels and axes.
+% There are errors in the parameter names.  What is 'alpha'?
+% remove the A,B,C,D,E.
+% replace I, II, III with A, B, C
+% what is parameter $\mu$? Is that an error?  Please use $\mu$ for reaction index as in the methods section.
+
+\subsection{Effects of Parameter Errors or Extrinsic Uncertainties}
+
+Exact parameters are never known for any real biological process, and even genetically identical single-cell systems are expected to exhibit cell-to-cell heterogeneity in parameters due to extrinsic variations outside of the specific pathways of interest.
+These parameter uncertainties or variations between cells could influence system performance, such that a control law that is optimized for one fixed, but incorrect, set of parameters may or may not succeed if used to break symmetry and track a desired target state when applied to a system with different parameters.
+To explore the sensitivity of control performance to unknown errors in the model parameters, sensitivity analysis was performed for each model parameter and for the situation where the corresponding parameter was modified in either one cell at a time (i.e., to simulate extrinsic noise) or in both cells simultaneously (i.e., to simulate parameter errors).
+For each perturbation analysis, both the $\mathbf{u}^{\rm FAC}$ and $\mathbf{u}^{\rm PAC}$ controller were analyzed over a range of parameters from one-half to two-fold of the original parameters.
+% XXX - How hard would it be to extend this to 10 fold in either direction?
+ Figure \ref{Parameter}A shows the results for these parameter sweeps when a single parameter in Cell 1 is change while the parameters of Cell 2 is held fixed, and Fig.\ \ref{Parameter}B shows the opposite when changes Cell 1 has fixed parameters and Cell 2 is modified.
+Finally, \ref{Parameter}B  shows the performance change when the corresponding parameter is changed in both Cell 1 and Cell 2 at the same time.
+
+Modification to parameters was found to produce a broad range of effects, where some changes increase performance, some decrease performance, and others have little or no effect.
+For example, increasing $\beta$ in Cell 1 worsens performance while increasing $\beta$ in Cell 2 improves performance (compare second column in Fig.\ \ref{Parameter}A and \ref{Parameter}B).
+In some cases, these effects were not monotonic; for example, increasing $\kappa$ in Cell 1 (Fig.\ \ref{Parameter}A, left column) would be highly advantageous up to a limit after which the control performance degrades rapidly.
+In other cases (such as for $k_0$), the effect of parameter perturbations on performance is insignificant even for relatively large ($s$=2) perturbations.Similarly, Fig.\ \ref{Parameter}C shows that when parameters of both Cell 1 and Cell 2 are jointly changed, these changes could also improve or detract from control performance.
+%It was found that the $\mathbf{u}^{\rm FAC}$ often outperformed the $\mathbf{u}^{\rm PAC}$, but for large $s$ in a few parameters the inverse was found (figure \ref{parameter} (Row III, E and At).
+
+When examining the effects of changing parameters when using the $\mathbf{u}^{\rm FAC}$ (solid lines) or the $\mathbf{u}^{\rm PAC}$ (dashed lines), general trends typically remained the same in that changes to a parameter which cause a decrease in the control performance of the FAC also tend to decrease the performance of the PAC.
+Some parameters (e.g., $\kappa$, $\gamma$) reached minima that performed better than the original, which suggests that the physical system itself has room for optimization beyond the original design that could lead to better control performance, {\em even without changing the control law}.
+One explanation for why changes to $\eta$ had no effect on control performance is that the original parameter was so small that it may require a larger scaling on a different magnitude before changes are observed.
+% XXX - what is \eta?  From the table this should be the cooperativity with is very large.  Is this a mistake?
+In a few circumstances, changes in the parameters $\kappa$ or $\gamma$ led to the surprising situation where the PAC controller, which only observes one cell at a time, was more robust to parameter errors than was the FAC controller, which can observe both cells, even to the extent that the PAC controller could sometimes outperform the FAC.
+We stress that this effect is due to the mismatch between the assumed and the actual parameters, and re-optimizing the two controllers for the new parameter combination will always yield better control results for the FAC compared to the PAC.
+
+\subsection{Effects of Time Delays on Controller Performance}
+
+\begin{figure*}
+\begin{center}
 \includegraphics[width=1\textwidth]{TimeDelayPerturbation.png}
 \vspace{-0.1in}
-\caption{Effects of time delays.
-(Ai-Fi) Joint (Ai,Ci,Di) and marginal (Bi,Di,Fi) probability distributions for the FAC control method at three different time delays of $\tau/\gamma$=0.01 (top), 0.1 (middle) and 1 (bottom) relative to the target point (blue circles).
-(Aii-Fii) Same as (Ai-Fi), but for PAC controller.
-(G) FAC PAC performance as function of time delay show that PAC outperforms the FAC when time delay is greater than $0.07 /\gamma$.
-In all analyses, the system model and PAC and FAC controllers are unchanged from \cite{May2021}.}
+\caption{Effects of time delays on control performance.
+(A-F) Joint (A,C,E) and marginal (B,D,F) probability distributions for the FAC control method at three different time delays of $\tau/\gamma$=0.01 (A,B), 0.1 (C,D) and 1.0 (E,F) relative to the target point (blue circles).
+(G-L) Same as (A-F), but for PAC controller.
+(M) FAC and PAC performance as function of time delay show that PAC outperforms the FAC when time delay is greater than $0.07 /\gamma$.
+In all analyses, the system model and PAC and FAC controllers are unchanged from \cite{May2021}; specifically, the protein degradation rate is $\gamma = 0.0203$ as identified from \cite{Baumchager}.}
 \label{Time}
@@
 \end{figure*}
+% XXX - This figure needs larger fonts for all the labels and axes. 
+% Update panel letters as written in caption
+% Add tic marks to left side of the A and C heat maps
+% Fix the ref to baumschlager paper.
 
-\subsection{Effects of Time Delays on Controller Performance}
+Feedback control can only be effective if one can quickly make measurements, compute adjustment to the control signal, and implement the needed change to the system.
+As the time required for any of these steps increases, control performance will be degraded, perhaps even leading to large fluctuations or instability.
+To explore how time delays would affect the noise-enhance controller $\mathbf{u}^{\rm FAC}$ and $\mathbf{u}^{\rm PAC}$, we generated large sets of time-delayed stochastic simulations (see Methods) for different lengths of the time delay.
+Each SSA was sub-sampled for 1000 times over 10000 minutes of simulation time after a burnin period of 10000 minutes.
 
-Time delay was hypothesized to adversely affect control performance of the $\mathbf{u}^{\rm FAC}$ and $\mathbf{u}^{\rm PAC}$ controllers.
-To test the effect of time delay on control performance, groups of time delayed SSA were simulated while changing controller and the level of time delay between each group.
-Each SSA was sampled for 1000 times over 10000 minutes after reaching a steady state by simulating the system for 10000 minutes.
+Figure \ref{Time} shows the joint and marginal distribution of each controller and time-delay pair and the resulting score, with panels A-F showing results for the FAC controller and panels G-L showing results for the PAC controller.
+Figure \ref{Time}M summarizes these results by plotting the score of both controllers versus the time delay.
+From the figures, it is clear that performance is rapidly degraded as the delay approaches and then exceeds the characteristic time scale of the process.
+At ver small time delays (below $\tau = 0.07/\gamma = XXX$ min), the FAC outperforms the PAC ($J=$87 versus 146 at $\tau = 0.01/\gamma = XXX$ min) but at moderate time delays (above $\tau = 0.07/\gamma = XXX$ min) the PAC outperforms the FAC ($J=$170 versus 219 at $\tau = 0.1/\gamma = XXX$ min).
+% XXX - you can use the value of gamma=0.203 in Table 1 to write these in units of minutes.
 
-Figure\ \ref{Time} shows the joint and marginal of each controller and time-delay pair and the resulting score, with panels Ai - Fi showing results for the FAC controller and panels Aii-Fii showing results for the PAC controller, and panel G showing the score of both controllers versus the time delay.
-From the figures, it is clear that performance is rapidly degraded as the delay approaches and then exceeds the characteristic time scale of the process.
-At low time delays (below $\tau = 0.07/\gamma$), the FAC outperforms the PAC ($J=$87 versus 146 at $\tau = 0.01/\gamma$) but at moderate time delays (above $\tau = 0.07/\gamma$) the PAC outperforms the FAC ($J=$170 versus 219 at $\tau = 0.1/\gamma$)).
-
-This data taken together show that time delays much larger that $0.07 / \gamma$ is detrimental to the FAC controller; delays beyond $0.2 / \gamma$ are detrimental for the PAC controller; and the best choice in controller is dependent the level of time delay in the system.
+This data taken together show that time much larger that $0.07 / \gamma = XXX min$ is detrimental to the FAC controller; delays beyond $0.2 / \gamma = XXX min$ are detrimental for the PAC controller; and the best choice in controller is dependent the level of time delay in the system.
 For extreme levels of time delay, both systems lose their asymmetry, and their scores become much worse ($J=$1041 and 931).
-
-\begin{figure*}
-\begin{center}
-\includegraphics[width=1\textwidth]{ParameterPerturbation.png}
-\caption{Effects of parametric variations to Cell 1 (top row), Cell 2 (middle row) or to both cells simultaneously (bottom row).
-Each column show the effect of changing a single model parameter (e.g., $\kappa$ in the leftmost row).
-Trends for the FAC (solid lines) and PAC (dashed lines), but the optimal controller can change under parameter variations.
-In all analyses, the system model and PAC and FAC controllers are unchanged from \cite{May2021}.}
-\label{Parameter}
-\end{center}
-\end{figure*}
-
-\subsection{Effects of Parameter Errors or Extrinsic Uncertainties}
-
-It is rare that any parameters are exactly known for biological processes, and real systems often exhibit slightly different parameters due to extrinsic variations.
-This parameter variation between cells is factor that could influence control performance.
-To test this, parameter sensitivity analysis was performed on the cell one while holding the parameters of cell two constant figure \ref{Parameter}(Row I).
-For each perturbation analysis, both the $\mathbf{u}^{\rm FAC}$ and $\mathbf{u}^{\rm PAC}$ controller was used over a range of $s$ between 0.5 and 2.
-The data for such perturbations are shown in figure \ref{Parameter}.
-The analysis was then repeated, this time holding the parameters of Cell 1 constant and perturbing only the parameters of Cell 2  (shown in figure\ \ref{Parameter}(Row II)).
-Finally, both cell one and cell two had their parameters jointly perturbed by the same amount (figure \ref{Parameter}(Row III)).
-
-Parameter changes were found to have a broad range of effects, including increasing performance, decreasing performance, or having little effect.
-Figure \ref{Parameter}(Bi) shows that increasing $\beta$ in Cell 1 worsens performance while increasing $\beta$ in Cell 2 improves performance.
-In some cases, these effects were not monotonic; for example increasing $\kappa$ in Cell 1 is highly advantageous up to a limit after which the control performance degrades rapidly.
-In other cases (such as for $k_0$), the effect of parameter perturbations on performance is insignificant even for relatively large ($s$=2) perturbations.
-When parameters of both Cell 1 and Cell 2 are jointly changed (Fig.\ \ref{Parameter} bottom row) some unique trends were observed.
-It was found that the $\mathbf{u}^{\rm FAC}$ often outperformed the $\mathbf{u}^{\rm PAC}$, but for large $s$ in a few parameters the inverse was found (figure \ref{parameter} (Row III, E and At).
-
-When examining the effects of changing parameters when using the $\mathbf{u}^{\rm FAC}$ (solid lines) or the $\mathbf{u}^{\rm PAC}$ (dashed lines), general trends typically remained the same in that changes to a parameter which caused a decrease in the control performance of the FAC also tended to decrease the performance of the PAC.
-Some parameters (e.g., $\kappa$, $\gamma$) reached minima that performed better than the original, which suggests that the physical system itself has room for optimization beyond the original design that could lead to better control performance, {\em even without changing the control law}.
-One explanation for why changes to $\eta$ had no effect on control performance is that the origonal parameter was so small that it may require a larger scaling on a different magnitude before changes are observed.
-For situations where the $\mathbf{u}^{\rm PAC}$ outperformed the $\mathbf{u}^{\rm FAC}$ at large $s$, this may be an indication that re-optimizing the controller for this parameter combination could yield better control results, since the controller is not re-optimized at each parameter combination.
 

### 2023-02-02 paper.tex
minor changes.

@@
 Parameter perturbation analysis showed that there is room to improve control performance by adjusting system parameters, and that joint optimization of the controller with the system parameters may lead to better control performance.
-It also shows that populations of cells where each cell mightt have its own parameter combination could still be reasonably controlled for small changes in parameter scale, but large changes could be detrimental.
+It also shows that populations of cells where each cell might have its own parameter combination could still be reasonably controlled for small changes in parameter scale, but large changes could be detrimental.
 

### 2023-02-09 paper.tex
updated

@@
 \begin{center}
-\includegraphics[width=\columnwidth]{CartoonAndControler.png}
+\includegraphics[width=\columnwidth]{CartoonAndControler.pdf}
 \caption{{\bf Optogenetic control of multiple cells using a single light course}
@@
 (4) Change title of panel C to "Feedback control law and resulting performance"
+% MMM - Made the edit
 
@@
 % XXX - I don't think that this is right.  Only the parameters u(UV) and gamma were fit to data.
+% MMM - yes that is right
 are presented in Table 1~\cite{May2021}.
@@
 I changed them to x1 and x2 here, but please chekc to make sure that this is sonsistent throughout text and figures.
+% MMM - Made the change to x1 x2
 where $x_1$ and $x_2$ are discrete integers greater than or equal to zero that represent the number of proteins in cell one and cell two.
@@
 First, from the perspective of numerical convenience when optimizing a controller, it can be advantageous to reduce the assumed range of protein levels in order to reduce the dimension of the CME and FSP analyses, thereby achieving a more tractable optimization problem.
-For example, the current model (see Methods) takes XXX seconds to solve the FSP analysis when $\alpha=1$ but the same analysis takes XXX seconds to solve when $\alpha=10$.
+For example, the current model (see Methods) takes 19.3 seconds to solve the FSP analysis when $\alpha=1$ but the same analysis takes 24619 seconds to solve when $\alpha=10$.
 Since numerical optimization of the controller can take many millions of calculation for different control laws, learning a controller using $\alpha=1$ that works for $\alpha=10$ would yield substantial savings in computational effort.
@@
 % XXX - Why only go up to 2?  Did it become intractable?
+% MMM - Non_Sparse Calculations are intractible after 2, but sparse versions of code will work fine
 Figure \ref{Volume} (Ai - Fi) shows the joint (left plots) and marginal (right plots) distributions of the two cells relative to the specified target position (circles in joint distributions and vertical dashed lines in marginal distributions) for the FAC controller and for $\mathbf{\alpha}$=.2 (A,B), 1 (C,D) and 2 (E,F).
@@
 % XXX - Please change the panel labels to A-F, G-L, and M.
-
+% MMM - made the change
 \subsection{Effects of changes to system granularity}
@@
 \begin{center}
-\includegraphics[width=1\textwidth]{GranularityPerturbation.png}
+\includegraphics[width=1\textwidth]{GranularityPerturbation.eps}
 \vspace{-0.1in}
@@
 % XXX - This figure needs larger fonts for all the labels and axes. 
+% MMM - made the change
 % Update panel letters as written in caption
@@
 % XXX - I would think you should be able to compute larger values of 10^6.  They would be slow, but should be doable in a few minutes per calculation.
+% MMM - constructing the FSP takes longer than solving it
 To bypass this limit in the FSP, sixteen SSA simulations were used to sample the CME of a system with a much larger volume of $\alpha$=100.
@@
 
+\begin{figure*}
+\begin{center}
+\includegraphics[width=1\textwidth]{GranularityScore.eps}
+\caption{}
+\label{}
+\end{center}
+\end{figure*}
+
+\begin{figure*}
+\begin{center}
+\includegraphics[width=1\textwidth]{CalculationSpeed.eps}
+\caption{}
+\label{Speed}
+\end{center}
+\end{figure*}
+
 We note however, that the improved steady state performance for systems with larger scales, comes at a cost that they also take longer to reach a steady state distribution.
-This tradeoff is illustrated in Fig.\ \ref{XXX} which shows the performance score $J(t)$ as the system adapts to a change from the original target point $\mathbf{T} = [XXX,XXX]$ to its mirror point $\mathbf{T}_m = [XXX,XXX]$.
+This tradeoff is illustrated in Fig.\ \ref{Score} which shows the performance score $J(t)$ as the system adapts to a change from the original target point $\mathbf{T} = [30,10]$ to its mirror point $\mathbf{T}_m = [10,30]$.
 % XXX - Please add a new figure to show this point.
@@
 Finally, make a plot of J(t) versus t for the three cases \alpha = 0.2, 1, 2.
- From the figure, it is clear that the smaller systems XXX.
+One hundred SSA simulations of the flip-mirroring process were simulated while measuring the average score all simulations over time.
+Fig.\ \ref{Score} (A and B) shows the average score and protien counts of the simulation when $\alpha=1$, Fig.
+\ \ref{Score} (C,D) when $\alpha = 10$, and Fig.
+\ \ref{Score} (E,F) when $\alpha = 100$.
+
+From the figures it is clear that the smaller $\alpha$ creates faster dynamic but also have higher scores due to inherit noise.
+When $\alpha = 10$, the system becomes much less noisy and have lower scores but takes some time to respond.
+When $\alpha$ is too large at 10, the system contains very small levels of noise, but the dynamics
  % XXX - update this to explain what you see in this new figure.
@@
 \begin{center}
-\includegraphics[width=1\textwidth]{ParameterPerturbation.png}
+\includegraphics[width=1\textwidth]{ParameterPerturbation.pdf}
 \caption{Effects of parametric variations on control performance.
-({\bf A}) Sensitivity of control performance score, $J$, to modifications in parameters applied to Cell 1 only, one at a time for parameters $\kappa$, XXX, XXX, XXX, and XXX (in order left to right).
+({\bf A}) Sensitivity of control performance score, $J$, to modifications in parameters applied to Cell 1 only, one at a time for parameters $\kappa$, $\beta$, $\eta$, $k_0$, and $\gamma$ (in order left to right).
 ({\bf B}) Same as {\bf A}, but where modifications are made to the corresponding parameter in Cell 2.
@@
 % XXX - This figure needs much larger fonts for all the labels and axes.
+% MMM - Did the change
 % There are errors in the parameter names.  What is 'alpha'?
@@
 % XXX - How hard would it be to extend this to 10 fold in either direction?
+% MMM-Changes to parameters can dramatically change the location of the probability distribution which is contained in a bounded (50,50) area.
+A 10 fold increase to a parameter can push all the probability to the top corner of the space
  Figure \ref{Parameter}A shows the results for these parameter sweeps when a single parameter in Cell 1 is change while the parameters of Cell 2 is held fixed, and Fig.\ \ref{Parameter}B shows the opposite when changes Cell 1 has fixed parameters and Cell 2 is modified.
@@
 \begin{center}
-\includegraphics[width=1\textwidth]{TimeDelayPerturbation.png}
+\includegraphics[width=1\textwidth]{TimeDelayPerturbation.eps}
 \vspace{-0.1in}
@@
 % XXX - This figure needs larger fonts for all the labels and axes. 
+% MMM- Did the change
 % Update panel letters as written in caption
@@
 From the figures, it is clear that performance is rapidly degraded as the delay approaches and then exceeds the characteristic time scale of the process.
-At ver small time delays (below $\tau = 0.07/\gamma = XXX$ min), the FAC outperforms the PAC ($J=$87 versus 146 at $\tau = 0.01/\gamma = XXX$ min) but at moderate time delays (above $\tau = 0.07/\gamma = XXX$ min) the PAC outperforms the FAC ($J=$170 versus 219 at $\tau = 0.1/\gamma = XXX$ min).
+At very small time delays (below $\tau = 0.07/\gamma = 0.0142$ min), the FAC outperforms the PAC ($J=$87 versus 146 at $\tau = 0.01/\gamma = 0.00203$ min) but at moderate time delays (above $\tau = 0.07/\gamma = 0.0142$ min) the PAC outperforms the FAC ($J=$170 versus 219 at $\tau = 0.1/\gamma = 0.0203$ min).
 % XXX - you can use the value of gamma=0.203 in Table 1 to write these in units of minutes.
 
-This data taken together show that time much larger that $0.07 / \gamma = XXX min$ is detrimental to the FAC controller; delays beyond $0.2 / \gamma = XXX min$ are detrimental for the PAC controller; and the best choice in controller is dependent the level of time delay in the system.
+This data taken together show that time much larger that $0.07 / \gamma = 0.0142 min$ is detrimental to the FAC controller; delays beyond $0.2 / \gamma = 0.0406 min$ are detrimental for the PAC controller; and the best choice in controller is dependent the level of time delay in the system.
 For extreme levels of time delay, both systems lose their asymmetry, and their scores become much worse ($J=$1041 and 931).
+A no-feedback controller was developed which required no inputs.
+Such a controller is immune to time delay, and received a score $J=402$.
+These data show that at large time delays above , attempting to improve control with feedback is worse than no-feedback.
 

### 2023-02-09 paper.tex
before meeting

@@
 For example, in measurements like those on \cite{Baumschlager} which are based on fluorescence protein intensities, it is not clear how many proteins correspond to what level of fluorescence intensity, and the scalar $\alpha$ is an unknown quantity that must be estimated (e.g., through calibration with single-molecule measurements or dilution experiment\cite{XXX - look for elowitz paper or something similar that tried to calibrate for single-molecule counts}).
-
-It is well established that the relative size of stochastic fluctuations (i.e., noise) in a chemical process decreases with the inverse square root of the process scale \cite{XXX van kampen?
-Gillespei look for this}.
+% MMM - Ding:2019??
+It is well established that the relative size of stochastic fluctuations (i.e., noise) in a chemical process decreases with the inverse square root of the process scale \cite{Kampen1961}.
 As the system size increases (with all species being scaled by the same factor), the dynamics approach those of a deterministic process with probability one (with zero-measure exceptions only for initial conditions lying on manifolds that separate attractors for different steady state behaviors)\cite{XXX}.
@@
 Some parameters (e.g., $\kappa$, $\gamma$) reached minima that performed better than the original, which suggests that the physical system itself has room for optimization beyond the original design that could lead to better control performance, {\em even without changing the control law}.
-One explanation for why changes to $\eta$ had no effect on control performance is that the original parameter was so small that it may require a larger scaling on a different magnitude before changes are observed.
+One explanation for why changes to $k_0$ had no effect on control performance is that the original parameter was so small that it may require a larger scaling on a different magnitude before changes are observed.
 % XXX - what is \eta?  From the table this should be the cooperativity with is very large.  Is this a mistake?
+% MMM - This was k_0, not eta
 In a few circumstances, changes in the parameters $\kappa$ or $\gamma$ led to the surprising situation where the PAC controller, which only observes one cell at a time, was more robust to parameter errors than was the FAC controller, which can observe both cells, even to the extent that the PAC controller could sometimes outperform the FAC.

### 2023-04-06 paper.tex
update

@@
 \includegraphics[width=1\textwidth]{GranularityScore.eps}
-\caption{}
+\caption{Effect of scale changes on control score.
+(A)Time series trajectory of score versus time shows that when the switch occurs at t=0, scores generally increase to a high value above 700, then drop over time.
+The time scale for $\alpha=1.0$ to reach steady state again is approx.
+1000, 2000, and 4000 minutes for $\alpha=[1.0, 2.0, 4.0]$ respectively.
+Additionally, the noise floor of each system drops as a function of alpha.
+(B and C) time series plots of simulated SSA trajectories show that time scale to reach equilibrium changes with $\alpha$, and that an intermediate transition state becomes more visible $\alpha=4.0$.
+The low noise floor seen in (A) can be confimed in the trajectories as well.
+}
 \label{}
-\end{center}
-\end{figure*}
-
-\begin{figure*}
-\begin{center}
-\includegraphics[width=1\textwidth]{CalculationSpeed.eps}
-\caption{}
-\label{Speed}
 \end{center}
@@
 This result implies that it may be possible to optimize controller using coarse-grained FSP analyses for small granularity problems and adapt these via interpolation for use with larger, more realistic systems, that exceed the computational limit of standard FSP computations.
+
+\subsection{Effects of Time Delays on Controller Performance}
+
+\begin{figure*}
+\begin{center}
+\includegraphics[width=1\textwidth]{TimeDelayPerturbation.eps}
+\vspace{-0.1in}
+\caption{Effects of time delays on control performance.
+(A-F) Joint (A,C,E) and marginal (B,D,F) probability distributions for the FAC control method at three different time delays of $\tau/\gamma$=0.01 (A,B), 0.1 (C,D) and 1.0 (E,F) relative to the target point (blue circles).
+(G-L) Same as (A-F), but for PAC controller.
+(M) FAC and PAC performance as function of time delay show that PAC outperforms the FAC when time delay is greater than $0.07 /\gamma$.
+In all analyses, the system model and PAC and FAC controllers are unchanged from \cite{May2021}; specifically, the protein degradation rate is $\gamma = 0.0203$ as identified from \cite{Baumchager}.}
+\label{Time}
+\end{center}
+\vspace{-0.2in}
+\end{figure*}
+% XXX - This figure needs larger fonts for all the labels and axes. 
+% MMM- Did the change
+% Update panel letters as written in caption
+% Add tic marks to left side of the A and C heat maps
+% Fix the ref to baumschlager paper.
+
+Feedback control can only be effective if one can quickly make measurements, compute adjustment to the control signal, and implement the needed change to the system.
+As the time required for any of these steps increases, control performance will be degraded, perhaps even leading to large fluctuations or instability.
+To explore how time delays would affect the noise-enhance controller $\mathbf{u}^{\rm FAC}$ and $\mathbf{u}^{\rm PAC}$, we generated large sets of time-delayed stochastic simulations (see Methods) for different lengths of the time delay.
+Each SSA was sub-sampled for 1000 times over 10000 minutes of simulation time after a burnin period of 10000 minutes.
+
+Figure \ref{Time} shows the joint and marginal distribution of each controller and time-delay pair and the resulting score, with panels A-F showing results for the FAC controller and panels G-L showing results for the PAC controller.
+Figure \ref{Time}M summarizes these results by plotting the score of both controllers versus the time delay.
+From the figures, it is clear that performance is rapidly degraded as the delay approaches and then exceeds the characteristic time scale of the process.
+At very small time delays (below $\tau = 0.07/\gamma = 0.0142$ min), the FAC outperforms the PAC ($J=$87 versus 146 at $\tau = 0.01/\gamma = 0.00203$ min) but at moderate time delays (above $\tau = 0.07/\gamma = 0.0142$ min) the PAC outperforms the FAC ($J=$170 versus 219 at $\tau = 0.1/\gamma = 0.0203$ min).
+% XXX - you can use the value of gamma=0.203 in Table 1 to write these in units of minutes.
+
+This data taken together show that time much larger that $0.07 / \gamma = 0.0142 min$ is detrimental to the FAC controller; delays beyond $0.2 / \gamma = 0.0406 min$ are detrimental for the PAC controller; and the best choice in controller is dependent the level of time delay in the system.
+For extreme levels of time delay, both systems lose their asymmetry, and their scores become much worse ($J=$1041 and 931).
+A no-feedback controller was developed which required no inputs.
+Such a controller is immune to time delay, and received a score $J=402$.
+These data show that at large time delays above , attempting to improve control with feedback is worse than no-feedback.
 
@@
 We stress that this effect is due to the mismatch between the assumed and the actual parameters, and re-optimizing the two controllers for the new parameter combination will always yield better control results for the FAC compared to the PAC.
-
-\subsection{Effects of Time Delays on Controller Performance}
-
-\begin{figure*}
-\begin{center}
-\includegraphics[width=1\textwidth]{TimeDelayPerturbation.eps}
-\vspace{-0.1in}
-\caption{Effects of time delays on control performance.
-(A-F) Joint (A,C,E) and marginal (B,D,F) probability distributions for the FAC control method at three different time delays of $\tau/\gamma$=0.01 (A,B), 0.1 (C,D) and 1.0 (E,F) relative to the target point (blue circles).
-(G-L) Same as (A-F), but for PAC controller.
-(M) FAC and PAC performance as function of time delay show that PAC outperforms the FAC when time delay is greater than $0.07 /\gamma$.
-In all analyses, the system model and PAC and FAC controllers are unchanged from \cite{May2021}; specifically, the protein degradation rate is $\gamma = 0.0203$ as identified from \cite{Baumchager}.}
-\label{Time}
-\end{center}
-\vspace{-0.2in}
-\end{figure*}
-% XXX - This figure needs larger fonts for all the labels and axes. 
-% MMM- Did the change
-% Update panel letters as written in caption
-% Add tic marks to left side of the A and C heat maps
-% Fix the ref to baumschlager paper.
-
-Feedback control can only be effective if one can quickly make measurements, compute adjustment to the control signal, and implement the needed change to the system.
-As the time required for any of these steps increases, control performance will be degraded, perhaps even leading to large fluctuations or instability.
-To explore how time delays would affect the noise-enhance controller $\mathbf{u}^{\rm FAC}$ and $\mathbf{u}^{\rm PAC}$, we generated large sets of time-delayed stochastic simulations (see Methods) for different lengths of the time delay.
-Each SSA was sub-sampled for 1000 times over 10000 minutes of simulation time after a burnin period of 10000 minutes.
-
-Figure \ref{Time} shows the joint and marginal distribution of each controller and time-delay pair and the resulting score, with panels A-F showing results for the FAC controller and panels G-L showing results for the PAC controller.
-Figure \ref{Time}M summarizes these results by plotting the score of both controllers versus the time delay.
-From the figures, it is clear that performance is rapidly degraded as the delay approaches and then exceeds the characteristic time scale of the process.
-At very small time delays (below $\tau = 0.07/\gamma = 0.0142$ min), the FAC outperforms the PAC ($J=$87 versus 146 at $\tau = 0.01/\gamma = 0.00203$ min) but at moderate time delays (above $\tau = 0.07/\gamma = 0.0142$ min) the PAC outperforms the FAC ($J=$170 versus 219 at $\tau = 0.1/\gamma = 0.0203$ min).
-% XXX - you can use the value of gamma=0.203 in Table 1 to write these in units of minutes.
-
-This data taken together show that time much larger that $0.07 / \gamma = 0.0142 min$ is detrimental to the FAC controller; delays beyond $0.2 / \gamma = 0.0406 min$ are detrimental for the PAC controller; and the best choice in controller is dependent the level of time delay in the system.
-For extreme levels of time delay, both systems lose their asymmetry, and their scores become much worse ($J=$1041 and 931).
-A no-feedback controller was developed which required no inputs.
-Such a controller is immune to time delay, and received a score $J=402$.
-These data show that at large time delays above , attempting to improve control with feedback is worse than no-feedback.
 

### 2023-04-26 paper.tex
update

--- 2023-02-09 paper.tex
@@
 \begin{abstract}
-Synthetic biology seeks to create modular components that generate complex and controllable biological behaviors, often at microscopic scales.
-In single-cell applications, where important regulatory RNA or proteins are present in small numbers, intrinsic fluctuations cause biochemical `noise' that frequently results in undesired effects, and this noise is usually treated as a nuisance that must be removed.
+
+Unnderstanding the effects of noise in control theory is important to ()
+Much work has been done to make control performance robust to noise, but the development of controllers which use noise to their advantage is not fully understood.
+Motivated by optogentics, we proposed a difficult control problem which requires the exploitation of noise in a stochastic system to break symmetry between two signals.
+in previous work, we found multiple such controllers which could exploit noise to break symmetry between two cells under a variety of system information.
+This work extends that analysis to include stochastic systems that trace a moving target, are affected by time delay, intrinsic noise, or measurement uncertainty.
+
+Synthetic biology seeks to create modular components that generate complex and controllable biological behaviors, often at microscopic scales where important regulatory RNA or proteins are present in small numbers.
+Intrinsic fluctuations at these scales cause biochemical `noise' that frequently results in undesired neussance effects.
 However, recent theoretical analyses have shown that noise can sometimes be used to one's advantage when seeking to manipulate nonlinear regulatory systems that would be difficult or impossible to control in a fully deterministic setting.
@@
 \includegraphics[width=\columnwidth]{CartoonAndControler.pdf}
-\caption{{\bf Optogenetic control of multiple cells using a single light course}
-({\bf A}) Schematic of the gene expression system.
-UV light activates T7 polymerase based transcription and protein expression from the first promoter-gene construct, and the protein actuates its own production via a second promoter-gene construct.
-({\bf B}) Schematic of the optogenetic control system.
-An observer is partially (PAC) or fully (FAC) aware of cell expression.
-The controller modulates a single spatially homogeneous light signal in attempt to drive each genetically identical cell to its own specified state.
-({\bf C}) The FAC (top) and PAC (bottom) control laws for two cells ($\mathbf{u}^{XAC}(x_1,x_2)$ shown on left) break the symmetry between Cell 1 and Cell 2 and result in distinct marginal distributions for the cells' responses (right).
-Figure adapted from \cite{May2021}.}
+\caption{{\bf Optogenetic control of multiple cells using a single light input.}
+({\bf A}) Schematic of the light activated genetic system with autoregulation.
+({\bf B}) Diagram of the stochastic SIMO control problem using two optogenetic cells sharing an input.
+({\bf C}) Adapted from (PAPER).
+Noise exploiting controllers were optimized to a fully aware control input (I), and partially aware control input(III).
+Marginal distributions are solved using the finite state projection and show a break in symmetry using both controllers (II and IV) }
 \label{cartoons}
@@
 New optogenetic tools have increased our ability to actuate embedded systems within cells reliably and with strong control performance\cite{Sheets2020,Baumschlager2017,Chen2020,Lillacci2018}.
-Great advances in these two fields have enabled computer programmable control of cellular protein production through the use of external optogenetic inputs and smart microscope techniques\cite{Fox2021,Baumschlager2021,Lugagne2017}.
+Advances in these two fields have enabled computer programmable control of cellular protein production through the use of external optogenetic inputs and smart microscope techniques\cite{Fox2021,Baumschlager2021,Lugagne2017}.
 These types of digital-synthetic actuators allow for fine-tuned, computer-modulated control of cellular systems that were previously impossible\cite{Rullan2018, Baumschlager2017} and with fast response times in comparison to chemical diffusion.
@@
 \section{Methods}
-In previous work~\cite{May2021}, a six-species gene regulation model was postulated for the light-activated association of two T7 split domains (species 1 and 2) to form active T7 polymerases (species 3) which could then in turn associate with inactive T7 promoters (species 4) to create an active gene construct (species 5) that would produce its eventual protein product (species 6).
-A second, far simpler single-species model was also postulated by assuming quasi-steady equilibrium for the first five species.
+In previous work~\cite{May2021}, a six-species gene regulation model was developed to describe light-activated association of two T7 split domains (species 1 and 2) to form active T7 polymerases (species 3) which could then in turn associate with inactive T7 promoters (species 4) to create an active gene construct (species 5) that would produce its eventual protein product (species 6).
+A second, far simpler single-species model was also developed by assuming quasi-steady equilibrium for the first five species.
 Both models were independently parametrized by fitting to the same experimental data from Baumschlager, et al.~\cite{XXX}, and then extended to include a secondary self-activated promoter-gene construct for which the expression rate was defined by a Hill function (see Eqn.\ \ref{prodRate} below).
@@
 
-To assess the potential robustness of noise-enhanced control of the proposed synthetic auto-repression toggle switch, we adopt the model and  parameters from \cite{May2021} as well as the fully aware and partially aware controllers, $u^{\rm FAC}(x_1,x_2)$ and  $u^{\rm FAC}(x_1)$, respectively as described in the methods section.
+\subsection{Effects of Parameter Errors or Extrinsic Uncertainties}
+\begin{figure*}
+\begin{center}
+\includegraphics[width=1\textwidth]{ParameterPerturbation.pdf}
+\caption{Parameter sweeps using the FAC and PAC show a broad range of control performance in cell 1 ({\bf A}),  in cell 2 ({\bf B}), and in both cells ({\bf C}). }
+\label{Parameter}
+\end{center}
+\end{figure*}
+
+Exact parameters values are typically unknown for real stochastic process, and even genetically identical single-cell systems can exhibit cell-to-cell heterogeneity in parameters due to extrinsic variations outside of the specific pathways of interest.
+Estimates of parameters can be taken, but not exact measurement.
+These parameter uncertainties could influence system performance.
+To explore the sensitivity of control performance to unknown errors in the model parameters, sensitivity analysis was performed for each model parameter and for the situation where the corresponding parameter was modified in either one cell at a time (i.e., to simulate extrinsic noise) or in both cells simultaneously (i.e., to simulate parameter errors).
+For each perturbation analysis, both the $\mathbf{u}^{\rm FAC}$ and $\mathbf{u}^{\rm PAC}$ controller were analyzed over a range of parameters from one-half to two-fold of the original parameters.
+% XXX - How hard would it be to extend this to 10 fold in either direction?
+% MMM-Changes to parameters can dramatically change the location of the probability distribution which is contained in a bounded (50,50) area.
+A 10 fold increase to a parameter can push all the probability to the top corner of the space
+ Figure \ref{Parameter}A shows the results for these parameter sweeps when a single parameter in Cell 1 is change while the parameters of Cell 2 is held fixed, and Fig.\ \ref{Parameter}B shows the opposite when changes Cell 1 has fixed parameters and Cell 2 is modified.
+Finally, \ref{Parameter}B  shows the performance change when the corresponding parameter is changed in both Cell 1 and Cell 2 at the same time.
+
+Modification to parameters was found to produce a broad range of effects, where some changes increase performance, some decrease performance, and others have little or no effect.
+For example, increasing $\beta$ in Cell 1 worsens performance while increasing $\beta$ in Cell 2 improves performance (compare second column in Fig.\ \ref{Parameter}A and \ref{Parameter}B).
+In some cases, these effects were not monotonic; for example, increasing $\kappa$ in Cell 1 (Fig.\ \ref{Parameter}A, left column) would be highly advantageous up to a limit after which the control performance degrades rapidly.
+In other cases (such as for $k_0$), the effect of parameter perturbations on performance is insignificant even for relatively large ($s$=2) perturbations.Similarly, Fig.\ \ref{Parameter}C shows that when parameters of both Cell 1 and Cell 2 are jointly changed, these changes could also improve or detract from control performance.
+%It was found that the $\mathbf{u}^{\rm FAC}$ often outperformed the $\mathbf{u}^{\rm PAC}$, but for large $s$ in a few parameters the inverse was found (figure \ref{parameter} (Row III, E and At).
+
+When examining the effects of changing parameters when using the $\mathbf{u}^{\rm FAC}$ (solid lines) or the $\mathbf{u}^{\rm PAC}$ (dashed lines), general trends typically remained the same in that changes to a parameter which cause a decrease in the control performance of the FAC also tend to decrease the performance of the PAC.
+Some parameters (e.g., $\kappa$, $\gamma$) reached minima that performed better than the original, which suggests that the physical system itself has room for optimization beyond the original design that could lead to better control performance, {\em even without changing the control law}.
+One explanation for why changes to $k_0$ had no effect on control performance is that the original parameter was so small that it may require a larger scaling on a different magnitude before changes are observed.
+% XXX - what is \eta?  From the table this should be the cooperativity with is very large.  Is this a mistake?
+% MMM - This was k_0, not eta
+In a few circumstances, changes in the parameters $\kappa$ or $\gamma$ led to the surprising situation where the PAC controller, which only observes one cell at a time, was more robust to parameter errors than was the FAC controller, which can observe both cells, even to the extent that the PAC controller could sometimes outperform the FAC.
+We stress that this effect is due to the mismatch between the assumed and the actual parameters, and re-optimizing the two controllers for the new parameter combination will always yield better control results for the FAC compared to the PAC.
+
+To assess the potential robustness of noise-enhanced control of the proposed synthetic auto-repression toggle switch, we adopted the model  from \cite{May2021} as well as the fully aware and partially aware controllers, $u^{\rm FAC}(x_1,x_2)$ and  $u^{\rm FAC}(x_1)$, respectively as described in the methods section.
 In the following subsections, we explore how well these controllers would perform in the more realistic setting where the approximate model used to define the controller no longer matches exactly to the dynamics of the process that the controller is being used to manipulate.
@@
 \begin{center}
-\includegraphics[width=1\textwidth]{GranularityPerturbation.eps}
+\includegraphics[width=1\textwidth]{GranularityPerturbation.pdf}
 \vspace{-0.1in}
-\caption{Effects of Scale Changes.
-(A-F) Joint (A,C,D,) and marginal (B,D,F) probability distributions for the FAC control method at $\alpha=0.2$ (A,B), $\alpha=1$ (C,D) and $\alpha=2.0$ (E,F).
-Target point shown as blue circle.
-(G-L) Same as (A-F), but for PAC controller.
-(M)  FAC (cyan) and PAC (magenta) controller performance as function of system volume.
-Horizontal dashed lines correspond to SSA analyses at $\alpha=100$.
-In all analyses, the system model and PAC and FAC controllers are unchanged from \cite{May2021} other than their adaptation to the new system volume size (see Methods).}
+\caption{Systems with high granularity were shown to have increased control performance but slower response times.}
 \label{Volume}
@@
 
-\begin{figure*}
-\begin{center}
-\includegraphics[width=1\textwidth]{GranularityScore.eps}
-\caption{}
-\label{}
-\end{center}
-\end{figure*}
-
-\begin{figure*}
-\begin{center}
-\includegraphics[width=1\textwidth]{CalculationSpeed.eps}
-\caption{}
-\label{Speed}
-\end{center}
-\end{figure*}
-
 We note however, that the improved steady state performance for systems with larger scales, comes at a cost that they also take longer to reach a steady state distribution.
@@
 One hundred SSA simulations of the flip-mirroring process were simulated while measuring the average score all simulations over time.
-Fig.\ \ref{Score} (A and B) shows the average score and protien counts of the simulation when $\alpha=1$, Fig.
+Fig.\ \ref{Score} (A and B) shows the average score and protein counts of the simulation when $\alpha=1$, Fig.
 \ \ref{Score} (C,D) when $\alpha = 10$, and Fig.
@@
 
+\subsection{Effects of Time Delays on Controller Performance}
 \begin{figure*}
 \begin{center}
-\includegraphics[width=1\textwidth]{ParameterPerturbation.pdf}
-\caption{Effects of parametric variations on control performance.
-({\bf A}) Sensitivity of control performance score, $J$, to modifications in parameters applied to Cell 1 only, one at a time for parameters $\kappa$, $\beta$, $\eta$, $k_0$, and $\gamma$ (in order left to right).
-({\bf B}) Same as {\bf A}, but where modifications are made to the corresponding parameter in Cell 2.
-({\bf C}) Same as ({\bf A}), but where the modification is made simultaneously to both cells.
-In all panels, trends lines are shown for the FAC (solid) and PAC (dashed) controllers.
-Controllers are unchanged from \cite{May2021}.}
-\label{Parameter}
-\end{center}
-\end{figure*}
-% XXX - This figure needs much larger fonts for all the labels and axes.
-% MMM - Did the change
-% There are errors in the parameter names.  What is 'alpha'?
-% remove the A,B,C,D,E.
-% replace I, II, III with A, B, C
-% what is parameter $\mu$? Is that an error?  Please use $\mu$ for reaction index as in the methods section.
-
-\subsection{Effects of Parameter Errors or Extrinsic Uncertainties}
-
-Exact parameters are never known for any real biological process, and even genetically identical single-cell systems are expected to exhibit cell-to-cell heterogeneity in parameters due to extrinsic variations outside of the specific pathways of interest.
-These parameter uncertainties or variations between cells could influence system performance, such that a control law that is optimized for one fixed, but incorrect, set of parameters may or may not succeed if used to break symmetry and track a desired target state when applied to a system with different parameters.
-To explore the sensitivity of control performance to unknown errors in the model parameters, sensitivity analysis was performed for each model parameter and for the situation where the corresponding parameter was modified in either one cell at a time (i.e., to simulate extrinsic noise) or in both cells simultaneously (i.e., to simulate parameter errors).
-For each perturbation analysis, both the $\mathbf{u}^{\rm FAC}$ and $\mathbf{u}^{\rm PAC}$ controller were analyzed over a range of parameters from one-half to two-fold of the original parameters.
-% XXX - How hard would it be to extend this to 10 fold in either direction?
-% MMM-Changes to parameters can dramatically change the location of the probability distribution which is contained in a bounded (50,50) area.
-A 10 fold increase to a parameter can push all the probability to the top corner of the space
- Figure \ref{Parameter}A shows the results for these parameter sweeps when a single parameter in Cell 1 is change while the parameters of Cell 2 is held fixed, and Fig.\ \ref{Parameter}B shows the opposite when changes Cell 1 has fixed parameters and Cell 2 is modified.
-Finally, \ref{Parameter}B  shows the performance change when the corresponding parameter is changed in both Cell 1 and Cell 2 at the same time.
-
-Modification to parameters was found to produce a broad range of effects, where some changes increase performance, some decrease performance, and others have little or no effect.
-For example, increasing $\beta$ in Cell 1 worsens performance while increasing $\beta$ in Cell 2 improves performance (compare second column in Fig.\ \ref{Parameter}A and \ref{Parameter}B).
-In some cases, these effects were not monotonic; for example, increasing $\kappa$ in Cell 1 (Fig.\ \ref{Parameter}A, left column) would be highly advantageous up to a limit after which the control performance degrades rapidly.
-In other cases (such as for $k_0$), the effect of parameter perturbations on performance is insignificant even for relatively large ($s$=2) perturbations.Similarly, Fig.\ \ref{Parameter}C shows that when parameters of both Cell 1 and Cell 2 are jointly changed, these changes could also improve or detract from control performance.
-%It was found that the $\mathbf{u}^{\rm FAC}$ often outperformed the $\mathbf{u}^{\rm PAC}$, but for large $s$ in a few parameters the inverse was found (figure \ref{parameter} (Row III, E and At).
-
-When examining the effects of changing parameters when using the $\mathbf{u}^{\rm FAC}$ (solid lines) or the $\mathbf{u}^{\rm PAC}$ (dashed lines), general trends typically remained the same in that changes to a parameter which cause a decrease in the control performance of the FAC also tend to decrease the performance of the PAC.
-Some parameters (e.g., $\kappa$, $\gamma$) reached minima that performed better than the original, which suggests that the physical system itself has room for optimization beyond the original design that could lead to better control performance, {\em even without changing the control law}.
-One explanation for why changes to $k_0$ had no effect on control performance is that the original parameter was so small that it may require a larger scaling on a different magnitude before changes are observed.
-% XXX - what is \eta?  From the table this should be the cooperativity with is very large.  Is this a mistake?
-% MMM - This was k_0, not eta
-In a few circumstances, changes in the parameters $\kappa$ or $\gamma$ led to the surprising situation where the PAC controller, which only observes one cell at a time, was more robust to parameter errors than was the FAC controller, which can observe both cells, even to the extent that the PAC controller could sometimes outperform the FAC.
-We stress that this effect is due to the mismatch between the assumed and the actual parameters, and re-optimizing the two controllers for the new parameter combination will always yield better control results for the FAC compared to the PAC.
-
-\subsection{Effects of Time Delays on Controller Performance}
-
-\begin{figure*}
-\begin{center}
-\includegraphics[width=1\textwidth]{TimeDelayPerturbation.eps}
+\includegraphics[width=1\textwidth]{TimeDelayPerturbation.pdf}
 \vspace{-0.1in}
-\caption{Effects of time delays on control performance.
-(A-F) Joint (A,C,E) and marginal (B,D,F) probability distributions for the FAC control method at three different time delays of $\tau/\gamma$=0.01 (A,B), 0.1 (C,D) and 1.0 (E,F) relative to the target point (blue circles).
-(G-L) Same as (A-F), but for PAC controller.
-(M) FAC and PAC performance as function of time delay show that PAC outperforms the FAC when time delay is greater than $0.07 /\gamma$.
-In all analyses, the system model and PAC and FAC controllers are unchanged from \cite{May2021}; specifically, the protein degradation rate is $\gamma = 0.0203$ as identified from \cite{Baumchager}.}
+\caption{Stochastic simulations of the time delayed system show decreasing control performance using either controller. }
 \label{Time}
@@
 These data show that at large time delays above , attempting to improve control with feedback is worse than no-feedback.
+
+% XXX - This figure needs much larger fonts for all the labels and axes.
+% MMM - Did the change
+% There are errors in the parameter names.  What is 'alpha'?
+% remove the A,B,C,D,E.
+% replace I, II, III with A, B, C
+% what is parameter $\mu$? Is that an error?  Please use $\mu$ for reaction index as in the methods section.
+
+\begin{figure*}
+\begin{center}
+\includegraphics[width=1\textwidth]{DelayAndGranularity.pdf}
+\vspace{-0.1in}
+\caption{Stochastic simulations of the FAC driven system (A) and PAC driven system (B) show different failure modes at at different $(\tau,\alpha)$ pairs (C).}
+\label{DG}
+\end{center}
+\vspace{-0.2in}
+\end{figure*}
+
+Joint analysis of time delay and granularity together was performed to better understand if granularity can undo the effect of time delay.
+System control performance was analyzed over a two dimensional domain of points $(\tau,\alpha)$ using sixteen stochastic simulations for XXX minutes.
+These simulations were driven using the FAC controller Fig \ref{DG} (A), and PAC controller Fig \ref{DG} (B).
+The landscape of the FAC and PAC control performance Fig \ref {CtrlP}(A and B)over the domain shows a valley at low $\tau$, and a plateau at large $\tau$.
+This suggests that increasing system scale at large time delays is not beneficial to control performance.
+To better understand the appearance of the landscape, trajectories in different regions were compared.
+Trajectories at the edge of the plateau region (blue star) showed bursty behavior Fig \ref {CtrlP}(C)(blue lines), while trajectories at the edge of the valley Fig \ref {CtrlP}(C) (red lines) showed oscillatory behavior.
+Despite the small different in time delay between the red and blue trajectory, a stark change to the overall behavior was observed.
+
+\begin{figure*}
+\begin{center}
+\includegraphics[width=1\textwidth]{ControlPoints.pdf}
+\vspace{-0.1in}
+\caption{Controllers optimized to a 2d domain of target points (A) show good control performance in the central region despite a broad range of targets (B and C). }
+\label{CtrlP}
+\end{center}
+\vspace{-0.2in}
+\end{figure*}
+
+Many controllers were optimized to different target points to determine if certain reference points were harder to control than others.
+Controllers were optimized over a two dimensional domain of target points $(T_1,T_2)$between between five and fourty-five.
+Figure \ref{CtrlP} (A) shows the control performance of the system after optimizing each controller to each point in the domain.
+Figure \ref{CtrlP} (B and C) show the optimized control input and the steady state of two points in the system.
+The unstable point in the rate equation matches the unstable point near $(20,20)$ in this domain.
+
+\begin{figure*}
+\begin{center}
+\includegraphics[width=1\textwidth]{ControlReference.pdf}
+\vspace{-0.1in}
+\caption{Finite state projections of the slowly driven system demonstrate good control performance under a variety of phase and frequency scenario. }
+\label{CR}
+\end{center}
+\vspace{-0.2in}
+\end{figure*}
+
+Dynamic reference targeting was performed by cycling though individual controllers which were optimized to a single target point at a fixed frequency.
+Thirty-two controllers were optimized along a pathway in order to target an in-sync reference point \ref{CR}(Synchronous), a phase lagged reference point \ref{CR}(Phase Lagged), or frequency separated reference point\ref{CR}(Phase Lagged and Frequency).
+The in phase signal is given by $r(t)=[10 sin (2\pi f t)+20, 10 sin (2\pi f t)+20]$, the phase-lagged signal is given by $r(t)=[10 sin (2\pi f t)+20, 10 cos (2\pi f t)+20]$, and the frequency driven signal is given by $r(t)=[10 sin (2\pi f t)+20, 10 sin (3\pi f t)+20]$ FSP simulations of cyclically driven systems were slowly driven at a frequency of 1/5000 seconds and their medians are shown in the Fig \ref{CR}(A4, B4,C4)(red cloud, blue cloud).
+Root mean squared error of each simulation are shown in Fig \ref{CR}(A5, B5,C5).
+Simulations show that in-phase control performed the best with an RSME of 7.0 despite moving over the unstable point in the system.
+When the system is driven with a phase-lag the RSME of the score increases to 8.3 and when driven at a different frequency the RMSE goes up to $9.1$.
+
+\begin{figure*}
+\begin{center}
+\includegraphics[width=1\textwidth]{ControlReferenceGranularity.pdf}
+\vspace{-0.1in}
+\caption{Stochastic simulations driven using a phased lagged controller show improved score at high granularity and low frequency.}
+\label{CRG}
+\end{center}
+\vspace{-0.2in}
+\end{figure*}
+Although the control of a dynamic reference point using a Piecewise controller is possible, the effect of cycling frequency is also important.
+16 SSA simulations of the phase-separated driving controller were performed on a system over a two dimensional domain of points $(\alpha,f)$ for a time of XXX minutes.
+Comparisons of control performance at $\alpha=0.5$ and $\alpha=2.0$ (Fig \ref}{CRG} (C))show that control performance improves with system scale.
+16 SSA simulations of the high granularity system driven by phase-lag were performed over a range of frequency values.
+These analysis show that at low frequency the systems have a noise floor which decreases with increasing alpha, but was the system were driven rapidly, all systems became equally uncontrollable.
+At high frequency, the scores never got worse than $J=650$.
+At high frequency the control performance plateaus, while at lower frequency the system improves with increasing $\alpha$ Fig \ref{CRG}(E).
 

### 2023-05-04 paper.tex
update

@@
 Thirty-two controllers were optimized along a pathway in order to target an in-sync reference point \ref{CR}(Synchronous), a phase lagged reference point \ref{CR}(Phase Lagged), or frequency separated reference point\ref{CR}(Phase Lagged and Frequency).
-The in phase signal is given by $r(t)=[10 sin (2\pi f t)+20, 10 sin (2\pi f t)+20]$, the phase-lagged signal is given by $r(t)=[10 sin (2\pi f t)+20, 10 cos (2\pi f t)+20]$, and the frequency driven signal is given by $r(t)=[10 sin (2\pi f t)+20, 10 sin (3\pi f t)+20]$ FSP simulations of cyclically driven systems were slowly driven at a frequency of 1/5000 seconds and their medians are shown in the Fig \ref{CR}(A4, B4,C4)(red cloud, blue cloud).
+The in phase signal is given by $r(t)=XXX$, the phase-lagged signal is given by $r(t)=XXX$, and the frequency driven signal is given by $r(t)=XXX$ FSP simulations of cyclically driven systems were slowly driven at a frequency of 1/5000 seconds and their medians are shown in the Fig \ref{CR}(A4, B4,C4)(red cloud, blue cloud).
 Root mean squared error of each simulation are shown in Fig \ref{CR}(A5, B5,C5).
-Simulations show that in-phase control performed the best with an RSME of 7.0 despite moving over the unstable point in the system.
-When the system is driven with a phase-lag the RSME of the score increases to 8.3 and when driven at a different frequency the RMSE goes up to $9.1$.
+Simulations show that in-phase control performed the best with an RSME of $7.0$ despite moving over the unstable point in the system.
+When the system is driven with a phase-lag the RSME of the score increases to $8.3$ and when driven at a different frequency the RMSE goes up to $9.1$.
 
@@
 \begin{center}
-\includegraphics[width=1\textwidth]{ControlReferenceGranularity.pdf}
+\includegraphics[width=1\textwidth]{ControlGranularityAndFrequency.pdf}
 \vspace{-0.1in}
@@
 Although the control of a dynamic reference point using a Piecewise controller is possible, the effect of cycling frequency is also important.
+
 16 SSA simulations of the phase-separated driving controller were performed on a system over a two dimensional domain of points $(\alpha,f)$ for a time of XXX minutes.
-Comparisons of control performance at $\alpha=0.5$ and $\alpha=2.0$ (Fig \ref}{CRG} (C))show that control performance improves with system scale.
+Comparisons of control performance at $\alpha=0.5$ and $\alpha=2.0$ (Fig (C))show that control performance improves with system scale.
 16 SSA simulations of the high granularity system driven by phase-lag were performed over a range of frequency values.
@@
 At high frequency, the scores never got worse than $J=650$.
-At high frequency the control performance plateaus, while at lower frequency the system improves with increasing $\alpha$ Fig \ref{CRG}(E).
+At high frequency the control performance plateaus, while at lower frequency the system improves with increasing $\alpha$ Fig(E).
 

### 2023-06-23 paper.tex
updated figs
before making major text edits'
update

@@
 Much work has been done to make control performance robust to noise, but the development of controllers which use noise to their advantage is not fully understood.
-Motivated by optogentics, we proposed a difficult control problem which requires the exploitation of noise in a stochastic system to break symmetry between two signals.
-in previous work, we found multiple such controllers which could exploit noise to break symmetry between two cells under a variety of system information.
+Motivated by optogentics, we previously proposed a difficult control problem which requires the exploitation of noise in a stochastic system to break symmetry between two signals.
+We found multiple such controllers which could exploit noise to break symmetry between two cells under a variety of system information.
 This work extends that analysis to include stochastic systems that trace a moving target, are affected by time delay, intrinsic noise, or measurement uncertainty.
@@
 \begin{center}
-\includegraphics[width=\columnwidth]{CartoonAndControler.pdf}
+\includegraphics[width=\columnwidth]{Cartoons.pdf}
 \caption{{\bf Optogenetic control of multiple cells using a single light input.}
@@
 
-Synthetic biology seeks to create modular\cite{Ng2019} and orthogonal\cite{Liu2018} components to sense and actuate\cite{Sheets2020} complex logical systems\cite{Groseclose2020}, and which are capable of performing a variety of advanced biological behaviors\cite{Shin2020}.
-New optogenetic tools have increased our ability to actuate embedded systems within cells reliably and with strong control performance\cite{Sheets2020,Baumschlager2017,Chen2020,Lillacci2018}.
-Advances in these two fields have enabled computer programmable control of cellular protein production through the use of external optogenetic inputs and smart microscope techniques\cite{Fox2021,Baumschlager2021,Lugagne2017}.
-These types of digital-synthetic actuators allow for fine-tuned, computer-modulated control of cellular systems that were previously impossible\cite{Rullan2018, Baumschlager2017} and with fast response times in comparison to chemical diffusion.
+Noise is a common theme throughout many areas of science, including control theory XXX, gene regulation XXX, mRNA dynamics XXX, and synthetic biology XXX.
+Excellent work has been done to enhance the robustness of control processes against noise.
+However, these systems are often controlled in spite of noise rather than working with it.
+Many stochastic systems can be modeled using a noise-free approach with additive noise.
+Nevertheless, a comprehensive analysis using the full master equation can uncover interesting properties, potentially leading to novel control mechanisms.
 
+The cell's drive towards homeostasis involves dealing with the inherently chaotic and noisy processes that reside within it.
+Brownian motion, stochastic gene regulation, and mRNA dynamics collectively generate a fundamentally noisy environment that cells must effectively manage.
+Despite these challenges, cells generally demonstrate strong capability to control these noisy processes.
+However, stochastic control theory is not well-developed, despite having substantial knowledge on how to model them.
+
+Synthetic biology aims to develop modular\cite{Ng2019} and orthogonal \cite{Liu2018} components to sense and manipulate \cite{Sheets2020} complex logical systems, enabling them to exhibit a wide range of advanced biological behaviors\cite{Shin2020}.
+Advancements in optogenetics have enhanced our ability to reliably actuate embedded systems within cells, offering precise control\cite{Sheets2020,Baumschlager2017,Chen2020,Lillacci2018}.
+These developments have facilitated computer-programmable regulation of cellular protein production through external optogenetic inputs and smart microscope techniques\cite{Fox2021,Baumschlager2021,Lugagne2017}.
+These digital-synthetic actuators enable fine-tuned, computer-modulated control of cellular systems, previously unattainable, with faster response times compared to chemical diffusion\cite{Rullan2018, Baumschlager2017} .
 Classical and modern control methods like PID control and model predictive control have been implemented in such systems\cite{} which can control synthetic systems to different stable points.
 % XXX Missing Citation above.
-Theoretical works have shown that stochastic systems can be controlled to unstable points by oscillating around those points \cite{Guarino2020} but new control techniques which exploit the full probability distribution information of the system can be used in certain scenarios to control simple systems to unstable points by exploiting the natural noise of protein production to its advantage.
-Previous control effort focus on deterministic ODE formulations of the control problem, but only a few studies seek to analyze the full chemical master equation to exploit noise to {\em improve} the control of synthetic biology processes \cite{Szymanska2015,May2021}.
+Theoretical works have shown that stochastic systems can be guided towards unstable points by oscillating around those points \cite{Guarino2020}.
+However, new control techniques that leverage the complete probability distribution information of the system can exploit the inherent noise of protein production to achieve better control.
+Previous control efforts primarily focused on deterministic ordinary differential equation (ODE) formulations of the control problem, only a limited number of studies have delved into analyzing the full chemical master equation to harness noise to improve the control of biological processes \cite{Szymanska2015,May2021}.
 
-Szymanska et al.~\cite{Szymanska2015} examined a theoretical system inspired by the genetic toggle switch from Kobayashi et al.\cite{Kobayashi:2004}, where two gene products repress one another, but where degradation of one protein was under the control of UV radiation via the XXX pathway.
-% XXX - add citation to Kobayashi paper, and update name of the pathway.  I think it was the SOS pathway.
-Their analysis showed that control was possible despite certain parameter errors and time delays due to maturation of fluorescent proteins and limited observation of the regulatory proteins. 
+inspired by the genetic toggle switch from Kobayashi et al.\cite{Kobayashi:2004}, Szymanska et al.~\cite{Szymanska2015} showed that stochastic control was possible despite certain parameter errors and time delays due to maturation of fluorescent proteins and limited observation of the regulatory proteins.
 To explore this concept in the context of more realistic models and parameters, May et al.~\cite{May2021} fit a simplified stochastic model to reproduce data measured in Baumschlager, et al.\cite{} for the expression from the XXX promoter under the optogenetic control of a UV-activated T7 polymerase (see model in Figure\ \ref{cartoons}A, top promoter).
-% XXX - add details about the promoter and citation.
-% XXX - let's change the order of panels A and B for this figure so that we can refer to them in the proper order.  I adjusted the text below to assume this change.
 An extended model showed that addition of a positive auto-regulation module (Figure\ \ref{cartoons}A, bottom promoter) could help to maintain an elevated expression phenotype in the presence of UV excitation.
-% XXX - Missing figure reference
-Using discrete stochastic models based on direct solutions to the chemical master equation, it was shown that the combination of biochemical noise, nonlinear auto-regulation, and a {\em single} optogenetic feedback could allow two genetically identical cells with arbitrary initial conditions to be controlled simultaneously to reach different specified phenotypes (Figure \ref{cartoons}B).
-For example, Figure\ \ref{cartoons}C(left) reproduces the heat maps for two control laws, where the level of a single external UV excitation signal was modulated depending on the observed cellular activity, and Figure\ \ref{cartoons}C(right) shows the corresponding marginal stationary distributions that could be achieved under these control laws.
-Effective control was possible not only for a Fully Aware Feedback Controller (FAC, Figure\ \ref{cartoons}C(top)) that observed the protein counts of all observed cells, but control could be achieved nearly as well for a Partially Aware Feedback Controller (PAC, Figure\ \ref{cartoons}C(bottom)) that observed the protein counts of only one cell.
-Moreover, it was again seen that these control laws remained effective even in the circumstance where the model used to design the control law was an inexact simplification of the process dynamics to which the controller was later applied.
+
+Using discrete stochastic models based on the chemical master equation, it was demonstrated that combining biochemical noise, nonlinear auto-regulation, and a single optogenetic feedback could control two genetically identical cells with different initial conditions to reach different desired signals at different frequency and phase (Figure \ref{cartoons}B).
+Figure\ \ref{cartoons}C(left) presents heat maps for two control laws, which a single external UV excitation signal based on observed cellular state.
+Figure\ \ref{cartoons}C(right) shows the corresponding achievable marginal stationary distributions.
+Effective control was possible with both Fully Aware Feedback Controllers (FAC, Figure\ \ref{cartoons}C(top)) and Partially Aware Feedback Controllers (PAC, Figure\ \ref{cartoons}C(bottom)), even when the control model was an approximate simplification of the actual process dynamics.
+
+Using discrete stochastic models based on the chemical master equation, this study demonstrates simultaneous control of two genetically identical cells with different initial conditions, by exploiting noise, nonlinear auto-regulation, and using a single input.
+Heat maps in Figure \ref{cartoons}C(left) illustrate two control laws based on observed cellular activity, while Figure\ \ref{cartoons}C(right) shows the corresponding achievable marginal stationary distributions.
+Effective control was achieved using both Fully Aware Feedback Controllers (FAC, Figure\ \ref{cartoons}C(top)) and Partially Aware Feedback Controllers (PAC, Figure\ \ref{cartoons}C(bottom)).
+
 %The solutions of the controllers was carried out by solving the chemical master equation (i.e., the forward Komogorov equation) to estimate the probability of possible cellular states.
@@
 
-In this paper, the analysis of this Single-Input-Multiple-Output (SIMO) multicellular control problem is extended to examine how the FAC and PAC controllers are affected by additional model uncertainties, including coarse grained approximations to the system dynamics, uncertainties or errors in system parameters, and delays in the biochemical sensors or optogenetic controllers through which the external feedback is implemented.
-In the next section, we briefly describe our methods to formulate the chemical master equation to analyze multiple individual cells in the same spatially homogeneous environment; we define the FAC and PAC controllers and their effects the dynamics of cells within that environment; and we explain how the control law is optimized to achieve desired performance objectives.
-Next, in the `Results' section, we explore how granularity of the model formulation, inaccuracy of parametric assumptions, or introduction of time delays would affect the FAC and PAC control performance.
-Finally, in `Conclusions' we summarize our findings and discuss the implications that these controllers, and their robustness to modeling errors, may have on future systems in synthetic biology.
+This paper extends the analysis of the Single-Input-Multiple-Output (SIMO) multicellular control problem by examining the impact of model uncertainties on FAC and PAC controllers.
+These uncertainties include coarse-grained system dynamics, parameter errors, and controller delays.
+Our methods involve formulating the chemical master equation to analyze cells in a state space, defining the controllers and their effects on cell dynamics, and optimizing the control input for desired performance.
+In the 'Results' section, we explore the effects of model granularity Figure \ref{cartoons}D parameter inaccuracies, and time delays Figure \ref{cartoons}E on FAC and PAC performance and add a simple scheme for controlling cells to dynamically changing point.
 
 \section{Methods}
-In previous work~\cite{May2021}, a six-species gene regulation model was developed to describe light-activated association of two T7 split domains (species 1 and 2) to form active T7 polymerases (species 3) which could then in turn associate with inactive T7 promoters (species 4) to create an active gene construct (species 5) that would produce its eventual protein product (species 6).
-A second, far simpler single-species model was also developed by assuming quasi-steady equilibrium for the first five species.
-Both models were independently parametrized by fitting to the same experimental data from Baumschlager, et al.~\cite{XXX}, and then extended to include a secondary self-activated promoter-gene construct for which the expression rate was defined by a Hill function (see Eqn.\ \ref{prodRate} below).
-In \cite{May2021}, we showed that a control law identified using the simple model would work to achieve differential control for the more complex model, thus motivating our current study's efforts to explore further how control performance would change upon additional simplifying assumptions or errors in model specification.
-In the following subsections, we discuss our methods to formulate this model, show how it is analyzed numerically, and discuss the specification of the control objective and the determination of an optimal controller.
+A gene regulation model consisting of six species was developed by May et al.
+\cite{May2021}.
+It described the light-activated association of two T7 split domains (species 1 and 2) to form an active T7 polymerase (species 3).
+These active polymerase could then associate with inactive T7 promoters (species 4), resulting in the formation of an active gene (species 5) that produced the desired protein product (species 6).
+
+Additionally, a simple single-species model was developed, assuming quasi-steady equilibrium for the first five species.
+Both models were independently parameterized using experimental data from Baumschlager et al.
+\cite{XXX}.
+Furthermore, an extension was made to incorporate a secondary self-activated promoter-gene construct, where the expression rate was determined by a Hill function \ref{prodRate}.
 
 \subsection{Model}
-To examine how model approximations would affect the application of noise-enhanced control strategies, we start with the one species model from \cite{May2021} (see Fig.\ \ref{cartoons}A), which consists of two reactions: protein production and degradation.
-The nonlinear light-activated protein production rate is given by
+To assess the impact of model approximations on the implementation of noise-enhanced control strategies, we begin with the one-species model proposed by May et al.
+\cite{May2021} (refer to Fig.
+\ref{cartoons}A).
+This model comprises of two reactions: nonlinear protein production and degradation and it is defined by the following equation:
  \begin{equation}
@@
 where $x$ is the instantaneous protein level; $\kappa$ is the maximum strength of the auto-regulation promoter; $\beta$ is the concentration at which auto-regulation promoter reaches half its strength; $\eta$ is the cooperativity in the auto-regulation promoter; $k_0$ is the promoter leakage rate; and $u(UV(t))$ is the T7 promoter strength.
-Through external feedback, the level of light excitation, and therefore the T7 promoter strength, $u(UV(t))$, can be modulated in response to the state of the system, thus removing explicit dependence on time.
-For example, in a system with the observation of two cells, one could define $u(UV(t)) = u^{\rm (FAC)}(x_1,x_2)$ as shown in Fig.\ \ref{cartoons}C(top left) for the fully aware control law, and for a system with the observation of only one cell, one could define $u^{\rm (PAC)}(UV(t)) = u(x_1)$ as shown in Fig.\ \ref{cartoons}C(bottom left) for the partially aware control law.
+Feedback enables the external modulation of the light input using the state of the system, thereby controlling the T7 promoter strength as a function of state rather than time and eliminating the explicit time dependance in $u(UV(t))$.
+
 Protein degradation is assumed to be a first order process with rate $\gamma$:
@@
 
-With these stoichiometry and propensities, we can formulate and run the Gillespie stochastic simulation algorithm\cite{Gillespie1992,Gillespie1977} to generate representative trajectories of the stochastic process.
-However, quantifying performance and designing an optimal controller requires a more direct analysis of the chemical master equation (CME).
-For the same specifications of the stoichiometry and propensity functions, the CME is a linear ODE that describes how probability mass changes in time due to the specified reaction propensities and stoichiometries, which can be written:
+The stoichiometry and propensities allow us to use the Gillespie stochastic simulation algorithm\cite{Gillespie1992, Gillespie1977} to generate representative trajectories of the stochastic process.
+However, a direct analysis of the chemical master equation (CME) is necessary for performance quantification and optimal controller design.
+The CME, based on the same stoichiometry and propensity functions, is a linear ODE that describes the time-dependent changes in probability mass due to the specified reaction propensities and stoichiometries.
+It can be expressed as:
 \begin{equation}
@@
 
-For the single-input-multiple-output control of the multiple cell system, any input signal that is applied to one cell in the population will be felt by all cells in the population.
-As a result, it is not possible to independently control any individual cell without introducing changes to the behavior of the others.
-Therefore, an effective controller must be able to strike a balance among the desired behaviors of all cells in the system.
+In the SIMO control of the multiple cell process, any input signal applied to one cell affects all cells in the population.
+Thus, it is not possible to independently control individual cells without impacting others.
+An effective controller must strike a balance among the desired behaviors of all cells in the system.
 
@@
 
-In this work, we explore two different controller designs that allow for the controller to use two different types of information.
-A fully aware controller ($\mathbf{u}^{FAC}$) uses direct observation of the protein count of two cells simultaneously while making its control input decision, while the partially aware controller $\mathbf{u}^{PAC}$) uses only the direct observation of a single cell:
+We studied two controller designs: the fully aware controller ($\mathbf{u}^{FAC}$) that utilizes simultaneous protein count observations from both cells, and the partially aware controller ($\mathbf{u}^{PAC}$) that relies on observations from a single cell.
 \begin{equation}
@@
 \begin{equation}
-\mathbf{u}^{PAC}=u(x_1)
+\mathbf{u}^{PAC}=u(x_2)
 \end{equation}
@@
 where $x_1$ and $x_2$ are discrete integers greater than or equal to zero that represent the number of proteins in cell one and cell two.
-In \cite{May2021}, the controllers $\mathbf{u}^{FAC}$ and $\mathbf{u}^{PAC}$ controllers were optimized to minimize $J$, and both $\mathbf{u}^{FAC}$ and $\mathbf{u}^{PAC}$ were saved as lookup tables after being optimized.
-In what follows, we will use these controllers exactly and without modification to explore how performance changes when the controllers are applied to systems that are approximations of the model under they were originally designed.
+In \cite{May2021}, we optimized the controllers $\mathbf{u}^{FAC}$ and $\mathbf{u}^{PAC}$ to minimize $J$ and saved them as lookup tables.
 
 \subsection{Scaling for system granularity}
-One of the major computational obstacles in using the CME or FSP formulations to analyze discrete stochastic chemical kinetics is that the dimension of the FSP scales linearly with each chemical species, and therefore exponentially with the number of independent species.
-To circumvent this issue, several previous studies have projected the FSP onto lower dimensional spaces based using various techniques such as time scale separations\cite{Peles2006}, Krylov subspaces\cite{Macnemara&Sidje?}, coarse meshes\cite{Munsky2008IEEE,Tapia2012CDC}, or principle orthogonal decompositions\cite{Vo2019}.
+Previous studies have addressed the computational challenge of analyzing discrete stochastic chemical kinetics by projecting the FSP onto lower-dimensional spaces using techniques such as time scale separations \cite{Peles2006}, Krylov subspaces \cite{Macnemara&Sidje?}, coarse meshes \cite{Munsky2008IEEE,Tapia2012CDC}, or principle orthogonal decompositions \cite{Vo2019}.
+However, using CME analyses is often hindered by an experimental limitations.
+Gene expression quantification based on fluorescence intensities, such as using fluorescent protein reporters, typically captures relative changes in expression over time rather than providing direct molecular counts.
 % XXX - would be good to look for a few other examples outside of our own papers where the CME dimension is reduced by linear projections.
-On the other hand, an experimental obstacle to using CME analyses is that gene expression quantification based on fluorescence intensities (e.g., using fluorescent protein reporters) usually does not provide molecular resolution, but rather measures only relative changes in expression over time.
 
-To address how variations in model or observation scales might impact the effectiveness of noise-based control, we use the controller optimized at one resolution and ask how effective it would be when applied to systems with the same dynamics but with different resolution.
-To accomplish this rescaling, let $M$ denote some metric for the size of the original system (e.g., the average number of proteins at steady state and maximal UV) and let $M'$ denote the same metric for the system at a different resolution.
-Toward this goal, we define a granularity parameter ($\alpha=M'/M$) that linearly scales the species populations to increase ($\alpha>1$) or decrease ($\alpha<1$) the maximum populations size, while maintaining the dynamics and general behavior of the model.
+It would be convenient to create a parameter that could increase or decrease the intrinsic noise of a system without changing the overall dynamics of the system.
+This would enable the analysis of models with more or less noise without the need of new unique models each time.
+A rescaling of the system could decrease the intrinsic noise while maintaining the same general behavior.
+To accomplish this rescaling, let $M$ denote the size of the origional unscaled system and let $M'$ denote the size of the newly scaled system.
+
+We define a granularity parameter ($\alpha=M'/M$) that linearly scales the species populations to increase ($\alpha>1$) or decrease ($\alpha<1$) the maximum populations size, while maintaining the dynamics and general behavior of the model.
 To apply the granularity parameter, we assume that each propensity function, $w_{\mu}$, from Eqns.\ \ref{Props} is rescaled to a different level of discreteness by substituting
@@
 % XXX - Please check this against what you did computationally.
-After this rescaling, if $\alpha<1 $ the system dynamics will now occur over a smaller range of protein counts, and if $\alpha >1$ the system dynamics occurs over a larger range of protein counts.
 
-We note that in transforming the propensity functions, the inputs to the controller have also be scaled by $1/\alpha$ before computing the appropriate functions of that state, e.g., ${u}^{\mathcal{C}} = {u}^{\mathcal{C}}(\mathbf{X}/\alpha)$.
+We note that in transforming the propensity functions, the inputs to the controller must also be scaled by $1/\alpha$ before computing the appropriate functions of that state, e.g., ${u}^{\mathcal{C}} = {u}^{\mathcal{C}}(\mathbf{X}/\alpha)$.
 Because identification of the original control formulation, $\mathbf{u}^{FAC}(x_1,x_2)$ and $\mathbf{u}^{PAC}(x_1)$ only considered integer values for $(x_1,x_2)$, fractional number inputs for after rescaling $(x_1/\alpha,x_2/\alpha)$ are handled by 2D cubic interpolation of the nearest values.
@@
 \subsection{Observation and actuation time delays}
-No real optogenetic control system can provide instantaneous feedback -- there will always be delays in biochemical reactions needed for observations (e.g., transcription and translation dynamics as well as folding and maturation of fluorescent proteins), delays in data analysis and decision making (e.g., image processing and calculation of the corresponding control signals), and delays in the actuation dynamics (e.g., activation of optogenetically excitable molecules).
-To explore the impacts of delays between the time a cellular fluctuation occurs and the time at which the corresponding control action can take effect, we developed a simple time-delay stochastic simulation algorithm.
-In this analysis, the state history is recorded following every stochastic event, allowing for reconstruction of the piecewise constant history of the species' populations, e.g., $x_1(s)$ and $x_2(s)$ for all $s\in [0,t]$.
+
+Delays are inherent in optogenetic control systems due to biochemical reactions involved in observations, data analysis, decision making, and actuation dynamics.
+To investigate the effects of time delay on control actuation, we devised a time-delay stochastic simulation algorithm.
+This algorithm records the state history after each stochastic jump, enabling reconstruction of the population history of the species and control input.
 Using this information, the time-delayed control signal at time $t$ can be specified as:
+
 \begin{equation}
@@
 
+\subsection{Dynamic setpoint control}
+
+static setpoint control is the control of a reference point which does not move over time. 
+dynamic setpont control targets a mvonig referenc peoint.
+synamis serpoint control is hard to deriuve, but the optimization of peicesice constangtof each static referenc pont is easy and can be stiched together.
+
 \section{Results}
@@
 \includegraphics[width=1\textwidth]{ParameterPerturbation.pdf}
-\caption{Parameter sweeps using the FAC and PAC show a broad range of control performance in cell 1 ({\bf A}),  in cell 2 ({\bf B}), and in both cells ({\bf C}). }
+\caption{Parameter sweeps using the FAC and PAC show a broad range of control performance in cell 1 ({\bf A}),  in cell 2 ({\bf B}), and in both cells ({\bf C}).
+Columns show each parameter in the model, rows show the cell which has its parameter perturbed.
+}
 \label{Parameter}
@@
 
-Exact parameters values are typically unknown for real stochastic process, and even genetically identical single-cell systems can exhibit cell-to-cell heterogeneity in parameters due to extrinsic variations outside of the specific pathways of interest.
-Estimates of parameters can be taken, but not exact measurement.
-These parameter uncertainties could influence system performance.
-To explore the sensitivity of control performance to unknown errors in the model parameters, sensitivity analysis was performed for each model parameter and for the situation where the corresponding parameter was modified in either one cell at a time (i.e., to simulate extrinsic noise) or in both cells simultaneously (i.e., to simulate parameter errors).
-For each perturbation analysis, both the $\mathbf{u}^{\rm FAC}$ and $\mathbf{u}^{\rm PAC}$ controller were analyzed over a range of parameters from one-half to two-fold of the original parameters.
+Real stochastic processes often have unknown parameter values, and even genetically identical single-cells often exhibit heterogeneity in parameters due to external variations.
+Parameter estimates can be obtained, but are not perfect measurements due to measurement limitations of stochastic processes.
+To asses the sensitivity of control performance to parameter errors, sensitivity analysis was performed for each model parameter of each cell, then on both cells simultaneously.
+For each parameter perturbation, both the $\mathbf{u}^{\rm FAC}$ and $\mathbf{u}^{\rm PAC}$ controllers were examined across a parameter range from one-tenth to ten-fold of the original values.
 % XXX - How hard would it be to extend this to 10 fold in either direction?
@@
 A 10 fold increase to a parameter can push all the probability to the top corner of the space
- Figure \ref{Parameter}A shows the results for these parameter sweeps when a single parameter in Cell 1 is change while the parameters of Cell 2 is held fixed, and Fig.\ \ref{Parameter}B shows the opposite when changes Cell 1 has fixed parameters and Cell 2 is modified.
-Finally, \ref{Parameter}B  shows the performance change when the corresponding parameter is changed in both Cell 1 and Cell 2 at the same time.
+ Figure \ref{Parameter}A shows the results for these parameter sweeps when a single parameter in Cell 1 is fixed, and Fig.\ \ref{Parameter}B shows the control performance as single parameters in Cell 2 are fixed while all other parameters are held constant.
+Finally, \ref{Parameter}B  shows the performance change when the corresponding parameters are changed in both Cell 1 and Cell 2 at the same time while keeping other parameters fixed.
 
-Modification to parameters was found to produce a broad range of effects, where some changes increase performance, some decrease performance, and others have little or no effect.
+Modification of parameters was found to produce a broad range of effects.
 For example, increasing $\beta$ in Cell 1 worsens performance while increasing $\beta$ in Cell 2 improves performance (compare second column in Fig.\ \ref{Parameter}A and \ref{Parameter}B).
@@
 In other cases (such as for $k_0$), the effect of parameter perturbations on performance is insignificant even for relatively large ($s$=2) perturbations.Similarly, Fig.\ \ref{Parameter}C shows that when parameters of both Cell 1 and Cell 2 are jointly changed, these changes could also improve or detract from control performance.
+
+Modification of parameters was found to produce a broad range of effects.
+Increasing $\beta$ in Cell 1 was found to worsen performance, while increasing $\beta$ in Cell 2 improved performance (Fig.
+\ref{Parameter}A, B, second column).
+The effects of parameter modification were not always monotonic.
+For example, increasing $\kappa$ in Cell 1 initially improved performance but reached a limit beyond which control performance degraded rapidly (Fig.
+\ref{Parameter}A, left column).
+In contrast, perturbations in the parameter $k_0$ had insignificant effects on performance, even for relatively large perturbations (s=10) (Fig.
+\ref{Parameter}A).
+Similarly, when parameters of both Cell 1 and Cell 2 were jointly changed, control performance could be either improved or decreased (Fig.
+\ref{Parameter}C).
+
 %It was found that the $\mathbf{u}^{\rm FAC}$ often outperformed the $\mathbf{u}^{\rm PAC}$, but for large $s$ in a few parameters the inverse was found (figure \ref{parameter} (Row III, E and At).
 
-When examining the effects of changing parameters when using the $\mathbf{u}^{\rm FAC}$ (solid lines) or the $\mathbf{u}^{\rm PAC}$ (dashed lines), general trends typically remained the same in that changes to a parameter which cause a decrease in the control performance of the FAC also tend to decrease the performance of the PAC.
-Some parameters (e.g., $\kappa$, $\gamma$) reached minima that performed better than the original, which suggests that the physical system itself has room for optimization beyond the original design that could lead to better control performance, {\em even without changing the control law}.
-One explanation for why changes to $k_0$ had no effect on control performance is that the original parameter was so small that it may require a larger scaling on a different magnitude before changes are observed.
+Changing parameters had similar effects on control performance for both $\mathbf{u}^{\rm FAC}$ (solid lines) and $\mathbf{u}^{\rm PAC}$ (dashed lines).
+Parameters such as $\kappa$ and $\gamma$ reached minima that performed better than the original values, indicating potential for system optimization beyond the original design.
+However, altering $k_0$ had no noticeable impact, possibly due to its small original value requiring a larger scaling for observable changes.
+
 % XXX - what is \eta?  From the table this should be the cooperativity with is very large.  Is this a mistake?
 % MMM - This was k_0, not eta
-In a few circumstances, changes in the parameters $\kappa$ or $\gamma$ led to the surprising situation where the PAC controller, which only observes one cell at a time, was more robust to parameter errors than was the FAC controller, which can observe both cells, even to the extent that the PAC controller could sometimes outperform the FAC.
-We stress that this effect is due to the mismatch between the assumed and the actual parameters, and re-optimizing the two controllers for the new parameter combination will always yield better control results for the FAC compared to the PAC.
+In some parameters like $\kappa$ or $\gamma$, the PAC controller showed greater robustness to parameter errors than the FAC controller, even outperforming it.
+However, this effect is due to the mismatch between assumed and actual parameters, and re-optimizing both controllers for the new parameters will always yield better control performance for the FAC compared to the PAC.
 
@@
 
-We first explore how the model defined for a given assumed system size (i.e., in which protein numbers have a range of $M$) when applied to a system with larger or smaller systems sizes (i.e., where protein numbers now have a range of $M' = \alpha M$.
+We first explore how the model defined for an origional system size (i.e., in which protein numbers have a range of $M$) when applied to a system with larger or smaller systems sizes (i.e., where protein numbers now have a range of $M' = \alpha M$.
 This modification could arise from two situations related to numerical convenience or inexact resolution in the measurement of concentrations.
@@
 For example, in measurements like those on \cite{Baumschlager} which are based on fluorescence protein intensities, it is not clear how many proteins correspond to what level of fluorescence intensity, and the scalar $\alpha$ is an unknown quantity that must be estimated (e.g., through calibration with single-molecule measurements or dilution experiment\cite{XXX - look for elowitz paper or something similar that tried to calibrate for single-molecule counts}).
+
 % MMM - Ding:2019??
-It is well established that the relative size of stochastic fluctuations (i.e., noise) in a chemical process decreases with the inverse square root of the process scale \cite{Kampen1961}.
-As the system size increases (with all species being scaled by the same factor), the dynamics approach those of a deterministic process with probability one (with zero-measure exceptions only for initial conditions lying on manifolds that separate attractors for different steady state behaviors)\cite{XXX}.
-For the purpose of noise-induced control, this introduces a tradeoff where noise is critical to break symmetry and enable differential control, yet noise is also deleterious to maintaining desired states once they have been achieved.
-Considering these competing objectives, it is not immediately clear how the reduction of noise through the addition of system granularity will affect control performance.
+Stochastic fluctuations in a chemical process decrease with the inverse square root of the process scale \cite{Kampen1961}.
+As the system size increases, the dynamics converge towards a deterministic process, except for certain exceptional initial conditions lying on manifolds that separate attractors for different steady state behaviors \cite{XXX}.
+This creates a tradeoff in noise-induced control, where noise enables differential control but is also deleterious to maintaining desired states once they have been achieved.
+The impact of reducing noise through increased system granularity on control performance remains unclear.
 
-To shed light on this tradeoff, control performance scores for the $\mathbf{u}^{\rm FAC}$ and $\mathbf{u}^{\rm PAC}$ controllers were calculated at different levels of granularity ($\alpha$) between 0.1 and 2.0.
+To shed light on this tradeoff, control performance scores for the $\mathbf{u}^{\rm FAC}$ and $\mathbf{u}^{\rm PAC}$ controllers were calculated at different levels of granularity ($\alpha$) between 0.2 and 2.0.
 % XXX - Why only go up to 2?  Did it become intractable?
 % MMM - Non_Sparse Calculations are intractible after 2, but sparse versions of code will work fine
-Figure \ref{Volume} (Ai - Fi) shows the joint (left plots) and marginal (right plots) distributions of the two cells relative to the specified target position (circles in joint distributions and vertical dashed lines in marginal distributions) for the FAC controller and for $\mathbf{\alpha}$=.2 (A,B), 1 (C,D) and 2 (E,F).
+Figure \ref{Volume} (A - F) shows the joint (left plots) and marginal (right plots) distributions of the two cells relative to the specified target position (circles in joint distributions and vertical dashed lines in marginal distributions) for the FAC controller and for $\mathbf{\alpha}$=.2 (A,B), 1 (C,D) and 2 (E,F).
 As the granularity changes from $\alpha$=0.2 , to $\alpha$=2, the distributions become focused at a point near the target, and the performance score improves from $J=$333 to 85 to 27, respectively.
@@
 \begin{center}
-\includegraphics[width=1\textwidth]{GranularityPerturbation.pdf}
+\includegraphics[width=1\textwidth]{Granularity.pdf}
 \vspace{-0.1in}
-\caption{Systems with high granularity were shown to have increased control performance but slower response times.}
+\caption{Systems with increased granularity are less noisy and have better control performance.
+(A-C) Joint (left) and marginal (right) distributions for the FAC shows increased control performance and tighter distributions as $\alpha$ increases from 0.2 (top row) to 2.0 (bottom row) .
+(D-F) Joint (left) and marginal (left) distributions using the PAC controller at different levels of granularity.}
 \label{Volume}
@@
 
-Figure \ref{Volume}M shows the trend of the performance score versus alpha for both the FAC (solid cyan line) and the PAC (solid magenta line) controller.
+Figure \ref{Volume}G shows the trend of the performance versus alpha for both the FAC (solid cyan line) and the PAC (solid magenta line) controller.
 This improvement in performance appears to approach a small value as the granularity goes to infinity, but since the size of the FSP increases with the square of the system size, systems much larger than $\alpha$=2 (where $\mathbf{A}_0\in \mathbb{R}^{10^4\times10^4}$) become more difficult to calculate.
@@
 
-We note however, that the improved steady state performance for systems with larger scales, comes at a cost that they also take longer to reach a steady state distribution.
-This tradeoff is illustrated in Fig.\ \ref{Score} which shows the performance score $J(t)$ as the system adapts to a change from the original target point $\mathbf{T} = [30,10]$ to its mirror point $\mathbf{T}_m = [10,30]$.
-% XXX - Please add a new figure to show this point.
-Start with an initial condition that is the steady state with target T and then flip it to T' = T(::-1).
-Then solve for P(t) and calculated J(t) from this.
-Finally, make a plot of J(t) versus t for the three cases \alpha = 0.2, 1, 2.
-One hundred SSA simulations of the flip-mirroring process were simulated while measuring the average score all simulations over time.
-Fig.\ \ref{Score} (A and B) shows the average score and protein counts of the simulation when $\alpha=1$, Fig.
-\ \ref{Score} (C,D) when $\alpha = 10$, and Fig.
-\ \ref{Score} (E,F) when $\alpha = 100$.
+To bypass this FSP limit, sixteen SSA simulations were simulated to steady state to sample the CME of a system with a significantly larger volume of $\alpha = 100$.
+Each SSA ran for $5\times10^7$ minutes, and the last $4\times10^7$ minutes were sampled for estimating the stationary distribution and calculating the performance score.
+The performance score estimates of this high granularity SSA, using the FAC and PAC controllers, were 4.13 and 27.5, respectively, shown as dashed lines in Fig.\ \ref{Volume}G.
+While it remains uncertain if further performance improvements can be achieved with larger system scales, these findings suggest that both controllers consistently improve with increased $\alpha$, with $\mathbf{u}^{\rm FAC}$ consistently outperforming $\mathbf{u}^{\rm PAC}$ in all cases examined.
+From figure H, it can be seen that a smaller $\alpha$ creates faster response but higher baseline score due to intrinsic noise.
+When $\alpha = 4$, the system becomes much less noisy and has lower scores but takes longer time to respond.
 
-From the figures it is clear that the smaller $\alpha$ creates faster dynamic but also have higher scores due to inherit noise.
-When $\alpha = 10$, the system becomes much less noisy and have lower scores but takes some time to respond.
-When $\alpha$ is too large at 10, the system contains very small levels of noise, but the dynamics
  % XXX - update this to explain what you see in this new figure.
@@
 \begin{center}
-\includegraphics[width=1\textwidth]{TimeDelayPerturbation.pdf}
+\includegraphics[width=1\textwidth]{TimeDelay.pdf}
 \vspace{-0.1in}
-\caption{Stochastic simulations of the time delayed system show decreasing control performance using either controller. }
+\caption{Stochastic simulations of the time delayed system show decreasing control performance using either the FAC or PAC.
+(A-C) Joint (left) and marginal distributions (right) of the at different levels of time delay for the FAC.
+(D-F) Joint (left) and marginal distributions(right) at three levels of time delay using the PAC.
+(G) plots of RMSE control performance versus time delay for both FAC and PAC.}
 \label{Time}
@@
 As the time required for any of these steps increases, control performance will be degraded, perhaps even leading to large fluctuations or instability.
-To explore how time delays would affect the noise-enhance controller $\mathbf{u}^{\rm FAC}$ and $\mathbf{u}^{\rm PAC}$, we generated large sets of time-delayed stochastic simulations (see Methods) for different lengths of the time delay.
+To explore how time delays would affect the noise-enhanced controller $\mathbf{u}^{\rm FAC}$ and $\mathbf{u}^{\rm PAC}$, we generated large sets of time-delayed stochastic simulations (see Methods) for different lengths of the time delay.
 Each SSA was sub-sampled for 1000 times over 10000 minutes of simulation time after a burnin period of 10000 minutes.
@@
 
-This data taken together show that time much larger that $0.07 / \gamma = 0.0142 min$ is detrimental to the FAC controller; delays beyond $0.2 / \gamma = 0.0406 min$ are detrimental for the PAC controller; and the best choice in controller is dependent the level of time delay in the system.
-For extreme levels of time delay, both systems lose their asymmetry, and their scores become much worse ($J=$1041 and 931).
-A no-feedback controller was developed which required no inputs.
-Such a controller is immune to time delay, and received a score $J=402$.
-These data show that at large time delays above , attempting to improve control with feedback is worse than no-feedback.
+This data show that time delays exceeding $0.07 / \gamma = 0.0142$ minutes is detrimental to the FAC control performance while delays beyond $0.2 / \gamma = 0.0406$ minutes are detrimental to the PAC controller.
+The choice of best controller depends on the level of time delay in the system.
+At extreme time delays, both controllers lose their asymmetry, resulting in significantly worse scores ($J = 1041$ and $931$).
+The control performance of a no-feedback controller was measured which required no inputs Figure\ref{Time}(G) (red line) and achieved a score of $J = 402$.
+These findings indicate that, for large time delays, attempting to improve control with feedback is less effective than using a no-feedback approach.
 
@@
 \vspace{-0.1in}
-\caption{Stochastic simulations of the FAC driven system (A) and PAC driven system (B) show different failure modes at at different $(\tau,\alpha)$ pairs (C).}
+\caption{Stochastic simulations of the FAC driven system and PAC driven system show different failure modes at at different $(\tau,\alpha)$ pairs.
+(A-I)Joint probability distributions at different $[\tau,\alpha]$ pairs.
+$\alpha$ across columns and $\tau$ across rows.
+(J) Control performance measured over a domain of $(\tau,\alpha)$ pairs show good control performance when both $\alpha$ is large and $\tau$ small.
+(K) Trajectories where good control was expected but not observed show oscillations instead.}
 \label{DG}
@@
 
-Joint analysis of time delay and granularity together was performed to better understand if granularity can undo the effect of time delay.
-System control performance was analyzed over a two dimensional domain of points $(\tau,\alpha)$ using sixteen stochastic simulations for XXX minutes.
-These simulations were driven using the FAC controller Fig \ref{DG} (A), and PAC controller Fig \ref{DG} (B).
-The landscape of the FAC and PAC control performance Fig \ref {CtrlP}(A and B)over the domain shows a valley at low $\tau$, and a plateau at large $\tau$.
+By increasing $\alpha$ we can change the characteristic of a system from being more noise-like to more deterministic-like.
+Similarly, increasing $\tau$ can decrease the quality of the control input.
+Generally, we would expect that increasing $\alpha$ would improve control performance, while increasing $\tau$ would decrease control performance.
+
+FAC steady state control performance was analyzed over a two dimensional domain of points $(\tau,\alpha)$ using thirty-two stochastic simulations simulated to steady state to see if the positive effect granularity could undo the effect of negative effect of time delay.
+Instead, noise induced oscillations were surprisingly found at certain $(\tau,\alpha)$ pairs and worse control performance was observed when system granularity increase for some $\tau$.
+Joint probability distributions at certain $(\tau,\alpha)$ pairs and their control performance can be seen in Fig \ref{DG}(A-I).
+Control performance reduced as $\alpha$ increased when $\tau=100$ but not at $\tau=1$.
+FAC control performance Fig \ref{DG}(J) over the domain of points shows a valley region at low $\tau$, and a plateau region at large $\tau$.
 This suggests that increasing system scale at large time delays is not beneficial to control performance.
-To better understand the appearance of the landscape, trajectories in different regions were compared.
-Trajectories at the edge of the plateau region (blue star) showed bursty behavior Fig \ref {CtrlP}(C)(blue lines), while trajectories at the edge of the valley Fig \ref {CtrlP}(C) (red lines) showed oscillatory behavior.
-Despite the small different in time delay between the red and blue trajectory, a stark change to the overall behavior was observed.
+Trajectories at the edge of the valleyFig \ref {DG}(J)(yellow star) showed oscillatory behavior Fig \ref {DG}(K)and trajectories within the plateau show bursty behavior.
+
+The control performance domain Fig \ref{DG}(J) shows that low noise and high quality control inputs are both required to achieve good control.
+When noise is improved by increasing $\alpha$ the trajectories were observed going from very noisy to very oscillating which prevented improved control performance.
+Decreases in $\alpha$ should make the system respond faster and be less susceptible to time delay, but also more noisy.
+These competing effects in control performance can be observed in Fig \ref{DG}(J) which show that control performance is always poor at low $\alpha$.
 
@@
 \begin{center}
-\includegraphics[width=1\textwidth]{ControlPoints.pdf}
+\includegraphics[width=1\textwidth]{StaticControl.pdf}
 \vspace{-0.1in}
-\caption{Controllers optimized to a 2d domain of target points (A) show good control performance in the central region despite a broad range of targets (B and C). }
+\caption{The FAC and PAC control methods were used to optimize control inputs and probability distributions for static set points in a 2D domain.
+Optimized control inputs and probability distributions were optimized using the FAC control method(A and B) or the PAC control method (C and D).
+(E) Domains of control performance for the FAC method over a region of static target points show regions of good control and symmetry along the dashed line.
+(F) Domain of control performance over a set of static control points for the PAC control method.}
 \label{CtrlP}
@@
 
-Many controllers were optimized to different target points to determine if certain reference points were harder to control than others.
-Controllers were optimized over a two dimensional domain of target points $(T_1,T_2)$between between five and fourty-five.
-Figure \ref{CtrlP} (A) shows the control performance of the system after optimizing each controller to each point in the domain.
-Figure \ref{CtrlP} (B and C) show the optimized control input and the steady state of two points in the system.
+We analyzed the control performance over a domain of target points to find if some regions are easier to control than others.
+Stability analysis of the auto-regulated system using a moderate light input can create up to four stable points and one unstable point in the state space.
+These points give rise to basins which may attract probability more easily than other points.
+Control performance of the stochastic system was analyzed over a domain of different set points and were optimized using more observable (FAC) or less observable (PAC) control.
+
+Controllers were optimized over a two dimensional domain of target points $[T_1,T_2]$ between between five and forty-five using the FAC input method or the PAC input method.
+Figure \ref{CtrlP} (A) shows the optimized control input of the system when the target is $T=[20,25]$ after optimizing the FAC.
+Figure \ref{CtrlP} (B) show the steady state probability distribution of the controlled system.
+Figure \ref{CtrlP} (C and D) show the optimized controller and probability distributions using the partially aware control method.
 The unstable point in the rate equation matches the unstable point near $(20,20)$ in this domain.
+
+FAC and PAC control performance over the domain of static set points Fig\ref{CtrlP}(E) and (F) showed regions of good control in the FAC between 10 and 30 but a smaller region in the PAC.
+The FAC control performance was found to be symmetric along the dashed black line.
+The cause was found to be that symmetric pairs had control inputs which were exact transposes of each other.
+The unstable point of the system is known to be near [20,20] which is also a point of poor control for the FAC system with respect to local control performance.
+Edge regions were difficult to control, likely due to the geometry of how control inputs move the probability.
 
@@
 \begin{center}
-\includegraphics[width=1\textwidth]{ControlReference.pdf}
+\includegraphics[width=1\textwidth]{DynamicControl.pdf}
 \vspace{-0.1in}
-\caption{Finite state projections of the slowly driven system demonstrate good control performance under a variety of phase and frequency scenario. }
+\caption{Finite state projections of the slowly driven system demonstrate good control performance under a variety of phase and frequency scenario.
+(A) Schematic of the control problem shows targeting the number of protein in two cells using a dynamically changing reference point by cycling through a set of individual stable points and under synchonous, phasse-lagged or frequency changes reference signal using only one control input.
+(B1-BN) shows the phase space of the reference phase space, error residuals, reference time series, FSP solution and RMSE respectively for the synchonous control method.
+(C1-C5) Shows the same data for the phase-lagged control input.
+(D1-D5) Shows the same analysis for the frequency-changed control.}
 \label{CR}
@@
 
-Dynamic reference targeting was performed by cycling though individual controllers which were optimized to a single target point at a fixed frequency.
-Thirty-two controllers were optimized along a pathway in order to target an in-sync reference point \ref{CR}(Synchronous), a phase lagged reference point \ref{CR}(Phase Lagged), or frequency separated reference point\ref{CR}(Phase Lagged and Frequency).
-The in phase signal is given by $r(t)=XXX$, the phase-lagged signal is given by $r(t)=XXX$, and the frequency driven signal is given by $r(t)=XXX$ FSP simulations of cyclically driven systems were slowly driven at a frequency of 1/5000 seconds and their medians are shown in the Fig \ref{CR}(A4, B4,C4)(red cloud, blue cloud).
-Root mean squared error of each simulation are shown in Fig \ref{CR}(A5, B5,C5).
-Simulations show that in-phase control performed the best with an RSME of $7.0$ despite moving over the unstable point in the system.
-When the system is driven with a phase-lag the RSME of the score increases to $8.3$ and when driven at a different frequency the RMSE goes up to $9.1$.
+We developed a method to control the dynamics of a system by optimizing steady state controllers for a predefined path $r(x)$ in a Piecewise manner and cycling through them at a fixed frequency.
+Thirty-two controllers were optimized along a pathway in order to target an in-sync reference point \ref{CR}, a phase lagged reference point \ref{CR}, or frequency separated reference point\ref{CR}(A).
+The synchronous signal is given by
+\begin{equation}
+ r_1(x)=[20 sin(2 \pi x)+10, 20 sin(2 \pi x)+10]
+\end{equation}
+the phase lagged signal is given by
+\begin{equation}
+ r_2(x)=[20 sin(2 \pi x)+10, 20 cos(2 \pi x)+10]
+\end{equation}
+and the frequency changed signal is given by
+\begin{equation}
+ r_3(x)=[20 sin(2 \pi x)+10, 20 sin(6 \pi x)+10]
+\end{equation}
+for $x\in[0,1]$ using 32 steps.
+FSP simulations of cyclically driven systems were slowly driven at a frequency of $1e-4$ seconds and their medians are shown in the Fig \ref{CR}(B4, C4, and D4)(red line, blue line).
+The time series probability distributions of $x_1$ and $x_2$ can be seen as red clouds and blue clouds respectively.
+Regions with purple clouds are the exact overlap of red and blue.
+Error distributions of the system were taken by shifting the probability distribution from the reference point to the origin Fig \ref{CR}(B2, C2, and D2).
+Root mean squared error of each simulation are shown in Fig \ref{CR}(B5, C5,D5).
+
+Suprisingly, the system is able to effectively control two signals to two different trajectories at different phase and frequency and using only a single control input despite the presence of noise Fig \ref{CR}(D4).
+Simulations show that the synchronous control performed the best with an RSME of $6.9$ despite moving over the unstable point in the system.
+When the system is driven with a phase-lag the RSME of the score increases to $8.2$ and when driven at a different frequency the RMSE goes up to $8.6$.
+Medians of the marginal distributions show good match with the desired reference signal in each control scheme Fig \ref{CR}(B4,C4, and D4).
+TIme series trajectories of the RMSE show spikes in the control performance using any control strategy.
 
@@
 \begin{center}
-\includegraphics[width=1\textwidth]{ControlGranularityAndFrequency.pdf}
+\includegraphics[width=1\textwidth]{Frequency.pdf}
 \vspace{-0.1in}
-\caption{Stochastic simulations driven using a phased lagged controller show improved score at high granularity and low frequency.}
+\caption{Control performance analyzed over a domain of $f,\alpha$ pairs show worst control performance at moderate frequencies near $1e-2$ Tracking reference signals when $\alpha=5$ span a range between 50 and 150 species (A).
+Stochastic simulations driven using a phased lagged controller at $\alpha = 5$ and low frequency show tighter control compared to $\alpha =1$ (B).
+Systems driven at moderate frequency show worse control performance than high frequency or low frequency (C).
+Control performance only of phase-lagged system only improves with increasing $\alpha$ and low $f$.}
 \label{CRG}
@@
 \end{figure*}
+
+systems driven at high frequency often have thier control performance reduced due to the response time scale of the system they are driving, but in a stochastic system the combination of high frequency dirving and noise could be catastriphic.
+
 Although the control of a dynamic reference point using a Piecewise controller is possible, the effect of cycling frequency is also important.
 
-16 SSA simulations of the phase-separated driving controller were performed on a system over a two dimensional domain of points $(\alpha,f)$ for a time of XXX minutes.
-Comparisons of control performance at $\alpha=0.5$ and $\alpha=2.0$ (Fig (C))show that control performance improves with system scale.
-16 SSA simulations of the high granularity system driven by phase-lag were performed over a range of frequency values.
-These analysis show that at low frequency the systems have a noise floor which decreases with increasing alpha, but was the system were driven rapidly, all systems became equally uncontrollable.
-At high frequency, the scores never got worse than $J=650$.
-At high frequency the control performance plateaus, while at lower frequency the system improves with increasing $\alpha$ Fig(E).
+Sixty-four SSA trajectories of the phase-separated dynamic controller were simulated over a two dimensional domain of points $(\alpha,f)$ for 16 cycles after reaching steady state.
+Increasing $\alpha$ while holding $f$ constant at $XXX$ showed improved control performance at low frequency but little change at high frequency Fig \ref{CRG}(D).
+Analysis of control performance using varying frequency but constant $\alpha$ showed the worst performance at moderate $f$.
 
 \section{Conclusion}
+
 Key results from \cite{May2021} showed that noise was a key requirement for the development of a noise-exploiting controller, and that deterministic systems could not be controlled to two different stable points if they both started at the same initial condition.

### 2023-06-26 paper.tex
update

@@
 
-Unnderstanding the effects of noise in control theory is important to ()
-Much work has been done to make control performance robust to noise, but the development of controllers which use noise to their advantage is not fully understood.
-Motivated by optogentics, we previously proposed a difficult control problem which requires the exploitation of noise in a stochastic system to break symmetry between two signals.
+Previous research has focused on robust control performance in the presence of noise, but understanding controllers that exploit noise remains incomplete. 
+Motivated by Maxwells-Demon, we previously proposed a difficult control problem which requires the exploitation of noise in a stochastic system to break symmetry between two signals.
 We found multiple such controllers which could exploit noise to break symmetry between two cells under a variety of system information.
 This work extends that analysis to include stochastic systems that trace a moving target, are affected by time delay, intrinsic noise, or measurement uncertainty.
-
-Synthetic biology seeks to create modular components that generate complex and controllable biological behaviors, often at microscopic scales where important regulatory RNA or proteins are present in small numbers.
-Intrinsic fluctuations at these scales cause biochemical `noise' that frequently results in undesired neussance effects.
-However, recent theoretical analyses have shown that noise can sometimes be used to one's advantage when seeking to manipulate nonlinear regulatory systems that would be difficult or impossible to control in a fully deterministic setting.
-Specifically, it was shown that a noise-exploiting optogenetic controller reminiscent of Maxwell's Demon could systematically force multiple identical systems to reach different independently-assigned stable points, even for a few special cases where the model is inexact or if observations were incomplete.
-This work expands upon those analyses to explore in greater detail how control performance is affected by errors or approximations in the model or by temporal delays in the observations.
  We find that noise-exploiting controllers can remain highly effective despite coarse approximations to the model's scale or incorrect estimations of key model parameters, and these controllers can even retain performance for significant time delays.
-Together, these findings suggest that noise-exploiting control should be possible in real experiments where models are always approximate, where parameters are always uncertain, and where observations are corrupted by errors.
+Together, these findings suggest that noise-exploiting control should be possible where models are always approximate, where parameters are always uncertain, and where observations are corrupted by errors.
 
 \end{abstract}
-Keywords: Optogenetic control, Stochastic gene regulation, Synthetic biology
+Keywords: Stochastic Control, Gene Regulation, Optogenetics
 
@@
 However, these systems are often controlled in spite of noise rather than working with it.
-Many stochastic systems can be modeled using a noise-free approach with additive noise.
-Nevertheless, a comprehensive analysis using the full master equation can uncover interesting properties, potentially leading to novel control mechanisms.
+Many stochastic systems can be modeled using a noise-free approach with additive noise but, a detailed analysis using the full master equation could uncover interesting properties, potentially leading to novel control mechanisms.
 
-The cell's drive towards homeostasis involves dealing with the inherently chaotic and noisy processes that reside within it.
+The cell's drive towards homeostasis requires dealing with the inherently chaotic and noisy processes that reside within it.
 Brownian motion, stochastic gene regulation, and mRNA dynamics collectively generate a fundamentally noisy environment that cells must effectively manage.
@@
 However, new control techniques that leverage the complete probability distribution information of the system can exploit the inherent noise of protein production to achieve better control.
-Previous control efforts primarily focused on deterministic ordinary differential equation (ODE) formulations of the control problem, only a limited number of studies have delved into analyzing the full chemical master equation to harness noise to improve the control of biological processes \cite{Szymanska2015,May2021}.
+Previous control efforts primarily focusted on deterministic ordinary differential equation (ODE) formulations of the control problem, only a limited number of studies have delved into analyzing the full chemical master equation to harness noise to improve the control of biological processes \cite{Szymanska2015,May2021}.
 
@@
 An extended model showed that addition of a positive auto-regulation module (Figure\ \ref{cartoons}A, bottom promoter) could help to maintain an elevated expression phenotype in the presence of UV excitation.
-
-Using discrete stochastic models based on the chemical master equation, it was demonstrated that combining biochemical noise, nonlinear auto-regulation, and a single optogenetic feedback could control two genetically identical cells with different initial conditions to reach different desired signals at different frequency and phase (Figure \ref{cartoons}B).
-Figure\ \ref{cartoons}C(left) presents heat maps for two control laws, which a single external UV excitation signal based on observed cellular state.
-Figure\ \ref{cartoons}C(right) shows the corresponding achievable marginal stationary distributions.
-Effective control was possible with both Fully Aware Feedback Controllers (FAC, Figure\ \ref{cartoons}C(top)) and Partially Aware Feedback Controllers (PAC, Figure\ \ref{cartoons}C(bottom)), even when the control model was an approximate simplification of the actual process dynamics.
-
-Using discrete stochastic models based on the chemical master equation, this study demonstrates simultaneous control of two genetically identical cells with different initial conditions, by exploiting noise, nonlinear auto-regulation, and using a single input.
-Heat maps in Figure \ref{cartoons}C(left) illustrate two control laws based on observed cellular activity, while Figure\ \ref{cartoons}C(right) shows the corresponding achievable marginal stationary distributions.
-Effective control was achieved using both Fully Aware Feedback Controllers (FAC, Figure\ \ref{cartoons}C(top)) and Partially Aware Feedback Controllers (PAC, Figure\ \ref{cartoons}C(bottom)).
 
@@
 In the 'Results' section, we explore the effects of model granularity Figure \ref{cartoons}D parameter inaccuracies, and time delays Figure \ref{cartoons}E on FAC and PAC performance and add a simple scheme for controlling cells to dynamically changing point.
+Using discrete stochastic models based on the chemical master equation, we demonstrate that combining biochemical noise, nonlinear auto-regulation, and a single optogenetic feedback could control two genetically identical cells with different initial conditions to reach different desired signals at different frequency and phase (Figure \ref{cartoons}B).
+
+Figure\ \ref{cartoons}C(left) presents heat maps for two control laws, which a single external UV excitation signal based on observed cellular state.
+Figure\ \ref{cartoons}C(right) shows the corresponding achievable marginal stationary distributions.
+Effective control was possible with both Fully Aware Feedback Controllers (FAC, Figure\ \ref{cartoons}C(top)) and Partially Aware Feedback Controllers (PAC, Figure\ \ref{cartoons}C(bottom)), even when the control model was an approximate simplification of the actual process dynamics.
 
@@
 
-\subsection{Dynamic setpoint control}
+\subsection{Dynamic Setpoint Control}
 
-static setpoint control is the control of a reference point which does not move over time. 
-dynamic setpont control targets a mvonig referenc peoint.
-synamis serpoint control is hard to deriuve, but the optimization of peicesice constangtof each static referenc pont is easy and can be stiched together.
+Although the control to a static set point is at the limits of currently feasible calculations, the extension to dynamically moving set point is much harder and requires untractible numerical simulation or further mathematical exploration.
+These solutions require numerical simulations which scale poorly with the number of states and quickly become impossible to perform.
+Therefore, a simple alternative is to sweep over a set of steady state controllers in Piecewise fashion that mimics a dynamically moving path.
+
+We developed a dynamic set point controller which periodically sweeps over a predefined path in state space in a discrete number of steps. Let the path $\bf{r}$ be given by
+\begin{equation}
+\bf{r}(x)=[r_1(x_n),r_2(x-N)]
+\end{equation}
+where $x_n \in [0,n]$, $r_1$ and $r_1$ an arbitrary periodic function that describes the desired target state for $x_1$ and $x_2$. 
+Then $x_n$ can be broken down into n steps
+\begin{equation}
+x_n(t)=mod(n*t/T,n)
+\end{equation}
+And the path can be walked over in time using
+\begin{equation}
+r(t)=\bf{r}(x_n(t))
+\end{equation}
 
@@
 
-systems driven at high frequency often have thier control performance reduced due to the response time scale of the system they are driving, but in a stochastic system the combination of high frequency dirving and noise could be catastriphic.
+High-frequency driven systems often experience reduced control performance due to the system's slow response.
+However, the combination of both high-frequency driving and noise could be catastrophic for system control performance.
+Conversely, increasing granularity may improve control performance but alters the system's time scale.
 
-Although the control of a dynamic reference point using a Piecewise controller is possible, the effect of cycling frequency is also important.
+To examine how driving frequency affects noise-exploiting control, Sixty-four SSA trajectories of the phase-separated dynamic controller were simulated over a two dimensional domain of points $(\alpha,f)$ for 16 cycles after reaching steady state.
+For each $(\alpha,f)$ pair the control performance of the steady state trajectories were calculated.
 
-Sixty-four SSA trajectories of the phase-separated dynamic controller were simulated over a two dimensional domain of points $(\alpha,f)$ for 16 cycles after reaching steady state.
-Increasing $\alpha$ while holding $f$ constant at $XXX$ showed improved control performance at low frequency but little change at high frequency Fig \ref{CRG}(D).
-Analysis of control performance using varying frequency but constant $\alpha$ showed the worst performance at moderate $f$.
+Fig \ref{CRG}(A) shows the desired reference of a system with $\alpha = 5.0$, and Fig \ref{CRG}(A) shows the mean of the system response when $f=1e-4$.
+Fig \ref{CRG}(C) Shows control performance over time as $\alpha$ is held at $5.0$ and $f$ increases from 1e-4 to 1e-2 to 1e-0 with an average RMSE of 6.4, 16.3, and 14.4 respectively.
+I creasing frequency while alpha is fixed at $0.5$, $1.0$ and $2.0$ shows that control performance from 12 to 6.3.
+
+Heat maps of control performance of $(f,\alpha)$ pairs showed the worst performance at moderate $f$ and that control performance was best when both granularity was high and frequency was low.
 

### 2023-06-26 paper.tex
update

@@
 
+The control of cells to two slowly changing dynamic reference signals using a single global input by the use of a noise-exploiting controller showed good control performance for a variety of signals.
+These analysis could be extended to include faster frequency by numerical calculation or by alternative error-probability adjustments.
+
 \begin{thebibliography}{10}

### 2023-06-27 paper.tex
Edits to Introduction

@@
 
-\title{Optogenetic Maxwell Demon to Exploit Intrinsic Noise and Control Cell Differentiation Despite Time Delays and Extrinsic Variability}
+\title{Exploiting Intrinsic Noise to Control Cell Differentiation Dynamics in the Presence of Time Delays, Inexact Models, and Extrinsic Variability}
 \maketitle
@@
 
-Noise is a common theme throughout many areas of science, including control theory XXX, gene regulation XXX, mRNA dynamics XXX, and synthetic biology XXX.
-Excellent work has been done to enhance the robustness of control processes against noise.
-However, these systems are often controlled in spite of noise rather than working with it.
-Many stochastic systems can be modeled using a noise-free approach with additive noise but, a detailed analysis using the full master equation could uncover interesting properties, potentially leading to novel control mechanisms.
+Uncertain fluctuations, or `noise,' is a common theme throughout many fields of engineering, and is a frequent concern when attempting to control any human-made system such as vehicles\cite{XXX}, production machinery\cite{XXX}, or chemical processes\cite{XXX}.
+Noise arrises from many sources, occasionally from quantum events and thermally-induced fluctuations but more commonly from unknown or uncharacterized physical processes, and as such, these fluctuations usually cannot be efficiently reproduced or predicted using purely physical or mechanistic models.
+Rather, statistical models are usually employed where key mechanisms or dynamics are subject to stochastic variations or inputs that act as a proxy for random or un-modeled fluctuations elsewhere in the system.
+In this context, much effort has been placed on enhancing the robustness of controlled processes, where uncertain or unpredictable variations in internal or external parameters are modeled as intrinsic or extrinsic noise \cite{XXX}.
+For many macroscopic, human-engineered systems, the resulting system dynamics can be modeled effectively simply by combining a deterministic (i.e., noise-free) model with additive noise (usually assumed to be Gaussian under arguments based on the central limit theorem)\ref{XXX}, and most current approaches seek to control the system to minimize variations that result from the noisy inputs.
 
-The cell's drive towards homeostasis requires dealing with the inherently chaotic and noisy processes that reside within it.
-Brownian motion, stochastic gene regulation, and mRNA dynamics collectively generate a fundamentally noisy environment that cells must effectively manage.
-Despite these challenges, cells generally demonstrate strong capability to control these noisy processes.
-However, stochastic control theory is not well-developed, despite having substantial knowledge on how to model them.
+Analyses of noise and control theory are equally relevant to understand the basic biological processes of gene transcription regulation\cite{XXX} or mRNA translation regulation \cite{XXX} or to modify these processes for practical use\cite{XXX}.
+However, at this mesoscopic scale of cellular biology, where transcription factors compete to activate or deactivate individual genes or where mRNA or protein molecules are present at just a few copies (or none at all) per cell, additive noise model is far less realistic.
+In this case, Brownian motion, discrete stochastic gene regulation, and mRNA dynamics collectively generate a fundamentally noisy environment that cells must effectively manage. 
+At this scale, the order or timing of a single reaction event (e.g., the binding of a transcription factor to a promoter) can have dramatic consequences that could last for several cellular generations (e.g., the activation or repression of a gene that promotes unfettered cell growth and differentiation).
+The cell's drive towards homeostasis requires dealing with the inherently chaotic and noisy processes that reside within it, and despite these challenges, cells generally demonstrate strong capability to survive these noisy processes.
+When seeking to understand how such systems evolve or react, the central limit theorem and the Gaussian noise may not apply, and a more detailed statistical analysis of probabilistic behavior is needed uncover hidden properties of cellular control mechanisms.
 
-Synthetic biology aims to develop modular\cite{Ng2019} and orthogonal \cite{Liu2018} components to sense and manipulate \cite{Sheets2020} complex logical systems, enabling them to exhibit a wide range of advanced biological behaviors\cite{Shin2020}.
-Advancements in optogenetics have enhanced our ability to reliably actuate embedded systems within cells, offering precise control\cite{Sheets2020,Baumschlager2017,Chen2020,Lillacci2018}.
+One emerging field that is particularly dependent on the integration of control and noise is synthetic biology, which aims to develop modular\cite{Ng2019} and orthogonal \cite{Liu2018} components to sense and manipulate \cite{Sheets2020} complex logical systems, enabling them to exhibit a wide range of advanced biological behaviors\cite{Shin2020}.
+Advances in optogenetics have enhanced the ability to reliably actuate embedded systems within cells, offering the potential to exert precise temporal and spatial control on cellular components\cite{Sheets2020,Baumschlager2017,Chen2020,Lillacci2018}.
 These developments have facilitated computer-programmable regulation of cellular protein production through external optogenetic inputs and smart microscope techniques\cite{Fox2021,Baumschlager2021,Lugagne2017}.
 These digital-synthetic actuators enable fine-tuned, computer-modulated control of cellular systems, previously unattainable, with faster response times compared to chemical diffusion\cite{Rullan2018, Baumschlager2017} .
-Classical and modern control methods like PID control and model predictive control have been implemented in such systems\cite{} which can control synthetic systems to different stable points.
+Classical and modern control methods like PID control and model predictive control have been implemented in such systems\cite{} to control synthetic systems to different stable points. 
 % XXX Missing Citation above.
-Theoretical works have shown that stochastic systems can be guided towards unstable points by oscillating around those points \cite{Guarino2020}.
-However, new control techniques that leverage the complete probability distribution information of the system can exploit the inherent noise of protein production to achieve better control.
-Previous control efforts primarily focusted on deterministic ordinary differential equation (ODE) formulations of the control problem, only a limited number of studies have delved into analyzing the full chemical master equation to harness noise to improve the control of biological processes \cite{Szymanska2015,May2021}.
 
-inspired by the genetic toggle switch from Kobayashi et al.\cite{Kobayashi:2004}, Szymanska et al.~\cite{Szymanska2015} showed that stochastic control was possible despite certain parameter errors and time delays due to maturation of fluorescent proteins and limited observation of the regulatory proteins.
-To explore this concept in the context of more realistic models and parameters, May et al.~\cite{May2021} fit a simplified stochastic model to reproduce data measured in Baumschlager, et al.\cite{} for the expression from the XXX promoter under the optogenetic control of a UV-activated T7 polymerase (see model in Figure\ \ref{cartoons}A, top promoter).
-An extended model showed that addition of a positive auto-regulation module (Figure\ \ref{cartoons}A, bottom promoter) could help to maintain an elevated expression phenotype in the presence of UV excitation.
+While most previous control efforts focussed on deterministic ordinary differential equation (ODE) formulations of the control problem, a few pioneering theoretical works have extended these analysis to include noise and have used stochastic simulations to show how systems can be guided towards unstable points by oscillating around those points \cite{Guarino2020}.
+It has recently been shown that new control techniques that leverage the complete probability distribution information of the system could actually harness the noise of single-cell gene regulation to achieve more complicated control objectives.
+For example, inspired by the genetic toggle switch from Kobayashi et al.\cite{Kobayashi:2004}, Szymanska et al.~\cite{Szymanska2015} showed that noise could be exploited to achieve independent control of multiple cells using a single input, even despite uncertain parameters or time delays due to maturation of fluorescent proteins or limited observation of the regulatory proteins.
+In May et al.~\cite{May2021}, we identified a simplified stochastic model to reproduce data measured in Baumschlager, et al.\cite{XXX} for the expression from the XXX promoter under the optogenetic control of a UV-activated T7 polymerase (see model in Figure\ \ref{cartoons}A, top promoter).
+We then proposed the addition of a positive auto-regulation (Figure\ \ref{cartoons}A, bottom promoter) to help maintain an elevated expression phenotype in the presence of UV excitation, and we demonstrated how a Single-Input-Multiple-Output (SIMO) multicellular controller could control multiple cells to arbitrary phenotypes using only a single input.
 
@@
 
-This paper extends the analysis of the Single-Input-Multiple-Output (SIMO) multicellular control problem by examining the impact of model uncertainties on FAC and PAC controllers.
-These uncertainties include coarse-grained system dynamics, parameter errors, and controller delays.
-Our methods involve formulating the chemical master equation to analyze cells in a state space, defining the controllers and their effects on cell dynamics, and optimizing the control input for desired performance.
-In the 'Results' section, we explore the effects of model granularity Figure \ref{cartoons}D parameter inaccuracies, and time delays Figure \ref{cartoons}E on FAC and PAC performance and add a simple scheme for controlling cells to dynamically changing point.
-Using discrete stochastic models based on the chemical master equation, we demonstrate that combining biochemical noise, nonlinear auto-regulation, and a single optogenetic feedback could control two genetically identical cells with different initial conditions to reach different desired signals at different frequency and phase (Figure \ref{cartoons}B).
+This paper extends the analysis of the SIMO multicellular control problem by examining the impact of model uncertainties and fluctuating control objectives on the control performance. 
+These uncertainties include coarse-grained approximations of the system dynamics, errors or extrinsic variations in the system parameter, and time delays between the observation of the cellular dynamics and actuation of the control process.
+In the following `Methods' we introduce our formulation of the chemical master equation to analyze the discrete stochastic distribution of cellular responses; we define multiple controllers and demonstrate the computation of their effects on cell dynamics; and we show how the control law can be optimized to improve performance.
+In the `Results' section, we explore the how model approximations, parameter inaccuracies, and time delays affect control performance, and we demonstrate a simple scheme for controlling cells to track a dynamically changing reference signal.
+Using discrete stochastic models based on the chemical master equation, we demonstrate that combining biochemical noise, nonlinear auto-regulation, and a single optogenetic feedback could control two genetically identical cells with different initial conditions to follow different desired trajectories at different frequencies and phases.
 
-Figure\ \ref{cartoons}C(left) presents heat maps for two control laws, which a single external UV excitation signal based on observed cellular state.
+%Figure\ \ref{cartoons}C(left) presents heat maps for two control laws, which a single external UV excitation signal based on observed cellular state.
 Figure\ \ref{cartoons}C(right) shows the corresponding achievable marginal stationary distributions.
@@
 \section{Methods}
-A gene regulation model consisting of six species was developed by May et al.
-\cite{May2021}.
-It described the light-activated association of two T7 split domains (species 1 and 2) to form an active T7 polymerase (species 3).
-These active polymerase could then associate with inactive T7 promoters (species 4), resulting in the formation of an active gene (species 5) that produced the desired protein product (species 6).
+In May et al.
+\cite{May2021}, we developed two models for the description of an optogenetically controlled gene expression system.
+These first model consisted of six species to describe the light-activated association of two T7 split domains (species 1 and 2) which combine to form an active T7 polymerase (species 3) under optogenetic excitation.
+These active polymerase could then associate with inactive T7 promoters (species 4), resulting in the formation of an active allele (species 5) that could then transcribe and translated to produce the desired protein product (species 6).
+The second, much simpler, single-species model was developed by assuming quasi-steady equilibrium for the first five species.
+Both models were independently parameterized using the same experimental data from Baumschlager et al.
+\cite{XXX}.
+Furthermore, an extension was made to each model to incorporate a secondary self-activated promoter-gene construct, where the expression rate was determined by a Hill function \ref{prodRate}.
+Through simulations, we showed that a feedback control law could be designed to force multiple cells to different and individually chosen equilibrium states using a single optogenetic control signal, and we also showed that although this control law was parametrized using the simple model, it could be used effectively to control the behavior of the more complicated system, thus demonstrating that control performance could remain high despite inaccuracies in the model.
+In the current work, we adopt the simpler of the two models and extend our analyses  to consider the effects of additional model inaccuracies, including coarse-grained model approximations, parameter errors, and time delays, and we also explore the possibility that a single optogenetic control signal could drive the system to track temporally-changing reference signals.
+Section \ref{sec:Model} introduces the model; Section \ref{sec:Stochastic} develops a Master Equation description of the models probabilistic dynamics; Section \ref{sec:Quantification} defines a control objective and optimizes that metric to obtain a baseline control law; and finally Sections \ref{sec:Scaling} and \ref{sec:Delay} introduces uncertainties into the model related t the granularity of the model approximation or the introduction of time delays, respectively.
 
-Additionally, a simple single-species model was developed, assuming quasi-steady equilibrium for the first five species.
-Both models were independently parameterized using experimental data from Baumschlager et al.
-\cite{XXX}.
-Furthermore, an extension was made to incorporate a secondary self-activated promoter-gene construct, where the expression rate was determined by a Hill function \ref{prodRate}.
-
-\subsection{Model}
+\subsection{Model}\label{sec:Model}
 To assess the impact of model approximations on the implementation of noise-enhanced control strategies, we begin with the one-species model proposed by May et al.
@@
 \ref{cartoons}A).
-This model comprises of two reactions: nonlinear protein production and degradation and it is defined by the following equation:
+This model comprises two reactions for production and degradation of the key protein.
+The nonlinear, UV-dependent production is defined by the following equation:
 \begin{equation}
@@
 %
-where $x$ is the instantaneous protein level; $\kappa$ is the maximum strength of the auto-regulation promoter; $\beta$ is the concentration at which auto-regulation promoter reaches half its strength; $\eta$ is the cooperativity in the auto-regulation promoter; $k_0$ is the promoter leakage rate; and $u(UV(t))$ is the T7 promoter strength.
+where $x$ is the instantaneous protein level; $\kappa$ is the maximum strength of the auto-regulation promoter; $\beta$ is the concentration at which auto-regulation promoter reaches its half maximal strength; $\eta$ is the cooperativity in the auto-regulation promoter; $k_0$ is the leakage rate from both promoters; and $u(UV(t))$ is the UV-dependent strength of the T7 promoter.
 Feedback enables the external modulation of the light input using the state of the system, thereby controlling the T7 promoter strength as a function of state rather than time and eliminating the explicit time dependance in $u(UV(t))$.
-
  Protein degradation is assumed to be a first order process with rate $\gamma$:
@@
 \end{equation}
-Parameters describing the auto-regulation promoter, $\kappa$, $\eta$, $\beta$, and $k_0$, and the degradation rate $\gamma$
-%were chosen to reproduce dynamics measured in \cite{Baumschlager2017} and 
-% XXX - I don't think that this is right.  Only the parameters u(UV) and gamma were fit to data.
-% MMM - yes that is right
-are presented in Table 1~\cite{May2021}.
+the baseline parameters describing the auto-regulation promoter, $\kappa$, $\eta$, $\beta$, and $k_0$, and the degradation rate $\gamma$ are presented in Table 1~\cite{May2021}.
 
 \begin{table}[]
-\caption{Model Parameters}
+\caption{Baseline Model Parameters}
 \begin{center}
@@
 \hline
-Parameter & Value   & units                     \\ \hline
-$\kappa$         & 0.406   & Molecules Per Minute \\
-$\beta$          & 20.0    & Molecules            \\
-$\eta $           & 8.00    & unit less          \\
-$k_0$           & 0.0001  & Molecules per Minute \\
-$\gamma$         & 0.0203  & per Minute         \\
+Parameter & Value   & units                     \\ \hline \hline
+$\kappa$         & 0.406   & Molecules Per Minute \\ \hline
+$\beta$          & 20.0    & Molecules            \\  \hline
+$\eta $           & 8.00    & unit less          \\  \hline
+$k_0$           & 0.0001  & Molecules per Minute \\  \hline
+$\gamma$         & 0.0203  & per Minute         \\  \hline
 $u$              & dynamic & Molecules per Minute \\ \hline
@@
 
-\subsection{Stochastic analyses of the model}
+\subsection{Stochastic analyses of the model}\label{sec:Stochastic}
 %Solving discrete stochastic representations of chemical dynamics is critical to the development of noise-exploiting controllers, and can be analyzed using the Stochastic Simulation Algorithm (SSA) , or the Finite State Projection (FSP).
@@
 To describe the discrete behavior of the above model for a population system with $N_{\rm c}$ cells, we define the current state of the system as the tuple of the non-negative numbers of proteins in each cell: $\mathbf{X}_i = [x_{1},x_{2},\ldots,x_{N_c}]_i\in \mathbb{Z}_{\ge 0}$, where the index $i$ denotes the enumeration of the state within the countably infinite set of all possible states, $\mathbf{X}_i\in\mathcal{X}=\{\mathbf{X}_1,\mathbf{X}_2,\ldots\}$.
-The stoichiometry vector for a reaction in any given cell is then defined as the change in state following that reaction event (e.g., $\mathbf{X}_i \rightarrow \mathbf{X}_i + \mathbf{s}_\mu$), where the $2N_c$ possible reactions are defined in pairs corresponding to production (odd indices) and degradation (even indices) as:
+The stoichiometry vector, $\mathbf{s}_{\mu}$, for reaction number $\mu$ is then defined as the change in state following that reaction event (e.g., $\mathbf{X}_i \rightarrow \mathbf{X}_i + \mathbf{s}_\mu$), where the $2N_c$ possible reactions are defined in pairs corresponding to production ($\mu\in\{1,3,5,\ldots\}$) and degradation ($\mu\in\{2,4,6,\ldots\}$) as:
 \begin{align}
@@
 
-\subsection{Quantification and optimization of control performance}
+\subsection{Quantification and optimization of control performance}\label{sec:Quantification}
 %To quantify the performance of the controlled system, Score metrics of steady state distributions were made in order to enable optimization of the system by adjusting controllers.
@@
 
-To quantify performance success, we define the performance error score, $J$, as the expected steady state squared Euclidean distance of the process from the specified target state, $\mathbf{T}$:
+To quantify performance success, we define the {\em steady state performance error}, $J$, as the expected steady state Euclidean distance of the process from the specified target state, $\mathbf{T}$:
  \begin{equation}
- J = E\{(\mathbf{X}-\mathbf{T})^2\}.
+ J = \mathbb{E}\{|\mathbf{X}-\mathbf{T}|_2\}.
  \end{equation}
-This score is easily calculated by applying a linear operation on steady state probability distribution $\mathbf{P}^*$ from Eq.\ \ref{SSDist}) as follows:
-{ \begin{align}
-J&= \lim_{t\rightarrow \infty}\mathbb{E}\{|\mathbf{X}(t)-\mathbf{T}|_2^2\}, \nonumber \\ 
-&=\sum_{i_1,i_2,\ldots} P^*(x_1=i_1,x_2=i_2,\ldots) \left[(i-\mathbf T_1)^2 + (j-\mathbf T_2)^2 +\ldots\right],\nonumber  \\
-&=\sum_{i,j,\ldots} P^*_{ij\ldots}(t)C_{ij\ldots} =\mathbf{C}\mathbf{P}^*,
+This score is easily calculated by applying a linear operation on steady state probability distribution $\mathbf{P}^*$ (Eq.\ \ref{SSDist}) as follows:
+\begin{align}
+J^2&= \lim_{t\rightarrow \infty}\mathbb{E}\{|\mathbf{X}(t)-\mathbf{T}|_2^2\}, \nonumber \\ 
+&=\sum_{i_1,i_2,\ldots} P^*(x_1=i_1,x_2=i_2,\ldots) \left[(i- T_1)^2 + (j- T_2)^2 +\ldots\right],\nonumber  \\
+&=\mathbf{C}\mathbf{P}^*,
 \label{Euclid}
-\end{align}} where $\mathbf{C}$ is simply a vector that contains the squared Euclidian distance of each state from the specified target T, i.e., $C_i = |\mathbf{X}_i-\mathbf{T}|_2^2$.
+\end{align}
+where $\mathbf{C}$ is simply a vector that contains the squared Euclidian distance of each state from the specified target T, i.e., $C_i = |\mathbf{X}_i-\mathbf{T}|_2^2$.
 As a result of this calculation, $J$ is a non-negative scalar that is zero if any only if $\mathbf{P}^*$ is a delta distribution located exactly at the target vector $\mathbf{T}$.
 
-We studied two controller designs: the fully aware controller ($\mathbf{u}^{FAC}$) that utilizes simultaneous protein count observations from both cells, and the partially aware controller ($\mathbf{u}^{PAC}$) that relies on observations from a single cell.
+We consider two controller designs: the fully aware controller ($\mathbf{u}^{FAC}$) that bases its control signal  on simultaneous protein count observations from both cells, and the partially aware controller ($\mathbf{u}^{PAC}$) that relies only on observations from a single cell:
 \begin{equation}
@@
 \end{equation}
-% XXX - Do we need to use 0 and 1.
-Why not use 1 and 2 so that this can match better to the formulaton above?
-I changed them to x1 and x2 here, but please chekc to make sure that this is sonsistent throughout text and figures.
-% MMM - Made the change to x1 x2
 where $x_1$ and $x_2$ are discrete integers greater than or equal to zero that represent the number of proteins in cell one and cell two.
-In \cite{May2021}, we optimized the controllers $\mathbf{u}^{FAC}$ and $\mathbf{u}^{PAC}$ to minimize $J$ and saved them as lookup tables.
+In \cite{May2021}, we optimized each of these controllers $\mathbf{u}^{FAC}$ and $\mathbf{u}^{PAC}$ to minimize $J$ for a single target state $\mathbf{T} = [XXX,XXX]$, and these controllers are assumed to be fixed for the current study unless noted otherwise.
 
-\subsection{Scaling for system granularity}
-Previous studies have addressed the computational challenge of analyzing discrete stochastic chemical kinetics by projecting the FSP onto lower-dimensional spaces using techniques such as time scale separations \cite{Peles2006}, Krylov subspaces \cite{Macnemara&Sidje?}, coarse meshes \cite{Munsky2008IEEE,Tapia2012CDC}, or principle orthogonal decompositions \cite{Vo2019}.
-However, using CME analyses is often hindered by an experimental limitations.
+\subsection{Scaling for system granularity}\label{sec:Scaling}
+%Previous studies have addressed the computational challenge of analyzing discrete stochastic chemical kinetics by projecting the FSP onto lower-dimensional spaces using techniques such as time scale separations \cite{Peles2006}, Krylov subspaces \cite{Macnemara&Sidje?}, coarse meshes \cite{Munsky2008IEEE,Tapia2012CDC}, or principle orthogonal decompositions \cite{Vo2019}.
+%
+%However, using CME analyses is often hindered by an experimental limitations.
 Gene expression quantification based on fluorescence intensities, such as using fluorescent protein reporters, typically captures relative changes in expression over time rather than providing direct molecular counts.
-% XXX - would be good to look for a few other examples outside of our own papers where the CME dimension is reduced by linear projections.
-
-It would be convenient to create a parameter that could increase or decrease the intrinsic noise of a system without changing the overall dynamics of the system.
+%% XXX - would be good to look for a few other examples outside of our own papers where the CME dimension is reduced by linear projections.
+%
+%It would be convenient to create a parameter that could increase or decrease the intrinsic noise of a system without changing the overall dynamics of the system.
 This would enable the analysis of models with more or less noise without the need of new unique models each time.
 A rescaling of the system could decrease the intrinsic noise while maintaining the same general behavior.
-To accomplish this rescaling, let $M$ denote the size of the origional unscaled system and let $M'$ denote the size of the newly scaled system.
+%To accomplish this rescaling, let $M$ denote the size of the origional unscaled system and let $M'$ denote the size of the newly scaled system.
 
-We define a granularity parameter ($\alpha=M'/M$) that linearly scales the species populations to increase ($\alpha>1$) or decrease ($\alpha<1$) the maximum populations size, while maintaining the dynamics and general behavior of the model.
+In realistic applications, models are never exact, but are often chosen as simplifications of known processes.
+For example, when analyzing discrete stochastic chemical kinetics, it is common to project the CME onto lower-dimensional spaces using finite state projection\cite{XXX}, time scale separations \cite{Peles2006}, Krylov subspaces \cite{Macnemara&Sidje?}, principle orthogonal decompositions \cite{Vo2019}, or other coarse meshes \cite{Munsky2008IEEE,Tapia2012CDC}.
+Similarly, measurements are also always inexact and in many cases may only provide information about relative change -- for example, although fluorescent proteins usually cannot be counted exactly, one may reasonably assume that a cell's total fluorescence intensity varies linearly with the fluorescent protein concentration.
+To explore how mismatches in assumed system scale (e.g., arising from model approximations or relative measurements) affect the controllability of the cellular process, we define a granularity parameter ($\alpha=M'/M$) that linearly scales each species' population to increase ($\alpha>1$) or decrease ($\alpha<1$), while maintaining the dynamics and general behavior of the model.
 To apply the granularity parameter, we assume that each propensity function, $w_{\mu}$, from Eqns.\ \ref{Props} is rescaled to a different level of discreteness by substituting
@@
 For example, the production and degradation of protein in cell one would become:
-\begin{equation}
-w'_1(\mathbf{X}) = u(\mathbf{X}/\alpha,t)  + \kappa \frac{(x_1/\alpha)^\eta}{(x_1/\alpha)^{\eta}+\beta^{\eta}} + k_0,\ w'_2(\mathbf{X}) = \gamma x_1/\alpha.
-\end{equation}
+\begin{align}
+w'_1(\mathbf{X}) &= u(\mathbf{X}/\alpha,t)  + \kappa \frac{(x_1/\alpha)^\eta}{(x_1/\alpha)^{\eta}+\beta^{\eta}} + k_0 \nonumber \\
+ w'_2(\mathbf{X}) &= \gamma x_1/\alpha.
+\end{align}
 % XXX - Please check this against what you did computationally. 
-
-We note that in transforming the propensity functions, the inputs to the controller must also be scaled by $1/\alpha$ before computing the appropriate functions of that state, e.g., ${u}^{\mathcal{C}} = {u}^{\mathcal{C}}(\mathbf{X}/\alpha)$.
+We note that in order to reuse a control law that has been defined for one level granularity and apply it to a system at another level of granularity, the inputs to the controller must also be scaled by $1/\alpha$ before computing the appropriate functions of that state, e.g., ${u}^{\mathcal{C}} = {u}^{\mathcal{C}}(\mathbf{X}/\alpha)$.
 Because identification of the original control formulation, $\mathbf{u}^{FAC}(x_1,x_2)$ and $\mathbf{u}^{PAC}(x_1)$ only considered integer values for $(x_1,x_2)$, fractional number inputs for after rescaling $(x_1/\alpha,x_2/\alpha)$ are handled by 2D cubic interpolation of the nearest values.
@@
 %\end{equation}
-Finally, to provide a consistency relative scoring, the definition of the performance score is also adjusted according to sale magnitudes, i.e.,:
-{\begin{align}
-J' &= \sum_{i=0}^{M'}  \sum_{j=0}^{M'}P^*(x_1=i,x_2=j) ((i/\alpha - \mathbf T_1)^2 + (j/\alpha -\mathbf T_2)^2),\nonumber \\
+Finally, to provide a consistent metric for relative scoring, the definition of the performance score is also adjusted according to scale magnitudes.
+For example, in the two cell system the new steady state performance error would become:
+\begin{align}
+(J^2)' &= \sum_{i=0}^{M'}  \sum_{j=0}^{M'}P^*(x_1=i,x_2=j) ((i/\alpha - \mathbf T_1)^2 + (j/\alpha -\mathbf T_2)^2),\nonumber \\
 & =\mathbf{C}'\mathbf{P}^*
 \label{EuclidV}
-\end{align}}
+\end{align}
 
-\subsection{Observation and actuation time delays}
+\subsection{Observation and actuation time delays}\label{sec:Delay}
 

### 2023-06-27 trash/paper2.tex
Edits to Introduction

--- 2023-06-26 paper.tex
@@
 
-Previous research has focused on robust control performance in the presence of noise, but understanding controllers that exploit noise remains incomplete. 
-Motivated by Maxwells-Demon, we previously proposed a difficult control problem which requires the exploitation of noise in a stochastic system to break symmetry between two signals.
-We found multiple such controllers which could exploit noise to break symmetry between two cells under a variety of system information.
+Much work has been done to make control performance robust to noise, but the development of controllers which use noise to their advantage is not fully understood.
+Motivated by optogentics, we proposed a difficult control problem which requires the exploitation of noise in a stochastic system to break symmetry between two signals.
+In previous work, we found multiple such controllers which could exploit noise to break symmetry between two cells under a variety of system information.
 This work extends that analysis to include stochastic systems that trace a moving target, are affected by time delay, intrinsic noise, or measurement uncertainty.
- We find that noise-exploiting controllers can remain highly effective despite coarse approximations to the model's scale or incorrect estimations of key model parameters, and these controllers can even retain performance for significant time delays.
-Together, these findings suggest that noise-exploiting control should be possible where models are always approximate, where parameters are always uncertain, and where observations are corrupted by errors.
-
 \end{abstract}
-Keywords: Stochastic Control, Gene Regulation, Optogenetics
+Keywords: Optogenetic control, Stochastic gene regulation, Synthetic biology
 
@@
 \section{Introduction}
+
+Excellent work has been done to make control problems robust to noise, but these control problems control their systems despite the noise rather than working with the noise.
+Many control problems set up in this way have less control performance than a noise free one, but It may be possible to use nonlinearity, and feedback control to improve control.
+
+robust control
+
+noise exploiting controlers were developed in may.2022 which use noise to thier advantage
+
+Maxwells demon
+
+optogenetics and noise in optogenetics
+Many biological systems are stochastic and require
+
+master equation
+
+SIMO control of optogentically motivated systems
+
 \begin{figure}
 \begin{center}
-\includegraphics[width=\columnwidth]{Cartoons.pdf}
+\includegraphics[width=\columnwidth]{CartoonAndController.pdf}
 \caption{{\bf Optogenetic control of multiple cells using a single light input.}
@@
 \end{figure}
-% XXX - Please modify figure as follows: (1) switch panels A and B to match caption.
-(2) Change title of panel B (previously panel A) to "External Observer, controller and actuator".
-(3) Change title of panel A (previously panel B) to "Internal gene regulation pathway".
-(4) Change title of panel C to "Feedback control law and resulting performance"
-% MMM - Made the edit
-
-Noise is a common theme throughout many areas of science, including control theory XXX, gene regulation XXX, mRNA dynamics XXX, and synthetic biology XXX.
-Excellent work has been done to enhance the robustness of control processes against noise.
-However, these systems are often controlled in spite of noise rather than working with it.
-Many stochastic systems can be modeled using a noise-free approach with additive noise but, a detailed analysis using the full master equation could uncover interesting properties, potentially leading to novel control mechanisms.
-
-The cell's drive towards homeostasis requires dealing with the inherently chaotic and noisy processes that reside within it.
-Brownian motion, stochastic gene regulation, and mRNA dynamics collectively generate a fundamentally noisy environment that cells must effectively manage.
-Despite these challenges, cells generally demonstrate strong capability to control these noisy processes.
-However, stochastic control theory is not well-developed, despite having substantial knowledge on how to model them.
-
-Synthetic biology aims to develop modular\cite{Ng2019} and orthogonal \cite{Liu2018} components to sense and manipulate \cite{Sheets2020} complex logical systems, enabling them to exhibit a wide range of advanced biological behaviors\cite{Shin2020}.
-Advancements in optogenetics have enhanced our ability to reliably actuate embedded systems within cells, offering precise control\cite{Sheets2020,Baumschlager2017,Chen2020,Lillacci2018}.
-These developments have facilitated computer-programmable regulation of cellular protein production through external optogenetic inputs and smart microscope techniques\cite{Fox2021,Baumschlager2021,Lugagne2017}.
-These digital-synthetic actuators enable fine-tuned, computer-modulated control of cellular systems, previously unattainable, with faster response times compared to chemical diffusion\cite{Rullan2018, Baumschlager2017} .
-Classical and modern control methods like PID control and model predictive control have been implemented in such systems\cite{} which can control synthetic systems to different stable points.
-% XXX Missing Citation above.
-Theoretical works have shown that stochastic systems can be guided towards unstable points by oscillating around those points \cite{Guarino2020}.
-However, new control techniques that leverage the complete probability distribution information of the system can exploit the inherent noise of protein production to achieve better control.
-Previous control efforts primarily focusted on deterministic ordinary differential equation (ODE) formulations of the control problem, only a limited number of studies have delved into analyzing the full chemical master equation to harness noise to improve the control of biological processes \cite{Szymanska2015,May2021}.
-
-inspired by the genetic toggle switch from Kobayashi et al.\cite{Kobayashi:2004}, Szymanska et al.~\cite{Szymanska2015} showed that stochastic control was possible despite certain parameter errors and time delays due to maturation of fluorescent proteins and limited observation of the regulatory proteins.
-To explore this concept in the context of more realistic models and parameters, May et al.~\cite{May2021} fit a simplified stochastic model to reproduce data measured in Baumschlager, et al.\cite{} for the expression from the XXX promoter under the optogenetic control of a UV-activated T7 polymerase (see model in Figure\ \ref{cartoons}A, top promoter).
-An extended model showed that addition of a positive auto-regulation module (Figure\ \ref{cartoons}A, bottom promoter) could help to maintain an elevated expression phenotype in the presence of UV excitation.
-
-%The solutions of the controllers was carried out by solving the chemical master equation (i.e., the forward Komogorov equation) to estimate the probability of possible cellular states.
-For a two cell, single-input-single output (SIMO) control problem, the FAC performed the best (figure\ \ref{cartoons}C(top)), and the PAC (figure\ \ref{cartoons}C(bottom)) 2nd, and pMPC controllers 3rd.
-In all cases, the controller's knowledge and explicit treatment of intrinsic noise was critical to improve the control performance, break symmetry, and enable the SIMO control of all cells.
-
-This paper extends the analysis of the Single-Input-Multiple-Output (SIMO) multicellular control problem by examining the impact of model uncertainties on FAC and PAC controllers.
-These uncertainties include coarse-grained system dynamics, parameter errors, and controller delays.
-Our methods involve formulating the chemical master equation to analyze cells in a state space, defining the controllers and their effects on cell dynamics, and optimizing the control input for desired performance.
-In the 'Results' section, we explore the effects of model granularity Figure \ref{cartoons}D parameter inaccuracies, and time delays Figure \ref{cartoons}E on FAC and PAC performance and add a simple scheme for controlling cells to dynamically changing point.
-Using discrete stochastic models based on the chemical master equation, we demonstrate that combining biochemical noise, nonlinear auto-regulation, and a single optogenetic feedback could control two genetically identical cells with different initial conditions to reach different desired signals at different frequency and phase (Figure \ref{cartoons}B).
-
-Figure\ \ref{cartoons}C(left) presents heat maps for two control laws, which a single external UV excitation signal based on observed cellular state.
-Figure\ \ref{cartoons}C(right) shows the corresponding achievable marginal stationary distributions.
-Effective control was possible with both Fully Aware Feedback Controllers (FAC, Figure\ \ref{cartoons}C(top)) and Partially Aware Feedback Controllers (PAC, Figure\ \ref{cartoons}C(bottom)), even when the control model was an approximate simplification of the actual process dynamics.
 
 \section{Methods}
-A gene regulation model consisting of six species was developed by May et al.
-\cite{May2021}.
-It described the light-activated association of two T7 split domains (species 1 and 2) to form an active T7 polymerase (species 3).
-These active polymerase could then associate with inactive T7 promoters (species 4), resulting in the formation of an active gene (species 5) that produced the desired protein product (species 6).
-
-Additionally, a simple single-species model was developed, assuming quasi-steady equilibrium for the first five species.
-Both models were independently parameterized using experimental data from Baumschlager et al.
-\cite{XXX}.
-Furthermore, an extension was made to incorporate a secondary self-activated promoter-gene construct, where the expression rate was determined by a Hill function \ref{prodRate}.
+In previous work~\cite{May2021}, a six-species gene regulation model was developed to describe light-activated association of two T7 split domains (species 1 and 2) to form active T7 polymerases (species 3) which could then in turn associate with inactive T7 promoters (species 4) to create an active gene construct (species 5) that would produce its eventual protein product (species 6).
+A second, far simpler single-species model was also developed by assuming quasi-steady equilibrium for the first five species.
+Both models were independently parametrized by fitting to the same experimental data from Baumschlager, et al.~\cite{XXX}, and then extended to include a secondary self-activated promoter-gene construct for which the expression rate was defined by a Hill function (see Eqn.\ \ref{prodRate} below).
+In \cite{May2021}, we showed that a control law identified using the simple model would work to achieve differential control for the more complex model, thus motivating our current study's efforts to explore further how control performance would change upon additional simplifying assumptions or errors in model specification.
+In the following subsections, we discuss our methods to formulate this model, show how it is analyzed numerically, and discuss the specification of the control objective and the determination of an optimal controller.
 
 \subsection{Model}
-To assess the impact of model approximations on the implementation of noise-enhanced control strategies, we begin with the one-species model proposed by May et al.
-\cite{May2021} (refer to Fig.
-\ref{cartoons}A).
-This model comprises of two reactions: nonlinear protein production and degradation and it is defined by the following equation:
+To examine how model approximations would affect the application of noise-enhanced control strategies, we start with the one species model from \cite{May2021} (see Fig.\ \ref{cartoons}A), which consists of two reactions: protein production and degradation.
+The nonlinear light-activated protein production rate is given by
  \begin{equation}
@@
 where $x$ is the instantaneous protein level; $\kappa$ is the maximum strength of the auto-regulation promoter; $\beta$ is the concentration at which auto-regulation promoter reaches half its strength; $\eta$ is the cooperativity in the auto-regulation promoter; $k_0$ is the promoter leakage rate; and $u(UV(t))$ is the T7 promoter strength.
-Feedback enables the external modulation of the light input using the state of the system, thereby controlling the T7 promoter strength as a function of state rather than time and eliminating the explicit time dependance in $u(UV(t))$.
-
+Through external feedback, the level of light excitation, and therefore the T7 promoter strength, $u(UV(t))$, can be modulated in response to the state of the system, thus removing explicit dependence on time.
+For example, in a system with the observation of two cells, one could define $u(UV(t)) = u^{\rm (FAC)}(x_1,x_2)$ as shown in Fig.\ \ref{cartoons}C(top left) for the fully aware control law, and for a system with the observation of only one cell, one could define $u^{\rm (PAC)}(UV(t)) = u(x_1)$ as shown in Fig.\ \ref{cartoons}C(bottom left) for the partially aware control law.
 Protein degradation is assumed to be a first order process with rate $\gamma$:
@@
 
-The stoichiometry and propensities allow us to use the Gillespie stochastic simulation algorithm\cite{Gillespie1992, Gillespie1977} to generate representative trajectories of the stochastic process.
-However, a direct analysis of the chemical master equation (CME) is necessary for performance quantification and optimal controller design.
-The CME, based on the same stoichiometry and propensity functions, is a linear ODE that describes the time-dependent changes in probability mass due to the specified reaction propensities and stoichiometries.
-It can be expressed as:
+With these stoichiometry and propensities, we can formulate and run the Gillespie stochastic simulation algorithm\cite{Gillespie1992,Gillespie1977} to generate representative trajectories of the stochastic process.
+However, quantifying performance and designing an optimal controller requires a more direct analysis of the chemical master equation (CME).
+For the same specifications of the stoichiometry and propensity functions, the CME is a linear ODE that describes how probability mass changes in time due to the specified reaction propensities and stoichiometries, which can be written:
 \begin{equation}
@@
 
-In the SIMO control of the multiple cell process, any input signal applied to one cell affects all cells in the population.
-Thus, it is not possible to independently control individual cells without impacting others.
-An effective controller must strike a balance among the desired behaviors of all cells in the system.
+For the single-input-multiple-output control of the multiple cell system, any input signal that is applied to one cell in the population will be felt by all cells in the population.
+As a result, it is not possible to independently control any individual cell without introducing changes to the behavior of the others.
+Therefore, an effective controller must be able to strike a balance among the desired behaviors of all cells in the system.
 
@@
 
-We studied two controller designs: the fully aware controller ($\mathbf{u}^{FAC}$) that utilizes simultaneous protein count observations from both cells, and the partially aware controller ($\mathbf{u}^{PAC}$) that relies on observations from a single cell.
+In this work, we explore two different controller designs that allow for the controller to use two different types of information.
+A fully aware controller ($\mathbf{u}^{FAC}$) uses direct observation of the protein count of two cells simultaneously while making its control input decision, while the partially aware controller $\mathbf{u}^{PAC}$) uses only the direct observation of a single cell:
 \begin{equation}
@@
 \begin{equation}
-\mathbf{u}^{PAC}=u(x_2)
+\mathbf{u}^{PAC}=u(x_1)
 \end{equation}
@@
 where $x_1$ and $x_2$ are discrete integers greater than or equal to zero that represent the number of proteins in cell one and cell two.
-In \cite{May2021}, we optimized the controllers $\mathbf{u}^{FAC}$ and $\mathbf{u}^{PAC}$ to minimize $J$ and saved them as lookup tables.
+In \cite{May2021}, the controllers $\mathbf{u}^{FAC}$ and $\mathbf{u}^{PAC}$ controllers were optimized to minimize $J$, and both $\mathbf{u}^{FAC}$ and $\mathbf{u}^{PAC}$ were saved as lookup tables after being optimized.
+In what follows, we will use these controllers exactly and without modification to explore how performance changes when the controllers are applied to systems that are approximations of the model under they were originally designed.
 
 \subsection{Scaling for system granularity}
-Previous studies have addressed the computational challenge of analyzing discrete stochastic chemical kinetics by projecting the FSP onto lower-dimensional spaces using techniques such as time scale separations \cite{Peles2006}, Krylov subspaces \cite{Macnemara&Sidje?}, coarse meshes \cite{Munsky2008IEEE,Tapia2012CDC}, or principle orthogonal decompositions \cite{Vo2019}.
-However, using CME analyses is often hindered by an experimental limitations.
-Gene expression quantification based on fluorescence intensities, such as using fluorescent protein reporters, typically captures relative changes in expression over time rather than providing direct molecular counts.
+One of the major computational obstacles in using the CME or FSP formulations to analyze discrete stochastic chemical kinetics is that the dimension of the FSP scales linearly with each chemical species, and therefore exponentially with the number of independent species.
+To circumvent this issue, several previous studies have projected the FSP onto lower dimensional spaces based using various techniques such as time scale separations\cite{Peles2006}, Krylov subspaces\cite{Macnemara&Sidje?}, coarse meshes\cite{Munsky2008IEEE,Tapia2012CDC}, or principle orthogonal decompositions\cite{Vo2019}.
 % XXX - would be good to look for a few other examples outside of our own papers where the CME dimension is reduced by linear projections.
+On the other hand, an experimental obstacle to using CME analyses is that gene expression quantification based on fluorescence intensities (e.g., using fluorescent protein reporters) usually does not provide molecular resolution, but rather measures only relative changes in expression over time.
 
-It would be convenient to create a parameter that could increase or decrease the intrinsic noise of a system without changing the overall dynamics of the system.
-This would enable the analysis of models with more or less noise without the need of new unique models each time.
-A rescaling of the system could decrease the intrinsic noise while maintaining the same general behavior.
-To accomplish this rescaling, let $M$ denote the size of the origional unscaled system and let $M'$ denote the size of the newly scaled system.
-
-We define a granularity parameter ($\alpha=M'/M$) that linearly scales the species populations to increase ($\alpha>1$) or decrease ($\alpha<1$) the maximum populations size, while maintaining the dynamics and general behavior of the model.
+To address how variations in model or observation scales might impact the effectiveness of noise-based control, we use the controller optimized at one resolution and ask how effective it would be when applied to systems with the same dynamics but with different resolution.
+To accomplish this rescaling, let $M$ denote some metric for the size of the original system (e.g., the average number of proteins at steady state and maximal UV) and let $M'$ denote the same metric for the system at a different resolution.
+Toward this goal, we define a granularity parameter ($\alpha=M'/M$) that linearly scales the species populations to increase ($\alpha>1$) or decrease ($\alpha<1$) the maximum populations size, while maintaining the dynamics and general behavior of the model.
 To apply the granularity parameter, we assume that each propensity function, $w_{\mu}$, from Eqns.\ \ref{Props} is rescaled to a different level of discreteness by substituting
@@
 % XXX - Please check this against what you did computationally. 
+After this rescaling, if $\alpha<1 $ the system dynamics will now occur over a smaller range of protein counts, and if $\alpha >1$ the system dynamics occurs over a larger range of protein counts.
 
-We note that in transforming the propensity functions, the inputs to the controller must also be scaled by $1/\alpha$ before computing the appropriate functions of that state, e.g., ${u}^{\mathcal{C}} = {u}^{\mathcal{C}}(\mathbf{X}/\alpha)$.
+We note that in transforming the propensity functions, the inputs to the controller have also be scaled by $1/\alpha$ before computing the appropriate functions of that state, e.g., ${u}^{\mathcal{C}} = {u}^{\mathcal{C}}(\mathbf{X}/\alpha)$.
 Because identification of the original control formulation, $\mathbf{u}^{FAC}(x_1,x_2)$ and $\mathbf{u}^{PAC}(x_1)$ only considered integer values for $(x_1,x_2)$, fractional number inputs for after rescaling $(x_1/\alpha,x_2/\alpha)$ are handled by 2D cubic interpolation of the nearest values.
@@
 \subsection{Observation and actuation time delays}
-
-Delays are inherent in optogenetic control systems due to biochemical reactions involved in observations, data analysis, decision making, and actuation dynamics.
-To investigate the effects of time delay on control actuation, we devised a time-delay stochastic simulation algorithm.
-This algorithm records the state history after each stochastic jump, enabling reconstruction of the population history of the species and control input.
+No real optogenetic control system can provide instantaneous feedback -- there will always be delays in biochemical reactions needed for observations (e.g., transcription and translation dynamics as well as folding and maturation of fluorescent proteins), delays in data analysis and decision making (e.g., image processing and calculation of the corresponding control signals), and delays in the actuation dynamics (e.g., activation of optogenetically excitable molecules).
+To explore the impacts of delays between the time a cellular fluctuation occurs and the time at which the corresponding control action can take effect, we developed a simple time-delay stochastic simulation algorithm.
+In this analysis, the state history is recorded following every stochastic event, allowing for reconstruction of the piecewise constant history of the species' populations, e.g., $x_1(s)$ and $x_2(s)$ for all $s\in [0,t]$.
 Using this information, the time-delayed control signal at time $t$ can be specified as:
-
 \begin{equation}
@@
 
-\subsection{Dynamic Setpoint Control}
-
-Although the control to a static set point is at the limits of currently feasible calculations, the extension to dynamically moving set point is much harder and requires untractible numerical simulation or further mathematical exploration.
-These solutions require numerical simulations which scale poorly with the number of states and quickly become impossible to perform.
-Therefore, a simple alternative is to sweep over a set of steady state controllers in Piecewise fashion that mimics a dynamically moving path.
-
-We developed a dynamic set point controller which periodically sweeps over a predefined path in state space in a discrete number of steps. Let the path $\bf{r}$ be given by
-\begin{equation}
-\bf{r}(x)=[r_1(x_n),r_2(x-N)]
-\end{equation}
-where $x_n \in [0,n]$, $r_1$ and $r_1$ an arbitrary periodic function that describes the desired target state for $x_1$ and $x_2$. 
-Then $x_n$ can be broken down into n steps
-\begin{equation}
-x_n(t)=mod(n*t/T,n)
-\end{equation}
-And the path can be walked over in time using
-\begin{equation}
-r(t)=\bf{r}(x_n(t))
-\end{equation}
-
 \section{Results}
@@
 \includegraphics[width=1\textwidth]{ParameterPerturbation.pdf}
-\caption{Parameter sweeps using the FAC and PAC show a broad range of control performance in cell 1 ({\bf A}),  in cell 2 ({\bf B}), and in both cells ({\bf C}).
-Columns show each parameter in the model, rows show the cell which has its parameter perturbed.
-}
+\caption{Parameter sweeps using the FAC and PAC show a broad range of control performance in cell 1 ({\bf A}),  in cell 2 ({\bf B}), and in both cells ({\bf C}). }
 \label{Parameter}
@@
 
-Real stochastic processes often have unknown parameter values, and even genetically identical single-cells often exhibit heterogeneity in parameters due to external variations.
-Parameter estimates can be obtained, but are not perfect measurements due to measurement limitations of stochastic processes.
-To asses the sensitivity of control performance to parameter errors, sensitivity analysis was performed for each model parameter of each cell, then on both cells simultaneously.
-For each parameter perturbation, both the $\mathbf{u}^{\rm FAC}$ and $\mathbf{u}^{\rm PAC}$ controllers were examined across a parameter range from one-tenth to ten-fold of the original values.
+Exact parameters values are typically unknown for real stochastic process, and even genetically identical single-cell systems can exhibit cell-to-cell heterogeneity in parameters due to extrinsic variations outside of the specific pathways of interest.
+Estimates of parameters can be taken, but not exact measurement.
+These parameter uncertainties could influence system performance.
+To explore the sensitivity of control performance to unknown errors in the model parameters, sensitivity analysis was performed for each model parameter and for the situation where the corresponding parameter was modified in either one cell at a time (i.e., to simulate extrinsic noise) or in both cells simultaneously (i.e., to simulate parameter errors).
+For each perturbation analysis, both the $\mathbf{u}^{\rm FAC}$ and $\mathbf{u}^{\rm PAC}$ controller were analyzed over a range of parameters from one-half to two-fold of the original parameters.
 % XXX - How hard would it be to extend this to 10 fold in either direction?
@@
 A 10 fold increase to a parameter can push all the probability to the top corner of the space
- Figure \ref{Parameter}A shows the results for these parameter sweeps when a single parameter in Cell 1 is fixed, and Fig.\ \ref{Parameter}B shows the control performance as single parameters in Cell 2 are fixed while all other parameters are held constant.
-Finally, \ref{Parameter}B  shows the performance change when the corresponding parameters are changed in both Cell 1 and Cell 2 at the same time while keeping other parameters fixed.
+ Figure \ref{Parameter}A shows the results for these parameter sweeps when a single parameter in Cell 1 is change while the parameters of Cell 2 is held fixed, and Fig.\ \ref{Parameter}B shows the opposite when changes Cell 1 has fixed parameters and Cell 2 is modified.
+Finally, \ref{Parameter}B  shows the performance change when the corresponding parameter is changed in both Cell 1 and Cell 2 at the same time.
 
-Modification of parameters was found to produce a broad range of effects.
+Modification to parameters was found to produce a broad range of effects, where some changes increase performance, some decrease performance, and others have little or no effect.
 For example, increasing $\beta$ in Cell 1 worsens performance while increasing $\beta$ in Cell 2 improves performance (compare second column in Fig.\ \ref{Parameter}A and \ref{Parameter}B).
@@
 In other cases (such as for $k_0$), the effect of parameter perturbations on performance is insignificant even for relatively large ($s$=2) perturbations.Similarly, Fig.\ \ref{Parameter}C shows that when parameters of both Cell 1 and Cell 2 are jointly changed, these changes could also improve or detract from control performance.
-
-Modification of parameters was found to produce a broad range of effects.
-Increasing $\beta$ in Cell 1 was found to worsen performance, while increasing $\beta$ in Cell 2 improved performance (Fig.
-\ref{Parameter}A, B, second column).
-The effects of parameter modification were not always monotonic.
-For example, increasing $\kappa$ in Cell 1 initially improved performance but reached a limit beyond which control performance degraded rapidly (Fig.
-\ref{Parameter}A, left column).
-In contrast, perturbations in the parameter $k_0$ had insignificant effects on performance, even for relatively large perturbations (s=10) (Fig.
-\ref{Parameter}A).
-Similarly, when parameters of both Cell 1 and Cell 2 were jointly changed, control performance could be either improved or decreased (Fig.
-\ref{Parameter}C).
-
 %It was found that the $\mathbf{u}^{\rm FAC}$ often outperformed the $\mathbf{u}^{\rm PAC}$, but for large $s$ in a few parameters the inverse was found (figure \ref{parameter} (Row III, E and At).
 
-Changing parameters had similar effects on control performance for both $\mathbf{u}^{\rm FAC}$ (solid lines) and $\mathbf{u}^{\rm PAC}$ (dashed lines).
-Parameters such as $\kappa$ and $\gamma$ reached minima that performed better than the original values, indicating potential for system optimization beyond the original design.
-However, altering $k_0$ had no noticeable impact, possibly due to its small original value requiring a larger scaling for observable changes.
-
+When examining the effects of changing parameters when using the $\mathbf{u}^{\rm FAC}$ (solid lines) or the $\mathbf{u}^{\rm PAC}$ (dashed lines), general trends typically remained the same in that changes to a parameter which cause a decrease in the control performance of the FAC also tend to decrease the performance of the PAC.
+Some parameters (e.g., $\kappa$, $\gamma$) reached minima that performed better than the original, which suggests that the physical system itself has room for optimization beyond the original design that could lead to better control performance, {\em even without changing the control law}.
+One explanation for why changes to $k_0$ had no effect on control performance is that the original parameter was so small that it may require a larger scaling on a different magnitude before changes are observed.
 % XXX - what is \eta?  From the table this should be the cooperativity with is very large.  Is this a mistake?
 % MMM - This was k_0, not eta
-In some parameters like $\kappa$ or $\gamma$, the PAC controller showed greater robustness to parameter errors than the FAC controller, even outperforming it.
-However, this effect is due to the mismatch between assumed and actual parameters, and re-optimizing both controllers for the new parameters will always yield better control performance for the FAC compared to the PAC.
+In a few circumstances, changes in the parameters $\kappa$ or $\gamma$ led to the surprising situation where the PAC controller, which only observes one cell at a time, was more robust to parameter errors than was the FAC controller, which can observe both cells, even to the extent that the PAC controller could sometimes outperform the FAC.
+We stress that this effect is due to the mismatch between the assumed and the actual parameters, and re-optimizing the two controllers for the new parameter combination will always yield better control results for the FAC compared to the PAC.
 
@@
 
-We first explore how the model defined for an origional system size (i.e., in which protein numbers have a range of $M$) when applied to a system with larger or smaller systems sizes (i.e., where protein numbers now have a range of $M' = \alpha M$.
+We first explore how the model defined for a given assumed system size (i.e., in which protein numbers have a range of $M$) when applied to a system with larger or smaller systems sizes (i.e., where protein numbers now have a range of $M' = \alpha M$.
 This modification could arise from two situations related to numerical convenience or inexact resolution in the measurement of concentrations.
@@
 For example, in measurements like those on \cite{Baumschlager} which are based on fluorescence protein intensities, it is not clear how many proteins correspond to what level of fluorescence intensity, and the scalar $\alpha$ is an unknown quantity that must be estimated (e.g., through calibration with single-molecule measurements or dilution experiment\cite{XXX - look for elowitz paper or something similar that tried to calibrate for single-molecule counts}).
+% MMM - Ding:2019??
+It is well established that the relative size of stochastic fluctuations (i.e., noise) in a chemical process decreases with the inverse square root of the process scale \cite{Kampen1961}.
+As the system size increases (with all species being scaled by the same factor), the dynamics approach those of a deterministic process with probability one (with zero-measure exceptions only for initial conditions lying on manifolds that separate attractors for different steady state behaviors)\cite{XXX}.
+For the purpose of noise-induced control, this introduces a tradeoff where noise is critical to break symmetry and enable differential control, yet noise is also deleterious to maintaining desired states once they have been achieved.
+Considering these competing objectives, it is not immediately clear how the reduction of noise through the addition of system granularity will affect control performance.
 
-% MMM - Ding:2019??
-Stochastic fluctuations in a chemical process decrease with the inverse square root of the process scale \cite{Kampen1961}.
-As the system size increases, the dynamics converge towards a deterministic process, except for certain exceptional initial conditions lying on manifolds that separate attractors for different steady state behaviors \cite{XXX}.
-This creates a tradeoff in noise-induced control, where noise enables differential control but is also deleterious to maintaining desired states once they have been achieved.
-The impact of reducing noise through increased system granularity on control performance remains unclear.
-
-To shed light on this tradeoff, control performance scores for the $\mathbf{u}^{\rm FAC}$ and $\mathbf{u}^{\rm PAC}$ controllers were calculated at different levels of granularity ($\alpha$) between 0.2 and 2.0.
+To shed light on this tradeoff, control performance scores for the $\mathbf{u}^{\rm FAC}$ and $\mathbf{u}^{\rm PAC}$ controllers were calculated at different levels of granularity ($\alpha$) between 0.1 and 2.0.
 % XXX - Why only go up to 2?  Did it become intractable?
 % MMM - Non_Sparse Calculations are intractible after 2, but sparse versions of code will work fine
-Figure \ref{Volume} (A - F) shows the joint (left plots) and marginal (right plots) distributions of the two cells relative to the specified target position (circles in joint distributions and vertical dashed lines in marginal distributions) for the FAC controller and for $\mathbf{\alpha}$=.2 (A,B), 1 (C,D) and 2 (E,F).
+Figure \ref{Volume} (Ai - Fi) shows the joint (left plots) and marginal (right plots) distributions of the two cells relative to the specified target position (circles in joint distributions and vertical dashed lines in marginal distributions) for the FAC controller and for $\mathbf{\alpha}$=.2 (A,B), 1 (C,D) and 2 (E,F).
 As the granularity changes from $\alpha$=0.2 , to $\alpha$=2, the distributions become focused at a point near the target, and the performance score improves from $J=$333 to 85 to 27, respectively.
@@
 \begin{center}
-\includegraphics[width=1\textwidth]{Granularity.pdf}
+\includegraphics[width=1\textwidth]{GranularityPerturbation.pdf}
 \vspace{-0.1in}
-\caption{Systems with increased granularity are less noisy and have better control performance.
-(A-C) Joint (left) and marginal (right) distributions for the FAC shows increased control performance and tighter distributions as $\alpha$ increases from 0.2 (top row) to 2.0 (bottom row) .
-(D-F) Joint (left) and marginal (left) distributions using the PAC controller at different levels of granularity.}
+\caption{Systems with high granularity were shown to have increased control performance but slower response times.}
 \label{Volume}
@@
 
-Figure \ref{Volume}G shows the trend of the performance versus alpha for both the FAC (solid cyan line) and the PAC (solid magenta line) controller.
+Figure \ref{Volume}M shows the trend of the performance score versus alpha for both the FAC (solid cyan line) and the PAC (solid magenta line) controller.
 This improvement in performance appears to approach a small value as the granularity goes to infinity, but since the size of the FSP increases with the square of the system size, systems much larger than $\alpha$=2 (where $\mathbf{A}_0\in \mathbb{R}^{10^4\times10^4}$) become more difficult to calculate.
@@
 
-To bypass this FSP limit, sixteen SSA simulations were simulated to steady state to sample the CME of a system with a significantly larger volume of $\alpha = 100$.
-Each SSA ran for $5\times10^7$ minutes, and the last $4\times10^7$ minutes were sampled for estimating the stationary distribution and calculating the performance score.
-The performance score estimates of this high granularity SSA, using the FAC and PAC controllers, were 4.13 and 27.5, respectively, shown as dashed lines in Fig.\ \ref{Volume}G.
-While it remains uncertain if further performance improvements can be achieved with larger system scales, these findings suggest that both controllers consistently improve with increased $\alpha$, with $\mathbf{u}^{\rm FAC}$ consistently outperforming $\mathbf{u}^{\rm PAC}$ in all cases examined.
-From figure H, it can be seen that a smaller $\alpha$ creates faster response but higher baseline score due to intrinsic noise.
-When $\alpha = 4$, the system becomes much less noisy and has lower scores but takes longer time to respond.
+We note however, that the improved steady state performance for systems with larger scales, comes at a cost that they also take longer to reach a steady state distribution.
+This tradeoff is illustrated in Fig.\ \ref{Score} which shows the performance score $J(t)$ as the system adapts to a change from the original target point $\mathbf{T} = [30,10]$ to its mirror point $\mathbf{T}_m = [10,30]$.
+% XXX - Please add a new figure to show this point.
+Start with an initial condition that is the steady state with target T and then flip it to T' = T(::-1).
+Then solve for P(t) and calculated J(t) from this.
+Finally, make a plot of J(t) versus t for the three cases \alpha = 0.2, 1, 2.
+One hundred SSA simulations of the flip-mirroring process were simulated while measuring the average score all simulations over time.
+Fig.\ \ref{Score} (A and B) shows the average score and protein counts of the simulation when $\alpha=1$, Fig.
+\ \ref{Score} (C,D) when $\alpha = 10$, and Fig.
+\ \ref{Score} (E,F) when $\alpha = 100$.
 
+From the figures it is clear that the smaller $\alpha$ creates faster dynamic but also have higher scores due to inherit noise.
+When $\alpha = 10$, the system becomes much less noisy and have lower scores but takes some time to respond.
+When $\alpha$ is too large at 10, the system contains very small levels of noise, but the dynamics
  % XXX - update this to explain what you see in this new figure.
@@
 \begin{center}
-\includegraphics[width=1\textwidth]{TimeDelay.pdf}
+\includegraphics[width=1\textwidth]{TimeDelayPerturbation.pdf}
 \vspace{-0.1in}
-\caption{Stochastic simulations of the time delayed system show decreasing control performance using either the FAC or PAC.
-(A-C) Joint (left) and marginal distributions (right) of the at different levels of time delay for the FAC.
-(D-F) Joint (left) and marginal distributions(right) at three levels of time delay using the PAC.
-(G) plots of RMSE control performance versus time delay for both FAC and PAC.}
+\caption{Stochastic simulations of the time delayed system show decreasing control performance using either controller. }
 \label{Time}
@@
 As the time required for any of these steps increases, control performance will be degraded, perhaps even leading to large fluctuations or instability.
-To explore how time delays would affect the noise-enhanced controller $\mathbf{u}^{\rm FAC}$ and $\mathbf{u}^{\rm PAC}$, we generated large sets of time-delayed stochastic simulations (see Methods) for different lengths of the time delay.
+To explore how time delays would affect the noise-enhance controller $\mathbf{u}^{\rm FAC}$ and $\mathbf{u}^{\rm PAC}$, we generated large sets of time-delayed stochastic simulations (see Methods) for different lengths of the time delay.
 Each SSA was sub-sampled for 1000 times over 10000 minutes of simulation time after a burnin period of 10000 minutes.
@@
 
-This data show that time delays exceeding $0.07 / \gamma = 0.0142$ minutes is detrimental to the FAC control performance while delays beyond $0.2 / \gamma = 0.0406$ minutes are detrimental to the PAC controller.
-The choice of best controller depends on the level of time delay in the system.
-At extreme time delays, both controllers lose their asymmetry, resulting in significantly worse scores ($J = 1041$ and $931$).
-The control performance of a no-feedback controller was measured which required no inputs Figure\ref{Time}(G) (red line) and achieved a score of $J = 402$.
-These findings indicate that, for large time delays, attempting to improve control with feedback is less effective than using a no-feedback approach.
+This data taken together show that time much larger that $0.07 / \gamma = 0.0142 min$ is detrimental to the FAC controller; delays beyond $0.2 / \gamma = 0.0406 min$ are detrimental for the PAC controller; and the best choice in controller is dependent the level of time delay in the system.
+For extreme levels of time delay, both systems lose their asymmetry, and their scores become much worse ($J=$1041 and 931).
+A no-feedback controller was developed which required no inputs.
+Such a controller is immune to time delay, and received a score $J=402$.
+These data show that at large time delays above , attempting to improve control with feedback is worse than no-feedback.
 
@@
 \vspace{-0.1in}
-\caption{Stochastic simulations of the FAC driven system and PAC driven system show different failure modes at at different $(\tau,\alpha)$ pairs.
-(A-I)Joint probability distributions at different $[\tau,\alpha]$ pairs.
-$\alpha$ across columns and $\tau$ across rows.
-(J) Control performance measured over a domain of $(\tau,\alpha)$ pairs show good control performance when both $\alpha$ is large and $\tau$ small.
-(K) Trajectories where good control was expected but not observed show oscillations instead.}
+\caption{Stochastic simulations of the FAC driven system (A) and PAC driven system (B) show different failure modes at at different $(\tau,\alpha)$ pairs (C).}
 \label{DG}
@@
 
-By increasing $\alpha$ we can change the characteristic of a system from being more noise-like to more deterministic-like.
-Similarly, increasing $\tau$ can decrease the quality of the control input.
-Generally, we would expect that increasing $\alpha$ would improve control performance, while increasing $\tau$ would decrease control performance.
-
-FAC steady state control performance was analyzed over a two dimensional domain of points $(\tau,\alpha)$ using thirty-two stochastic simulations simulated to steady state to see if the positive effect granularity could undo the effect of negative effect of time delay.
-Instead, noise induced oscillations were surprisingly found at certain $(\tau,\alpha)$ pairs and worse control performance was observed when system granularity increase for some $\tau$.
-Joint probability distributions at certain $(\tau,\alpha)$ pairs and their control performance can be seen in Fig \ref{DG}(A-I).
-Control performance reduced as $\alpha$ increased when $\tau=100$ but not at $\tau=1$.
-FAC control performance Fig \ref{DG}(J) over the domain of points shows a valley region at low $\tau$, and a plateau region at large $\tau$.
+Joint analysis of time delay and granularity together was performed to better understand if granularity can undo the effect of time delay.
+System control performance was analyzed over a two dimensional domain of points $(\tau,\alpha)$ using sixteen stochastic simulations for XXX minutes.
+These simulations were driven using the FAC controller Fig \ref{DG} (A), and PAC controller Fig \ref{DG} (B).
+The landscape of the FAC and PAC control performance Fig \ref {CtrlP}(A and B)over the domain shows a valley at low $\tau$, and a plateau at large $\tau$.
 This suggests that increasing system scale at large time delays is not beneficial to control performance.
-Trajectories at the edge of the valleyFig \ref {DG}(J)(yellow star) showed oscillatory behavior Fig \ref {DG}(K)and trajectories within the plateau show bursty behavior.
-
-The control performance domain Fig \ref{DG}(J) shows that low noise and high quality control inputs are both required to achieve good control.
-When noise is improved by increasing $\alpha$ the trajectories were observed going from very noisy to very oscillating which prevented improved control performance.
-Decreases in $\alpha$ should make the system respond faster and be less susceptible to time delay, but also more noisy.
-These competing effects in control performance can be observed in Fig \ref{DG}(J) which show that control performance is always poor at low $\alpha$.
+To better understand the appearance of the landscape, trajectories in different regions were compared.
+Trajectories at the edge of the plateau region (blue star) showed bursty behavior Fig \ref {CtrlP}(C)(blue lines), while trajectories at the edge of the valley Fig \ref {CtrlP}(C) (red lines) showed oscillatory behavior.
+Despite the small different in time delay between the red and blue trajectory, a stark change to the overall behavior was observed.
 
@@
 \vspace{-0.1in}
-\caption{The FAC and PAC control methods were used to optimize control inputs and probability distributions for static set points in a 2D domain.
-Optimized control inputs and probability distributions were optimized using the FAC control method(A and B) or the PAC control method (C and D).
-(E) Domains of control performance for the FAC method over a region of static target points show regions of good control and symmetry along the dashed line.
-(F) Domain of control performance over a set of static control points for the PAC control method.}
+\caption{Controllers optimized to a 2d domain of target points (A) show good control performance in the central region despite a broad range of targets (B and C). }
 \label{CtrlP}
@@
 
-We analyzed the control performance over a domain of target points to find if some regions are easier to control than others.
-Stability analysis of the auto-regulated system using a moderate light input can create up to four stable points and one unstable point in the state space.
-These points give rise to basins which may attract probability more easily than other points.
-Control performance of the stochastic system was analyzed over a domain of different set points and were optimized using more observable (FAC) or less observable (PAC) control.
-
-Controllers were optimized over a two dimensional domain of target points $[T_1,T_2]$ between between five and forty-five using the FAC input method or the PAC input method.
-Figure \ref{CtrlP} (A) shows the optimized control input of the system when the target is $T=[20,25]$ after optimizing the FAC.
-Figure \ref{CtrlP} (B) show the steady state probability distribution of the controlled system.
-Figure \ref{CtrlP} (C and D) show the optimized controller and probability distributions using the partially aware control method.
+Many controllers were optimized to different target points to determine if certain reference points were harder to control than others.
+Controllers were optimized over a two dimensional domain of target points $(T_1,T_2)$between between five and fourty-five.
+Figure \ref{CtrlP} (A) shows the control performance of the system after optimizing each controller to each point in the domain.
+Figure \ref{CtrlP} (B and C) show the optimized control input and the steady state of two points in the system.
 The unstable point in the rate equation matches the unstable point near $(20,20)$ in this domain.
-
-FAC and PAC control performance over the domain of static set points Fig\ref{CtrlP}(E) and (F) showed regions of good control in the FAC between 10 and 30 but a smaller region in the PAC.
-The FAC control performance was found to be symmetric along the dashed black line.
-The cause was found to be that symmetric pairs had control inputs which were exact transposes of each other.
-The unstable point of the system is known to be near [20,20] which is also a point of poor control for the FAC system with respect to local control performance.
-Edge regions were difficult to control, likely due to the geometry of how control inputs move the probability.
 
@@
 \vspace{-0.1in}
-\caption{Finite state projections of the slowly driven system demonstrate good control performance under a variety of phase and frequency scenario.
-(A) Schematic of the control problem shows targeting the number of protein in two cells using a dynamically changing reference point by cycling through a set of individual stable points and under synchonous, phasse-lagged or frequency changes reference signal using only one control input.
-(B1-BN) shows the phase space of the reference phase space, error residuals, reference time series, FSP solution and RMSE respectively for the synchonous control method.
-(C1-C5) Shows the same data for the phase-lagged control input.
-(D1-D5) Shows the same analysis for the frequency-changed control.}
+\caption{Finite state projections of the slowly driven system demonstrate good control performance under a variety of phase and frequency scenario. }
 \label{CR}
@@
 
-We developed a method to control the dynamics of a system by optimizing steady state controllers for a predefined path $r(x)$ in a Piecewise manner and cycling through them at a fixed frequency.
-Thirty-two controllers were optimized along a pathway in order to target an in-sync reference point \ref{CR}, a phase lagged reference point \ref{CR}, or frequency separated reference point\ref{CR}(A).
-The synchronous signal is given by
-\begin{equation}
- r_1(x)=[20 sin(2 \pi x)+10, 20 sin(2 \pi x)+10]
-\end{equation}
-the phase lagged signal is given by
-\begin{equation}
- r_2(x)=[20 sin(2 \pi x)+10, 20 cos(2 \pi x)+10]
-\end{equation}
-and the frequency changed signal is given by
-\begin{equation}
- r_3(x)=[20 sin(2 \pi x)+10, 20 sin(6 \pi x)+10]
-\end{equation}
-for $x\in[0,1]$ using 32 steps.
-FSP simulations of cyclically driven systems were slowly driven at a frequency of $1e-4$ seconds and their medians are shown in the Fig \ref{CR}(B4, C4, and D4)(red line, blue line).
-The time series probability distributions of $x_1$ and $x_2$ can be seen as red clouds and blue clouds respectively.
-Regions with purple clouds are the exact overlap of red and blue.
-Error distributions of the system were taken by shifting the probability distribution from the reference point to the origin Fig \ref{CR}(B2, C2, and D2).
-Root mean squared error of each simulation are shown in Fig \ref{CR}(B5, C5,D5).
-
-Suprisingly, the system is able to effectively control two signals to two different trajectories at different phase and frequency and using only a single control input despite the presence of noise Fig \ref{CR}(D4).
-Simulations show that the synchronous control performed the best with an RSME of $6.9$ despite moving over the unstable point in the system.
-When the system is driven with a phase-lag the RSME of the score increases to $8.2$ and when driven at a different frequency the RMSE goes up to $8.6$.
-Medians of the marginal distributions show good match with the desired reference signal in each control scheme Fig \ref{CR}(B4,C4, and D4).
-TIme series trajectories of the RMSE show spikes in the control performance using any control strategy.
+Dynamic reference targeting was performed by cycling though individual controllers which were optimized to a single target point at a fixed frequency.
+Thirty-two controllers were optimized along a pathway in order to target an in-sync reference point \ref{CR}(Synchronous), a phase lagged reference point \ref{CR}(Phase Lagged), or frequency separated reference point\ref{CR}(Phase Lagged and Frequency).
+The in phase signal is given by $r(t)=XXX$, the phase-lagged signal is given by $r(t)=XXX$, and the frequency driven signal is given by $r(t)=XXX$ FSP simulations of cyclically driven systems were slowly driven at a frequency of 1/5000 seconds and their medians are shown in the Fig \ref{CR}(A4, B4,C4)(red cloud, blue cloud).
+Root mean squared error of each simulation are shown in Fig \ref{CR}(A5, B5,C5).
+Simulations show that in-phase control performed the best with an RSME of $7.0$ despite moving over the unstable point in the system.
+When the system is driven with a phase-lag the RSME of the score increases to $8.3$ and when driven at a different frequency the RMSE goes up to $9.1$.
 
@@
 \begin{center}
-\includegraphics[width=1\textwidth]{Frequency.pdf}
+\includegraphics[width=1\textwidth]{FrequencyControl.pdf}
 \vspace{-0.1in}
-\caption{Control performance analyzed over a domain of $f,\alpha$ pairs show worst control performance at moderate frequencies near $1e-2$ Tracking reference signals when $\alpha=5$ span a range between 50 and 150 species (A).
-Stochastic simulations driven using a phased lagged controller at $\alpha = 5$ and low frequency show tighter control compared to $\alpha =1$ (B).
-Systems driven at moderate frequency show worse control performance than high frequency or low frequency (C).
-Control performance only of phase-lagged system only improves with increasing $\alpha$ and low $f$.}
+\caption{Stochastic simulations driven using a phased lagged controller show improved score at high granularity and low frequency.}
 \label{CRG}
@@
 \end{figure*}
+Although the control of a dynamic reference point using a Piecewise controller is possible, the effect of cycling frequency is also important.
 
-High-frequency driven systems often experience reduced control performance due to the system's slow response.
-However, the combination of both high-frequency driving and noise could be catastrophic for system control performance.
-Conversely, increasing granularity may improve control performance but alters the system's time scale.
-
-To examine how driving frequency affects noise-exploiting control, Sixty-four SSA trajectories of the phase-separated dynamic controller were simulated over a two dimensional domain of points $(\alpha,f)$ for 16 cycles after reaching steady state.
-For each $(\alpha,f)$ pair the control performance of the steady state trajectories were calculated.
-
-Fig \ref{CRG}(A) shows the desired reference of a system with $\alpha = 5.0$, and Fig \ref{CRG}(A) shows the mean of the system response when $f=1e-4$.
-Fig \ref{CRG}(C) Shows control performance over time as $\alpha$ is held at $5.0$ and $f$ increases from 1e-4 to 1e-2 to 1e-0 with an average RMSE of 6.4, 16.3, and 14.4 respectively.
-I creasing frequency while alpha is fixed at $0.5$, $1.0$ and $2.0$ shows that control performance from 12 to 6.3.
-
-Heat maps of control performance of $(f,\alpha)$ pairs showed the worst performance at moderate $f$ and that control performance was best when both granularity was high and frequency was low.
+16 SSA simulations of the phase-separated driving controller were performed on a system over a two dimensional domain of points $(\alpha,f)$ for a time of XXX minutes.
+Comparisons of control performance at $\alpha=0.5$ and $\alpha=2.0$ (Fig (C))show that control performance improves with system scale.
+16 SSA simulations of the high granularity system driven by phase-lag were performed over a range of frequency values.
+These analysis show that at low frequency the systems have a noise floor which decreases with increasing alpha, but was the system were driven rapidly, all systems became equally uncontrollable.
+At high frequency, the scores never got worse than $J=650$.
+At high frequency the control performance plateaus, while at lower frequency the system improves with increasing $\alpha$ Fig(E).
 
 \section{Conclusion}
-
 Key results from \cite{May2021} showed that noise was a key requirement for the development of a noise-exploiting controller, and that deterministic systems could not be controlled to two different stable points if they both started at the same initial condition.

### 2023-06-27 paper.tex
edits made to methods section.

@@
 
+\def\brian[#1]{{\color{red} #1 }}
+
 \title{Exploiting Intrinsic Noise to Control Cell Differentiation Dynamics in the Presence of Time Delays, Inexact Models, and Extrinsic Variability}
@@
 \begin{abstract}
-
 Previous research has focused on robust control performance in the presence of noise, but understanding controllers that exploit noise remains incomplete. 
@@
 \section{Methods}
-In May et al.
-\cite{May2021}, we developed two models for the description of an optogenetically controlled gene expression system.
+In May et al. \cite{May2021}, we developed two models for the description of an optogenetically controlled gene expression system.
 These first model consisted of six species to describe the light-activated association of two T7 split domains (species 1 and 2) which combine to form an active T7 polymerase (species 3) under optogenetic excitation.
-These active polymerase could then associate with inactive T7 promoters (species 4), resulting in the formation of an active allele (species 5) that could then transcribe and translated to produce the desired protein product (species 6).
+The active polymerase could then associate with inactive T7 promoters (species 4), resulting in the formation of an active allele (species 5) that could then transcribe and translated to produce the desired protein product (species 6).
 The second, much simpler, single-species model was developed by assuming quasi-steady equilibrium for the first five species. 
-Both models were independently parameterized using the same experimental data from Baumschlager et al.
-\cite{XXX}.
+Both models were independently parameterized using the same experimental data from Baumschlager et al. \cite{XXX}. 
 Furthermore, an extension was made to each model to incorporate a secondary self-activated promoter-gene construct, where the expression rate was determined by a Hill function \ref{prodRate}.
-Through simulations, we showed that a feedback control law could be designed to force multiple cells to different and individually chosen equilibrium states using a single optogenetic control signal, and we also showed that although this control law was parametrized using the simple model, it could be used effectively to control the behavior of the more complicated system, thus demonstrating that control performance could remain high despite inaccuracies in the model.
+Through simulations, we showed that a feedback control law could be designed to force multiple cells to different and individually chosen equilibrium states using a single optogenetic control signal. 
+We also showed that when this control law was parametrized using the simple model, it could be used effectively to control the behavior of the more complicated system, thus demonstrating that control performance could remain high despite inaccuracies in the model.
 In the current work, we adopt the simpler of the two models and extend our analyses to consider the effects of additional model inaccuracies, including coarse-grained model approximations, parameter errors, and time delays, and we also explore the possibility that a single optogenetic control signal could drive the system to track temporally-changing reference signals.
@@
 To assess the impact of model approximations on the implementation of noise-enhanced control strategies, we begin with the one-species model proposed by May et al.
-\cite{May2021} (refer to Fig.
+\cite{May2021} (Fig.
 \ref{cartoons}A).
@@
 \end{equation}
-the baseline parameters describing the auto-regulation promoter, $\kappa$, $\eta$, $\beta$, and $k_0$, and the degradation rate $\gamma$ are presented in Table 1~\cite{May2021}.
+All baseline parameters describing the auto-regulation promoter, $\kappa$, $\eta$, $\beta$, and $k_0$, and the degradation rate $\gamma$ are presented in Table 1~\cite{May2021} and are fixed throughout the current study.
 
@@
 \subsection{Stochastic analyses of the model}\label{sec:Stochastic}
-%Solving discrete stochastic representations of chemical dynamics is critical to the development of noise-exploiting controllers, and can be analyzed using the Stochastic Simulation Algorithm (SSA) , or the Finite State Projection (FSP).
-%To recast the above ODE formulation into the discrete stochastic representation, 
-To describe the discrete behavior of the above model for a population system with $N_{\rm c}$ cells, we define the current state of the system as the tuple of the non-negative numbers of proteins in each cell: $\mathbf{X}_i = [x_{1},x_{2},\ldots,x_{N_c}]_i\in \mathbb{Z}_{\ge 0}$, where the index $i$ denotes the enumeration of the state within the countably infinite set of all possible states, $\mathbf{X}_i\in\mathcal{X}=\{\mathbf{X}_1,\mathbf{X}_2,\ldots\}$.
-The stoichiometry vector, $\mathbf{s}_{\mu}$, for reaction number $\mu$ is then defined as the change in state following that reaction event (e.g., $\mathbf{X}_i \rightarrow \mathbf{X}_i + \mathbf{s}_\mu$), where the $2N_c$ possible reactions are defined in pairs corresponding to production ($\mu\in\{1,3,5,\ldots\}$) and degradation ($\mu\in\{2,4,6,\ldots\}$) as:
+To describe the discrete stochastic behavior of the above model for a population system with $N_{\rm c}$ cells, we define the current state of the system as the tuple of the non-negative numbers of proteins in each cell: $\mathbf{X}_i = [x_{1},x_{2},\ldots,x_{N_c}]_i\in \mathbb{Z}_{\ge 0}$, where the index $i$ denotes the enumeration of the state within the countably infinite set of all possible states, $\mathbf{X}_i\in\mathcal{X}=\{\mathbf{X}_1,\mathbf{X}_2,\ldots\}$.
+The stoichiometry vector, $\mathbf{s}_{\mu}$, for reaction number $\mu$ is then defined as the change in state following that reaction event (e.g., $\mathbf{X}_i \rightarrow \mathbf{X}_i + \mathbf{s}_\mu$).
+Specifically, the $2N_c$ possible reactions are defined in pairs corresponding to production ($\mu\in\{1,3,5,\ldots\}$) and degradation ($\mu\in\{2,4,6,\ldots\}$) as:
 \begin{align}
@@
 
-The stoichiometry and propensities allow us to use the Gillespie stochastic simulation algorithm\cite{Gillespie1992, Gillespie1977} to generate representative trajectories of the stochastic process.
-However, a direct analysis of the chemical master equation (CME) is necessary for performance quantification and optimal controller design.
-The CME, based on the same stoichiometry and propensity functions, is a linear ODE that describes the time-dependent changes in probability mass due to the specified reaction propensities and stoichiometries.
-It can be expressed as:
+These definitions of the stoichiometry and propensity functions allow us to implement the Gillespie stochastic simulation algorithm\cite{Gillespie1992, Gillespie1977} to generate representative trajectories of the stochastic process.
+
+\brian[Add a short paragraph explaining the basic SSA.
+This should include explicit statement of how the two RVs are generated: the time of the next reaction and the choice of which reaction occurs.
+This is needed since you will modify it below for the delayed reaction SSA.]
+
+However, a more direct analysis of the chemical master equation (CME) is necessary for performance quantification and optimal controller design.
+The high dimensional CME is a linear ODE that describes the time-dependent changes in probability mass of all possible states.
+Using the specified reaction propensities and stoichiometries, the CME can be expressed as:
 \begin{equation}
@@
 
-More specifically, the open-loop infinitesimal generator, $\mathbf{A}_0$, is constructed according to:
+More specifically, the zero-control infinitesimal generator, $\mathbf{A}_0$, is constructed according to:
 \begin{equation}
@@
 \end{equation}
-where ${u}^{\mathcal{C}}(\mathbf{X}_j)$ is the specification of the $\mathcal{C}$ controller in terms of the state, or its partial observations.
+where ${u}^{\mathcal{C}}(\mathbf{X}_j)$ is the specification of the controller, $\mathcal{C}$, in terms of the current instantaneous state, or its partial observations.
 
@@
 
+\brian[Add a paragraph introducing the FSP truncation and include some justification.]
+
 \subsection{Quantification and optimization of control performance}\label{sec:Quantification}
-%To quantify the performance of the controlled system, Score metrics of steady state distributions were made in order to enable optimization of the system by adjusting controllers.
-
-In the SIMO control of the multiple cell process, any input signal applied to one cell affects all cells in the population.
-Thus, it is not possible to independently control individual cells without impacting others.
-An effective controller must strike a balance among the desired behaviors of all cells in the system.
+Using optogenetic control, it is assumed that only a single light input would be applied at a given time and that it would be applied uniformly to multiple cells at the same intensity. 
+In this single-input-multiple-output (SIMO) control process, any input signal applied to one cell affects all cells in the population. 
+Thus, it is not possible to independently control individual cells without impacting others, so an effective controller must strike a balance among the desired behaviors of all cells in the system.
 
@@
 \begin{equation}
-\mathbf{u}^{FAC}=u(x_1,x_2)
+\mathbf{u}^{FAC}=u(x_1,x_2),
 \end{equation}
 \begin{equation}
-\mathbf{u}^{PAC}=u(x_2)
+\mathbf{u}^{PAC}=u(x_2),
 \end{equation}
 where $x_1$ and $x_2$ are discrete integers greater than or equal to zero that represent the number of proteins in cell one and cell two. 
-In \cite{May2021}, we optimized each of these controllers $\mathbf{u}^{FAC}$ and $\mathbf{u}^{PAC}$ to minimize $J$ for a single target state $\mathbf{T} = [XXX,XXX]$, and these controllers are assumed to be fixed for the current study unless noted otherwise.
+In \cite{May2021}, we optimized each of these two controllers to find the control laws $\mathbf{u}^{FAC}$ and $\mathbf{u}^{PAC}$ that minimize $J$ for a single target state $\mathbf{T} = [XXX,XXX]$. 
+These controllers are assumed to be fixed for the current study unless noted otherwise.
 
@@
 For example, when analyzing discrete stochastic chemical kinetics, it is common to project the CME onto lower-dimensional spaces using finite state projection\cite{XXX}, time scale separations \cite{Peles2006}, Krylov subspaces \cite{Macnemara&Sidje?}, principle orthogonal decompositions \cite{Vo2019}, or other coarse meshes \cite{Munsky2008IEEE,Tapia2012CDC}.
-Similarly, measurements are also always inexact and in many cases may only provide information about relative change -- for example, although fluorescent proteins usually cannot be counted exactly, one may reasonably assume that a cell's total fluorescence intensity varies linearly with the fluorescent protein concentration.
+Similarly, measurements are also always inexact and in many cases may only provide information about relative changes -- for example, although fluorescent proteins usually cannot be counted exactly, one may reasonably assume that a cell's total fluorescence intensity varies linearly with the fluorescent protein concentration.
 To explore how mismatches in assumed system scale (e.g., arising from model approximations or relative measurements) affect the controllability of the cellular process, we define a granularity parameter ($\alpha=M'/M$) that linearly scales each species' population to increase ($\alpha>1$) or decrease ($\alpha<1$), while maintaining the dynamics and general behavior of the model.
-To apply the granularity parameter, we assume that each propensity function, $w_{\mu}$, from Eqns.\ \ref{Props} is rescaled to a different level of discreteness by substituting
+To apply this granularity parameter, we assume that each propensity function, $w_{\mu}$ (Eqns.\ \ref{Props}) is rescaled to a different level of discreteness by substituting
 \begin{equation}
@@
 \end{align}
-% XXX - Please check this against what you did computationally. 
-We note that in order to reuse a control law that has been defined for one level granularity and apply it to a system at another level of granularity, the inputs to the controller must also be scaled by $1/\alpha$ before computing the appropriate functions of that state, e.g., ${u}^{\mathcal{C}} = {u}^{\mathcal{C}}(\mathbf{X}/\alpha)$.
-Because identification of the original control formulation, $\mathbf{u}^{FAC}(x_1,x_2)$ and $\mathbf{u}^{PAC}(x_1)$ only considered integer values for $(x_1,x_2)$, fractional number inputs for after rescaling $(x_1/\alpha,x_2/\alpha)$ are handled by 2D cubic interpolation of the nearest values.
+We note that in order to reuse a control law that has been defined for one level granularity and apply it to a system at another level of granularity, the inputs to the controller must also be scaled by $1/\alpha$ before computing the  control level assigned to the current state, e.g., ${u}^{\mathcal{C}} = {u}^{\mathcal{C}}(\mathbf{X}/\alpha)$.
+Because identification of the original control formulation, $\mathbf{u}^{FAC}(x_1,x_2)$ and $\mathbf{u}^{PAC}(x_1)$ only considered integer values for $(x_1,x_2)$, control signal values at fractional state values after rescaling $(x_1/\alpha,x_2/\alpha)$ are calculated using 2D cubic interpolation from the control values at the nearest integer state values.
 %\begin{equation}
@@
 (J^2)' &= \sum_{i=0}^{M'}  \sum_{j=0}^{M'}P^*(x_1=i,x_2=j) ((i/\alpha - \mathbf T_1)^2 + (j/\alpha -\mathbf T_2)^2),\nonumber \\
-& =\mathbf{C}'\mathbf{P}^*
+& =\mathbf{C}'\mathbf{P}^*.
 \label{EuclidV}
 \end{align}
+We reiterate that the system parameters and control law were defined and fixed using the the base granularity ($\alpha=1$), and these are not recomputed or refit upon changing the system granularity.
 
@@
 
-Delays are inherent in optogenetic control systems due to biochemical reactions involved in observations, data analysis, decision making, and actuation dynamics.
-To investigate the effects of time delay on control actuation, we devised a time-delay stochastic simulation algorithm.
+Delays are inherent in any realistic control systems, and it this case delays would be expected to arise due to the time needed for various biochemical reactions such as formation of complete polymerases, activation of promoters, transcription and transport of mature mRNA, and the translation and maturation of protein.
+Additional delays would also arise from data analysis, decision making, and actuation dynamics. 
+To investigate the effects of time delay on control actuation, we devised a simple time-delay stochastic simulation algorithm. 
 This algorithm records the state history after each stochastic jump, enabling reconstruction of the population history of the species and control input.  
@@
 where $\tau$ is the time delay between observation and actuation, and ${u}^{\mathcal{C}}$ is the previously optimized control law (e.g., $\mathbf{u}^{\rm FAC}$ or $\mathbf{u}^{\rm PAC}$). 
+When running the stochastic simulation algorithm under this fluctuating control signal, a slight modification is needed to the SSA.
+\brian[Explain how you modify the SSA for the delayed reaction SSA.  Look through the literature to cite someone else who has implenented a similar method.]
+
 We note that the time delay stochastic process was only simulated using the SSA because to our knowledge an appropriate direct FSP/CME integration procedure has not yet been developed.
@@
 
-\subsection{Dynamic Setpoint Control}
+\subsection{Dynamic Setpoint Control}\label{sec:Dynamic}
+Because optimizing the control law for a static set point as described in Section \ref{sec:Quantification} requires differentiation of the CME (Eq.
+\ref{XXX}) with respect to the control signal at each state, this calculation is approaching the limits of currently feasibility.
+Extending these calculations to optimize controllers for a dynamically moving set point is much harder and would likely require intractable numerical simulation or development of new mathematical approaches that are beyond the scope of the current study.
+To circumvent this challenge, we instead propose a simple alternative to sweep over a set of many different steady state controllers in piecewise fashion to mimic a dynamically moving path.
 
-Although the control to a static set point is at the limits of currently feasible calculations, the extension to dynamically moving set point is much harder and requires untractible numerical simulation or further mathematical exploration.
-These solutions require numerical simulations which scale poorly with the number of states and quickly become impossible to perform.
-Therefore, a simple alternative is to sweep over a set of steady state controllers in Piecewise fashion that mimics a dynamically moving path.
+\brian[I cannot understand the following section -- please rewrite.]
 

### 2023-06-27 paper.tex
started on results.

@@
 
-\subsection{Effects of Parameter Errors or Extrinsic Uncertainties}
+\brian[Add a paragraph or two to explain the big picture here.
+Use figure 1 to remind the reader of (1) the control goals, (2) the control law, (3) the resulting distributions, (4) the objective function and (5) the optimization of the control law to minimize the error.
+Refer to the methods for each such calculation.
+This section is critical to make sure the reader understands what is coming next.
+Then provide an outline of the remaining subsections of the results section so that the reader knows what to expect.]
+
+\subsection{Stochastic SIMO Optogenetic Control can Remain Effective Despite Small Parameter Errors or Extrinsic Uncertainties}
 \begin{figure*}
@@
 
-Real stochastic processes often have unknown parameter values, and even genetically identical single-cells often exhibit heterogeneity in parameters due to external variations.
-Parameter estimates can be obtained, but are not perfect measurements due to measurement limitations of stochastic processes.
-To asses the sensitivity of control performance to parameter errors, sensitivity analysis was performed for each model parameter of each cell, then on both cells simultaneously.
-For each parameter perturbation, both the $\mathbf{u}^{\rm FAC}$ and $\mathbf{u}^{\rm PAC}$ controllers were examined across a parameter range from one-tenth to ten-fold of the original values.
-% XXX - How hard would it be to extend this to 10 fold in either direction?
-% MMM-Changes to parameters can dramatically change the location of the probability distribution which is contained in a bounded (50,50) area.
-A 10 fold increase to a parameter can push all the probability to the top corner of the space
- Figure \ref{Parameter}A shows the results for these parameter sweeps when a single parameter in Cell 1 is fixed, and Fig.\ \ref{Parameter}B shows the control performance as single parameters in Cell 2 are fixed while all other parameters are held constant.
-Finally, \ref{Parameter}B  shows the performance change when the corresponding parameters are changed in both Cell 1 and Cell 2 at the same time while keeping other parameters fixed.
+Real stochastic processes always have unknown or uncertain mechanisms and parameter values, and although model structures and parameter estimates can often be obtained through fitting to training data, these estimates will never be perfect due to unavoidable measurement errors or limited data sets.
+Furthermore, even with plentiful and precise training data, model and parameter uncertainties are inevitable due to the fact that even genetically identical single-cells exhibit heterogeneity in parameters due to extrinsic variations.
+To asses the control performance to be expected under such parameter errors, we performed sensitivity analysis on each model parameter for each cell, then on both cells simultaneously. 
+For each parameter, both the $\mathbf{u}^{\rm FAC}$ and $\mathbf{u}^{\rm PAC}$ controllers were examined across a parameter perturbation range from one-tenth to ten-fold of the original values.
+Figure \ref{Parameter}A shows the results for these parameter sweeps when a single parameter in Cell 1 is varied (all other parameters are fixed), and Fig.\ \ref{Parameter}B shows the control performance as single parameters in Cell 2 is varied.
+Finally, \ref{Parameter}B shows the performance change when a single parameter is changed simultaneously in both Cell 1 and Cell 2 at the same time while keeping other parameters fixed.
+In each case the baseline parameters is denoted as a vertical dashed line, and lower values of the RMSE ($J$ in Eq. \ref{XXX}) denote improved performance. 
+Solid lines are used to show the performance using the fully aware controller (FAC) that has access to both cells, and the dashed lines show the performance for the partially aware controller (PAC) that only has access to Cell 1.
 
 Modification of parameters was found to produce a broad range of effects. 
-For example, increasing $\beta$ in Cell 1 worsens performance while increasing $\beta$ in Cell 2 improves performance (compare second column in Fig.\ \ref{Parameter}A and \ref{Parameter}B).
-In some cases, these effects were not monotonic; for example, increasing $\kappa$ in Cell 1 (Fig.\ \ref{Parameter}A, left column) would be highly advantageous up to a limit after which the control performance degrades rapidly.
-In other cases (such as for $k_0$), the effect of parameter perturbations on performance is insignificant even for relatively large ($s$=2) perturbations.Similarly, Fig.\ \ref{Parameter}C shows that when parameters of both Cell 1 and Cell 2 are jointly changed, these changes could also improve or detract from control performance.
-
-Modification of parameters was found to produce a broad range of effects.
-Increasing $\beta$ in Cell 1 was found to worsen performance, while increasing $\beta$ in Cell 2 improved performance (Fig.
-\ref{Parameter}A, B, second column).
-The effects of parameter modification were not always monotonic.
-For example, increasing $\kappa$ in Cell 1 initially improved performance but reached a limit beyond which control performance degraded rapidly (Fig.
-\ref{Parameter}A, left column).
-In contrast, perturbations in the parameter $k_0$ had insignificant effects on performance, even for relatively large perturbations (s=10) (Fig.
-\ref{Parameter}A).
-Similarly, when parameters of both Cell 1 and Cell 2 were jointly changed, control performance could be either improved or decreased (Fig.
-\ref{Parameter}C).
-
+For example, increasing $\beta$ in Cell 1 quickly worsens performance while increasing $\beta$ in Cell 2 improves performance (compare second column in Fig.\ \ref{Parameter}A and \ref{Parameter}B). 
+In some cases, the effects on performance are not monotonic; for example, increasing $\kappa$ in Cell 1 (Fig.\ \ref{Parameter}A, leftmost column) would be highly advantageous up to a limit after which the control performance degrades rapidly.
+In other cases (such as for the promoter leakage rate, $k_0$), the effect of parameter perturbations on performance is insignificant even for relatively large ($s$=10) perturbations, suggesting that this parameter is not important to control performance.
+Figure \ref{Parameter}C shows that even when parameters of both Cell 1 and Cell 2 are jointly changed, these changes could also improve or detract from control performance. 
+In particular, the analysis shows that control performance could be improved by modifying the system design either to increase the auto-regulation promoter strength ($\kappa$) or its cooperativity ($\eta$) or to decrease the promoter binding constant ($\beta$) or the protein degradation rate ($\gamma$).
 %It was found that the $\mathbf{u}^{\rm FAC}$ often outperformed the $\mathbf{u}^{\rm PAC}$, but for large $s$ in a few parameters the inverse was found (figure \ref{parameter} (Row III, E and At).
 
-Changing parameters had similar effects on control performance for both $\mathbf{u}^{\rm FAC}$ (solid lines) and $\mathbf{u}^{\rm PAC}$ (dashed lines).
-Parameters such as $\kappa$ and $\gamma$ reached minima that performed better than the original values, indicating potential for system optimization beyond the original design.
-However, altering $k_0$ had no noticeable impact, possibly due to its small original value requiring a larger scaling for observable changes.
+\brian[The remainder of this section is confusing and seems out of place.  Please revise to remove redundancies.]
 
-% XXX - what is \eta?  From the table this should be the cooperativity with is very large.  Is this a mistake?
-% MMM - This was k_0, not eta
-In some parameters like $\kappa$ or $\gamma$, the PAC controller showed greater robustness to parameter errors than the FAC controller, even outperforming it.
-However, this effect is due to the mismatch between assumed and actual parameters, and re-optimizing both controllers for the new parameters will always yield better control performance for the FAC compared to the PAC.
+In most cases, changing parameters had similar effects on control performance for both $\mathbf{u}^{\rm FAC}$ (solid lines) and $\mathbf{u}^{\rm PAC}$ (dashed lines).
+However, for some parameters like $\kappa$ or $\gamma$, the PAC controller showed greater robustness to parameter errors than the FAC controller, and the PAC outperforms the FAC under several parameter perturbation conditions (compare dashed and solid lines).
+We note that this effect is due entirely to the mismatch between the assumed and actual parameters - if one were to recalibrate the model and re-optimizing the controllers for the new parameters, the FAC controller would always yield better control performance compared to the PAC.
 

### 2023-07-06 paper.tex
update

@@
 However, at this mesoscopic scale of cellular biology, where transcription factors compete to activate or deactivate individual genes or where mRNA or protein molecules are present at just a few copies (or none at all) per cell, additive noise model is far less realistic.
-In this case, Brownian motion, discrete stochastic gene regulation, and mRNA dynamics collectively generate a fundamentally noisy environment that cells must effectively manage. 
+In this case, Brownian motion, discrete stochastic gene regulation, and noisy mRNA dynamics collectively generate a fundamentally stocastic environment that cells must effectively manage. 
 At this scale, the order or timing of a single reaction event (e.g., the binding of a transcription factor to a promoter) can have dramatic consequences that could last for several cellular generations (e.g., the activation or repression of a gene that promotes unfettered cell growth and differentiation).
@@
 
-These definitions of the stoichiometry and propensity functions allow us to implement the Gillespie stochastic simulation algorithm\cite{Gillespie1992, Gillespie1977} to generate representative trajectories of the stochastic process.
-
-\brian[Add a short paragraph explaining the basic SSA.
-This should include explicit statement of how the two RVs are generated: the time of the next reaction and the choice of which reaction occurs.
-This is needed since you will modify it below for the delayed reaction SSA.]
+These definitions of the stoichiometry and propensity functions allow us to implement the Gillespie stochastic simulation algorithm (SSA) \cite{Gillespie1992, Gillespie1977} to generate representative trajectories of the stochastic process.This algorithm uses discrete reactions and discrete molecule counts to sample the chemical master equation by utilizing an exponential random variable for time intervals and a categorical random variable for reaction selection.
+The exponential random variable is given by
+\begin{equation}
+\delta t(\mathbf{X},t)=log(rand \sum w_i(\mathbf{X},t) )
+\end{equation}
+where $\delta t$ is the time at which the next reaction occurs, and rand is a randomly generated number between 0 and 1.
+The probability of an individual reaction to fire from all possible reactions is given by
+\begin{equation}
+\zeta_i(\mathbf{X},t)=\frac{w_i(\mathbf{X},t)}{\sum w_i(\mathbf{X},t)}
+\end{equation}
+Where $\zeta_i(\mathbf{X},t)$ is the probability of the ith reaction to occur at state $\mathbf{X}$ and time $t$.
+Then the reaction that fired at time $t$ is a categorical random variable pulled from $\zeta_i(\mathbf{X},t)$.
+Then the SSA can be created by walking through time jumps and updating the state of the system by adding the stiochiometry of the reaction to the state of the system.
 
@@
 \end{equation}
-For convenience, the CME can also be formulated more compactly in matrix/vector form as:
+For convenience, the CME can also be formulated more compactly in matrix-vector form as:
 \begin{equation}
@@
 \end{array}\right.
-\label{InfB}
+\label{InfB}x
 \end{equation}
@@
 \brian[Add a paragraph introducing the FSP truncation and include some justification.]
+Finite state projections are a powerful way to simulate master equation analysis which enables the assessment of simulation accuracy, but these simulations have challenges measuring the steady state of the system since all probability will always enter error at long enough time scales.
+To enable control of steady state systems, a truncated FSP which reflects probability back into the system enables this calculation.
 
 \subsection{Quantification and optimization of control performance}\label{sec:Quantification}
-Using optogenetic control, it is assumed that only a single light input would be applied at a given time and that it would be applied uniformly to multiple cells at the same intensity. 
-In this single-input-multiple-output (SIMO) control process, any input signal applied to one cell affects all cells in the population. 
-Thus, it is not possible to independently control individual cells without impacting others, so an effective controller must strike a balance among the desired behaviors of all cells in the system.
+In the SIMO control of stochastic processes, a single input is applied simultaneously to all systems at once and a control signal that acts beneficially on one system destabilizes other systems.
+Stochastic SIMO control inputs which seek to control cells one-cell-at-a-time create increased instability and response variability in the other signals not being controlled.
+Thus, even in systems which do a good job at controlling a system cannot independently control individual cells to a desired target point without impacting the variance of others, so an effective controller must strike a balance among the desired behaviors of all cells in the system.
 
@@
  \end{equation}
-This score is easily calculated by applying a linear operation on steady state probability distribution $\mathbf{P}^*$ (Eq.\ \ref{SSDist}) as follows:
+This score is calculated by applying a linear operator on steady state probability distribution $\mathbf{P}^*$ (Eq.\ \ref{SSDist}) as follows:
 \begin{align}
-J^2&= \lim_{t\rightarrow \infty}\mathbb{E}\{|\mathbf{X}(t)-\mathbf{T}|_2^2\}, \nonumber \\ 
+J&= \lim_{t\rightarrow \infty}\mathbb{E}\{|\mathbf{X}(t)-\mathbf{T}|_2^2\}, \nonumber \\ 
 &=\sum_{i_1,i_2,\ldots} P^*(x_1=i_1,x_2=i_2,\ldots) \left[(i- T_1)^2 + (j- T_2)^2 +\ldots\right],\nonumber  \\
@@
 where $x_1$ and $x_2$ are discrete integers greater than or equal to zero that represent the number of proteins in cell one and cell two.
-In \cite{May2021}, we optimized each of these two controllers to find the control laws $\mathbf{u}^{FAC}$ and $\mathbf{u}^{PAC}$ that minimize $J$ for a single target state $\mathbf{T} = [XXX,XXX]$. 
-These controllers are assumed to be fixed for the current study unless noted otherwise.
+
+Optimization of $\mathbf{u}^{FAC}$ was performed using a gradient descent method to minimize $J$.
+This method requires calculating the negative gradient $-\frac{dJ}{d\mathbf{u}^{FAC}}$ and walking in a small step $du$ in that direction.
+The calculation of the gradient is given by
+\begin{equation}
+\frac{dJ}{d\mathbf{u}^{FAC}}=C \frac{dP}{d\mathbf{u}^{FAC}}
+\end{equation}
+Where $\frac{dP}{d\mathbf{u}^{FAC}}$ can be solved using general minimized residual calculation of
+\begin{equation}
+A \frac{dP}{d\mathbf{u}^{FAC}}=B \mathbf{u}^{FAC}
+\end{equation}
+to find $\frac{dP}{d\mathbf{u}^{FAC}}$.
+Optimization of $\mathbf{u}^{PAC}$ was done by restricting $\frac{dJ}{d\mathbf{u}^{FAC}}$ in ambient space to search a low dimensional subspace with a linear lift operator D.
+Then the subspace search is given by.
+\begin{equation}
+\frac{dJ}{d\mathbf{u}^{PAC}}=C \frac{dP}{d\mathbf{u}^{PAC}} = C D \frac{dP}{d\mathbf{u}^{FAC}}  
+\end{equation}
+Root mean squared error was calculated by taking the square root of the score.
+\begin{equation}
+RMSE=\sqrt{J}
+\end{equation}
+Since the square root is a monotonically increasing function, minimizing J is the same as minimizing RMSE.
+
+In \cite{May2021}, we optimized each of these two controllers to find the control laws $\mathbf{u}^{FAC}$ and $\mathbf{u}^{PAC}$ that minimize $J$ for a single target state $\mathbf{T} = [10,30]$, but in this work extend it to ant target point in the state space.
 
@@
 
-Delays are inherent in any realistic control systems, and it this case delays would be expected to arise due to the time needed for various biochemical reactions such as formation of complete polymerases, activation of promoters, transcription and transport of mature mRNA, and the translation and maturation of protein.
+Delays are inherent in any realistic control systems, and it this case delays would be expected to arise due to the time needed for various biochemical reactions such as formation of complete polymerases, activation of promoters, transcription and transport of mature mRNA, and the translation and maturation of protein \cite{Cai2007}.
 Additional delays would also arise from data analysis, decision making, and actuation dynamics. 
@@
 where $\tau$ is the time delay between observation and actuation, and ${u}^{\mathcal{C}}$ is the previously optimized control law (e.g., $\mathbf{u}^{\rm FAC}$ or $\mathbf{u}^{\rm PAC}$). 
-When running the stochastic simulation algorithm under this fluctuating control signal, a slight modification is needed to the SSA.
 \brian[Explain how you modify the SSA for the delayed reaction SSA.  Look through the literature to cite someone else who has implenented a similar method.]
-
 We note that the time delay stochastic process was only simulated using the SSA because to our knowledge an appropriate direct FSP/CME integration procedure has not yet been developed.
@@
 Because optimizing the control law for a static set point as described in Section \ref{sec:Quantification} requires differentiation of the CME (Eq.
-\ref{XXX}) with respect to the control signal at each state, this calculation is approaching the limits of currently feasibility.
+\ref{XXX}) with respect to the control signal at each state, this calculation is approaching the limits of current feasibility.
 Extending these calculations to optimize controllers for a dynamically moving set point is much harder and would likely require intractable numerical simulation or development of new mathematical approaches that are beyond the scope of the current study.
-To circumvent this challenge, we instead propose a simple alternative to sweep over a set of many different steady state controllers in piecewise fashion to mimic a dynamically moving path.
+To circumvent this challenge, we instead propose a simple alternative to sweep though a set of controllers over time which target different static points in piecewise fashion to mimic a dynamically moving path.
 
-\brian[I cannot understand the following section -- please rewrite.]
-
-We developed a dynamic set point controller which periodically sweeps over a predefined path in state space in a discrete number of steps. Let the path $\bf{r}$ be given by
+ We start by describing a path $\bf{p}_n(x)$ in state space which walks over a two dimensional path in a discrete number of steps.
+Then, optimize a controller $\bf{c}_n(x)$for each target point and cycling through the controllers over time to create $\mathbf{r}(x,t)$.
+Let  $\bf{p}_n(x)$ be a path taken in $n$ steps, given by
 \begin{equation}
-\bf{r}(x)=[r_1(x_n),r_2(x-N)]
+\bf{p}_n(x)=[p_1,...,p_n]
 \end{equation}
-where $x_n \in [0,n]$, $r_1$ and $r_1$ an arbitrary periodic function that describes the desired target state for $x_1$ and $x_2$. 
-Then $x_n$ can be broken down into n steps
+where each $p_i$ is a pair of points on a 2D domain and $\bf{p}_n$ is a set of point pairs. Then, we optimize a set of controllers for each pair in the path.
 \begin{equation}
-x_n(t)=mod(n*t/T,n)
+\bf{c}_n(x)=[c_1,...,c_n]
 \end{equation}
-And the path can be walked over in time using
+where each $\mathbf{c}_i$ is a matrix of control inputs which targets $\bf{p}_i(x)$.
+Then, we cycle through each controller at a fixed frequency by changing the controller index over time. The indexing function is given by 
 \begin{equation}
-r(t)=\bf{r}(x_n(t))
+i(t)=\lceil \frac{n}{T} mod(t,T)\rceil = \lceil n \ f \ mod(t,\frac{1}{f})\rceil
 \end{equation}
-
+where $T$ is the cycling period, $f$ the cycling frequency and $n$ the number of points in the path. Then the cycling controller can be created by 
+ \begin{equation}
+\mathbf{r}(\mathbf{x},t)=\bf{c}_{i(t)}(x)
+\end{equation}
+Where $\mathbf{r}(\mathbf{x},t)$ is a function of state and time. This arbitrary form of $\mathbf{r}(\mathbf{x},t)$, enables cycling through controllers along different paths at different frequencies.
 \section{Results}
@@
 Then provide an outline of the remaining subsections of the results section so that the reader knows what to expect.]
+
+We developed a method to create a noise-exploiting controller by taking full advantage of master equation analysis.
+This method uses non-linearity in a model, feedback control of the system and the exploitation of noise to optimize a Markovian controller which minimizes the expected squared Euclidian distance from the target position.
+Despite the advantages of full master equation analysis, good control of stochastic systems would ideally be resistant to parameter variability from heterogenous cell populations, time delay in controller actuation, and robust to systems with intrinsically more or less noise.
+To answer these questions, we will examine how parameter changes, time delay, and system scale influence the control performance of the controllers.
+Then, we extend these analyses to control a dynamically moving set point which changes its phase and frequency.
 
@@
 Furthermore, even with plentiful and precise training data, model and parameter uncertainties are inevitable due to the fact that even genetically identical single-cells exhibit heterogeneity in parameters due to extrinsic variations.
-To asses the control performance to be expected under such parameter errors, we performed sensitivity analysis on each model parameter for each cell, then on both cells simultaneously. 
+To asses the control performance under such parameter errors, we performed sensitivity analysis on each model parameter for each cell, then on both cells simultaneously. 
 For each parameter, both the $\mathbf{u}^{\rm FAC}$ and $\mathbf{u}^{\rm PAC}$ controllers were examined across a parameter perturbation range from one-tenth to ten-fold of the original values.
@@
 Finally, \ref{Parameter}B shows the performance change when a single parameter is changed simultaneously in both Cell 1 and Cell 2 at the same time while keeping other parameters fixed.
-In each case the baseline parameters is denoted as a vertical dashed line, and lower values of the RMSE ($J$ in Eq. \ref{XXX}) denote improved performance. 
+In each case the baseline parameters is denoted as a vertical dashed line, and lower values of the RMSE ( Eq. \ref{XXX}) denote improved performance. 
 Solid lines are used to show the performance using the fully aware controller (FAC) that has access to both cells, and the dashed lines show the performance for the partially aware controller (PAC) that only has access to Cell 1.
@@
 In particular, the analysis shows that control performance could be improved by modifying the system design either to increase the auto-regulation promoter strength ($\kappa$) or its cooperativity ($\eta$) or to decrease the promoter binding constant ($\beta$) or the protein degradation rate ($\gamma$).
-%It was found that the $\mathbf{u}^{\rm FAC}$ often outperformed the $\mathbf{u}^{\rm PAC}$, but for large $s$ in a few parameters the inverse was found (figure \ref{parameter} (Row III, E and At).
 
-\brian[The remainder of this section is confusing and seems out of place.  Please revise to remove redundancies.]
-
-In most cases, changing parameters had similar effects on control performance for both $\mathbf{u}^{\rm FAC}$ (solid lines) and $\mathbf{u}^{\rm PAC}$ (dashed lines).
-However, for some parameters like $\kappa$ or $\gamma$, the PAC controller showed greater robustness to parameter errors than the FAC controller, and the PAC outperforms the FAC under several parameter perturbation conditions (compare dashed and solid lines).
-We note that this effect is due entirely to the mismatch between the assumed and actual parameters - if one were to recalibrate the model and re-optimizing the controllers for the new parameters, the FAC controller would always yield better control performance compared to the PAC.
-
-To assess the potential robustness of noise-enhanced control of the proposed synthetic auto-repression toggle switch, we adopted the model  from \cite{May2021} as well as the fully aware and partially aware controllers, $u^{\rm FAC}(x_1,x_2)$ and  $u^{\rm FAC}(x_1)$, respectively as described in the methods section.
-In the following subsections, we explore how well these controllers would perform in the more realistic setting where the approximate model used to define the controller no longer matches exactly to the dynamics of the process that the controller is being used to manipulate.
-
-We first explore how the model defined for an origional system size (i.e., in which protein numbers have a range of $M$) when applied to a system with larger or smaller systems sizes (i.e., where protein numbers now have a range of $M' = \alpha M$.
-This modification could arise from two situations related to numerical convenience or inexact resolution in the measurement of concentrations.
-First, from the perspective of numerical convenience when optimizing a controller, it can be advantageous to reduce the assumed range of protein levels in order to reduce the dimension of the CME and FSP analyses, thereby achieving a more tractable optimization problem.
-For example, the current model (see Methods) takes 19.3 seconds to solve the FSP analysis when $\alpha=1$ but the same analysis takes 24619 seconds to solve when $\alpha=10$.
-Since numerical optimization of the controller can take many millions of calculation for different control laws, learning a controller using $\alpha=1$ that works for $\alpha=10$ would yield substantial savings in computational effort.
-The second situation under which sensitivity to system scaling could become important is when measurements are relative and do not provide single-molecule resolution.
-For example, in measurements like those on \cite{Baumschlager} which are based on fluorescence protein intensities, it is not clear how many proteins correspond to what level of fluorescence intensity, and the scalar $\alpha$ is an unknown quantity that must be estimated (e.g., through calibration with single-molecule measurements or dilution experiment\cite{XXX - look for elowitz paper or something similar that tried to calibrate for single-molecule counts}).
-
-% MMM - Ding:2019??
-Stochastic fluctuations in a chemical process decrease with the inverse square root of the process scale \cite{Kampen1961}.
-As the system size increases, the dynamics converge towards a deterministic process, except for certain exceptional initial conditions lying on manifolds that separate attractors for different steady state behaviors \cite{XXX}.
-This creates a tradeoff in noise-induced control, where noise enables differential control but is also deleterious to maintaining desired states once they have been achieved.
-The impact of reducing noise through increased system granularity on control performance remains unclear.
-
-To shed light on this tradeoff, control performance scores for the $\mathbf{u}^{\rm FAC}$ and $\mathbf{u}^{\rm PAC}$ controllers were calculated at different levels of granularity ($\alpha$) between 0.2 and 2.0.
-% XXX - Why only go up to 2?  Did it become intractable?
-% MMM - Non_Sparse Calculations are intractible after 2, but sparse versions of code will work fine
-Figure \ref{Volume} (A - F) shows the joint (left plots) and marginal (right plots) distributions of the two cells relative to the specified target position (circles in joint distributions and vertical dashed lines in marginal distributions) for the FAC controller and for $\mathbf{\alpha}$=.2 (A,B), 1 (C,D) and 2 (E,F).
-As the granularity changes from $\alpha$=0.2 , to $\alpha$=2, the distributions become focused at a point near the target, and the performance score improves from $J=$333 to 85 to 27, respectively.
-The PAC controller performance also improves considerably with alpha as shown in Fig.\ \ref{Volume} (G-L).
-% XXX - Please change the panel labels to A-F, G-L, and M.
-% MMM - made the change
 \subsection{Effects of changes to system granularity}
@@
 % Add tic marks to left side of the A and C heat maps
+Stochastic fluctuations in a chemical process decrease with the inverse square root of the process scale \cite{Kampen1961}.
+As the system size increases, the dynamics converge towards a deterministic process, except for certain exceptional initial conditions lying on manifolds that separate attractors for different steady state behaviors \cite{XXX}.
+This creates a tradeoff in noise-induced control, where noise enables differential control but is also deleterious to maintaining desired states once they have been achieved.
+The impact of reducing noise through increased system granularity on control performance remains unclear.
 
+To gain insight on this tradeoff, control performance scores for the $\mathbf{u}^{\rm FAC}$ and $\mathbf{u}^{\rm PAC}$ controllers were calculated at different levels of granularity ($\alpha$) between 0.2 and 2.0.
+% XXX - Why only go up to 2?  Did it become intractable?
+% MMM - Non_Sparse Calculations are intractible after 2, but sparse versions of code will work fine
+Figure \ref{Volume} (A - F) shows the joint probability distributions (left plots) and marginal probability distributions (right plots) of the system at low granularity ($\alpha=0.2$), original granularity ($\alpha=1.0$), and increased granularity ($\alpha=2.0$).
+As the system increases, the scale the size of the dynamics of the system becomes streched over a larger region in state space.
+This effect can be observed in the smoothness of C with respect to A despite using the same system and the same controller.
+
+The noise reducing effect of granularity also improves the control performance of both controllers.
 Figure \ref{Volume}G shows the trend of the performance versus alpha for both the FAC (solid cyan line) and the PAC (solid magenta line) controller.
@@
 
-To bypass this FSP limit, sixteen SSA simulations were simulated to steady state to sample the CME of a system with a significantly larger volume of $\alpha = 100$.
-Each SSA ran for $5\times10^7$ minutes, and the last $4\times10^7$ minutes were sampled for estimating the stationary distribution and calculating the performance score.
-The performance score estimates of this high granularity SSA, using the FAC and PAC controllers, were 4.13 and 27.5, respectively, shown as dashed lines in Fig.\ \ref{Volume}G.
-While it remains uncertain if further performance improvements can be achieved with larger system scales, these findings suggest that both controllers consistently improve with increased $\alpha$, with $\mathbf{u}^{\rm FAC}$ consistently outperforming $\mathbf{u}^{\rm PAC}$ in all cases examined.
-From figure H, it can be seen that a smaller $\alpha$ creates faster response but higher baseline score due to intrinsic noise.
-When $\alpha = 4$, the system becomes much less noisy and has lower scores but takes longer time to respond.
+The last property of system scale is the extended time dynamics.
+Fig \ref{Volume}(H) shows the time taken for control performance to reach steady state as $\alpha$ increases, we can see the control performance floor decrease due to lower intrinsic noise but the time scale taken to reach steady state increases with $\alpha$.
 
@@
 
-Figure \ref{Time} shows the joint and marginal distribution of each controller and time-delay pair and the resulting score, with panels A-F showing results for the FAC controller and panels G-L showing results for the PAC controller.
-Figure \ref{Time}M summarizes these results by plotting the score of both controllers versus the time delay.
+Figure \ref{Time} shows the joint distributions (left) and marginal distributions (right) at varying levels of time-delay, with panels A-C showing results for the FAC controller and panels D-F showing results for the PAC controller.
+Figure \ref{Time}G summarizes these results by plotting the score of both controllers versus the time delay.
 From the figures, it is clear that performance is rapidly degraded as the delay approaches and then exceeds the characteristic time scale of the process.
@@
 Generally, we would expect that increasing $\alpha$ would improve control performance, while increasing $\tau$ would decrease control performance.
+Therefore, joint changes in $\tau$ and $\alpha$ might reveal how the positive effect of system scale might be undone by the negative effect of time delay.
 
@@
 Joint probability distributions at certain $(\tau,\alpha)$ pairs and their control performance can be seen in Fig \ref{DG}(A-I).
-Control performance reduced as $\alpha$ increased when $\tau=100$ but not at $\tau=1$.
+Control performance measured by RMSE increased as $\alpha$ increased when $\tau=100$ but not at $\tau=1$.
 FAC control performance Fig \ref{DG}(J) over the domain of points shows a valley region at low $\tau$, and a plateau region at large $\tau$.
 This suggests that increasing system scale at large time delays is not beneficial to control performance.
-Trajectories at the edge of the valleyFig \ref {DG}(J)(yellow star) showed oscillatory behavior Fig \ref {DG}(K)and trajectories within the plateau show bursty behavior.
+Trajectories at the edge of the valley Fig \ref {DG}(J)(yellow star) showed oscillatory behavior Fig \ref {DG}(K) and trajectories within the plateau show bursty behavior.
 
@@
 Figure \ref{CtrlP} (B) show the steady state probability distribution of the controlled system.
-Figure \ref{CtrlP} (C and D) show the optimized controller and probability distributions using the partially aware control method.
+Figure \ref{CtrlP} (C and D) show the optimized controller and probability distributions using the PAC method.
 The unstable point in the rate equation matches the unstable point near $(20,20)$ in this domain.
@@
 \caption{Finite state projections of the slowly driven system demonstrate good control performance under a variety of phase and frequency scenario.
-(A) Schematic of the control problem shows targeting the number of protein in two cells using a dynamically changing reference point by cycling through a set of individual stable points and under synchonous, phasse-lagged or frequency changes reference signal using only one control input.
-(B1-BN) shows the phase space of the reference phase space, error residuals, reference time series, FSP solution and RMSE respectively for the synchonous control method.
+(A) Schematic of the control problem shows targeting the number of protein in two cells using a dynamically changing reference point by cycling through a set of individual stable points and under synchronous, phase-lagged or frequency changes reference signal using only one control input.
+(B1-BN) shows the phase space of the reference time series, FSP solution, RMSE, reference phase space, and error residuals respectively for the synchronous control method.
 (C1-C5) Shows the same data for the phase-lagged control input.
@@
 
-We developed a method to control the dynamics of a system by optimizing steady state controllers for a predefined path $r(x)$ in a Piecewise manner and cycling through them at a fixed frequency.
+In \ref{Methods}, we developed a method to control the dynamics of a slowly changing system on a predefined path $r(x)$ by cycling through controllers at a fixed frequency.
 Thirty-two controllers were optimized along a pathway in order to target an in-sync reference point \ref{CR}, a phase lagged reference point \ref{CR}, or frequency separated reference point\ref{CR}(A).
-The synchronous signal is given by
-\begin{equation}
- r_1(x)=[20 sin(2 \pi x)+10, 20 sin(2 \pi x)+10]
-\end{equation}
-the phase lagged signal is given by
-\begin{equation}
- r_2(x)=[20 sin(2 \pi x)+10, 20 cos(2 \pi x)+10]
-\end{equation}
-and the frequency changed signal is given by
-\begin{equation}
- r_3(x)=[20 sin(2 \pi x)+10, 20 sin(6 \pi x)+10]
-\end{equation}
-for $x\in[0,1]$ using 32 steps.
-FSP simulations of cyclically driven systems were slowly driven at a frequency of $1e-4$ seconds and their medians are shown in the Fig \ref{CR}(B4, C4, and D4)(red line, blue line).
+The synchronous signal is given by Fig \ref{CR}(B1) $r_1(x)=[20 sin(2 \pi x)+10, 20 sin(2 \pi x)+10]$, the phase lagged signal Fig \ref{CR}(C1) is given by $r_2(x)=[20 sin(2 \pi x)+10, 20 cos(2 \pi x)+10]$, and the frequency changed signal Fig \ref{CR}(D1) is given by $r_3(x)=[20 sin(2 \pi x)+10, 20 sin(6 \pi x)+10]$ using 32 steps for each controller.
+
+FSP simulations of cyclically driven systems were slowly driven at a frequency of $1e-4$ seconds and their medians are shown in the Fig \ref{CR}(B2, C2, and D2)(red line, blue line).
 The time series probability distributions of $x_1$ and $x_2$ can be seen as red clouds and blue clouds respectively.
@@
 Error distributions of the system were taken by shifting the probability distribution from the reference point to the origin Fig \ref{CR}(B2, C2, and D2).
-Root mean squared error of each simulation are shown in Fig \ref{CR}(B5, C5,D5).
+Root mean squared error of each simulation are shown in Fig \ref{CR}(B3, C3,D3).
 
-Suprisingly, the system is able to effectively control two signals to two different trajectories at different phase and frequency and using only a single control input despite the presence of noise Fig \ref{CR}(D4).
-Simulations show that the synchronous control performed the best with an RSME of $6.9$ despite moving over the unstable point in the system.
-When the system is driven with a phase-lag the RSME of the score increases to $8.2$ and when driven at a different frequency the RMSE goes up to $8.6$.
-Medians of the marginal distributions show good match with the desired reference signal in each control scheme Fig \ref{CR}(B4,C4, and D4).
-TIme series trajectories of the RMSE show spikes in the control performance using any control strategy.
+Simulations show that the synchronous control performed the best with an average RSME of $6.9$ despite moving over the unstable point in the system.
+When the system is driven with a phase-lag the average RSME of the score increases to $8.2$ and when driven at a different frequency the average RMSE goes up to $8.6$.
+Suprisingly, the system is able to effectively control two signals to two different trajectories at different phase and frequency and using only a single control input despite the presence of noise Fig \ref{CR}(D3).
+Medians of the marginal distributions show good match with the desired reference signal in each control scheme Fig \ref{CR}(B4,C4, and D4), and time series trajectories of the RMSE show spikes in the control performance using any control strategy.
 
@@
 \vspace{-0.1in}
-\caption{Control performance analyzed over a domain of $f,\alpha$ pairs show worst control performance at moderate frequencies near $1e-2$ Tracking reference signals when $\alpha=5$ span a range between 50 and 150 species (A).
+\caption{Control performance analyzed over a domain of $f,\alpha$ pairs show worst control performance at moderate frequencies near $1e-2$ due to phase lag.
+Tracking reference signals when $\alpha=5$ span a range between 50 and 150 species (A).
 Stochastic simulations driven using a phased lagged controller at $\alpha = 5$ and low frequency show tighter control compared to $\alpha =1$ (B).
@@
 
-High-frequency driven systems often experience reduced control performance due to the system's slow response.
-However, the combination of both high-frequency driving and noise could be catastrophic for system control performance.
-Conversely, increasing granularity may improve control performance but alters the system's time scale.
+Slowly responding systems driven at high frequency often exhibit poor control performance which improves as the driving frequency matches the time scale of the system.
+The combination of both frequency driving and noise could be catastrophic for system control performance.
+Conversely, increasing granularity($\alpha$) may improve control performance but alters the system's time scale.
+Therefore these competing effects of system scale has not yet been analyzed.
 
-To examine how driving frequency affects noise-exploiting control, Sixty-four SSA trajectories of the phase-separated dynamic controller were simulated over a two dimensional domain of points $(\alpha,f)$ for 16 cycles after reaching steady state.
-For each $(\alpha,f)$ pair the control performance of the steady state trajectories were calculated.
+To examine how driving frequency and system scale affects control performance, Sixty-four SSA trajectories of the phase-separated dynamic controller were simulated over a two dimensional domain of points $(\alpha,f)$ for 16 cycles after reaching steady state.
+For each $(\alpha,f)$ pair the average RMSE of the steady state trajectories were calculated.
 
-Fig \ref{CRG}(A) shows the desired reference of a system with $\alpha = 5.0$, and Fig \ref{CRG}(A) shows the mean of the system response when $f=1e-4$.
-Fig \ref{CRG}(C) Shows control performance over time as $\alpha$ is held at $5.0$ and $f$ increases from 1e-4 to 1e-2 to 1e-0 with an average RMSE of 6.4, 16.3, and 14.4 respectively.
-I creasing frequency while alpha is fixed at $0.5$, $1.0$ and $2.0$ shows that control performance from 12 to 6.3.
+Fig \ref{CRG}(A) shows the desired signal of a system when $\alpha = 4.0$, and Fig \ref{CRG}(B) shows the mean of the system response when $f=10^{-4}$.
+Fig \ref{CRG}(C) Shows control performance over normalized periodic time as $\alpha$ is held at $4.0$ and $f$ increases from $f=10^{-4}$ to $f=10^{-2}$ to $f=10^{-0}$ with an average RMSE of 6.2, 16.1, and 15.0 respectively.
+Heat maps of control performance over $(\alpha,f)$ pairs can be seen in Fig \ref{CRG} (D).
+RMSE control performance of 128 simulations were simulated over a range of frequency between $10^-3$ and $10^-1$ Fig \ref{CRG}(D) for 32 cycles to obtain a better measurement of the true RMSE Fig \ref{CRG}(D) at $\alpha=2.0$, $\alpha=4.0$, and $\alpha=8.0$.
 
-Heat maps of control performance of $(f,\alpha)$ pairs showed the worst performance at moderate $f$ and that control performance was best when both granularity was high and frequency was low.
+Heat maps of control performance over $(f,\alpha)$ pairs showed the worst performance at moderate $f$ near $f=10^{-2}$ and that control performance was best when both granularity was high and frequency was low.
+When examining poorly performing trajectories from moderate frequencies, a phase lag was observed in the system, while at very large frequencies the system did not respond at all.
+This phase las resulted in worse RMSE than no response at all.
+Since the RMSE of the system at $\alpha=4$ is less than the RMSE of the $\alpha=1$ system, control performance improved with $\alpha$.
 

### 2023-07-07 paper.tex
Update format.  Made edits to methods.

@@
 
-%\documentclass[12pt]{iopart}
-\documentclass[12pt]{article}
+\documentclass[12pt]{iopart}
+%\documentclass[12pt]{article}
 %\usepackage{amsmath}
-\usepackage{cite}
-\usepackage{amsmath,amssymb,amsfonts}
-\usepackage{algorithmic}
+%\usepackage{cite}
+\usepackage{amssymb,amsfonts}
+%\usepackage{algorithmic}
 \usepackage{graphicx}
-\usepackage{textcomp}
+%\usepackage{textcomp}
 \usepackage{xcolor}
-\usepackage{graphicx}
+%\usepackage{graphicx}
 %Uncomment next line if AMS fonts required
@@
 
-\title{Exploiting Intrinsic Noise to Control Cell Differentiation Dynamics in the Presence of Time Delays, Inexact Models, and Extrinsic Variability}
+\title[Heterogeneous Control of Cell Differentiation Dynamics]{Exploiting Intrinsic Noise to Achieve Heterogeneous Control of Cell Differentiation Dynamics in the Presence of Time Delays, Model Uncertainties, and Extrinsic Variability}
 \maketitle
@@
 
-%\address{$^1$ School of Bioengineering, Colorado State University, Fort Collins, Co}
-%\address{$^2$ Dept. of Chemical and Biological Engineering, Colorado State University, Fort Collins, Co}
-
-%\vspace{10pt}
-%\begin{indented}
-%\item[]August 2017
-%\end{indented}
+\address{$^1$ School of Bioengineering, Colorado State University, Fort Collins, CO, USA}
+\address{$^2$ Department of Chemical and Biological Engineering, Colorado State University, Fort Collins, CO, USA}
 
@@
 Parameter & Value   & units                     \\ \hline \hline
-$\kappa$         & 0.406   & Molecules Per Minute \\ \hline
+$\kappa$         & 0.406   & 1/min \\ \hline
 $\beta$          & 20.0    & Molecules            \\  \hline
-$\eta $           & 8.00    & unit less          \\  \hline
-$k_0$           & 0.0001  & Molecules per Minute \\  \hline
-$\gamma$         & 0.0203  & per Minute         \\  \hline
-$u$              & dynamic & Molecules per Minute \\ \hline
+$\eta $           & 8.00    & unitless          \\  \hline
+$k_0$           & 0.0001  & 1/min \\  \hline
+$\gamma$         & 0.0203  & 1/min         \\  \hline
+$u$              & dynamic & 1/min \\ \hline
 \end{tabular}
@@
 Specifically, the $2N_c$ possible reactions are defined in pairs corresponding to production ($\mu\in\{1,3,5,\ldots\}$) and degradation ($\mu\in\{2,4,6,\ldots\}$) as:
-\begin{align}
+\begin{eqnarray}
 \mathbf{s}_1 &= \mathbf{e}_1,\ \mathbf{s}_2 = -\mathbf{e}_1,\nonumber\\ 
@@
 \mathbf{s}_{\rm 2N_c-1} &= \mathbf{e}_{\rm N_c},\ \mathbf{s}_{\rm 2N_c} = -\mathbf{e}_{\rm N_c}, \label{Stochs}
- \end{align}
+ \end{eqnarray}
 where each $\mathbf{e}_{i}\in\mathbb{Z}^{{\rm N_ c}}$ is is a Euclidean vector (i.e., unity for the $i^{\rm th}$ entry and otherwise zero). The corresponding propensity functions are:
-\begin{align}
+\begin{eqnarray}
 w_1(\mathbf{X}) &= u(\mathbf{X},t)  + \kappa \frac{x_1^\eta}{x_1^{\eta}+\beta^{\eta}} + k_0,\ w_2(\mathbf{X}) = \gamma x_1,\nonumber\\
@@
 w_{\rm 2N_c-1}(\mathbf{X}) &= u(\mathbf{X},t)  + \kappa \frac{x_N^\eta}{x_N^{\eta}+\beta^{\eta}} + k_0,\ w_{\rm 2N_c}(\mathbf{X}) = \gamma x_N.\label{Props}
- \end{align}
+ \end{eqnarray}
 
-These definitions of the stoichiometry and propensity functions allow us to implement the Gillespie stochastic simulation algorithm (SSA) \cite{Gillespie1992, Gillespie1977} to generate representative trajectories of the stochastic process.This algorithm uses discrete reactions and discrete molecule counts to sample the chemical master equation by utilizing an exponential random variable for time intervals and a categorical random variable for reaction selection.
-The exponential random variable is given by
+These definitions of the stoichiometry and propensity functions allow us to implement the Gillespie stochastic simulation algorithm (SSA) \cite{Gillespie1992, Gillespie1977} to generate representative trajectories of the stochastic process.
+At each step two random numbers are generated to determine the time and the type of the next reaction.
+Given the current state $\mathbf{X}$, the time until the next reaction is distributed according to an exponential random with rate parameter equal to the inverse of the sum of the propensity functions:
 \begin{equation}
-\delta t(\mathbf{X},t)=log(rand \sum w_i(\mathbf{X},t) )
+Pr(\delta t = \tau) = \sum w_i(\mathbf{X}) \exp\left(- \sum{\mu=1}^{2N_c-1} w_\mu(\mathbf{X})\tau\right),
 \end{equation}
-where $\delta t$ is the time at which the next reaction occurs, and rand is a randomly generated number between 0 and 1.
-The probability of an individual reaction to fire from all possible reactions is given by
+and an instance of this random variable, $\delta t$,  can be sampled from this distribution using the expression:
 \begin{equation}
-\zeta_i(\mathbf{X},t)=\frac{w_i(\mathbf{X},t)}{\sum w_i(\mathbf{X},t)}
+\delta t = -\log\left(r_1 \sum_{\mu=1}^{2N_c-1}  w_\mu(\mathbf{X})\right),
 \end{equation}
-Where $\zeta_i(\mathbf{X},t)$ is the probability of the ith reaction to occur at state $\mathbf{X}$ and time $t$.
-Then the reaction that fired at time $t$ is a categorical random variable pulled from $\zeta_i(\mathbf{X},t)$.
-Then the SSA can be created by walking through time jumps and updating the state of the system by adding the stiochiometry of the reaction to the state of the system.
+where $r_1$ is a uniform random variable between zero and one.
+The probability of for the specific individual reaction $R_k$ to fire from all possible reactions is given by:
+\begin{equation}
+Pr(R_k)=\frac{w_k}{\sum w_{\mu=1}^{2N_c-1} (\mathbf{X})},
+\end{equation}
+and thus the reaction that fires at time $t+\tau$ is a categorical random variable pulled from $Pr(R_k)$.
+The SSA is then simulated by stepping through time one reaction at a time and updating the state of the system by adding the stoichiometry of the reaction to the state of the system.
 
@@
 \begin{array}{rl}
--\sum_{\mu =1}^{2N_{\rm c}} w_{\mu}(\mathbf{X}_j), &\text{for }i=j,\\
-w_{\mu}(\mathbf{X}_j), &\text{for }\mathbf{X}_i = \mathbf{X}_j+\mathbf{s}_\mu\\
-0, & \text{otherwise,}
+-\sum_{\mu =1}^{2N_{\rm c}} w_{\mu}(\mathbf{X}_j), &\textrm{for }i=j,\\
+w_{\mu}(\mathbf{X}_j), &\textrm{for }\mathbf{X}_i = \mathbf{X}_j+\mathbf{s}_\mu\\
+0, & \textrm{otherwise,}
 \end{array}\right. 
@@
 \begin{array}{rl}
-- N_{\rm c}{u}^{\mathcal{C}}(\mathbf{X}_j), &\text{for }i=j\\
-{u}^{\mathcal{C}}(\mathbf{X}_j), &\text{for }\begin{array}{ll}\mathbf{X}_i =\mathbf{X}_j + \mathbf{e}_{i_{\rm c}}, \\ \text{and } i_{\rm c} = 1,\ldots,N_{\rm c}\end{array},\\
-0, & \text{otherwise,}
+- N_{\rm c}{u}^{\mathcal{C}}(\mathbf{X}_j), &\textrm{for }i=j\\
+{u}^{\mathcal{C}}(\mathbf{X}_j), &\textrm{for }\begin{array}{ll}\mathbf{X}_i =\mathbf{X}_j + \mathbf{e}_{i_{\rm c}}, \\ \textrm{and } i_{\rm c} = 1,\ldots,N_{\rm c}\end{array},\\
+0, & \textrm{otherwise,}
 \end{array}\right.
@@
 
-\brian[Add a paragraph introducing the FSP truncation and include some justification.]
-Finite state projections are a powerful way to simulate master equation analysis which enables the assessment of simulation accuracy, but these simulations have challenges measuring the steady state of the system since all probability will always enter error at long enough time scales.
+In principle, the master equation in Eq.\ \ref{CME} could contain an uncountably infinite number of states, and therefore the exact solution as well as the null vector in Eq.\ \ref{SSDist} may not be computable exactly.
+To address this issue, we first truncate the system at a finite number for each species and then apply a reflecting boundary condition, resulting in a finite dimensional master equation.
+To assess the time interval over which this truncation is valid, we also solve the Finite State Projection\cite{Munsky:2006} for the same truncation, which allows us to compute an upper bound on the truncation error as a function of time.
+
+\brian[We should use the FSP to compute the error of the truncation for a couple cases.  We can talk about how to do this when we meet.]
+
+%Finite state projections are a powerful way to simulate master equation analysis which enables the assessment of simulation accuracy, but these simulations have challenges measuring the steady state of the system since all probability will always enter error at long enough time scales.
 To enable control of steady state systems, a truncated FSP which reflects probability back into the system enables this calculation.
@@
  \end{equation}
-This score is calculated by applying a linear operator on steady state probability distribution $\mathbf{P}^*$ (Eq.\ \ref{SSDist}) as follows:
-\begin{align}
-J&= \lim_{t\rightarrow \infty}\mathbb{E}\{|\mathbf{X}(t)-\mathbf{T}|_2^2\}, \nonumber \\ 
+The squared score is easily calculated by applying a linear operator on steady state probability distribution $\mathbf{P}^*$ (Eq.\ \ref{SSDist}) as follows:
+\begin{eqnarray}
+J^2&= \lim_{t\rightarrow \infty}\mathbb{E}\{|\mathbf{X}(t)-\mathbf{T}|_2^2\}, \nonumber \\ 
 &=\sum_{i_1,i_2,\ldots} P^*(x_1=i_1,x_2=i_2,\ldots) \left[(i- T_1)^2 + (j- T_2)^2 +\ldots\right],\nonumber  \\
@@
 \label{Euclid}
-\end{align}
+\end{eqnarray}
 where $\mathbf{C}$ is simply a vector that contains the squared Euclidian distance of each state from the specified target T, i.e., $C_i = |\mathbf{X}_i-\mathbf{T}|_2^2$. 
@@
 Optimization of $\mathbf{u}^{FAC}$ was performed using a gradient descent method to minimize $J$.
-This method requires calculating the negative gradient $-\frac{dJ}{d\mathbf{u}^{FAC}}$ and walking in a small step $du$ in that direction.
+Since the square root is a monotonically increasing function, minimizing $J^2$ results in the same control as would minimizing $J$ directly.
+Therefore, we calculate the negative gradient $-\frac{d(J^2)}{d\mathbf{u}^{FAC}}$ and walking in a small step $du$ in that direction.
 The calculation of the gradient is given by
 \begin{equation}
-\frac{dJ}{d\mathbf{u}^{FAC}}=C \frac{dP}{d\mathbf{u}^{FAC}}
+\frac{d(J^2)}{d\mathbf{u}^{FAC}}=C \frac{dP}{d\mathbf{u}^{FAC}}
 \end{equation}
@@
 to find $\frac{dP}{d\mathbf{u}^{FAC}}$.
-Optimization of $\mathbf{u}^{PAC}$ was done by restricting $\frac{dJ}{d\mathbf{u}^{FAC}}$ in ambient space to search a low dimensional subspace with a linear lift operator D.
+Optimization of $\mathbf{u}^{PAC}$ was done by restricting $\frac{d(J^2)}{d\mathbf{u}^{FAC}}$ in ambient space to search a low dimensional subspace with a linear lift operator D.
 Then the subspace search is given by.
 \begin{equation}
-\frac{dJ}{d\mathbf{u}^{PAC}}=C \frac{dP}{d\mathbf{u}^{PAC}} = C D \frac{dP}{d\mathbf{u}^{FAC}}  
+\frac{d(J^2)}{d\mathbf{u}^{PAC}}=C \frac{dP}{d\mathbf{u}^{PAC}} = C D \frac{dP}{d\mathbf{u}^{FAC}}  
 \end{equation}
-Root mean squared error was calculated by taking the square root of the score.
-\begin{equation}
-RMSE=\sqrt{J}
-\end{equation}
-Since the square root is a monotonically increasing function, minimizing J is the same as minimizing RMSE.
+%Root mean squared error was calculated by taking the square root of the score.
+%\begin{equation}
+%RMSE=\sqrt{J}
+%\end{equation}
 
-In \cite{May2021}, we optimized each of these two controllers to find the control laws $\mathbf{u}^{FAC}$ and $\mathbf{u}^{PAC}$ that minimize $J$ for a single target state $\mathbf{T} = [10,30]$, but in this work extend it to ant target point in the state space.
+In \cite{May2021}, we optimized each of these two controllers to find the control laws $\mathbf{u}^{FAC}$ and $\mathbf{u}^{PAC}$ that minimize $J^2$ for a single target state $\mathbf{T} = [10,30]$, but in this work extend it to any target point in the state space.
 
@@
 For example, the production and degradation of protein in cell one would become:
-\begin{align}
+\begin{eqnarray}
 w'_1(\mathbf{X}) &= u(\mathbf{X}/\alpha,t)  + \kappa \frac{(x_1/\alpha)^\eta}{(x_1/\alpha)^{\eta}+\beta^{\eta}} + k_0 \nonumber \\
  w'_2(\mathbf{X}) &= \gamma x_1/\alpha.
-\end{align}
+\end{eqnarray}
 We note that in order to reuse a control law that has been defined for one level granularity and apply it to a system at another level of granularity, the inputs to the controller must also be scaled by $1/\alpha$ before computing the  control level assigned to the current state, e.g., ${u}^{\mathcal{C}} = {u}^{\mathcal{C}}(\mathbf{X}/\alpha)$.
@@
 For example, in the two cell system the new steady state performance error would become:
-\begin{align}
+\begin{eqnarray}
 (J^2)' &= \sum_{i=0}^{M'}  \sum_{j=0}^{M'}P^*(x_1=i,x_2=j) ((i/\alpha - \mathbf T_1)^2 + (j/\alpha -\mathbf T_2)^2),\nonumber \\
@@
 \label{EuclidV}
-\end{align}
+\end{eqnarray}
 We reiterate that the system parameters and control law were defined and fixed using the the base granularity ($\alpha=1$), and these are not recomputed or refit upon changing the system granularity.
@@
 \begin{array}{rl}
-      0 ,&\text{ for }  t \leq \tau, \\
-      {u}^{\mathcal{C}}(x_1(t-\tau), x_2(t-\tau)) , &\text{ for }   t > \tau,\\
+      0 ,&\textrm{ for }  t \leq \tau, \\
+      {u}^{\mathcal{C}}(x_1(t-\tau), x_2(t-\tau)) , &\textrm{ for }   t > \tau,\\
 \end{array}\right. 
@@
  We start by describing a path $\bf{p}_n(x)$ in state space which walks over a two dimensional path in a discrete number of steps.
-Then, optimize a controller $\bf{c}_n(x)$for each target point and cycling through the controllers over time to create $\mathbf{r}(x,t)$.
+  Then, we optimize controllers $\bf{c}_n(x)$for each target point and cycling through the controllers over time to create $\mathbf{r}(x,t)$.
 Let  $\bf{p}_n(x)$ be a path taken in $n$ steps, given by
@@
 Where $\mathbf{r}(\mathbf{x},t)$ is a function of state and time. This arbitrary form of $\mathbf{r}(\mathbf{x},t)$, enables cycling through controllers along different paths at different frequencies.
+ \brian[This is not clear.  Please re-write carefully and think about what the reader needs to know to understand and reproduce the method.]
+
 \section{Results}
-
-\brian[Add a paragraph or two to explain the big picture here.
-Use figure 1 to remind the reader of (1) the control goals, (2) the control law, (3) the resulting distributions, (4) the objective function and (5) the optimization of the control law to minimize the error.
-Refer to the methods for each such calculation.
-This section is critical to make sure the reader understands what is coming next.
-Then provide an outline of the remaining subsections of the results section so that the reader knows what to expect.]
 
@@
 Despite the advantages of full master equation analysis, good control of stochastic systems would ideally be resistant to parameter variability from heterogenous cell populations, time delay in controller actuation, and robust to systems with intrinsically more or less noise.
-To answer these questions, we will examine how parameter changes, time delay, and system scale influence the control performance of the controllers.
-Then, we extend these analyses to control a dynamically moving set point which changes its phase and frequency.
+To answer these questions, in the following subsections, we examine how parameter changes (e.g., incorrect models or extrinsic noise), time delays, and mismatches between system and models scale influence the performance of the controllers.
+Then, we extend these analyses to control a dynamically moving set point that seeks to control different identical cells to achieve different fluctuation phases and frequencies.
 

### 2023-07-07 paper.tex
some edits

@@
 
+\brian[{\bf We still need this.} Add a paragraph or two to explain the big picture here.
+Use figure 1 to remind the reader of (1) the control goals, (2) the control law, (3) the resulting distributions, (4) the objective function and (5) the optimization of the control law to minimize the error.
+Refer to the methods for each such calculation.
+This section is critical to make sure the reader understands what is coming next.
+Then provide an outline of the remaining subsections of the results section so that the reader knows what to expect.]
+
 We developed a method to create a noise-exploiting controller by taking full advantage of master equation analysis.
@@
 Finally, \ref{Parameter}B shows the performance change when a single parameter is changed simultaneously in both Cell 1 and Cell 2 at the same time while keeping other parameters fixed.
-In each case the baseline parameters is denoted as a vertical dashed line, and lower values of the RMSE ( Eq. \ref{XXX}) denote improved performance. 
+In each case the baseline parameters is denoted as a vertical dashed line, and lower values of J (Eq. \ref{XXX}) denote improved performance. 
 Solid lines are used to show the performance using the fully aware controller (FAC) that has access to both cells, and the dashed lines show the performance for the partially aware controller (PAC) that only has access to Cell 1.

### 2023-07-11 paper.tex
Small edits.

@@
 where $x_1$ and $x_2$ are discrete integers greater than or equal to zero that represent the number of proteins in cell one and cell two.
+Despite their differences in their observation data, both the FAC and the simpler PAC are optimized to achieve the same goal, namely to drive both cells to their respective set points.
 
@@
 To apply this granularity parameter, we assume that each propensity function, $w_{\mu}$ (Eqns.\ \ref{Props}) is rescaled to a different level of discreteness by substituting
-\begin{equation}
+\begin{equation}\label{eq:granscale}
 w'(\mathbf{X})=w(\mathbf{X}/\alpha).
@@
 
-\subsection{Effects of changes to system granularity}
+\subsection{Controllers trained using one assumed level of granularity can remain effective at other levels of granularity.}
+
+We next explored how the granularity of the system would affect the differential control of the process.
+Specifically, we define the level of granularity using the parameter $\alpha$ as described in Section \ref{sec:Scaling} and which affects the propensity functions and scoring according to equations \ref{eq:granscale}-\ref{EuclidV}.
+Effectively, larger $\alpha$ corresponds to systems where larger numbers of individual molecules are needed to achieve the same concentration (e.g., larger volumes), while smaller $\alpha$ corresponds to smaller population sizes (e.g., smaller volumes).
+We optimized the controller based on a the default assumption that $\alpha=1$, and we wished to know what would be the consequences if this same controller were to be applied to a system that has a different granularity.
+
+To explore this tradeoff, control performance scores for the $\mathbf{u}^{\rm FAC}$ and $\mathbf{u}^{\rm PAC}$ controllers were calculated at different levels of granularity ($\alpha$) between 0.2 and 2.0.
+Figure \ref{Volume} (A - F) shows the joint probability distributions (left plots) and marginal probability distributions (right plots) of the system at a low granularity ($\alpha=0.2$), the original granularity ($\alpha=1.0$), and at an increased granularity ($\alpha=2.0$).
+
+As the system increases, the scale the size of the dynamics of the system becomes stretched over a larger region in state space. 
+This effect can be observed in the smoothness of C with respect to A despite using the same system and the same controller. 
+\brian[the previous two sentences do not make sense to me.]
+
+The effect that granularity has on the control performance depends on two competing phenomena: First, because the controller was optimized for one granularity ($\alpha = 1$), one would expect that the controller would become worse if that granularity were in correct.
+However, at large granularity, we find that the opposite is true - the controller actually improves when applied to an incorrect model.
+The reason for this potentially surprising result, is that relative amplitude of stochastic fluctuations (i.e., the standard deviation divided by the mean) in a chemical process decrease with the inverse square root of the process scale \cite{Kampen1961}.
+At an extreme, as the system size increases, the dynamics converge towards a deterministic process, except for certain exceptional initial conditions lying on manifolds that separate different steady state behaviors \cite{XXX}.
+This creates a tradeoff in noise-induced control, where noise enables differential control but is also deleterious to maintaining desired states once they have been achieved.
+
 \begin{figure*}[t!]
@@
 \end{figure*}
-% XXX - This figure needs larger fonts for all the labels and axes. 
-% MMM - made the change
-% Update panel letters as written in caption
-% Add tic marks to left side of the A and C heat maps
-Stochastic fluctuations in a chemical process decrease with the inverse square root of the process scale \cite{Kampen1961}.
-As the system size increases, the dynamics converge towards a deterministic process, except for certain exceptional initial conditions lying on manifolds that separate attractors for different steady state behaviors \cite{XXX}.
-This creates a tradeoff in noise-induced control, where noise enables differential control but is also deleterious to maintaining desired states once they have been achieved.
-The impact of reducing noise through increased system granularity on control performance remains unclear.
-
-To gain insight on this tradeoff, control performance scores for the $\mathbf{u}^{\rm FAC}$ and $\mathbf{u}^{\rm PAC}$ controllers were calculated at different levels of granularity ($\alpha$) between 0.2 and 2.0.
-% XXX - Why only go up to 2?  Did it become intractable?
-% MMM - Non_Sparse Calculations are intractible after 2, but sparse versions of code will work fine
-Figure \ref{Volume} (A - F) shows the joint probability distributions (left plots) and marginal probability distributions (right plots) of the system at low granularity ($\alpha=0.2$), original granularity ($\alpha=1.0$), and increased granularity ($\alpha=2.0$).
-As the system increases, the scale the size of the dynamics of the system becomes streched over a larger region in state space.
-This effect can be observed in the smoothness of C with respect to A despite using the same system and the same controller.
 
@@
 This improvement in performance appears to approach a small value as the granularity goes to infinity, but since the size of the FSP increases with the square of the system size, systems much larger than $\alpha$=2 (where $\mathbf{A}_0\in \mathbb{R}^{10^4\times10^4}$) become more difficult to calculate.
-% XXX - I would think you should be able to compute larger values of 10^6.  They would be slow, but should be doable in a few minutes per calculation.
-% MMM - constructing the FSP takes longer than solving it
 To bypass this limit in the FSP, sixteen SSA simulations were used to sample the CME of a system with a much larger volume of $\alpha$=100.

### 2023-07-11 paper.tex
Edits to granularity section.

@@
 Effectively, larger $\alpha$ corresponds to systems where larger numbers of individual molecules are needed to achieve the same concentration (e.g., larger volumes), while smaller $\alpha$ corresponds to smaller population sizes (e.g., smaller volumes).
-We optimized the controller based on a the default assumption that $\alpha=1$, and we wished to know what would be the consequences if this same controller were to be applied to a system that has a different granularity.
+We previously optimized the controllers $\mathbf{u}^{\rm FAC}$ and $\mathbf{u}^{\rm PAC}$ based on a the default assumption that $\alpha=1$, and we wished to know what would be the consequences if this same controller were to be applied to a system that has a different granularity.
 
@@
 Figure \ref{Volume} (A - F) shows the joint probability distributions (left plots) and marginal probability distributions (right plots) of the system at a low granularity ($\alpha=0.2$), the original granularity ($\alpha=1.0$), and at an increased granularity ($\alpha=2.0$).
+%As the system increases, the scale the size of the dynamics of the system becomes stretched over a larger region in state space. 
+%This effect can be observed in the smoothness of C with respect to A despite using the same system and the same controller. 
+%\brian[the previous two sentences do not make sense to me.]
+Figure \ref{Volume}G shows the trend of the performance versus alpha for both the FAC (solid cyan line) and the PAC (solid magenta line) controller.
+This improvement in performance appears to approach a small value as the granularity goes to infinity, but since the size of the FSP increases with the square of the system size, systems much larger than $\alpha$=2 (where $\mathbf{A}_0\in \mathbb{R}^{10^4\times10^4}$) become more difficult to calculate using the current methods.
+To bypass this limit in the FSP, sixteen SSA simulations were used to sample the CME of a system with a much larger volume of $\alpha$=100.
+Each SSA was ran for $5\times10^7$ minutes and only the last $4\times10^7$ minutes were sampled to estimate the stationary distribution and calculate the performance score.
+The performance score estimates of this high granularity SSA using the FAC and PAC were 4.13 and 27.5 respectively, which are plotted as dashed lines in Fig.\ \ref{Volume}G.
+Although it is unclear if further performance improvements could be obtained with further increases to the system scale, for all cases considered so far, we found that both controllers monotonically improved with increased $\alpha$ and that $\mathbf{u}^{\rm FAC}$ always outperforms the $\mathbf{u}^{\rm PAC}$.
 
-As the system increases, the scale the size of the dynamics of the system becomes stretched over a larger region in state space. 
-This effect can be observed in the smoothness of C with respect to A despite using the same system and the same controller. 
-\brian[the previous two sentences do not make sense to me.]
+The effect that granularity has on the control performance depends on two competing phenomena: First, because the controller was optimized for one level of granularity ($\alpha = 1$), one might expect that the controller would become worse if the granularity were incorrect.
+However, at large granularity, we find that the opposite is true - the control performance actually improves when applied to an incorrect model.
+The reason for this surprising result is that relative amplitude of stochastic fluctuations (i.e., the standard deviation divided by the mean) in a chemical process decreases with the inverse square root of the process scale \cite{Kampen1961}.
+In other words, the process become more predictable and therefore more controllable.
+At the extreme as the system size increases, the dynamics converge towards a deterministic process, except for certain exceptional initial conditions lying on manifolds that would separate different steady state behaviors \cite{XXX}.
 
-The effect that granularity has on the control performance depends on two competing phenomena: First, because the controller was optimized for one granularity ($\alpha = 1$), one would expect that the controller would become worse if that granularity were in correct.
-However, at large granularity, we find that the opposite is true - the controller actually improves when applied to an incorrect model.
-The reason for this potentially surprising result, is that relative amplitude of stochastic fluctuations (i.e., the standard deviation divided by the mean) in a chemical process decrease with the inverse square root of the process scale \cite{Kampen1961}.
-At an extreme, as the system size increases, the dynamics converge towards a deterministic process, except for certain exceptional initial conditions lying on manifolds that separate different steady state behaviors \cite{XXX}.
-This creates a tradeoff in noise-induced control, where noise enables differential control but is also deleterious to maintaining desired states once they have been achieved.
+However, the improvement in the steady state performance does not come without a cost.
+Although higher granularity reduces noise and makes it easier to maintain desired states once they have been achieved, noise is necessary to break symmetry between the two cells' dynamics to achieve those states in the first place.
+This tradeoff is illustrated in Fig \ref{Volume}(H), which plots the control performance over time after the changing the controller goal to exchange the low- and high-target cells with one another.
+From the figure, we can see although steady state control performance improves for higher values of $\alpha$, the time taken to reach that steady state performance increases also increases $\alpha$.
 
@@
 \end{figure*}
-
-The noise reducing effect of granularity also improves the control performance of both controllers.
-Figure \ref{Volume}G shows the trend of the performance versus alpha for both the FAC (solid cyan line) and the PAC (solid magenta line) controller.
-This improvement in performance appears to approach a small value as the granularity goes to infinity, but since the size of the FSP increases with the square of the system size, systems much larger than $\alpha$=2 (where $\mathbf{A}_0\in \mathbb{R}^{10^4\times10^4}$) become more difficult to calculate.
-To bypass this limit in the FSP, sixteen SSA simulations were used to sample the CME of a system with a much larger volume of $\alpha$=100.
-Each SSA was ran for $5\times10^7$ minutes and only the last $4\times10^7$ minutes were sampled to estimate the stationary distribution and the calculate the performance score.
-The performance score estimates of this high granularity SSA using the FAC and PAC were 4.13 and 27.5 respectively, which are plotted as dashed lines in Fig.\ \ref{Volume}G.
-Although it is unclear if further performance improvements could be obtained with further increases to the system scale, for all cases considered so far, we found that both controllers monotonically improved with increased $\alpha$ and that $\mathbf{u}^{\rm FAC}$ always outperforms the $\mathbf{u}^{\rm PAC}$.
-
-The last property of system scale is the extended time dynamics.
-Fig \ref{Volume}(H) shows the time taken for control performance to reach steady state as $\alpha$ increases, we can see the control performance floor decrease due to lower intrinsic noise but the time scale taken to reach steady state increases with $\alpha$.
 

### 2023-07-11 paper.tex
edit first half of the time delay section.

@@
 
-\subsection{Effects of Time Delays on Controller Performance}
+\subsection{Heterogeneous control can remain effective despite moderate time delays}
 \begin{figure*}
@@
 
-Feedback control can only be effective if one can quickly make measurements, compute adjustment to the control signal, and implement the needed change to the system.
+In general, feedback control can only be effective if one can quickly make measurements, compute adjustments to the control signal, and implement the needed changes within an appropriate amount of time relative to the characteristic times scales of the system.
 As the time required for any of these steps increases, control performance will be degraded, perhaps even leading to large fluctuations or instability.
-To explore how time delays would affect the noise-enhanced controller $\mathbf{u}^{\rm FAC}$ and $\mathbf{u}^{\rm PAC}$, we generated large sets of time-delayed stochastic simulations (see Methods) for different lengths of the time delay.
+To explore how time delays would affect the noise-enhanced controllers $\mathbf{u}^{\rm FAC}$ and $\mathbf{u}^{\rm PAC}$, we generated large sets of time-delayed stochastic simulations (see Methods) for different lengths of the time delay.
 Each SSA was sub-sampled for 1000 times over 10000 minutes of simulation time after a burnin period of 10000 minutes.
@@
 Figure \ref{Time}G summarizes these results by plotting the score of both controllers versus the time delay.
-From the figures, it is clear that performance is rapidly degraded as the delay approaches and then exceeds the characteristic time scale of the process.
+From the figures, it is clear that performance is rapidly degraded as the delay approaches and then exceeds the characteristic time scale ($1/\gamma$) set by the degradation rate of the process.
 At very small time delays (below $\tau = 0.07/\gamma = 0.0142$ min), the FAC outperforms the PAC ($J=$87 versus 146 at $\tau = 0.01/\gamma = 0.00203$ min) but at moderate time delays (above $\tau = 0.07/\gamma = 0.0142$ min) the PAC outperforms the FAC ($J=$170 versus 219 at $\tau = 0.1/\gamma = 0.0203$ min).
-% XXX - you can use the value of gamma=0.203 in Table 1 to write these in units of minutes.
+
+\brian[Are these times correct?  Isn't gamma = 0.0203 in Table 1?  tau = 0.07 / 0.0203 (1/min) = 3.44 min. Please check this carefully.]
+
+\brian[What are the units in Figure 4G?]
 
@@
 
-By increasing $\alpha$ we can change the characteristic of a system from being more noise-like to more deterministic-like.
+% Brian got here.
+
+By increasing the granularity $\alpha$ we can change the characteristic of a system from being more noise-like to more deterministic-like.
 Similarly, increasing $\tau$ can decrease the quality of the control input.

### 2023-07-12 paper.tex
Finished adding comments to time delay section

@@
 w'_1(\mathbf{X}) &= u(\mathbf{X}/\alpha,t)  + \kappa \frac{(x_1/\alpha)^\eta}{(x_1/\alpha)^{\eta}+\beta^{\eta}} + k_0 \nonumber \\
- w'_2(\mathbf{X}) &= \gamma x_1/\alpha.
+ w'_2(\mathbf{X}) &= \gamma x_1/\alpha. \label{eq:rescaling}
 \end{eqnarray}
@@
 
-This data show that time delays exceeding $0.07 / \gamma = 0.0142$ minutes is detrimental to the FAC control performance while delays beyond $0.2 / \gamma = 0.0406$ minutes are detrimental to the PAC controller.
-The choice of best controller depends on the level of time delay in the system.
+This data show that time delays exceeding $\tau_c^{\rm FAC} = 0.07 / \gamma = 0.0142$ minutes is detrimental to the FAC control performance while delays beyond $\tau_c^{\rm PAC} = 0.2 / \gamma = 0.0406$ minutes are detrimental to the PAC controller.
+\brian[Check these times.] The choice of best controller depends on the level of time delay in the system.
 At extreme time delays, both controllers lose their asymmetry, resulting in significantly worse scores ($J = 1041$ and $931$).
@@
 
-% Brian got here.
-
-By increasing the granularity $\alpha$ we can change the characteristic of a system from being more noise-like to more deterministic-like.
-Similarly, increasing $\tau$ can decrease the quality of the control input.
-Generally, we would expect that increasing $\alpha$ would improve control performance, while increasing $\tau$ would decrease control performance.
-Therefore, joint changes in $\tau$ and $\alpha$ might reveal how the positive effect of system scale might be undone by the negative effect of time delay.
-
-FAC steady state control performance was analyzed over a two dimensional domain of points $(\tau,\alpha)$ using thirty-two stochastic simulations simulated to steady state to see if the positive effect granularity could undo the effect of negative effect of time delay.
-Instead, noise induced oscillations were surprisingly found at certain $(\tau,\alpha)$ pairs and worse control performance was observed when system granularity increase for some $\tau$.
-Joint probability distributions at certain $(\tau,\alpha)$ pairs and their control performance can be seen in Fig \ref{DG}(A-I).
-Control performance measured by RMSE increased as $\alpha$ increased when $\tau=100$ but not at $\tau=1$.
-FAC control performance Fig \ref{DG}(J) over the domain of points shows a valley region at low $\tau$, and a plateau region at large $\tau$.
-This suggests that increasing system scale at large time delays is not beneficial to control performance.
-Trajectories at the edge of the valley Fig \ref {DG}(J)(yellow star) showed oscillatory behavior Fig \ref {DG}(K) and trajectories within the plateau show bursty behavior.
-
-The control performance domain Fig \ref{DG}(J) shows that low noise and high quality control inputs are both required to achieve good control.
-When noise is improved by increasing $\alpha$ the trajectories were observed going from very noisy to very oscillating which prevented improved control performance.
-Decreases in $\alpha$ should make the system respond faster and be less susceptible to time delay, but also more noisy.
-These competing effects in control performance can be observed in Fig \ref{DG}(J) which show that control performance is always poor at low $\alpha$.
+As discussed in the previous section, increasing the granularity of the system $\alpha$ reduces the randomness to improve the steady state performance but at the cost of slowing down the controller response.
+To explore the joint effects of time delays and granularity, Fig,\ \ref{DG}J shows the FAC steady state control performance as a function of the time delay and system granularity $(\tau,\alpha)$ using thirty-two stochastic simulations simulated to steady state at each combination.
+Control performance errors measured by RMSE improved as $\alpha$ increased when the delays were small ($\tau=1$ and $\tau=10$) but not when $\tau=100$.
+\brian[What are the units of tau in fig 5?
+This section is confusing in that it is not clear to me how tau and gamma are changing relative to one another when alpha is changed.
+From Eq 21 it seems that gamma changes with 1/alpha.
+If this is correct, does that mean that the critical time changes with $\tau_c = 1/gamma = alpha/gamma ~= 5*alpha$?
+If you plot that as a thick dashed dark line in Fig 5J,  would it match to the contours?
+]
+Recalling that under the system granularity rescaling (Eq.\ \ref{eq:rescaling}), as $\alpha$ changes, the effective degradation rate scales according to $\gamma' = \gamma/\alpha$.
+Therefore, the critical limit for the delays should also change with the system scale according to $\tau_c^{\rm FAC} = 0.07\alpha/\gamma = XXX\alpha$.
+Fig \ref{DG}(J) depicts this critical line and shows that as the time delay approaches and then exceeds this level, the steady state performance become dramatically worse.
+Finally, to understand the model of failure at these longer delays, it is interesting to examine trajectories induced by the controller just above this critical delay.
+For parameter set denoted by the yellow start in Fig.\ \ref {DG}(J), Fig.\ \ref {DG}(K) shows the controlled response after a long burn in period to achieve steady state.
+In this case, the application of feedback after a delay leads to a strong oscillatory behavior and worse performance than that achieved without any control at all.
+This observation further stresses the importance of considering time delays when designing such controllers.
 
@@
 \end{figure*}
+
+\subsection{With noise-induced control, a single input can drive multiple cells to follow different temporal trajectories.}
 

### 2023-07-12 paper.tex
edit first half of trajectory section.

@@
 
-We analyzed the control performance over a domain of target points to find if some regions are easier to control than others.
-Stability analysis of the auto-regulated system using a moderate light input can create up to four stable points and one unstable point in the state space.
+%We next analyzed the control performance over a domain of target points to find if some states are easier to achieve than others.
+For a two-cell system, stability analysis of the auto-regulated system using a moderate light input can create up to four stable points and one unstable point in the state space.
 These points give rise to basins which may attract probability more easily than other points.
-Control performance of the stochastic system was analyzed over a domain of different set points and were optimized using more observable (FAC) or less observable (PAC) control.
+We next optimized the FAC controller and calculated its performance for the two-cell system over a two dimensional domain of target points $[T_1,T_2]$ between between five and forty-five.
+For example, Fig \ref{CtrlP} (A) shows the optimized FAC control input of the system when the target is $\mathbf{T}=[20,25]$, and Fig.\ \ref{CtrlP} (B) shows the corresponding steady state probability distribution of the controlled system.
+Figure \ref{CtrlP} (C and D) show the the same thing but for a different target state of $\mathbf{T}=[XXX,XXX]$.
 
-Controllers were optimized over a two dimensional domain of target points $[T_1,T_2]$ between between five and forty-five using the FAC input method or the PAC input method.
-Figure \ref{CtrlP} (A) shows the optimized control input of the system when the target is $T=[20,25]$ after optimizing the FAC.
-Figure \ref{CtrlP} (B) show the steady state probability distribution of the controlled system.
-Figure \ref{CtrlP} (C and D) show the optimized controller and probability distributions using the PAC method.
-The unstable point in the rate equation matches the unstable point near $(20,20)$ in this domain.
+\brian[Let's remove the PAC from this section.
+Show only the FAC and and replace the Fig 6C/D with the FAC at a different target point.
+Put a new star of a different color to match to the new 6C/D.
+Remove Fig 6F.
+Make sure to change all the figure labels and titles and captions.]
 
-FAC and PAC control performance over the domain of static set points Fig\ref{CtrlP}(E) and (F) showed regions of good control in the FAC between 10 and 30 but a smaller region in the PAC.
-The FAC control performance was found to be symmetric along the dashed black line.
-The cause was found to be that symmetric pairs had control inputs which were exact transposes of each other.
-The unstable point of the system is known to be near [20,20] which is also a point of poor control for the FAC system with respect to local control performance.
-Edge regions were difficult to control, likely due to the geometry of how control inputs move the probability.
+Figure \ref{CtrlP}(E) shows the FAC control performance over the entire domain of static set points, and illustrates that some regions are easier to obtain than others.
+For example, near the two stable fixed points at [XXX,XXX] and [XXX,XXX], the control performance is quite strong (white regions), whereas near the unstable point [20,20], the FAC results in relatively worse control.
+
+Edge regions were difficult to control, likely due to the geometry of how control inputs move the probability. \brian[I don't understand this statement.]
 
@@
 
-In \ref{Methods}, we developed a method to control the dynamics of a slowly changing system on a predefined path $r(x)$ by cycling through controllers at a fixed frequency.
-Thirty-two controllers were optimized along a pathway in order to target an in-sync reference point \ref{CR}, a phase lagged reference point \ref{CR}, or frequency separated reference point\ref{CR}(A).
+We developed a method (Section \ref{sec:Dynamic}) to control the system dynamics to follow a predefined path $r(x)$ by cycling through controllers optimized for individual states along that path.
+Specifically, we found 32 controllers along each three representative pathways, including an in-sync reference point (Fig.\ \ref{CR}B1), a phase lagged reference point (Fig.\ \ref{CR}C1),  and a  frequency separated reference point (Fig.\ \ref{CR}D1).
 The synchronous signal is given by Fig \ref{CR}(B1) $r_1(x)=[20 sin(2 \pi x)+10, 20 sin(2 \pi x)+10]$, the phase lagged signal Fig \ref{CR}(C1) is given by $r_2(x)=[20 sin(2 \pi x)+10, 20 cos(2 \pi x)+10]$, and the frequency changed signal Fig \ref{CR}(D1) is given by $r_3(x)=[20 sin(2 \pi x)+10, 20 sin(6 \pi x)+10]$ using 32 steps for each controller.
 
-FSP simulations of cyclically driven systems were slowly driven at a frequency of $1e-4$ seconds and their medians are shown in the Fig \ref{CR}(B2, C2, and D2)(red line, blue line).
-The time series probability distributions of $x_1$ and $x_2$ can be seen as red clouds and blue clouds respectively.
-Regions with purple clouds are the exact overlap of red and blue.
-Error distributions of the system were taken by shifting the probability distribution from the reference point to the origin Fig \ref{CR}(B2, C2, and D2).
-Root mean squared error of each simulation are shown in Fig \ref{CR}(B3, C3,D3).
-
-Simulations show that the synchronous control performed the best with an average RSME of $6.9$ despite moving over the unstable point in the system.
-When the system is driven with a phase-lag the average RSME of the score increases to $8.2$ and when driven at a different frequency the average RMSE goes up to $8.6$.
-Suprisingly, the system is able to effectively control two signals to two different trajectories at different phase and frequency and using only a single control input despite the presence of noise Fig \ref{CR}(D3).
-Medians of the marginal distributions show good match with the desired reference signal in each control scheme Fig \ref{CR}(B4,C4, and D4), and time series trajectories of the RMSE show spikes in the control performance using any control strategy.
+FSP simulations for these cyclically driven systems were slowly driven at a frequency of $1e-4$ seconds \brian[Check units.
+This doesn't make sense.
+did you mean period?
+If so, this is very fast.
+Or did you mean s$^{-1}$?] and Fig.\ \ref{CR}(B2, C2, and D2) show their response distributions (shading) and median responses (lines) for the two cells x1 and x2 in red and blue, respectively.
+Regions with purple shading depict the exact overlap of red and blue when both cells have the same distribution of response.
+We observe that the single-input control system is able to effectively control the system to follow all three input trajectories, including when the two reference trajectories have different phases and frequencies (Fig \ref{CR}D3).
+For all reference trajectories, the medians of the marginal distributions show good match with the desired reference signal (Fig \ref{CR}B4,C4, and D4).
+To quantify the overall performance, Figs.\ \ref{CR}(B3, C3,D3) show the root mean squared error as a function of time,
+and Figs.\ \ref{CR}(B2, C2, and D2) show the time-averaged error distribution for the system response relative to the time varying target.
+From these figures, we observe that all trajectories result in short RMSE spikes at times, and these correspond to short, transient periods when the controller passes through the regions of poor control depicted in Fig.\ \ref{CtrlP}(E).
+Overall, the simulations show that the synchronous control performed best with an average RSME of $6.9$ despite moving over the unstable point in the system.
+When the system is driven with a phase-lag, the average RSME of the score increases to $8.2$, and when driven at a different frequency, the average RMSE goes up to $8.6$.
 
@@
 
-Slowly responding systems driven at high frequency often exhibit poor control performance which improves as the driving frequency matches the time scale of the system.
+%Slowly responding systems driven at high frequency often exhibit poor control performance which improves as the driving frequency matches the time scale of the system.
 The combination of both frequency driving and noise could be catastrophic for system control performance.
-Conversely, increasing granularity($\alpha$) may improve control performance but alters the system's time scale.
-Therefore these competing effects of system scale has not yet been analyzed.
+A given control system generally cannot effectively respond to reference signals that change faster that the system's natural time scale.
+Since we saw previously that increasing granularity ($\alpha$) improves steady state control performance (Fig.\ \ref{Time}G) but also lengthens the systems time scale (Fig.\ \ref{Time}H), one should expect that granularity would also affect the types and speeds of signals to which the system can respond.
+To examine how driving frequency and system scale affects control performance, 64 SSA trajectories of the phase-separated dynamic controller were simulated over a two dimensional domain of points $(\alpha,f)$ for 16 cycles after reaching steady state.
+For each $(\alpha,f)$ pair, the average RMSE of the steady state trajectories were calculated.
 
-To examine how driving frequency and system scale affects control performance, Sixty-four SSA trajectories of the phase-separated dynamic controller were simulated over a two dimensional domain of points $(\alpha,f)$ for 16 cycles after reaching steady state.
-For each $(\alpha,f)$ pair the average RMSE of the steady state trajectories were calculated.
+%Brian got here
 

### 2023-07-12 paper.tex
completed edits to results section.

@@
 The combination of both frequency driving and noise could be catastrophic for system control performance.
-A given control system generally cannot effectively respond to reference signals that change faster that the system's natural time scale.
-Since we saw previously that increasing granularity ($\alpha$) improves steady state control performance (Fig.\ \ref{Time}G) but also lengthens the systems time scale (Fig.\ \ref{Time}H), one should expect that granularity would also affect the types and speeds of signals to which the system can respond.
-To examine how driving frequency and system scale affects control performance, 64 SSA trajectories of the phase-separated dynamic controller were simulated over a two dimensional domain of points $(\alpha,f)$ for 16 cycles after reaching steady state.
-For each $(\alpha,f)$ pair, the average RMSE of the steady state trajectories were calculated.
+In general, a given control system cannot effectively track a reference signal that changes faster that the system's natural time scale.
+Since we saw previously that increasing granularity ($\alpha$) improves steady state control performance (Fig.\ \ref{Time}G) but also lengthens this time scale (Fig.\ \ref{Time}H), one should expect that granularity would also affect the types and speeds of signals to which the system can respond.
+To examine how driving frequency and system scale affects control performance, 64 or 128 SSA trajectories of the phase-separated dynamic controller were simulated over a two dimensional domain of points $(\alpha,f)$ for 16 or 32 cycles after reaching steady state.
+Figure \ref{CRG}(A) shows the desired reference signal for the system, and Fig \ref{CRG}(B) shows the mean of the system response when the frequency is $f=10^{-4}$ \brian[Need units] and the granularity is $\alpha = 4.0$.
+Figure \ref{CRG}(C) shows the corresponding control performance over normalized periodic time as $\alpha$ is held at $4.0$ and $f$ is increased from $f=10^{-4}$ \brian[Units] to $f=10^{-2}$ \brian[Units] to $f=10^{-0}$ \brian[Units] leading to average control performances of 6.2, 16.1, and 15.0 respectively.
+Figure \ref{CRG}D,E extends this analysis to plot the average control performance over different combinations of $\alpha$ and $f$. 
+%RMSE control performance of 128 simulations were simulated over a range of frequency between $10^-3$ and $10^-1$ Fig \ref{CRG}(D) for 32 cycles to obtain a better measurement of the true RMSE Fig \ref{CRG}(D) at $\alpha=2.0$, $\alpha=4.0$, and $\alpha=8.0$.
+In particular, the heat map of control performance over $(f,\alpha)$ shows that the worst performance occurs at moderate $f$ near $f=10^{-2}$ and that control performance is best when granularity is high and frequency is low.
+ When examining poorly performing trajectories at moderate frequencies, a phase lag was observed in the system.
+In this case, although the qualitative form of the response may match to the reference signal, this phase lag resulted in worse RMSE than no controller at all.
+At very large frequencies, the effect of the fast fluctuating control input is averaged, and the system barely responds.
 
-%Brian got here
+\brian[ try adding the line at $f = 1/tau_c =  \gamma/(0.7*\alpha)$ to this plot.  Does it match the contours separating good and poor control?]
 
-Fig \ref{CRG}(A) shows the desired signal of a system when $\alpha = 4.0$, and Fig \ref{CRG}(B) shows the mean of the system response when $f=10^{-4}$.
-Fig \ref{CRG}(C) Shows control performance over normalized periodic time as $\alpha$ is held at $4.0$ and $f$ increases from $f=10^{-4}$ to $f=10^{-2}$ to $f=10^{-0}$ with an average RMSE of 6.2, 16.1, and 15.0 respectively.
-Heat maps of control performance over $(\alpha,f)$ pairs can be seen in Fig \ref{CRG} (D).
-RMSE control performance of 128 simulations were simulated over a range of frequency between $10^-3$ and $10^-1$ Fig \ref{CRG}(D) for 32 cycles to obtain a better measurement of the true RMSE Fig \ref{CRG}(D) at $\alpha=2.0$, $\alpha=4.0$, and $\alpha=8.0$.
-
-Heat maps of control performance over $(f,\alpha)$ pairs showed the worst performance at moderate $f$ near $f=10^{-2}$ and that control performance was best when both granularity was high and frequency was low.
-When examining poorly performing trajectories from moderate frequencies, a phase lag was observed in the system, while at very large frequencies the system did not respond at all.
-This phase las resulted in worse RMSE than no response at all.
-Since the RMSE of the system at $\alpha=4$ is less than the RMSE of the $\alpha=1$ system, control performance improved with $\alpha$.
+\brian[Go through every figure and make sure that the units are shown and that they are correct. If quantities are unitless, make sure that the form of the normalization is clearly expressed.]
 

### 2023-07-28 paper.tex
update

@@
 \def\brian[#1]{{\color{red} #1 }}
+\def\michael[#1]{{\color{green} #1 }}
 
@@
 \brian[We should use the FSP to compute the error of the truncation for a couple cases.  We can talk about how to do this when we meet.]
-
+\michael[ok!]
 %Finite state projections are a powerful way to simulate master equation analysis which enables the assessment of simulation accuracy, but these simulations have challenges measuring the steady state of the system since all probability will always enter error at long enough time scales.
@@
 
- We start by describing a path $\bf{p}_n(x)$ in state space which walks over a two dimensional path in a discrete number of steps.
-  Then, we optimize controllers $\bf{c}_n(x)$for each target point and cycling through the controllers over time to create $\mathbf{r}(x,t)$.
-Let  $\bf{p}_n(x)$ be a path taken in $n$ steps, given by
+ We start by describing a discrete path  of target points in state space which cycle at fixed frequency.
+  Then, we optimize controllers for each target point and cycling through the controllers over time to create the final controller $\mathbf{R}(\mathbf{x},t)$.
+Let $\mathbf{p}(t)$ be a continuous and cyclicly moving point in state space
 \begin{equation}
-\bf{p}_n(x)=[p_1,...,p_n]
+\mathbf{p(t)}=[x_1(t),x_2(t)]
 \end{equation}
-where each $p_i$ is a pair of points on a 2D domain and $\bf{p}_n$ is a set of point pairs. Then, we optimize a set of controllers for each pair in the path.
+where $x_1(t)$, and $x_2(t)$ are both cyclic functions of time $\in \mathbf{R}$.
+Then, $\mathbf{q}$ is a collection of points $q_i$ which evaluate the cyclic function over its natural period (T) in n steps. 
 \begin{equation}
-\bf{c}_n(x)=[c_1,...,c_n]
+\mathbf{q}=[q_1,\ldots ,q_n]=[\mathbf{p}(t_1),\ldots,\mathbf{p}(t_n)]
 \end{equation}
-where each $\mathbf{c}_i$ is a matrix of control inputs which targets $\bf{p}_i(x)$.
-Then, we cycle through each controller at a fixed frequency by changing the controller index over time. The indexing function is given by 
+where each $t_i$ comes from linearly spaced points between $0$ and $T$, in $n$ steps 
 \begin{equation}
-i(t)=\lceil \frac{n}{T} mod(t,T)\rceil = \lceil n \ f \ mod(t,\frac{1}{f})\rceil
+t_i = [t_1,\ldots,t_n]=[0,\frac{1}{n} T,\ldots,\frac{n}{n}T]
 \end{equation}
-where $T$ is the cycling period, $f$ the cycling frequency and $n$ the number of points in the path. Then the cycling controller can be created by 
+Then, we optimize a set of controllers which target each $q_i$ in the discretized path.
 \begin{equation}
-\mathbf{r}(\mathbf{x},t)=\bf{c}_{i(t)}(x)
+\mathbf{U}_i(x)=[\mathbf{u}_1(\mathbf{x}),...,\mathbf{u}_n(\mathbf{x})]
 \end{equation}
-Where $\mathbf{r}(\mathbf{x},t)$ is a function of state and time. This arbitrary form of $\mathbf{r}(\mathbf{x},t)$, enables cycling through controllers along different paths at different frequencies.
- \brian[This is not clear.  Please re-write carefully and think about what the reader needs to know to understand and reproduce the method.]
+where $\mathbf{U}_i(x)$
+
+Then, we cycle through each controller at a fixed frequency by using a function which the cycles the controller index ($\phi(t)$) over time. \begin{equation}
+\phi(t)=\lceil \frac{n}{T} mod(t,T)\rceil = \lceil n \ f \ mod(t,\frac{1}{f})\rceil
+\end{equation}
+where $f$ the cycling frequency given by $1/T$. Then the final control equation is given by 
+ \begin{equation}
+\mathbf{R}(\mathbf{x},t)=\bf{U}_{\phi(t)}(x)
+\end{equation}
+Where $\mathbf{R}(\mathbf{x},t)$ returns the control input scalar as a function of state and time. 
+The dynamical system equation becomes
+\begin{equation}
+\frac{d}{dt}\mathbf{P}=\mathbf{A}+\mathbf{B}\bf{U}_{\phi(t)}(x)
+\end{equation}
 
@@
 \brian[What are the units in Figure 4G?]
+\michael[minutes]
 
@@
 To explore the joint effects of time delays and granularity, Fig,\ \ref{DG}J shows the FAC steady state control performance as a function of the time delay and system granularity $(\tau,\alpha)$ using thirty-two stochastic simulations simulated to steady state at each combination.
-Control performance errors measured by RMSE improved as $\alpha$ increased when the delays were small ($\tau=1$ and $\tau=10$) but not when $\tau=100$.
+Control performance errors measured by RMSE improved as $\alpha$ increased when the delays were small ($\tau=1 \ min$ and $\tau=10 \ min$) but not when $\tau=100 \ min$.
 \brian[What are the units of tau in fig 5?
@@
 If you plot that as a thick dashed dark line in Fig 5J,  would it match to the contours?
-]
+] \michael[$tc=1/(0.7*alpha)$ matches the counters well]
 Recalling that under the system granularity rescaling (Eq.\ \ref{eq:rescaling}), as $\alpha$ changes, the effective degradation rate scales according to $\gamma' = \gamma/\alpha$.
@@
 Optimized control inputs and probability distributions were optimized using the FAC control method(A and B) or the PAC control method (C and D).
-(E) Domains of control performance for the FAC method over a region of static target points show regions of good control and symmetry along the dashed line.
-(F) Domain of control performance over a set of static control points for the PAC control method.}
+(E) Domains of control performance for the FAC method over a region of static target points show regions of good control and symmetry along the dashed line.}
 \label{CtrlP}
@@
 For example, Fig \ref{CtrlP} (A) shows the optimized FAC control input of the system when the target is $\mathbf{T}=[20,25]$, and Fig.\ \ref{CtrlP} (B) shows the corresponding steady state probability distribution of the controlled system.
-Figure \ref{CtrlP} (C and D) show the the same thing but for a different target state of $\mathbf{T}=[XXX,XXX]$.
+Figure \ref{CtrlP} (C and D) show the the same thing but for a different target state of $\mathbf{T}=[10, 25]$.
 
@@
 Remove Fig 6F.
-Make sure to change all the figure labels and titles and captions.]
+Make sure to change all the figure labels and titles and captions.] \michael[ok, I also moved the settpoint score heatmap to the right]
 
 Figure \ref{CtrlP}(E) shows the FAC control performance over the entire domain of static set points, and illustrates that some regions are easier to obtain than others.
-For example, near the two stable fixed points at [XXX,XXX] and [XXX,XXX], the control performance is quite strong (white regions), whereas near the unstable point [20,20], the FAC results in relatively worse control.
+For example, in the space contained in the green box which also hold the four stable points of the system, the control performance is quite strong (white regions), whereas near the unstable point [20,20], the FAC results in relatively worse control.
 
-Edge regions were difficult to control, likely due to the geometry of how control inputs move the probability. \brian[I don't understand this statement.]
+The top left and bottom right edge regions were the most difficult to control, since increases the controller input does not push probability in these directions.
+\brian[I don't understand this statement.] \michael[editied it better]
 
@@
 
-FSP simulations for these cyclically driven systems were slowly driven at a frequency of $1e-4$ seconds \brian[Check units.
+FSP simulations for these cyclically driven systems were slowly driven at a frequency of $1e^{-4} min^{-1}$  \brian[Check units.
 This doesn't make sense.
@@
 If so, this is very fast.
-Or did you mean s$^{-1}$?] and Fig.\ \ref{CR}(B2, C2, and D2) show their response distributions (shading) and median responses (lines) for the two cells x1 and x2 in red and blue, respectively.
+Or did you mean s$^{-1}$?] \michael[it was $min^{-1}$ not s] and Fig.\ \ref{CR}(B2, C2, and D2) show their response distributions (shading) and median responses (lines) for the two cells x1 and x2 in red and blue, respectively.
 Regions with purple shading depict the exact overlap of red and blue when both cells have the same distribution of response.
@@
 Since we saw previously that increasing granularity ($\alpha$) improves steady state control performance (Fig.\ \ref{Time}G) but also lengthens this time scale (Fig.\ \ref{Time}H), one should expect that granularity would also affect the types and speeds of signals to which the system can respond.
-To examine how driving frequency and system scale affects control performance, 64 or 128 SSA trajectories of the phase-separated dynamic controller were simulated over a two dimensional domain of points $(\alpha,f)$ for 16 or 32 cycles after reaching steady state.
-Figure \ref{CRG}(A) shows the desired reference signal for the system, and Fig \ref{CRG}(B) shows the mean of the system response when the frequency is $f=10^{-4}$ \brian[Need units] and the granularity is $\alpha = 4.0$.
-Figure \ref{CRG}(C) shows the corresponding control performance over normalized periodic time as $\alpha$ is held at $4.0$ and $f$ is increased from $f=10^{-4}$ \brian[Units] to $f=10^{-2}$ \brian[Units] to $f=10^{-0}$ \brian[Units] leading to average control performances of 6.2, 16.1, and 15.0 respectively.
+To examine how driving frequency and system scale affects control performance, 64 SSA trajectories of the phase-separated dynamic controller were simulated over a two dimensional domain of points $(\alpha,f)$ for 16 or 32 cycles after reaching steady state.
+Figure \ref{CRG}(A) shows the desired reference signal for the system, and Fig \ref{CRG}(B) shows the mean of the system response when the frequency is $f=10^{-4} min^{-1}$ \brian[Need units] \michael[added them]and the granularity is $\alpha = 4.0$.
+Figure \ref{CRG}(C) shows the corresponding control performance over normalized periodic time as $\alpha$ is held at $4.0$ and $f$ is increased from $f=10^{-4}min^{-1}$ \brian[Units]  \michael[added them]to $f=10^{-2}min^{-1}$ \brian[Units]  \michael[added them]to $f=10^{-0}min^{-1}$ \brian[Units] \michael[added them] leading to average control performances of 6.2, 16.1, and 15.0 respectively.
 Figure \ref{CRG}D,E extends this analysis to plot the average control performance over different combinations of $\alpha$ and $f$. 
 %RMSE control performance of 128 simulations were simulated over a range of frequency between $10^-3$ and $10^-1$ Fig \ref{CRG}(D) for 32 cycles to obtain a better measurement of the true RMSE Fig \ref{CRG}(D) at $\alpha=2.0$, $\alpha=4.0$, and $\alpha=8.0$.
-In particular, the heat map of control performance over $(f,\alpha)$ shows that the worst performance occurs at moderate $f$ near $f=10^{-2}$ and that control performance is best when granularity is high and frequency is low.
+In particular, the heat map of control performance over $(f,\alpha)$ shows that the worst performance occurs at moderate $f$ near $f=10^{-2}min^{-1}$ and that control performance is best when granularity is high and frequency is low.
  When examining poorly performing trajectories at moderate frequencies, a phase lag was observed in the system.
@@
 \brian[ try adding the line at $f = 1/tau_c =  \gamma/(0.7*\alpha)$ to this plot.  Does it match the contours separating good and poor control?]
+\michael [it didnt fit well, but ~$f = 1/tau_c =  (0.7*\alpha)/\gamma$ fits well]
 

### 2023-07-28 paper.tex
updated again

@@
 
-We developed a method to create a noise-exploiting controller by taking full advantage of master equation analysis.
-This method uses non-linearity in a model, feedback control of the system and the exploitation of noise to optimize a Markovian controller which minimizes the expected squared Euclidian distance from the target position.
-Despite the advantages of full master equation analysis, good control of stochastic systems would ideally be resistant to parameter variability from heterogenous cell populations, time delay in controller actuation, and robust to systems with intrinsically more or less noise.
-To answer these questions, in the following subsections, we examine how parameter changes (e.g., incorrect models or extrinsic noise), time delays, and mismatches between system and models scale influence the performance of the controllers.
-Then, we extend these analyses to control a dynamically moving set point that seeks to control different identical cells to achieve different fluctuation phases and frequencies.
+\michael[new paragraph here]
 
+ In this section, we seek to show that noise-exploiting controllers are effective at controlling a noisy system to a dynamically moving target point.
+To do this, we show that control performance is robust to a range of parameter uncertainties despite not making adjustments to the controler.
+Then, we extend the control performance analysis to understand how time delay and system scale alters control performance.
+Then, we find regions in target space which are easiest to control by optimizing individual controllers for each target.
+We extend these analysis to change the target point over time in a Peicewise fashion.
  \subsection{Stochastic SIMO Optogenetic Control can Remain Effective Despite Small Parameter Errors or Extrinsic Uncertainties}
@@
 
-Real stochastic processes always have unknown or uncertain mechanisms and parameter values, and although model structures and parameter estimates can often be obtained through fitting to training data, these estimates will never be perfect due to unavoidable measurement errors or limited data sets.
-Furthermore, even with plentiful and precise training data, model and parameter uncertainties are inevitable due to the fact that even genetically identical single-cells exhibit heterogeneity in parameters due to extrinsic variations.
+Real stochastic processes always have unknown or uncertain mechanisms and parameter values.
+Although model structures and parameter estimates can often be obtained through fitting to training data, these estimates will never be perfect due to unavoidable measurement errors or limited data sets.
+Even with plentiful and precise training data, model and parameter uncertainties are inevitable due to the fact that even genetically identical single-cells exhibit heterogeneity in parameters due to extrinsic variations.
 To asses the control performance under such parameter errors, we performed sensitivity analysis on each model parameter for each cell, then on both cells simultaneously. 
@@
 
-Modification of parameters was found to produce a broad range of effects. 
+Modification of parameters were found to produce a broad range of effects. 
 For example, increasing $\beta$ in Cell 1 quickly worsens performance while increasing $\beta$ in Cell 2 improves performance (compare second column in Fig.\ \ref{Parameter}A and \ref{Parameter}B). 
@@
 Figure \ref{Volume}G shows the trend of the performance versus alpha for both the FAC (solid cyan line) and the PAC (solid magenta line) controller.
-This improvement in performance appears to approach a small value as the granularity goes to infinity, but since the size of the FSP increases with the square of the system size, systems much larger than $\alpha$=2 (where $\mathbf{A}_0\in \mathbb{R}^{10^4\times10^4}$) become more difficult to calculate using the current methods.
+This improvement in performance appears to approach a small value as the granularity goes to infinity, but since the size of the FSP increases with the square of the system size, systems much larger than $\alpha$=2 (where $\mathbf{A}_0\in \mathbb{R}^{10^4\times10^4}$) become more difficult to calculate using master equation techniques.
 To bypass this limit in the FSP, sixteen SSA simulations were used to sample the CME of a system with a much larger volume of $\alpha$=100.
 Each SSA was ran for $5\times10^7$ minutes and only the last $4\times10^7$ minutes were sampled to estimate the stationary distribution and calculate the performance score.
-The performance score estimates of this high granularity SSA using the FAC and PAC were 4.13 and 27.5 respectively, which are plotted as dashed lines in Fig.\ \ref{Volume}G.
+The performance score estimates of this high granularity SSA using the FAC and PAC were $2.03$ and $5.24$ respectively, which are plotted as dashed lines in Fig.\ \ref{Volume}G.
 Although it is unclear if further performance improvements could be obtained with further increases to the system scale, for all cases considered so far, we found that both controllers monotonically improved with increased $\alpha$ and that $\mathbf{u}^{\rm FAC}$ always outperforms the $\mathbf{u}^{\rm PAC}$.
@@
 Figure \ref{Time}G summarizes these results by plotting the score of both controllers versus the time delay.
-From the figures, it is clear that performance is rapidly degraded as the delay approaches and then exceeds the characteristic time scale ($1/\gamma$) set by the degradation rate of the process.
-At very small time delays (below $\tau = 0.07/\gamma = 0.0142$ min), the FAC outperforms the PAC ($J=$87 versus 146 at $\tau = 0.01/\gamma = 0.00203$ min) but at moderate time delays (above $\tau = 0.07/\gamma = 0.0142$ min) the PAC outperforms the FAC ($J=$170 versus 219 at $\tau = 0.1/\gamma = 0.0203$ min).
+From the figures, it is clear that performance is rapidly degraded as the delay approaches and then exceeds the characteristic time scale ($\tau_c = 1 / \gamma = 49 min$) set by the degradation rate of the process (yellow dashed line).
+At very small time delays ($\tau< 3.4 min$), the FAC outperforms the PAC  but at moderate time delays ($\tau > 3.4 min$) the PAC outperforms the FAC.
 
 \brian[Are these times correct?  Isn't gamma = 0.0203 in Table 1?  tau = 0.07 / 0.0203 (1/min) = 3.44 min. Please check this carefully.]
-
+\michael[I changed tau crit to be 0.1 since 0.07 is arbirarty anyway]
 \brian[What are the units in Figure 4G?]
@@
 
-This data show that time delays exceeding $\tau_c^{\rm FAC} = 0.07 / \gamma = 0.0142$ minutes is detrimental to the FAC control performance while delays beyond $\tau_c^{\rm PAC} = 0.2 / \gamma = 0.0406$ minutes are detrimental to the PAC controller.
-\brian[Check these times.] The choice of best controller depends on the level of time delay in the system.
-At extreme time delays, both controllers lose their asymmetry, resulting in significantly worse scores ($J = 1041$ and $931$).
-The control performance of a no-feedback controller was measured which required no inputs Figure\ref{Time}(G) (red line) and achieved a score of $J = 402$.
+This data show that time delays exceeding $\tau_c = 1 / \gamma = 49$ minutes begins to degrade control performance of both controllers.
+\brian[Check these times.] \michael[rewrote the sentece]The choice of best controller depends on the level of time delay in the system.
+At extreme time delays, both controllers lose their asymmetry, resulting in significantly worse scores ($RMSE = 32.2$ and $23.5$).
+The control performance of a no-feedback controller was measured which required no inputs Figure\ref{Time}(G) (red line) and achieved a score of $J = 20.0$.
 These findings indicate that, for large time delays, attempting to improve control with feedback is less effective than using a no-feedback approach.
@@
 Control performance errors measured by RMSE improved as $\alpha$ increased when the delays were small ($\tau=1 \ min$ and $\tau=10 \ min$) but not when $\tau=100 \ min$.
+
 \brian[What are the units of tau in fig 5?
@@
 ] \michael[$tc=1/(0.7*alpha)$ matches the counters well]
+
 Recalling that under the system granularity rescaling (Eq.\ \ref{eq:rescaling}), as $\alpha$ changes, the effective degradation rate scales according to $\gamma' = \gamma/\alpha$.
-Therefore, the critical limit for the delays should also change with the system scale according to $\tau_c^{\rm FAC} = 0.07\alpha/\gamma = XXX\alpha$.
+Therefore, the critical limit for the delays should also change with the system scale according to $\tau_c = 1\alpha/\gamma$.
 Fig \ref{DG}(J) depicts this critical line and shows that as the time delay approaches and then exceeds this level, the steady state performance become dramatically worse.
+\michael[edited this section]
+
 Finally, to understand the model of failure at these longer delays, it is interesting to examine trajectories induced by the controller just above this critical delay.
@@
 Figure \ref{CtrlP}(E) shows the FAC control performance over the entire domain of static set points, and illustrates that some regions are easier to obtain than others.
-For example, in the space contained in the green box which also hold the four stable points of the system, the control performance is quite strong (white regions), whereas near the unstable point [20,20], the FAC results in relatively worse control.
+For example, in the space contained in the dashed-green box which also hold the four stable points of the system, the control performance is quite strong, whereas near the unstable point (dashed-blue box), the FAC results in relatively worse control.
 

### 2023-07-31 paper.tex
update

@@
 For example, inspired by the genetic toggle switch from Kobayashi et al.\cite{Kobayashi:2004}, Szymanska et al.~\cite{Szymanska2015} showed that noise could be exploited to achieve independent control of multiple cells using a single input, even despite uncertain parameters or time delays due to maturation of fluorescent proteins or limited observation of the regulatory proteins.
-In May et al.~\cite{May2021}, we identified a simplified stochastic model to reproduce data measured in Baumschlager, et al.\cite{XXX} for the expression from the XXX promoter under the optogenetic control of a UV-activated T7 polymerase (see model in Figure\ \ref{cartoons}A, top promoter).
-We then proposed the addition of a positive auto-regulation (Figure\ \ref{cartoons}A, bottom promoter) to help maintain an elevated expression phenotype in the presence of UV excitation, and we demonstrated how a Single-Input-Multiple-Output (SIMO) multicellular controller could control multiple cells to arbitrary phenotypes using only a single input.
+In May et al.~\cite{May2021}, we identified a simplified stochastic model to reproduce data measured in Baumschlager, et al.\cite{XXX} for the expression from the XXX promoter under the optogenetic control of a UV-activated T7 polymerase (see model in Figure \ref{cartoons}A, top promoter).
+We then proposed the addition of a positive auto-regulation (Figure \ref{cartoons}A, bottom promoter) to help maintain an elevated expression phenotype in the presence of UV excitation, and we demonstrated how a Single-Input-Multiple-Output (SIMO) multicellular controller could control multiple cells to arbitrary phenotypes using only a single input.
 
@@
 To assess the impact of model approximations on the implementation of noise-enhanced control strategies, we begin with the one-species model proposed by May et al.
-\cite{May2021} (Fig.
+\cite{May2021} (Figure.
 \ref{cartoons}A).
@@
 to find $\frac{dP}{d\mathbf{u}^{FAC}}$.
-Optimization of $\mathbf{u}^{PAC}$ was done by restricting $\frac{d(J^2)}{d\mathbf{u}^{FAC}}$ in ambient space to search a low dimensional subspace with a linear lift operator D.
+
+%Optimization of $\mathbf{u}^{PAC}$ was done by restricting $\frac{d(J^2)}{d\mathbf{u}^{FAC}}$ in ambient space to search a low dimensional subspace with a linear lift operator D.
 Then the subspace search is given by.
-\begin{equation}
-\frac{d(J^2)}{d\mathbf{u}^{PAC}}=C \frac{dP}{d\mathbf{u}^{PAC}} = C D \frac{dP}{d\mathbf{u}^{FAC}}  
-\end{equation}
+%\begin{equation}
+%\frac{d(J^2)}{d\mathbf{u}^{PAC}}=C \frac{dP}{d\mathbf{u}^{PAC}} = C D \frac{dP}{d\mathbf{u}^{FAC}}  
+%\end{equation}
+
 %Root mean squared error was calculated by taking the square root of the score.
@@
 
-In \cite{May2021}, we optimized each of these two controllers to find the control laws $\mathbf{u}^{FAC}$ and $\mathbf{u}^{PAC}$ that minimize $J^2$ for a single target state $\mathbf{T} = [10,30]$, but in this work extend it to any target point in the state space.
+In \cite{May2021}, controllers were optimized to minimize $J^2$ for a single target state $\mathbf{T} = [10,30]$, but in this work they are extend to different arbitrary target points.
 
@@
 To investigate the effects of time delay on control actuation, we devised a simple time-delay stochastic simulation algorithm. 
-This algorithm records the state history after each stochastic jump, enabling reconstruction of the population history of the species and control input.  
+This algorithm records the state history after each stochastic jump, enabling reconstruction of the population history of the species and the time delayed control input propensity.  
 Using this information, the time-delayed control signal at time $t$ can be specified as:
@@
 \end{equation}
-where $\tau$ is the time delay between observation and actuation, and ${u}^{\mathcal{C}}$ is the previously optimized control law (e.g., $\mathbf{u}^{\rm FAC}$ or $\mathbf{u}^{\rm PAC}$). 
+where $\tau$ is the time delay between observation and actuation, ${u}^{\mathcal{C}}$ is the previously optimized control law (e.g., $\mathbf{u}^{\rm FAC}$ or $\mathbf{u}^{\rm PAC}$), and $x$ gathered from the state history.
 \brian[Explain how you modify the SSA for the delayed reaction SSA.  Look through the literature to cite someone else who has implenented a similar method.]
@@
 Because optimizing the control law for a static set point as described in Section \ref{sec:Quantification} requires differentiation of the CME (Eq.
-\ref{XXX}) with respect to the control signal at each state, this calculation is approaching the limits of current feasibility.
+\ref{CME}) with respect to the control signal at each state, this calculation is approaching the limits of current feasibility.
 Extending these calculations to optimize controllers for a dynamically moving set point is much harder and would likely require intractable numerical simulation or development of new mathematical approaches that are beyond the scope of the current study.
@@
 \begin{equation}
-\mathbf{U}_i(x)=[\mathbf{u}_1(\mathbf{x}),...,\mathbf{u}_n(\mathbf{x})]
+\mathbf{U}_i(\mathbf{x})=[\mathbf{u}_1(\mathbf{x}),...,\mathbf{u}_n(\mathbf{x})]
 \end{equation}
-where $\mathbf{U}_i(x)$
-
-Then, we cycle through each controller at a fixed frequency by using a function which the cycles the controller index ($\phi(t)$) over time. \begin{equation}
+Next, we cycle through each controller at a fixed frequency by using ($\phi(t)$). 
+\begin{equation}
 \phi(t)=\lceil \frac{n}{T} mod(t,T)\rceil = \lceil n \ f \ mod(t,\frac{1}{f})\rceil
@@
 \begin{equation}
-\frac{d}{dt}\mathbf{P}=\mathbf{A}+\mathbf{B}\bf{U}_{\phi(t)}(x)
+\frac{d}{dt}\mathbf{P}=\mathbf{A}+\mathbf{B}\mathbf{R}(\mathbf{x},t)
 \end{equation}
-
+Since $\mathbf{R}(\mathbf{x},t)$ is a function of time, the system is no longer time invariant. 
 \section{Results}
@@
 In this section, we seek to show that noise-exploiting controllers are effective at controlling a noisy system to a dynamically moving target point.
-To do this, we show that control performance is robust to a range of parameter uncertainties despite not making adjustments to the controler.
+To do this, we show that control performance is robust to a range of parameter uncertainties without the needing adjustments to the controller.
 Then, we extend the control performance analysis to understand how time delay and system scale alters control performance.
-Then, we find regions in target space which are easiest to control by optimizing individual controllers for each target.
+Furthermore, we find regions in target space which are easiest to control by optimizing individual controllers for each target.
 We extend these analysis to change the target point over time in a Peicewise fashion.
@@
 To asses the control performance under such parameter errors, we performed sensitivity analysis on each model parameter for each cell, then on both cells simultaneously. 
-For each parameter, both the $\mathbf{u}^{\rm FAC}$ and $\mathbf{u}^{\rm PAC}$ controllers were examined across a parameter perturbation range from one-tenth to ten-fold of the original values.
-Figure \ref{Parameter}A shows the results for these parameter sweeps when a single parameter in Cell 1 is varied (all other parameters are fixed), and Fig.\ \ref{Parameter}B shows the control performance as single parameters in Cell 2 is varied.
-Finally, \ref{Parameter}B shows the performance change when a single parameter is changed simultaneously in both Cell 1 and Cell 2 at the same time while keeping other parameters fixed.
-In each case the baseline parameters is denoted as a vertical dashed line, and lower values of J (Eq. \ref{XXX}) denote improved performance. 
-Solid lines are used to show the performance using the fully aware controller (FAC) that has access to both cells, and the dashed lines show the performance for the partially aware controller (PAC) that only has access to Cell 1.
+For each parameter, $\mathbf{u}^{\rm FAC}$ (solid lines) and $\mathbf{u}^{\rm PAC}$(dashed lines) control performance was examined across a parameter perturbation range from one-tenth to ten-fold of the original values.
+Figure \ref{Parameter}A shows the results for these parameter sweeps when a single parameter in Cell 1 is varied (all other parameters are fixed), and Figure.\ \ref{Parameter}B shows the control performance as single parameters in Cell 2 is varied.
+Finally, \ref{Parameter}C shows the performance change when a single parameter is changed simultaneously in both Cell 1 and Cell 2 at the same time.
 
 Modification of parameters were found to produce a broad range of effects. 
-For example, increasing $\beta$ in Cell 1 quickly worsens performance while increasing $\beta$ in Cell 2 improves performance (compare second column in Fig.\ \ref{Parameter}A and \ref{Parameter}B). 
-In some cases, the effects on performance are not monotonic; for example, increasing $\kappa$ in Cell 1 (Fig.\ \ref{Parameter}A, leftmost column) would be highly advantageous up to a limit after which the control performance degrades rapidly.
+For example, increasing $\beta$ in Cell 1 quickly worsens performance while increasing $\beta$ in Cell 2 improves performance (compare second column in Figure.\ \ref{Parameter}A and \ref{Parameter}B).
+In some cases, the effects on performance are not monotonic; for example, increasing $\kappa$ in Cell 1 (Figure.\ \ref{Parameter}A, leftmost column) would be highly advantageous up to a limit after which the control performance degrades rapidly.
 In other cases (such as for the promoter leakage rate, $k_0$), the effect of parameter perturbations on performance is insignificant even for relatively large ($s$=10) perturbations, suggesting that this parameter is not important to control performance.
@@
 Each SSA was ran for $5\times10^7$ minutes and only the last $4\times10^7$ minutes were sampled to estimate the stationary distribution and calculate the performance score.
-The performance score estimates of this high granularity SSA using the FAC and PAC were $2.03$ and $5.24$ respectively, which are plotted as dashed lines in Fig.\ \ref{Volume}G.
+The performance score estimates of this high granularity SSA using the FAC and PAC were $2.03$ and $5.24$ respectively, which are plotted as dashed lines in Figure.\ \ref{Volume}G.
 Although it is unclear if further performance improvements could be obtained with further increases to the system scale, for all cases considered so far, we found that both controllers monotonically improved with increased $\alpha$ and that $\mathbf{u}^{\rm FAC}$ always outperforms the $\mathbf{u}^{\rm PAC}$.
@@
 Although higher granularity reduces noise and makes it easier to maintain desired states once they have been achieved, noise is necessary to break symmetry between the two cells' dynamics to achieve those states in the first place.
-This tradeoff is illustrated in Fig \ref{Volume}(H), which plots the control performance over time after the changing the controller goal to exchange the low- and high-target cells with one another.
+This tradeoff is illustrated in Figure\ref{Volume}(H), which plots the control performance over time after the changing the controller goal to exchange the low- and high-target cells with one another.
 From the figure, we can see although steady state control performance improves for higher values of $\alpha$, the time taken to reach that steady state performance increases also increases $\alpha$.
@@
 Figure \ref{Time}G summarizes these results by plotting the score of both controllers versus the time delay.
-From the figures, it is clear that performance is rapidly degraded as the delay approaches and then exceeds the characteristic time scale ($\tau_c = 1 / \gamma = 49 min$) set by the degradation rate of the process (yellow dashed line).
-At very small time delays ($\tau< 3.4 min$), the FAC outperforms the PAC  but at moderate time delays ($\tau > 3.4 min$) the PAC outperforms the FAC.
+From the figures, it is clear that performance is rapidly degraded as the delay approaches and then exceeds the characteristic time scale ($\tau_c = 1 / \gamma = 49$ min) set by the degradation rate of the process (yellow dashed line).
+At very small time delays ($\tau< 3.4$ min), the FAC outperforms the PAC  but at moderate time delays ($\tau > 3.4$) the PAC outperforms the FAC.
 
 \brian[Are these times correct?  Isn't gamma = 0.0203 in Table 1?  tau = 0.07 / 0.0203 (1/min) = 3.44 min. Please check this carefully.]
-\michael[I changed tau crit to be 0.1 since 0.07 is arbirarty anyway]
+\michael[I changed characteristic tau to be 1 since it fits all the curves the best]
 \brian[What are the units in Figure 4G?]
@@
 
-This data show that time delays exceeding $\tau_c = 1 / \gamma = 49$ minutes begins to degrade control performance of both controllers.
-\brian[Check these times.] \michael[rewrote the sentece]The choice of best controller depends on the level of time delay in the system.
-At extreme time delays, both controllers lose their asymmetry, resulting in significantly worse scores ($RMSE = 32.2$ and $23.5$).
-The control performance of a no-feedback controller was measured which required no inputs Figure\ref{Time}(G) (red line) and achieved a score of $J = 20.0$.
+These data show that time delays greater than $\tau = 0.1 / \gamma = 4.9$ min begin to degrade FAC control performance and plateau at $\tau_c = 1.0 / \gamma = 49$ min.
+\brian[Check these times.] \michael[rewrote the sentence]The choice of best controller depends on the level of time delay in the system.
+At extreme time delays, both controllers lose their asymmetry, resulting in significantly worse scores ($RMSE = 32$ and $23$).
+The control performance of simple constant control input without feedback was better than both controlers at very large time delays.
+Figure\ref{Time}(G) (red line) and achieved a score of $J = 20.0$.
 These findings indicate that, for large time delays, attempting to improve control with feedback is less effective than using a no-feedback approach.
@@
 To explore the joint effects of time delays and granularity, Fig,\ \ref{DG}J shows the FAC steady state control performance as a function of the time delay and system granularity $(\tau,\alpha)$ using thirty-two stochastic simulations simulated to steady state at each combination.
-Control performance errors measured by RMSE improved as $\alpha$ increased when the delays were small ($\tau=1 \ min$ and $\tau=10 \ min$) but not when $\tau=100 \ min$.
+Control performance errors measured by RMSE improved as $\alpha$ increased when the delays were small ($\tau=1 \ $ min and $\tau=10 \ $ min) but not when $\tau=100 \ $ min.
 
-\brian[What are the units of tau in fig 5?
+\brian[What are the units of tau in Figure5?
 This section is confusing in that it is not clear to me how tau and gamma are changing relative to one another when alpha is changed.
@@
 If this is correct, does that mean that the critical time changes with $\tau_c = 1/gamma = alpha/gamma ~= 5*alpha$?
-If you plot that as a thick dashed dark line in Fig 5J,  would it match to the contours?
+If you plot that as a thick dashed dark line in Figure5J,  would it match to the contours?
 ] \michael[$tc=1/(0.7*alpha)$ matches the counters well]
@@
 Therefore, the critical limit for the delays should also change with the system scale according to $\tau_c = 1\alpha/\gamma$.
-Fig \ref{DG}(J) depicts this critical line and shows that as the time delay approaches and then exceeds this level, the steady state performance become dramatically worse.
+Figure\ref{DG}(J) depicts this characteristic line and shows that as the time delay approaches and then exceeds this level, the steady state performance become dramatically worse.
 \michael[edited this section]
 
-Finally, to understand the model of failure at these longer delays, it is interesting to examine trajectories induced by the controller just above this critical delay.
-For parameter set denoted by the yellow start in Fig.\ \ref {DG}(J), Fig.\ \ref {DG}(K) shows the controlled response after a long burn in period to achieve steady state.
+Finally, to understand the model of failure at these longer delays, it is interesting to examine trajectories induced by the controller just above this characteristic delay.
+For parameter set denoted by the yellow start in Figure.\ \ref {DG}(J), Figure.\ \ref {DG}(K) shows the controlled response after a long burn in period to achieve steady state.
 In this case, the application of feedback after a delay leads to a strong oscillatory behavior and worse performance than that achieved without any control at all.
@@
 We next optimized the FAC controller and calculated its performance for the two-cell system over a two dimensional domain of target points $[T_1,T_2]$ between between five and forty-five.
-For example, Fig \ref{CtrlP} (A) shows the optimized FAC control input of the system when the target is $\mathbf{T}=[20,25]$, and Fig.\ \ref{CtrlP} (B) shows the corresponding steady state probability distribution of the controlled system.
+For example, Figure\ref{CtrlP} (A) shows the optimized FAC control input of the system when the target is $\mathbf{T}=[20,25]$, and Figure.\ \ref{CtrlP} (B) shows the corresponding steady state probability distribution of the controlled system.
 Figure \ref{CtrlP} (C and D) show the the same thing but for a different target state of $\mathbf{T}=[10, 25]$.
@@
 \brian[Let's remove the PAC from this section.
-Show only the FAC and and replace the Fig 6C/D with the FAC at a different target point.
+Show only the FAC and and replace the Figure6C/D with the FAC at a different target point.
 Put a new star of a different color to match to the new 6C/D.
-Remove Fig 6F.
+Remove Figure6F.
 Make sure to change all the figure labels and titles and captions.] \michael[ok, I also moved the settpoint score heatmap to the right]
@@
 
-The top left and bottom right edge regions were the most difficult to control, since increases the controller input does not push probability in these directions.
+The top left and bottom right edge regions were the most difficult to control, since the control input does not push probability in these directions.
 \brian[I don't understand this statement.] \michael[editied it better]
@@
 
-We developed a method (Section \ref{sec:Dynamic}) to control the system dynamics to follow a predefined path $r(x)$ by cycling through controllers optimized for individual states along that path.
-Specifically, we found 32 controllers along each three representative pathways, including an in-sync reference point (Fig.\ \ref{CR}B1), a phase lagged reference point (Fig.\ \ref{CR}C1),  and a  frequency separated reference point (Fig.\ \ref{CR}D1).
-The synchronous signal is given by Fig \ref{CR}(B1) $r_1(x)=[20 sin(2 \pi x)+10, 20 sin(2 \pi x)+10]$, the phase lagged signal Fig \ref{CR}(C1) is given by $r_2(x)=[20 sin(2 \pi x)+10, 20 cos(2 \pi x)+10]$, and the frequency changed signal Fig \ref{CR}(D1) is given by $r_3(x)=[20 sin(2 \pi x)+10, 20 sin(6 \pi x)+10]$ using 32 steps for each controller.
+We developed a method (Section \ref{sec:Dynamic}) to control the system dynamics to follow a predefined path $p(t)$.
+Specifically, three controllers were optimized along three representative pathways in thirty-two steps each, which include an in-sync reference point (Figure.\ \ref{CR}B1), a phase lagged reference point (Figure.\ \ref{CR}C1),  and a  frequency separated reference point (Figure.\ \ref{CR}D1).
+The synchronous signal is given by Figure \ref{CR}(B1) $p_1(t)=[20 sin(2 \pi f t)+10, 20 sin(2 \pi f t)+10]$, the phase lagged signal Figure \ref{CR}(C1) is given by $p_2(t)=[20 sin(2 \pi f t)+10, 20 cos(2 \pi f t)+10]$, and the frequency changed signal Figure \ref{CR}(D1) is given by $p_3(t)=[20 sin(2 \pi f t)+10, 20 sin(6 \pi f t+3\pi)+10]$ using 32 steps for each controller.
 
-FSP simulations for these cyclically driven systems were slowly driven at a frequency of $1e^{-4} min^{-1}$  \brian[Check units.
+All FSP simulations were slowly driven at a frequency of $1e^{-4} min^{-1}$  \brian[Check units.
 This doesn't make sense.
@@
 If so, this is very fast.
-Or did you mean s$^{-1}$?] \michael[it was $min^{-1}$ not s] and Fig.\ \ref{CR}(B2, C2, and D2) show their response distributions (shading) and median responses (lines) for the two cells x1 and x2 in red and blue, respectively.
+Or did you mean s$^{-1}$?] \michael[it was $min^{-1}$ not s] and Figure.\ \ref{CR}(B2, C2, and D2) show their response distributions (shading) and median responses (lines) for the two cells $x_1$ and $x_2$ in red and blue, respectively.
 Regions with purple shading depict the exact overlap of red and blue when both cells have the same distribution of response.
-We observe that the single-input control system is able to effectively control the system to follow all three input trajectories, including when the two reference trajectories have different phases and frequencies (Fig \ref{CR}D3).
-For all reference trajectories, the medians of the marginal distributions show good match with the desired reference signal (Fig \ref{CR}B4,C4, and D4).
+We observe that the single-input control system is able to effectively control the system to follow all three input trajectories, including when the two reference trajectories have different phases and frequencies (Figure \ref{CR}D2).
+For all reference trajectories, the medians of the marginal distributions show good match with the desired reference signal (Figure \ref{CR}B2,C2, and D2).
 To quantify the overall performance, Figs.\ \ref{CR}(B3, C3,D3) show the root mean squared error as a function of time,
-and Figs.\ \ref{CR}(B2, C2, and D2) show the time-averaged error distribution for the system response relative to the time varying target.
-From these figures, we observe that all trajectories result in short RMSE spikes at times, and these correspond to short, transient periods when the controller passes through the regions of poor control depicted in Fig.\ \ref{CtrlP}(E).
-Overall, the simulations show that the synchronous control performed best with an average RSME of $6.9$ despite moving over the unstable point in the system.
+and Figs.\ \ref{CR}(B4, C4, and D4) show the time-averaged error distribution for the system response relative to the time varying target.
+From these figures, we observe that all trajectories result in short RMSE spikes at times, and these correspond to short, transient periods when the controller passes through the regions of poor control depicted in Figure.\ \ref{CtrlP}(E).
+Overall, the simulations show that the synchronous control performed best with an average RSME of $6.9$.
 When the system is driven with a phase-lag, the average RSME of the score increases to $8.2$, and when driven at a different frequency, the average RMSE goes up to $8.6$.
@@
 In general, a given control system cannot effectively track a reference signal that changes faster that the system's natural time scale.
-Since we saw previously that increasing granularity ($\alpha$) improves steady state control performance (Fig.\ \ref{Time}G) but also lengthens this time scale (Fig.\ \ref{Time}H), one should expect that granularity would also affect the types and speeds of signals to which the system can respond.
-To examine how driving frequency and system scale affects control performance, 64 SSA trajectories of the phase-separated dynamic controller were simulated over a two dimensional domain of points $(\alpha,f)$ for 16 or 32 cycles after reaching steady state.
-Figure \ref{CRG}(A) shows the desired reference signal for the system, and Fig \ref{CRG}(B) shows the mean of the system response when the frequency is $f=10^{-4} min^{-1}$ \brian[Need units] \michael[added them]and the granularity is $\alpha = 4.0$.
+Since it was shown that increasing granularity ($\alpha$) improves steady state control performance (Figure.\ \ref{Time}G) but also lengthens this time scale (Figure.\ \ref{Time}H), one should expect that these competing effects of granularity would also affect the types and speeds of signals to which the system can respond.
+To examine how driving frequency and system scale affects control performance, 64 SSA trajectories of the phase-separated dynamic controller were simulated over a two dimensional domain of points $(\alpha,f)$ for 32 cycles after reaching steady state.
+Figure \ref{CRG}(A) shows the desired reference signal for the system, and Figure \ref{CRG}(B) shows the mean of the system response when the frequency is $f=10^{-4} min^{-1}$ \brian[Need units] \michael[added them]and the granularity is $\alpha = 4.0$.
 Figure \ref{CRG}(C) shows the corresponding control performance over normalized periodic time as $\alpha$ is held at $4.0$ and $f$ is increased from $f=10^{-4}min^{-1}$ \brian[Units]  \michael[added them]to $f=10^{-2}min^{-1}$ \brian[Units]  \michael[added them]to $f=10^{-0}min^{-1}$ \brian[Units] \michael[added them] leading to average control performances of 6.2, 16.1, and 15.0 respectively.
 Figure \ref{CRG}D,E extends this analysis to plot the average control performance over different combinations of $\alpha$ and $f$.
-%RMSE control performance of 128 simulations were simulated over a range of frequency between $10^-3$ and $10^-1$ Fig \ref{CRG}(D) for 32 cycles to obtain a better measurement of the true RMSE Fig \ref{CRG}(D) at $\alpha=2.0$, $\alpha=4.0$, and $\alpha=8.0$.
-In particular, the heat map of control performance over $(f,\alpha)$ shows that the worst performance occurs at moderate $f$ near $f=10^{-2}min^{-1}$ and that control performance is best when granularity is high and frequency is low.
+The characteristic frequency (given by the predicted $\tau_c^{-1}$ over a range of $\alpha$) is plotted as the yellow dashed line.
+%RMSE control performance of 128 simulations were simulated over a range of frequency between $10^-3$ and $10^-1$ Figure\ref{CRG}(D) for 32 cycles to obtain a better measurement of the true RMSE Figure\ref{CRG}(D) at $\alpha=2.0$, $\alpha=4.0$, and $\alpha=8.0$.
+In particular, the heat map of control performance over $(f,\alpha)$ shows that the worst performance occurs at moderate $f$ near $f=10^{-2}min^{-1}$ and that control performance is best when granularity is high and frequency is low \ref{CRG}(E).
+Plots of the characteristic frequency given by the inverse of characteristic time \ref{CRG}(yellow line), show good match with the heat maps at high granularity but not at low granularity.
+These data
  When examining poorly performing trajectories at moderate frequencies, a phase lag was observed in the system.
-In this case, although the qualitative form of the response may match to the reference signal, this phase lag resulted in worse RMSE than no controller at all.
+In this case, although the qualitative form of the response may match to the reference signal, this phase lag resulted in worse RMSE than no system response at all.
 At very large frequencies, the effect of the fast fluctuating control input is averaged, and the system barely responds.
@@
 \brian[ try adding the line at $f = 1/tau_c =  \gamma/(0.7*\alpha)$ to this plot.  Does it match the contours separating good and poor control?]
-\michael [it didnt fit well, but ~$f = 1/tau_c =  (0.7*\alpha)/\gamma$ fits well]
+\michael[added $\tau_c^{-1}$. using $\tau_c=1/ \gamma$]
 

### 2023-07-31 backup/paper_BrianJuly31.tex
Update text.

--- 2023-07-31 paper.tex
@@
 \usepackage{xcolor}
+\usepackage{hyperref}
 %\usepackage{graphicx}
@@
 %\usepackage{iopams}
+
 \begin{document}
-
 \def\brian[#1]{{\color{red} #1 }}
@@
 
-\title[Heterogeneous Control of Cell Differentiation Dynamics]{Exploiting Intrinsic Noise to Achieve Heterogeneous Control of Cell Differentiation Dynamics in the Presence of Time Delays, Model Uncertainties, and Extrinsic Variability}
+\title[Heterogeneous Control of Cell Differentiation Dynamics]{Exploiting Intrinsic Noise for Heterogeneous Cell Control Under Time Delays and Model Uncertainties}
 \maketitle
@@
 \begin{abstract}
-Previous research has focused on robust control performance in the presence of noise, but understanding controllers that exploit noise remains incomplete. 
-Motivated by Maxwells-Demon, we previously proposed a difficult control problem which requires the exploitation of noise in a stochastic system to break symmetry between two signals.
-We found multiple such controllers which could exploit noise to break symmetry between two cells under a variety of system information.
-This work extends that analysis to include stochastic systems that trace a moving target, are affected by time delay, intrinsic noise, or measurement uncertainty.
- We find that noise-exploiting controllers can remain highly effective despite coarse approximations to the model's scale or incorrect estimations of key model parameters, and these controllers can even retain performance for significant time delays.
-Together, these findings suggest that noise-exploiting control should be possible where models are always approximate, where parameters are always uncertain, and where observations are corrupted by errors.
+Previous research has focused to enable robust control performance despite the presence of noise, but understanding how controllers may exploit that noise remains incomplete. 
+Motivated by Maxwells-Demon, we previously proposed a cellular control regime in which the exploitation of stochastic noise can break symmetry between and allow for specific control of multiple cells using a single input signal (i.e., single-input-multiple-output or SIMO control).
+The current work extends that analysis to include uncertain stochastic systems where system dynamics are are affected by time delays, intrinsic noises, and model uncertainty. 
+We find that noise-exploiting controllers can remain highly effective despite coarse approximations to the model's scale or incorrect estimations or extrinsic noise in key model parameters, and these controllers can even retain performance under substantial observer or actuator time delays.
+We also demonstrate how SIMO controllers could drive multi-cell systems to follow different trajectories with different phases and frequencies.
+Together, these findings suggest that noise-exploiting control should be possible even in the practical case where models are aways approximate, where parameters are always uncertain, and where observations are corrupted by errors.
 
@@
 \includegraphics[width=\columnwidth]{Cartoons.pdf}
-\caption{{\bf Optogenetic control of multiple cells using a single light input.}
-({\bf A}) Schematic of the light activated genetic system with autoregulation.
-({\bf B}) Diagram of the stochastic SIMO control problem using two optogenetic cells sharing an input.
-({\bf C}) Adapted from (PAPER).
-Noise exploiting controllers were optimized to a fully aware control input (I), and partially aware control input(III).
-Marginal distributions are solved using the finite state projection and show a break in symmetry using both controllers (II and IV) }
+\caption{{\bf Single-input-multiple-output (SIMO) control of multiple cells using a single optogenetic input.}
+({\bf A}) Schematic of the light-activated genetic system with auto-regulation.
+({\bf B}) Diagram of the stochastic SIMO control problem using two optogenetic cells sharing a single input.
+({\bf C}) Noise exploiting controllers were optimized to define a fully aware control input (I) and a partially aware control input (III), where the control signal (color scale) depends on the observed expression within the cell or cells (x- and y-axes).
+(II and IV) Corresponding steady state marginal distributions for the different cells (red and blue) under these controllers demonstrate a clear a break in symmetry.
+Dashed lines represent the control target objective.
+Control performance (RMSE) is show above each distribution.}
 \label{cartoons}
@@
 \end{figure}
-% XXX - Please modify figure as follows: (1) switch panels A and B to match caption.
-(2) Change title of panel B (previously panel A) to "External Observer, controller and actuator".
-(3) Change title of panel A (previously panel B) to "Internal gene regulation pathway".
-(4) Change title of panel C to "Feedback control law and resulting performance"
-% MMM - Made the edit
 
-Uncertain fluctuations, or `noise,' is a common theme throughout many fields of engineering, and is a frequent concern when attempting to control any human-made system such as vehicles\cite{XXX}, production machinery\cite{XXX}, or chemical processes\cite{XXX}.
-Noise arrises from many sources, occasionally from quantum events and thermally-induced fluctuations but more commonly from unknown or uncharacterized physical processes, and as such, these fluctuations usually cannot be efficiently reproduced or predicted using purely physical or mechanistic models.
+Uncertain fluctuations, or `noise,' is a common theme throughout many fields of engineering, and is a frequent concern when attempting to control any human-made system such as vehicles \brian[\cite{XXX}], production machinery \brian[\cite{XXX}], or chemical processes \brian[\cite{XXX}].
+\brian[Please look for some relevant references or a review of control under noise. I highlight several missing refs in red.]
+Noise arrises from many sources, occasionally from quantum events and thermally-induced fluctuations, but more commonly from unknown or uncharacterized physical processes, and as such, these fluctuations usually cannot be efficiently reproduced or predicted using purely physical or mechanistic models.
 Rather, statistical models are usually employed where key mechanisms or dynamics are subject to stochastic variations or inputs that act as a proxy for random or un-modeled fluctuations elsewhere in the system.
-In this context, much effort has been placed on enhancing the robustness of controlled processes, where uncertain or unpredictable variations in internal or external parameters are modeled as intrinsic or extrinsic noise \cite{XXX}.
-For many macroscopic, human-engineered systems, the resulting system dynamics can be modeled effectively simply by combining a deterministic (i.e., noise-free) model with additive noise (usually assumed to be Gaussian under arguments based on the central limit theorem)\ref{XXX}, and most current approaches seek to control the system to minimize variations that result from the noisy inputs.
+In this context, much effort has been placed on enhancing the robustness of controlled processes, where uncertain or unpredictable variations in internal or external parameters are modeled as intrinsic or extrinsic noise \brian[\cite{XXX}].
+For many macroscopic, human-engineered systems, the resulting system dynamics can be modeled effectively simply by combining a deterministic (i.e., noise-free) model with additive noise (usually assumed to be Gaussian under arguments based on the central limit theorem) \brian[\cite{XXX}], and most current approaches seek to control the system to minimize variations that result from the noisy inputs.
 
-Analyses of noise and control theory are equally relevant to understand the basic biological processes of gene transcription regulation\cite{XXX} or mRNA translation regulation \cite{XXX} or to modify these processes for practical use\cite{XXX}.
-However, at this mesoscopic scale of cellular biology, where transcription factors compete to activate or deactivate individual genes or where mRNA or protein molecules are present at just a few copies (or none at all) per cell, additive noise model is far less realistic.
-In this case, Brownian motion, discrete stochastic gene regulation, and noisy mRNA dynamics collectively generate a fundamentally stocastic environment that cells must effectively manage. 
+Analyses of noise and control theory are equally relevant to understand the basic biological processes of gene transcription regulation \brian[\cite{XXX}] or mRNA translation regulation \brian[\cite{XXX}]or to modify these processes for practical use \brian[\cite{XXX}].
+However, at this mesoscopic scale of cellular biology, where transcription factors compete to activate or deactivate individual genes or where mRNA or protein molecules are present at just a few copies (or none at all) per cell, the additive noise model is much less realistic.
+In this case, Brownian motion, discrete stochastic gene regulation, and noisy mRNA dynamics collectively generate a fundamentally stochastic environment that cells must effectively manage. 
 At this scale, the order or timing of a single reaction event (e.g., the binding of a transcription factor to a promoter) can have dramatic consequences that could last for several cellular generations (e.g., the activation or repression of a gene that promotes unfettered cell growth and differentiation).
@@
 These digital-synthetic actuators enable fine-tuned, computer-modulated control of cellular systems, previously unattainable, with faster response times compared to chemical diffusion\cite{Rullan2018, Baumschlager2017} .
-Classical and modern control methods like PID control and model predictive control have been implemented in such systems\cite{} to control synthetic systems to different stable points. 
-% XXX Missing Citation above.
+Classical and modern control methods like PID control and model predictive control have been implemented in such systems \brian[\cite{XXX}] to control synthetic systems to different stable points.
 
@@
 For example, inspired by the genetic toggle switch from Kobayashi et al.\cite{Kobayashi:2004}, Szymanska et al.~\cite{Szymanska2015} showed that noise could be exploited to achieve independent control of multiple cells using a single input, even despite uncertain parameters or time delays due to maturation of fluorescent proteins or limited observation of the regulatory proteins.
-In May et al.~\cite{May2021}, we identified a simplified stochastic model to reproduce data measured in Baumschlager, et al.\cite{XXX} for the expression from the XXX promoter under the optogenetic control of a UV-activated T7 polymerase (see model in Figure \ref{cartoons}A, top promoter).
+In May et al.~\cite{May2021}, we identified a simplified stochastic model to reproduce data measured in Baumschlager, et al.
+\brian[\cite{XXX}] for the expression from a transcription promoter under the optogenetic control of a UV-activated T7 polymerase (see model in Figure \ref{cartoons}A, top promoter).
 We then proposed the addition of a positive auto-regulation (Figure \ref{cartoons}A, bottom promoter) to help maintain an elevated expression phenotype in the presence of UV excitation, and we demonstrated how a Single-Input-Multiple-Output (SIMO) multicellular controller could control multiple cells to arbitrary phenotypes using only a single input.
-
-%The solutions of the controllers was carried out by solving the chemical master equation (i.e., the forward Komogorov equation) to estimate the probability of possible cellular states.
-For a two cell, single-input-single output (SIMO) control problem, the FAC performed the best (figure\ \ref{cartoons}C(top)), and the PAC (figure\ \ref{cartoons}C(bottom)) 2nd, and pMPC controllers 3rd.
-In all cases, the controller's knowledge and explicit treatment of intrinsic noise was critical to improve the control performance, break symmetry, and enable the SIMO control of all cells.
 
@@
 These uncertainties include coarse-grained approximations of the system dynamics, errors or extrinsic variations in the system parameter, and time delays between the observation of the cellular dynamics and actuation of the control process.
-In the following `Methods' we introduce our formulation of the chemical master equation to analyze the discrete stochastic distribution of cellular responses; we define multiple controllers and demonstrate the computation of their effects on cell dynamics; and we show how the control law can be optimized to improve performance.
+In the following `Methods' we introduce our formulation of the chemical master equation (CME) to analyze the discrete stochastic distribution of cellular responses; we define multiple controllers and demonstrate the computation of their effects on cell dynamics; and we show how the control law can be optimized to improve performance.
 In the `Results' section, we explore the how model approximations, parameter inaccuracies, and time delays affect control performance, and we demonstrate a simple scheme for controlling cells to track a dynamically changing reference signal.
 Using discrete stochastic models based on the chemical master equation, we demonstrate that combining biochemical noise, nonlinear auto-regulation, and a single optogenetic feedback could control two genetically identical cells with different initial conditions to follow different desired trajectories at different frequencies and phases.
-
-%Figure\ \ref{cartoons}C(left) presents heat maps for two control laws, which a single external UV excitation signal based on observed cellular state.
-Figure\ \ref{cartoons}C(right) shows the corresponding achievable marginal stationary distributions.
-Effective control was possible with both Fully Aware Feedback Controllers (FAC, Figure\ \ref{cartoons}C(top)) and Partially Aware Feedback Controllers (PAC, Figure\ \ref{cartoons}C(bottom)), even when the control model was an approximate simplification of the actual process dynamics.
 
@@
 The second, much simpler, single-species model was developed by assuming quasi-steady equilibrium for the first five species. 
-Both models were independently parameterized using the same experimental data from Baumschlager et al. \cite{XXX}. 
+Both models were independently parameterized using the same experimental data from Baumschlager et al. \brian[\cite{XXX}]. 
 Furthermore, an extension was made to each model to incorporate a secondary self-activated promoter-gene construct, where the expression rate was determined by a Hill function \ref{prodRate}.
@@
 In the current work, we adopt the simpler of the two models and extend our analyses to consider the effects of additional model inaccuracies, including coarse-grained model approximations, parameter errors, and time delays, and we also explore the possibility that a single optogenetic control signal could drive the system to track temporally-changing reference signals.
-Section \ref{sec:Model} introduces the model; Section \ref{sec:Stochastic} develops a Master Equation description of the models probabilistic dynamics; Section \ref{sec:Quantification} defines a control objective and optimizes that metric to obtain a baseline control law; and finally Sections \ref{sec:Scaling} and \ref{sec:Delay} introduces uncertainties into the model related t the granularity of the model approximation or the introduction of time delays, respectively.
+Section \ref{sec:Model} introduces the model; Section \ref{sec:Stochastic} develops a Master Equation description of the models probabilistic dynamics; Section \ref{sec:Quantification} defines a control objective and optimizes that metric to obtain a baseline control law; and Sections \ref{sec:Scaling} and \ref{sec:Delay} introduce uncertainties into the model related to the granularity of the model approximation or the introduction of time delays, respectively.
 
@@
 \end{equation}
-%
-%These model dynamics were then simplified to a single ODE with auto-regulation describing the protein accumulation as: 
-%\begin{equation}
-%\frac{dp}{dt}=u(x_1,x_2)  + \kappa\frac{p^\eta}{p^\eta+\beta^\eta}+ k_0 -\gamma p,
-%\label{rateeq}
-%\end{equation}
-%
 where $x$ is the instantaneous protein level; $\kappa$ is the maximum strength of the auto-regulation promoter; $\beta$ is the concentration at which auto-regulation promoter reaches its half maximal strength; $\eta$ is the cooperativity in the auto-regulation promoter; $k_0$ is the leakage rate from both promoters; and $u(UV(t))$ is the UV-dependent strength of the T7 promoter.
@@
 \begin{center}
-\begin{tabular}{|c|c|c|}
+\begin{tabular}{|c|c|c|l|}
 \hline
-Parameter & Value   & units                     \\ \hline \hline
-$\kappa$         & 0.406   & 1/min \\ \hline
-$\beta$          & 20.0    & Molecules            \\  \hline
-$\eta $           & 8.00    & unitless          \\  \hline
-$k_0$           & 0.0001  & 1/min \\  \hline
-$\gamma$         & 0.0203  & 1/min         \\  \hline
-$u$              & dynamic & 1/min \\ \hline
+Parameter & Value   & Units & Meaning                     \\ \hline \hline
+$\kappa$         & 0.406   & 1/min & maximal production rate \\ \hline
+$\beta$          & 20.0    & molecules  & concentration at half-max activation        \\  \hline
+$\eta $           & 8.00    & unitless  & cooperativity factor        \\  \hline
+$k_0$           & 0.0001  & 1/min & promote leakage rate \\  \hline
+$\gamma$         & 0.0203  & 1/min & degradation/dilution rate         \\  \hline
+$u(t)$              & variable & 1/min & applied control signal  \\ \hline
 \end{tabular}
@@
 \subsection{Stochastic analyses of the model}\label{sec:Stochastic}
-To describe the discrete stochastic behavior of the above model for a population system with $N_{\rm c}$ cells, we define the current state of the system as the tuple of the non-negative numbers of proteins in each cell: $\mathbf{X}_i = [x_{1},x_{2},\ldots,x_{N_c}]_i\in \mathbb{Z}_{\ge 0}$, where the index $i$ denotes the enumeration of the state within the countably infinite set of all possible states, $\mathbf{X}_i\in\mathcal{X}=\{\mathbf{X}_1,\mathbf{X}_2,\ldots\}$.
+To describe the discrete stochastic behavior of the above model for a population of $N_{\rm c}$ cells, we define the current state of the system as the tuple of the non-negative numbers of proteins in each cell: $\mathbf{X}_i = [x_{1},x_{2},\ldots,x_{N_c}]_i\in \mathbb{Z}_{\ge 0}$, where the index $i$ denotes the enumeration of the state within the countably infinite set of all possible states, i.e., $\mathbf{X}_i\in\mathcal{X}=\{\mathbf{X}_1,\mathbf{X}_2,\ldots\}$.
 The stoichiometry vector, $\mathbf{s}_{\mu}$, for reaction number $\mu$ is then defined as the change in state following that reaction event (e.g., $\mathbf{X}_i \rightarrow \mathbf{X}_i + \mathbf{s}_\mu$).
@@
 These definitions of the stoichiometry and propensity functions allow us to implement the Gillespie stochastic simulation algorithm (SSA) \cite{Gillespie1992, Gillespie1977} to generate representative trajectories of the stochastic process.
-At each step two random numbers are generated to determine the time and the type of the next reaction.
+At each step, two random numbers are generated to determine the time and the type of the next reaction.
 Given the current state $\mathbf{X}$, the time until the next reaction is distributed according to an exponential random with rate parameter equal to the inverse of the sum of the propensity functions:
@@
 where $r_1$ is a uniform random variable between zero and one.
-The probability of for the specific individual reaction $R_k$ to fire from all possible reactions is given by:
+The probability for the specific individual reaction $R_k$ to fire from all possible reactions is given by:
 \begin{equation}
@@
 \end{equation}
-and thus the reaction that fires at time $t+\tau$ is a categorical random variable pulled from $Pr(R_k)$.
+and which specific reaction that fires at time $t+\tau$ is a categorical random variable pulled from $Pr(R_k)$.
 The SSA is then simulated by stepping through time one reaction at a time and updating the state of the system by adding the stoichiometry of the reaction to the state of the system.
 
-However, a more direct analysis of the chemical master equation (CME) is necessary for performance quantification and optimal controller design.
+However, a more direct analysis of the chemical master equation (CME) is necessary to quantify performance and optimize the controller design.
 The high dimensional CME is a linear ODE that describes the time-dependent changes in probability mass of all possible states.
@@
 \end{array}\right.
-\label{InfB}x
+\label{InfB}
 \end{equation}
@@
 To address this issue, we first truncate the system at a finite number for each species and then apply a reflecting boundary condition, resulting in a finite dimensional master equation.
-To assess the time interval over which this truncation is valid, we also solve the Finite State Projection\cite{Munsky:2006} for the same truncation, which allows us to compute an upper bound on the truncation error as a function of time.
+To assess the time interval over which this truncation is valid, we also solve the Finite State Projection \brian[\cite{Munsky:2006}] for the same truncation, which allows us to compute an upper bound on the truncation error as a function of time.
 
@@
 \michael[ok!]
-%Finite state projections are a powerful way to simulate master equation analysis which enables the assessment of simulation accuracy, but these simulations have challenges measuring the steady state of the system since all probability will always enter error at long enough time scales.
-To enable control of steady state systems, a truncated FSP which reflects probability back into the system enables this calculation.
 
 \subsection{Quantification and optimization of control performance}\label{sec:Quantification}
-In the SIMO control of stochastic processes, a single input is applied simultaneously to all systems at once and a control signal that acts beneficially on one system destabilizes other systems.
-Stochastic SIMO control inputs which seek to control cells one-cell-at-a-time create increased instability and response variability in the other signals not being controlled.
-Thus, even in systems which do a good job at controlling a system cannot independently control individual cells to a desired target point without impacting the variance of others, so an effective controller must strike a balance among the desired behaviors of all cells in the system.
-
-To quantify performance success, we define the {\em steady state performance error}, $J$, as the expected steady state Euclidean distance of the process from the specified target state, $\mathbf{T}$:
+In the SIMO control of stochastic processes, a single input is applied simultaneously to all systems at once, and therefore a control signal that acts beneficially on one cell may destabilize other cells.
+An effective controller must strike a balance among the desired behaviors of all cells in the system. 
+To quantify overall performance success, we define the {\em steady state performance error}, $J$, as the expected steady state Euclidean distance of the process from the specified target state, $\mathbf{T}$:
  \begin{equation}
@@
  \end{equation}
-The squared score is easily calculated by applying a linear operator on steady state probability distribution $\mathbf{P}^*$ (Eq.\ \ref{SSDist}) as follows:
+The squared score is easily calculated by applying a linear operator to the steady state probability distribution $\mathbf{P}^*$ (Eq.\ \ref{SSDist}) as follows:
 \begin{eqnarray}
@@
 \end{eqnarray}
-where $\mathbf{C}$ is simply a vector that contains the squared Euclidian distance of each state from the specified target T, i.e., $C_i = |\mathbf{X}_i-\mathbf{T}|_2^2$. 
-As a result of this calculation, $J$ is a non-negative scalar that is zero if any only if $\mathbf{P}^*$ is a delta distribution located exactly at the target vector $\mathbf{T}$.
+where $\mathbf{C}$ is simply a vector that contains the squared Euclidian distance of each state from the specified target $\mathbf{T}$, i.e., $C_i = |\mathbf{X}_i-\mathbf{T}|_2^2$. 
+As a result of this calculation, $J$ is a non-negative scalar that is zero only if $\mathbf{P}^*$ is a delta distribution located exactly at the target vector $\mathbf{T}$.
 
@@
 \end{equation}
-where $x_1$ and $x_2$ are discrete integers greater than or equal to zero that represent the number of proteins in cell one and cell two.
+where $x_1$ and $x_2$ are discrete integers greater than or equal to zero that represent the instantaneous number of proteins in cell one and cell two.
 Despite their differences in their observation data, both the FAC and the simpler PAC are optimized to achieve the same goal, namely to drive both cells to their respective set points.
@@
 Since the square root is a monotonically increasing function, minimizing $J^2$ results in the same control as would minimizing $J$ directly.
-Therefore, we calculate the negative gradient $-\frac{d(J^2)}{d\mathbf{u}^{FAC}}$ and walking in a small step $du$ in that direction.
-The calculation of the gradient is given by
+Therefore, we calculate the negative gradient $-\frac{d(J^2)}{d\mathbf{u}^{FAC}}$ and adjust parameters a small step $d\mathbf{u}$ in that direction.
+The calculation of the gradient is given by:
 \begin{equation}
-\frac{d(J^2)}{d\mathbf{u}^{FAC}}=C \frac{dP}{d\mathbf{u}^{FAC}}
+\frac{d(J^2)}{d\mathbf{u}^{FAC}}=C \frac{dP}{d\mathbf{u}^{FAC}},
 \end{equation}
-Where $\frac{dP}{d\mathbf{u}^{FAC}}$ can be solved using general minimized residual calculation of
+where $\frac{dP}{d\mathbf{u}^{FAC}}$ can be solved using general minimized residual calculation of
 \begin{equation}
-A \frac{dP}{d\mathbf{u}^{FAC}}=B \mathbf{u}^{FAC}
+A \frac{dP}{d\mathbf{u}^{FAC}}=B \mathbf{u}^{FAC}.
 \end{equation}
-to find $\frac{dP}{d\mathbf{u}^{FAC}}$.
-
-%Optimization of $\mathbf{u}^{PAC}$ was done by restricting $\frac{d(J^2)}{d\mathbf{u}^{FAC}}$ in ambient space to search a low dimensional subspace with a linear lift operator D.
-Then the subspace search is given by.
-%\begin{equation}
-%\frac{d(J^2)}{d\mathbf{u}^{PAC}}=C \frac{dP}{d\mathbf{u}^{PAC}} = C D \frac{dP}{d\mathbf{u}^{FAC}}  
-%\end{equation}
-
-%Root mean squared error was calculated by taking the square root of the score.
-%\begin{equation}
-%RMSE=\sqrt{J}
-%\end{equation}
 
@@
 In realistic applications, models are never exact, but are often chosen as simplifications of known processes. 
-For example, when analyzing discrete stochastic chemical kinetics, it is common to project the CME onto lower-dimensional spaces using finite state projection\cite{XXX}, time scale separations \cite{Peles2006}, Krylov subspaces \cite{Macnemara&Sidje?}, principle orthogonal decompositions \cite{Vo2019}, or other coarse meshes \cite{Munsky2008IEEE,Tapia2012CDC}.
+For example, when analyzing discrete stochastic chemical kinetics, it is common to project the CME onto lower-dimensional spaces using finite state projection \brian[\cite{XXX}], time scale separations \brian[\cite{Peles2006}], Krylov subspaces \brian[\cite{Macnemara&Sidje?}], principle orthogonal decompositions \brian[\cite{Vo2019}], or other coarse meshes \brian[\cite{Munsky2008IEEE,Tapia2012CDC}].
 Similarly, measurements are also always inexact and in many cases may only provide information about relative changes -- for example, although fluorescent proteins usually cannot be counted exactly, one may reasonably assume that a cell's total fluorescence intensity varies linearly with the fluorescent protein concentration.
-To explore how mismatches in assumed system scale (e.g., arising from model approximations or relative measurements) affect the controllability of the cellular process, we define a granularity parameter ($\alpha=M'/M$) that linearly scales each species' population to increase ($\alpha>1$) or decrease ($\alpha<1$), while maintaining the dynamics and general behavior of the model.
+To explore how mismatches in the assumed system scale (e.g., arising from model approximations or relative measurements) affect the controllability of the cellular process, we define a granularity parameter ($\alpha=M'/M$) that linearly scales each species' population to increase ($\alpha>1$) or decrease ($\alpha<1$), while maintaining the dynamics and general behavior of the model.
 To apply this granularity parameter, we assume that each propensity function, $w_{\mu}$ (Eqns.\ \ref{Props}) is rescaled to a different level of discreteness by substituting
@@
 \begin{eqnarray}
-w'_1(\mathbf{X}) &= u(\mathbf{X}/\alpha,t)  + \kappa \frac{(x_1/\alpha)^\eta}{(x_1/\alpha)^{\eta}+\beta^{\eta}} + k_0 \nonumber \\
+w'_1(\mathbf{X}) &= u(\mathbf{X}/\alpha,t)  + \kappa \frac{(x_1/\alpha)^\eta}{(x_1/\alpha)^{\eta}+\beta^{\eta}} + k_0, \nonumber \\
  w'_2(\mathbf{X}) &= \gamma x_1/\alpha. \label{eq:rescaling}
@@
 Because identification of the original control formulation, $\mathbf{u}^{FAC}(x_1,x_2)$ and $\mathbf{u}^{PAC}(x_1)$ only considered integer values for $(x_1,x_2)$, control signal values at fractional state values after rescaling $(x_1/\alpha,x_2/\alpha)$ are calculated using 2D cubic interpolation from the control values at the nearest integer state values.
-%\begin{equation}
-%u'=u(x_0 / \alpha, x_1 / \alpha) 
-%\end{equation}
-%\begin{equation}
-%u'=u(\alpha x_0)
-%\end{equation}
 Finally, to provide a consistent metric for relative scoring, the definition of the performance score is also adjusted according to scale magnitudes.
@@
 \end{eqnarray}
-We reiterate that the system parameters and control law were defined and fixed using the the base granularity ($\alpha=1$), and these are not recomputed or refit upon changing the system granularity.
+We reiterate that the system parameters and control law were defined and fixed using the the base granularity ($\alpha=1$), and to simulate a practical application where scales may be unknown or variable, these are not recomputed or refit upon changing the system granularity.
 
@@
 
-Delays are inherent in any realistic control systems, and it this case delays would be expected to arise due to the time needed for various biochemical reactions such as formation of complete polymerases, activation of promoters, transcription and transport of mature mRNA, and the translation and maturation of protein \cite{Cai2007}.
+Delays are inherent to any realistic control system, and in this case delays would be expected to arise due to the time needed for various biochemical reactions such as the formation of complete polymerases, activation of promoters, transcription and transport of mature mRNA, and the translation and maturation of protein \cite{Cai2007}.
 Additional delays would also arise from data analysis, decision making, and actuation dynamics. 
-To investigate the effects of time delay on control actuation, we devised a simple time-delay stochastic simulation algorithm. 
-This algorithm records the state history after each stochastic jump, enabling reconstruction of the population history of the species and the time delayed control input propensity.  
+To investigate the effects of observation or actuation time delays on control performance, we devised a simple time-delay stochastic simulation algorithm. 
+This algorithm records the state history after each stochastic reaction, enabling reconstruction of the population history of the species and the time delayed control input propensity.  
 Using this information, the time-delayed control signal at time $t$ can be specified as:
@@
 where $\tau$ is the time delay between observation and actuation, ${u}^{\mathcal{C}}$ is the previously optimized control law (e.g., $\mathbf{u}^{\rm FAC}$ or $\mathbf{u}^{\rm PAC}$), and $x$ gathered from the state history.
-\brian[Explain how you modify the SSA for the delayed reaction SSA.  Look through the literature to cite someone else who has implenented a similar method.]
 We note that the time delay stochastic process was only simulated using the SSA because to our knowledge an appropriate direct FSP/CME integration procedure has not yet been developed.
+\brian[Still missing reference to the extrande method at this website
+\url{https://journals.plos.org/ploscompbiol/article?id=10.1371/journal.pcbi.1004923}].
 
-%\subsection{Parameter Scalings and Sensitivity}
-%Finally, to explore the sensitivity of the control performance to inaccuracies in model parameters for one or more cells, we calculate the control performance score ($J$), over a range for each parameter scaled linearly with a scaling parameter ($\zeta$) while holding all other parameters constant.
-%%Any new arbitrary parameter $\lambda\in \{ \kappa,\beta,k_0,\eta,\gamma\}$ can be rescaled from an old one ($\lambda$) according to
-%%\begin{equation}
-%%\lambda' = s \lambda
-%%\end{equation}
-%We explore two types of sensitivity analysis: one which performs the analysis on one cell at a time while keeping the other constant, and another which scales similar parameters in both cells at the same time.
-%%In the former case, it should be noted that each cell receives its own set of parameters, so that the number of parameters double during sensitivity analysis of a two cell control process. 
-%In either case, steady state distributions of the perturbed models are solved using the FSP.
-% XXX - I don't think that we need this section in Methods -- but perhaps some of the text should be moved to the results if needed.
-
-\subsection{Dynamic Setpoint Control}\label{sec:Dynamic}
+\subsection{Tracking Time-Varying Trajectories}\label{sec:Dynamic}
 Because optimizing the control law for a static set point as described in Section \ref{sec:Quantification} requires differentiation of the CME (Eq.
@@
 Extending these calculations to optimize controllers for a dynamically moving set point is much harder and would likely require intractable numerical simulation or development of new mathematical approaches that are beyond the scope of the current study.
-To circumvent this challenge, we instead propose a simple alternative to sweep though a set of controllers over time which target different static points in piecewise fashion to mimic a dynamically moving path.
+To circumvent this challenge, we instead propose a simple alternative in which the controller sweep though a piecewise constant set of controllers each designed for a specific static target point along the desired trajectory.
 
- We start by describing a discrete path  of target points in state space which cycle at fixed frequency.
-  Then, we optimize controllers for each target point and cycling through the controllers over time to create the final controller $\mathbf{R}(\mathbf{x},t)$.
-Let $\mathbf{p}(t)$ be a continuous and cyclicly moving point in state space
+Our goal is to control the system to follow a specific target trajectory, $\mathbf{T}(t)$. We choose a discrete set of $K$ target points along this trajectory:
 \begin{equation}
-\mathbf{p(t)}=[x_1(t),x_2(t)]
+\left\{\mathbf{T}_1,\mathbf{T}_2,\ldots,\mathbf{T}_K\right\},
 \end{equation}
-where $x_1(t)$, and $x_2(t)$ are both cyclic functions of time $\in \mathbf{R}$.
-Then, $\mathbf{q}$ is a collection of points $q_i$ which evaluate the cyclic function over its natural period (T) in n steps. 
+and for each individual target $\mathbf{T}_i$, we optimize to find a corresponding controller, $\mathbf{u}_i=\mathbf{u}(\mathbf{T}_i)$.
+Finally, to implement the control at any given time, $t$, we find the index of the nearest precomputed target state:
 \begin{equation}
-\mathbf{q}=[q_1,\ldots ,q_n]=[\mathbf{p}(t_1),\ldots,\mathbf{p}(t_n)]
+\hat{i}(t) = \textrm{argmin}_i \left| \mathbf{T}(t) -  \mathbf{T}_i\right|_2,
 \end{equation}
-where each $t_i$ comes from linearly spaced points between $0$ and $T$, in $n$ steps 
+and assign that controller.
+%
+%controllers for each target point and cycling through the controllers over time to create the final controller $\mathbf{R}(\mathbf{x},t)$.
+%Let $\mathbf{p}(t)$ be a continuous and cyclicly moving point in state space
+%\begin{equation}
+%\mathbf{p(t)}=[x_1(t),x_2(t)]
+%\end{equation}
+%where $x_1(t)$, and $x_2(t)$ are both cyclic functions of time $\in \mathbf{R}$.
+%Then, $\mathbf{q}$ is a collection of points $q_i$ which evaluate the cyclic function over its natural period (T) in n steps. 
+%\begin{equation}
+%\mathbf{q}=[q_1,\ldots ,q_n]=[\mathbf{p}(t_1),\ldots,\mathbf{p}(t_n)]
+%\end{equation}
+%where each $t_i$ comes from linearly spaced points between $0$ and $T$, in $n$ steps 
+%\begin{equation}
+%t_i = [t_1,\ldots,t_n]=[0,\frac{1}{n} T,\ldots,\frac{n}{n}T]
+%\end{equation}
+%Then, we optimize a set of controllers which target each $q_i$ in the discretized path.
+%\begin{equation}
+%\mathbf{U}_i(\mathbf{x})=[\mathbf{u}_1(\mathbf{x}),...,\mathbf{u}_n(\mathbf{x})]
+%\end{equation}
+%Next, we cycle through each controller at a fixed frequency by using ($\phi(t)$). 
+%\begin{equation}
+%\phi(t)=\lceil \frac{n}{T} mod(t,T)\rceil = \lceil n \ f \ mod(t,\frac{1}{f})\rceil
+%\end{equation}
+%where $f$ the cycling frequency given by $1/T$. 
+%Then the final control equation is given by 
+% \begin{equation}
+%\mathbf{R}(\mathbf{x},t)=\bf{U}_{\phi(t)}(x)
+%\end{equation}
+%Where $\mathbf{R}(\mathbf{x},t)$ returns the control input scalar as a function of state and time. 
+Under this piecewise constant controller law, the full time-varying CME becomes
 \begin{equation}
-t_i = [t_1,\ldots,t_n]=[0,\frac{1}{n} T,\ldots,\frac{n}{n}T]
+\frac{d}{dt}\mathbf{P}=\mathbf{A}+\mathbf{B}\mathbf{u}_{\hat{i}(t)}(\mathbf{x}).
 \end{equation}
-Then, we optimize a set of controllers which target each $q_i$ in the discretized path.
-\begin{equation}
-\mathbf{U}_i(\mathbf{x})=[\mathbf{u}_1(\mathbf{x}),...,\mathbf{u}_n(\mathbf{x})]
-\end{equation}
-Next, we cycle through each controller at a fixed frequency by using ($\phi(t)$). 
-\begin{equation}
-\phi(t)=\lceil \frac{n}{T} mod(t,T)\rceil = \lceil n \ f \ mod(t,\frac{1}{f})\rceil
-\end{equation}
-where $f$ the cycling frequency given by $1/T$. Then the final control equation is given by 
- \begin{equation}
-\mathbf{R}(\mathbf{x},t)=\bf{U}_{\phi(t)}(x)
-\end{equation}
-Where $\mathbf{R}(\mathbf{x},t)$ returns the control input scalar as a function of state and time. 
-The dynamical system equation becomes
-\begin{equation}
-\frac{d}{dt}\mathbf{P}=\mathbf{A}+\mathbf{B}\mathbf{R}(\mathbf{x},t)
-\end{equation}
-Since $\mathbf{R}(\mathbf{x},t)$ is a function of time, the system is no longer time invariant. 
 \section{Results}
 
-\brian[{\bf We still need this.} Add a paragraph or two to explain the big picture here.
+%\brian[{\bf We still need this.} Add a paragraph or two to explain the big picture here.
 Use figure 1 to remind the reader of (1) the control goals, (2) the control law, (3) the resulting distributions, (4) the objective function and (5) the optimization of the control law to minimize the error.
@@
 
-\michael[new paragraph here]
+As developed in the Methods section, we focus on a SIMO controller where {\em all cells receive the same input} at every instant of time.
+For this, the control signal depends upon the observed state, e.g., $u(t) = \mathbf{u}(\mathbf{x})$ as developed above and illustrated in Fig.
+\ref{cartoons}(I,III).
+These controllers have been optimized to break symmetry so that multiple cells can be controlled to different targets as illustrated in Fig.
+\ref{cartoons}(II,IV), and we quantify performance according to the RMSE error introduced in Eq.
+\ref{Euclid}.
+To explore how such controllers may perform in realistic settings where the models are approximate and parameter are unknown or extrinsically variable, we now fix the parameters of those controllers and explore performance robustness to different types of model uncertainties, including parameter errors or variations (Section \ref{sec:ResPars}), incorrect assumptions on system scales (Section \ref{sec:ResGran}) and time delays (Section \ref{sec:ResTime}).
+Finally, in Section \ref{sec:ResTraj}, we extend the control analysis to consider the performance of the controller for tracking variable reference signals with different frequencies and phases.
 
-In this section, we seek to show that noise-exploiting controllers are effective at controlling a noisy system to a dynamically moving target point.
-To do this, we show that control performance is robust to a range of parameter uncertainties without the needing adjustments to the controller.
-Then, we extend the control performance analysis to understand how time delay and system scale alters control performance.
-Furthermore, we find regions in target space which are easiest to control by optimizing individual controllers for each target.
-We extend these analysis to change the target point over time in a Peicewise fashion.
- \subsection{Stochastic SIMO Optogenetic Control can Remain Effective Despite Small Parameter Errors or Extrinsic Uncertainties}
+ \subsection{Stochastic SIMO Optogenetic Control can Remain Effective Despite Small Parameter Errors or Extrinsic Uncertainties}\label{sec:ResPars}
 \begin{figure*}
@@
 For each parameter, $\mathbf{u}^{\rm FAC}$ (solid lines) and $\mathbf{u}^{\rm PAC}$(dashed lines) control performance was examined across a parameter perturbation range from one-tenth to ten-fold of the original values.
-Figure \ref{Parameter}A shows the results for these parameter sweeps when a single parameter in Cell 1 is varied (all other parameters are fixed), and Figure.\ \ref{Parameter}B shows the control performance as single parameters in Cell 2 is varied.
+Figure \ref{Parameter}A shows the results for these parameter sweeps when a single parameter in Cell 1 is varied (all other parameters are fixed), and Figure.\ \ref{Parameter}B shows the control performance when a single parameter in Cell 2 is varied.
 Finally, \ref{Parameter}C shows the performance change when a single parameter is changed simultaneously in both Cell 1 and Cell 2 at the same time.
 
-Modification of parameters were found to produce a broad range of effects. 
+Modifications of parameters were found to produce a broad range of effects. 
 For example, increasing $\beta$ in Cell 1 quickly worsens performance while increasing $\beta$ in Cell 2 improves performance (compare second column in Figure.\ \ref{Parameter}A and \ref{Parameter}B).
@@
 In particular, the analysis shows that control performance could be improved by modifying the system design either to increase the auto-regulation promoter strength ($\kappa$) or its cooperativity ($\eta$) or to decrease the promoter binding constant ($\beta$) or the protein degradation rate ($\gamma$).
+We note that co-optimization of both the system and the controller law would allow for further improvements to the performance.
 
-\subsection{Controllers trained using one assumed level of granularity can remain effective at other levels of granularity.}
+\subsection{Controllers trained using one assumed level of granularity can remain effective at other levels of granularity.}\label{sec:ResGran}
 
-We next explored how the granularity of the system would affect the differential control of the process.
-Specifically, we define the level of granularity using the parameter $\alpha$ as described in Section \ref{sec:Scaling} and which affects the propensity functions and scoring according to equations \ref{eq:granscale}-\ref{EuclidV}.
-Effectively, larger $\alpha$ corresponds to systems where larger numbers of individual molecules are needed to achieve the same concentration (e.g., larger volumes), while smaller $\alpha$ corresponds to smaller population sizes (e.g., smaller volumes).
+We next explored how the granularity of the system would affect the differential control of the multi-cell system.
+Specifically, we define the level of granularity using the parameter $\alpha$ as described in Section \ref{sec:Scaling} and which affects the propensity functions and scoring according to Eqns.\ \ref{eq:granscale}-\ref{EuclidV}.
+Effectively, larger $\alpha$ corresponds to systems where larger numbers of individual molecules are needed to achieve the same concentration (e.g., larger volumes), while smaller $\alpha$ corresponds to situations where smaller population sizes can achieve that concentration (e.g., smaller volumes).
 We previously optimized the controllers $\mathbf{u}^{\rm FAC}$ and $\mathbf{u}^{\rm PAC}$ based on a the default assumption that $\alpha=1$, and we wished to know what would be the consequences if this same controller were to be applied to a system that has a different granularity.
@@
 To bypass this limit in the FSP, sixteen SSA simulations were used to sample the CME of a system with a much larger volume of $\alpha$=100.
-Each SSA was ran for $5\times10^7$ minutes and only the last $4\times10^7$ minutes were sampled to estimate the stationary distribution and calculate the performance score.
+Each SSA was run for $5\times10^7$ minutes and only the last $4\times10^7$ minutes were sampled to estimate the stationary distribution and calculate the performance score.
 The performance score estimates of this high granularity SSA using the FAC and PAC were $2.03$ and $5.24$ respectively, which are plotted as dashed lines in Figure.\ \ref{Volume}G.
@@
 However, at large granularity, we find that the opposite is true - the control performance actually improves when applied to an incorrect model.
-The reason for this surprising result is that relative amplitude of stochastic fluctuations (i.e., the standard deviation divided by the mean) in a chemical process decreases with the inverse square root of the process scale \cite{Kampen1961}.
-In other words, the process become more predictable and therefore more controllable.
-At the extreme as the system size increases, the dynamics converge towards a deterministic process, except for certain exceptional initial conditions lying on manifolds that would separate different steady state behaviors \cite{XXX}.
+The reason for this surprising result is that relative amplitude of stochastic fluctuations (i.e., the standard deviation divided by the mean) in a chemical process decreases with the inverse square root of the process scale \brian[\cite{VanKampen}].
+In other words, the process becomes more predictable and therefore more controllable.
+At the extreme as the system size increases, the dynamics converge towards a deterministic process, except for certain exceptional initial conditions lying on manifolds that would separate different steady state behaviors \brian[\cite{XXX}].
 
-However, the improvement in the steady state performance does not come without a cost.
-Although higher granularity reduces noise and makes it easier to maintain desired states once they have been achieved, noise is necessary to break symmetry between the two cells' dynamics to achieve those states in the first place.
-This tradeoff is illustrated in Figure\ref{Volume}(H), which plots the control performance over time after the changing the controller goal to exchange the low- and high-target cells with one another.
-From the figure, we can see although steady state control performance improves for higher values of $\alpha$, the time taken to reach that steady state performance increases also increases $\alpha$.
+However, this improvement in the steady state performance does not come without a cost.
+Although higher granularity reduces noise and makes it easier to maintain desired states once they have been achieved, noise is necessary to break symmetry between the two cells' dynamics in order to achieve those states in the first place.
+This tradeoff is illustrated in Fig.\ \ref{Volume}(H), which plots the control performance over time after changing the control goal to exchange the low- and high-target cells with one another.
+From the figure, we can see although steady state control performance improves for higher values of $\alpha$, the time taken to reach that steady state performance  increases with $\alpha$, suggesting that larger-volume system may become more susceptible to times delays or less able to track variable reference trajectories.
 
@@
 
-\subsection{Heterogeneous control can remain effective despite moderate time delays}
+\subsection{Heterogeneous control can remain effective despite moderate time delays}\label{sec:ResTime}
 \begin{figure*}
@@
 
-In general, feedback control can only be effective if one can quickly make measurements, compute adjustments to the control signal, and implement the needed changes within an appropriate amount of time relative to the characteristic times scales of the system.
+In general, feedback control can only be effective if one can quickly make measurements, compute adjustments to the control signal, and implement the needed changes within an appropriate amount of time relative to the characteristic timescale of the system.
 As the time required for any of these steps increases, control performance will be degraded, perhaps even leading to large fluctuations or instability.
-To explore how time delays would affect the noise-enhanced controllers $\mathbf{u}^{\rm FAC}$ and $\mathbf{u}^{\rm PAC}$, we generated large sets of time-delayed stochastic simulations (see Methods) for different lengths of the time delay.
+To explore how time delays affect the noise-enhanced controllers $\mathbf{u}^{\rm FAC}$ and $\mathbf{u}^{\rm PAC}$, we generated large sets of time-delayed stochastic simulations (see Section \ref{sec:Delay}) for different lengths of the time delay.
 Each SSA was sub-sampled for 1000 times over 10000 minutes of simulation time after a burnin period of 10000 minutes.
@@
 Figure \ref{Time}G summarizes these results by plotting the score of both controllers versus the time delay.
-From the figures, it is clear that performance is rapidly degraded as the delay approaches and then exceeds the characteristic time scale ($\tau_c = 1 / \gamma = 49$ min) set by the degradation rate of the process (yellow dashed line).
-At very small time delays ($\tau< 3.4$ min), the FAC outperforms the PAC  but at moderate time delays ($\tau > 3.4$) the PAC outperforms the FAC.
-
-\brian[Are these times correct?  Isn't gamma = 0.0203 in Table 1?  tau = 0.07 / 0.0203 (1/min) = 3.44 min. Please check this carefully.]
-\michael[I changed characteristic tau to be 1 since it fits all the curves the best]
-\brian[What are the units in Figure 4G?]
-\michael[minutes]
-
-These data show that time delays greater than $\tau = 0.1 / \gamma = 4.9$ min begin to degrade FAC control performance and plateau at $\tau_c = 1.0 / \gamma = 49$ min.
+From the figures, it is clear that performance is rapidly degraded as the delay approaches and then exceeds the characteristic time ($\tau_c = 1 / \gamma = 49$ min) set by the degradation rate of the process (yellow dashed line).
+At very small time delays ($\tau< 3.4$ min), the FAC outperforms the PAC  but at moderate time delays ($\tau > 3.4$ min) the PAC outperforms the FAC.
+%
+%These data show that time delays greater than $\tau = 0.1 / \gamma = 4.9$ min begin to degrade FAC control performance and plateau at $\tau_c = 1.0 / \gamma = 49$ min.
 \brian[Check these times.] \michael[rewrote the sentence]The choice of best controller depends on the level of time delay in the system.
-At extreme time delays, both controllers lose their asymmetry, resulting in significantly worse scores ($RMSE = 32$ and $23$).
-The control performance of simple constant control input without feedback was better than both controlers at very large time delays.
-Figure\ref{Time}(G) (red line) and achieved a score of $J = 20.0$.
-These findings indicate that, for large time delays, attempting to improve control with feedback is less effective than using a no-feedback approach.
-
+At high time delays, both controllers lose their asymmetry, resulting in significantly worse performances ($RMSE = 32$ and $23$) and even perform worse than under a simple constant control input without feedback (depicted by a horizontal red line in Fig.\ \ref{Time}G).
 % XXX - This figure needs much larger fonts for all the labels and axes.
@@
 (J) Control performance measured over a domain of $(\tau,\alpha)$ pairs show good control performance when both $\alpha$ is large and $\tau$ small.
+\brian[Label 'J' is missing from the figure.]
 (K) Trajectories where good control was expected but not observed show oscillations instead.}
@@
 Control performance errors measured by RMSE improved as $\alpha$ increased when the delays were small ($\tau=1 \ $ min and $\tau=10 \ $ min) but not when $\tau=100 \ $ min.
-
-\brian[What are the units of tau in Figure5?
+%
+%\brian[What are the units of tau in Figure5?
 This section is confusing in that it is not clear to me how tau and gamma are changing relative to one another when alpha is changed.
@@
 Therefore, the critical limit for the delays should also change with the system scale according to $\tau_c = 1\alpha/\gamma$.
-Figure\ref{DG}(J) depicts this characteristic line and shows that as the time delay approaches and then exceeds this level, the steady state performance become dramatically worse.
-\michael[edited this section]
+Figure\ref{DG}(J) depicts this characteristic line and shows that as the time delay approaches and then exceeds this level, the steady state performance becomes dramatically worse.
 
-Finally, to understand the model of failure at these longer delays, it is interesting to examine trajectories induced by the controller just above this characteristic delay.
-For parameter set denoted by the yellow start in Figure.\ \ref {DG}(J), Figure.\ \ref {DG}(K) shows the controlled response after a long burn in period to achieve steady state.
+Finally, to understand the model of failure at these longer delays, it is interesting to examine trajectories induced by the controller just below this characteristic delay.
+For parameter set denoted by the red star in Figure.\ \ref {DG}(J), Figure.\ \ref {DG}(K) shows the controlled response after a long burn in period to achieve steady state.
 In this case, the application of feedback after a delay leads to a strong oscillatory behavior and worse performance than that achieved without any control at all.
@@
 
-\subsection{With noise-induced control, a single input can drive multiple cells to follow different temporal trajectories.}
+\subsection{With noise-induced control, a single input can drive multiple cells to follow different temporal trajectories.}\label{sec:ResTraj}
 
@@
 These points give rise to basins which may attract probability more easily than other points.
-We next optimized the FAC controller and calculated its performance for the two-cell system over a two dimensional domain of target points $[T_1,T_2]$ between between five and forty-five.
-For example, Figure\ref{CtrlP} (A) shows the optimized FAC control input of the system when the target is $\mathbf{T}=[20,25]$, and Figure.\ \ref{CtrlP} (B) shows the corresponding steady state probability distribution of the controlled system.
+We next optimized the FAC controller and calculated its performance for the two-cell system over a two dimensional domain of discrete target points $\mathbf{T}_i = [T_{i1},T_{i2}]$ between between five and forty-five.
+For example, Fig.\ \ref{CtrlP} (A) shows the optimized FAC control input of the system when the target is $\mathbf{T}=[20,25]$, and Fig.\ \ref{CtrlP} (B) shows the corresponding steady state probability distribution of the controlled system.
 Figure \ref{CtrlP} (C and D) show the the same thing but for a different target state of $\mathbf{T}=[10, 25]$.
-
-\brian[Let's remove the PAC from this section.
+%\brian[Let's remove the PAC from this section.
 Show only the FAC and and replace the Figure6C/D with the FAC at a different target point.
@@
 Make sure to change all the figure labels and titles and captions.] \michael[ok, I also moved the settpoint score heatmap to the right]
-
-Figure \ref{CtrlP}(E) shows the FAC control performance over the entire domain of static set points, and illustrates that some regions are easier to obtain than others.
-For example, in the space contained in the dashed-green box which also hold the four stable points of the system, the control performance is quite strong, whereas near the unstable point (dashed-blue box), the FAC results in relatively worse control.
-
-The top left and bottom right edge regions were the most difficult to control, since the control input does not push probability in these directions.
+Figure \ref{CtrlP}(E) shows the overall steady state FAC control performance over the entire domain of static set points, and illustrates that some regions are easier to attain than others. 
+%For example, in the space contained in the dashed-green box which also hold the four stable points of the system, the control performance is quite strong, whereas near the unstable point (dashed-blue box), the FAC results in relatively worse control.
+%For example, the top left and bottom right edge regions were the most difficult to control, since the control input does not push probability in these directions.
 \brian[I don't understand this statement.] \michael[editied it better]
@@
 
-We developed a method (Section \ref{sec:Dynamic}) to control the system dynamics to follow a predefined path $p(t)$.
-Specifically, three controllers were optimized along three representative pathways in thirty-two steps each, which include an in-sync reference point (Figure.\ \ref{CR}B1), a phase lagged reference point (Figure.\ \ref{CR}C1),  and a  frequency separated reference point (Figure.\ \ref{CR}D1).
-The synchronous signal is given by Figure \ref{CR}(B1) $p_1(t)=[20 sin(2 \pi f t)+10, 20 sin(2 \pi f t)+10]$, the phase lagged signal Figure \ref{CR}(C1) is given by $p_2(t)=[20 sin(2 \pi f t)+10, 20 cos(2 \pi f t)+10]$, and the frequency changed signal Figure \ref{CR}(D1) is given by $p_3(t)=[20 sin(2 \pi f t)+10, 20 sin(6 \pi f t+3\pi)+10]$ using 32 steps for each controller.
+We developed a method (Section \ref{sec:Dynamic}) to control the system dynamics to follow a predefined path $\mathbf{T}(t)$ by alternating between 32 different pre-computed controllers along each path.
+We considered three representative pathways, including an in-sync reference point (Fig.\ \ref{CR}B1) where
+$$
+T^{\rm B}_1(t) =20 \sin(2 \pi f t)+10,\textrm{ and } T^{\rm B}_2(t) = 20 \sin(2 \pi f t)+10;
+$$
+a phase lagged reference point (Fig.\ \ref{CR}C1) where
+$$
+T^{\rm C}_1(t)=20 \sin(2 \pi f t)+10,\textrm{ and } T^{\rm C}_2(t) =20 \cos(2 \pi f t)+10;
+$$
+and a frequency separated reference point (Fig.\ \ref{CR}D1) where
+$$
+T^{\rm D}_1(t)=20 \sin(2 \pi f t)+10,\textrm{ and } T^{\rm D}_2(t) =20 \cos(2 \pi f t)+10.
+$$
+For each reference signal, the driving frequency was defined as $f=10^{-4}$ cycles per minute.
 
-All FSP simulations were slowly driven at a frequency of $1e^{-4} min^{-1}$  \brian[Check units.
-This doesn't make sense.
-did you mean period?
-If so, this is very fast.
-Or did you mean s$^{-1}$?] \michael[it was $min^{-1}$ not s] and Figure.\ \ref{CR}(B2, C2, and D2) show their response distributions (shading) and median responses (lines) for the two cells $x_1$ and $x_2$ in red and blue, respectively.
-Regions with purple shading depict the exact overlap of red and blue when both cells have the same distribution of response.
-We observe that the single-input control system is able to effectively control the system to follow all three input trajectories, including when the two reference trajectories have different phases and frequencies (Figure \ref{CR}D2).
-For all reference trajectories, the medians of the marginal distributions show good match with the desired reference signal (Figure \ref{CR}B2,C2, and D2).
-To quantify the overall performance, Figs.\ \ref{CR}(B3, C3,D3) show the root mean squared error as a function of time,
-and Figs.\ \ref{CR}(B4, C4, and D4) show the time-averaged error distribution for the system response relative to the time varying target.
-From these figures, we observe that all trajectories result in short RMSE spikes at times, and these correspond to short, transient periods when the controller passes through the regions of poor control depicted in Figure.\ \ref{CtrlP}(E).
+All FSP simulations were calculated under the time varying control law, and Figs.\ \ref{CR}(B2, C2, and D2) show the corresponding response distributions (shading) and median responses (lines) for two cells $x_1$ and $x_2$ in red and blue, respectively.
+Regions with purple shading depict the exact overlap of red and blue, when both cells have the same distribution of response. 
+From Figs.\ \ref{CR}(B2, C2, and D2), we observe that the SIMO control system can effectively drive the system to follow all three input trajectories, including when the two reference trajectories have different phases and frequencies (panel D2).
+To quantify the overall performance, Figs.\ \ref{CR}(B3, C3, and D3) show the root mean squared error as a function of time, and Figs.\ \ref{CR}(B4, C4, and D4) show the time-averaged error distribution for the system response relative to the time varying target.
+From these figures, we observe that all trajectories result in short RMSE spikes during short transient periods when the controller passes through the regions of poor control (i.e., through regions found in Fig.\ \ref{CtrlP}E to have high RMSE errors).
 Overall, the simulations show that the synchronous control performed best with an average RSME of $6.9$.
@@
 
-%Slowly responding systems driven at high frequency often exhibit poor control performance which improves as the driving frequency matches the time scale of the system.
-The combination of both frequency driving and noise could be catastrophic for system control performance.
 In general, a given control system cannot effectively track a reference signal that changes faster that the system's natural time scale.
-Since it was shown that increasing granularity ($\alpha$) improves steady state control performance (Figure.\ \ref{Time}G) but also lengthens this time scale (Figure.\ \ref{Time}H), one should expect that these competing effects of granularity would also affect the types and speeds of signals to which the system can respond.
+Since it was shown that increasing granularity ($\alpha$) improves steady state control performance (Fig.\ \ref{Time}G) but also lengthens this time scale (Fig.\ \ref{Time}H), one should expect that these competing effects of granularity would also affect the types and speeds of signals to which the system can respond.
 To examine how driving frequency and system scale affects control performance, 64 SSA trajectories of the phase-separated dynamic controller were simulated over a two dimensional domain of points $(\alpha,f)$ for 32 cycles after reaching steady state.
-Figure \ref{CRG}(A) shows the desired reference signal for the system, and Figure \ref{CRG}(B) shows the mean of the system response when the frequency is $f=10^{-4} min^{-1}$ \brian[Need units] \michael[added them]and the granularity is $\alpha = 4.0$.
-Figure \ref{CRG}(C) shows the corresponding control performance over normalized periodic time as $\alpha$ is held at $4.0$ and $f$ is increased from $f=10^{-4}min^{-1}$ \brian[Units]  \michael[added them]to $f=10^{-2}min^{-1}$ \brian[Units]  \michael[added them]to $f=10^{-0}min^{-1}$ \brian[Units] \michael[added them] leading to average control performances of 6.2, 16.1, and 15.0 respectively.
+Figure \ref{CRG}(A) shows the desired reference signal for the system, and Fig.\ \ref{CRG}(B) shows the mean of the system response when the frequency is $f=10^{-4}$ cycles/min and the granularity is $\alpha = 4.0$.
+Figure \ref{CRG}(C) shows the corresponding control performance over normalized periodic time as $\alpha$ is held at $4.0$ and $f$ is increased from $f=10^{-4}$ cycles/min to $f=10^{-2}$ cycles/min and $f=10^{-0}$ cycles/min leading to average control performances of 6.2, 16.1, and 15.0 respectively.
 Figure \ref{CRG}D,E extends this analysis to plot the average control performance over different combinations of $\alpha$ and $f$. 
@@
 %RMSE control performance of 128 simulations were simulated over a range of frequency between $10^-3$ and $10^-1$ Figure\ref{CRG}(D) for 32 cycles to obtain a better measurement of the true RMSE Figure\ref{CRG}(D) at $\alpha=2.0$, $\alpha=4.0$, and $\alpha=8.0$.
-In particular, the heat map of control performance over $(f,\alpha)$ shows that the worst performance occurs at moderate $f$ near $f=10^{-2}min^{-1}$ and that control performance is best when granularity is high and frequency is low \ref{CRG}(E).
-Plots of the characteristic frequency given by the inverse of characteristic time \ref{CRG}(yellow line), show good match with the heat maps at high granularity but not at low granularity.
-These data
- When examining poorly performing trajectories at moderate frequencies, a phase lag was observed in the system.
-In this case, although the qualitative form of the response may match to the reference signal, this phase lag resulted in worse RMSE than no system response at all.
-At very large frequencies, the effect of the fast fluctuating control input is averaged, and the system barely responds.
-
-\brian[ try adding the line at $f = 1/tau_c =  \gamma/(0.7*\alpha)$ to this plot.  Does it match the contours separating good and poor control?]
-\michael[added $\tau_c^{-1}$. using $\tau_c=1/ \gamma$]
-
-\brian[Go through every figure and make sure that the units are shown and that they are correct. If quantities are unitless, make sure that the form of the normalization is clearly expressed.]
+In particular, the heat map of control performance over $(f,\alpha)$ shows that the worst performance occurs at moderate $f$ near $f=10^{-2}$ cycles/min and that control performance is best when granularity is high and frequency is low \ref{CRG}(E).
+More specifically, we find that strong performance was attainable only if the driving frequency was kept lower than a characteristic frequency $f_c \equiv 1 / \tau_c = \gamma/\alpha$, which is denoted by the dashed line in Fig.
+\ref{CRG}E.
 
@@
 
-Key results from \cite{May2021} showed that noise was a key requirement for the development of a noise-exploiting controller, and that deterministic systems could not be controlled to two different stable points if they both started at the same initial condition.
-As a stochastic system becomes increasingly more granular, it also becomes less noisy and more similar to its ODE solution which is known to be impossible to control to two different fates, but here, we have shown that the removal of noise through system granularity led to better control performance.
-Therefore, stochastic controllers may perform work best on models which are nearly deterministic, but not quite.
-Unfortunately, such systems were found to take longer time to achieve steady state.
+Noise, whether it arises from inherently stochastic processes or from unknown or unmodeled interactions, can play a critical role in the performance of feedback control.
+For the field of synthetic biology, this noise has typically been avoided and genetic systems have primarily been engineered to be as robust as possible to these uncertain fluctuations.
+In contrast, many natural cellular processes exist and thrive in settings where single-molecule events such as gene activation leads to large relative fluctuations, and where response heterogeneity is unavoidable.
+Key results from \cite{May2021} showed that certain controllers could exploit this noise to achieve objectives that would not be possible in a deterministic setting.
+With such controllers, a single regulatory signal, such as an optogenetic input, could drive two or more two genetically identical cells to different, arbiltrarilly chosen fates using just a single input signal and irrespective of the cells' initial conditions.
 
-The ability to analyze the model at one system granularity apply it elsewhere might help analyze models for which the computational effort may be too large.
+The effectiveness of any model-based controller depends upon the accuracy of the model with which that controller has been optimized, and as the real system deviates from its idealized model, the control performance will naturally be affected.
+In this work, we explore the effects of several such deviations, including uncertainties or errors in parameters, mismatches in assumptions for systems scale, and times delays.
+
+Regarding parameter values, our perturbation analyses (Fig.\ \ref{Parameter}) showed that control performance is strongly affected by the system parameters. 
+We found that whether parameters were incorrect in all cells (e.g., due to systematic errors in the model) or for just one cell at a time (e.g., due to extrinsic noise in the cells themselves) affected the control performance in different way.
+In most cases, small changes could be tolerated, whereas large changes, especially to certain key parameters, could be catastrophic. 
+Moreover, we found that there can be room to improve control performance by adjusting system parameters, suggesting that joint optimization of the controller with the system itself could lead to even stronger performance.
+Armed with such insight into which parameters are the most sensitive and which can safely be ignored, one could in principle focus measurement efforts to more precisely quantify the critical parameters and focus design efforts to reduce variability in key aspects of modular parts.
+
+The size of the system also plays an important role in its control performance. 
+For a fixed concentration, as the volume of a chemical reaction system increases, it becomes less noisy, and its dynamics approach that of deterministic process. 
+At this limit, symmetry can no longer be broken, and feedback control cannot independently drive different cells to different fates.  
+On the other hand, deleterious fluctuations also become smaller, so larger systems can more easily maintained maintain their desired phenotypes.
+Overall, we have shown (Fig.\ \ref{Volume}A-G) that the removal of noise through system granularity led to better steady state control performance, but  such systems were found to take a much longer time to achieve steady state (Fig.\ \ref{Volume}H).
+Interestingly, this result could have implications on the malleability of cells at different stages of their growth cycle, where differentiation of smaller cells (e.g., those immediately after division) may be more susceptible to control signals, while larger cells (e.g., mature cells that have already established their phenotypes) may be relatively impervious to external signals.
+
+Although our objective was to determine how well control performance would be maintained under different system sizes, we were surprised to find that controllers designed at one level of granularity (e.g., $\alpha$ =1) worked surprisingly well to control systems at much larger granularities.
+From a practical perspective, the ability to analyze a model at one system scale and then effectively apply it to another could be highly beneficial. 
 Since the computation time of the FSP solution to the CME grows with the square of the number of states, this can cause an explosion in computational requirements for large systems. 
-One alternative is to learn a controller at a computationally feasible number of states and apply them to large systems which cannot be solves for using the FSP.
+Our results suggest a promising alternative in which one could learn or optimize a controller using FSP analyses for a computationally feasible number of states and later apply them to larger systems that cannot be solved using current techniques.
 
-Parameter perturbation analysis showed that there is room to improve control performance by adjusting system parameters, and that joint optimization of the controller with the system parameters may lead to better control performance.
-It also shows that populations of cells where each cell might have its own parameter combination could still be reasonably controlled for small changes in parameter scale, but large changes could be detrimental.
-
-Time delay analysis showed that increasing time delay decreased control performance.
-It was also found that a controller with less control information outperformed a controller with more information at higher levels of time delay.
+Time delay analysis (Fig.
+\ref{Time}) showed that increasing time delay decreased control performance.
+However, we also found that not every controller was equally affected by time delays.
+In particular, we found that at intermediate and larger time delays, a partially aware controller that has  less information can outperform a fully aware controller (Fig.
+\ref{Time}G).
 We believe this is happening because a controller with more information can afford to be more aggressive to implement its control, and time delay can cause this aggression to backfire.
-A controller was optimized which required no information in May et al which had a score of 402.
-This controller should be immune to time delay since no feedback is required for this controller.
-Here we saw both controllers did performed worse than 402 at very large time delay.
 

### 2023-07-31 paper.tex
Update to captions.

@@
 \end{equation}
-where $x_1$ and $x_2$ are discrete integers greater than or equal to zero that represent the number of proteins in cell one and cell two.
+where $x_1$ and $x_2$ are discrete integers greater than or equal to zero that represent the instantaneous number of proteins in cell one and cell two.
 Despite their differences in their observation data, both the FAC and the simpler PAC are optimized to achieve the same goal, namely to drive both cells to their respective set points.
 
-Optimization of $\mathbf{u}^{FAC}$ was performed using a gradient descent method to minimize $J$.
+Optimization of $\mathbf{u}^{FAC}$ and $\mathbf{u}^{PAC}$ were performed using a gradient descent method to minimize $J$.
 Since the square root is a monotonically increasing function, minimizing $J^2$ results in the same control as would minimizing $J$ directly.
-Therefore, we calculate the negative gradient $-\frac{d(J^2)}{d\mathbf{u}^{FAC}}$ and walking in a small step $du$ in that direction.
-The calculation of the gradient is given by
+Therefore, we calculate the negative gradient $-\frac{d(J^2)}{d\mathbf{u}^{(.)}}$ and adjust parameters a small step $d\mathbf{u}^{(.)}$ in that direction.
+For example, the calculation of the gradient for the FAC controller is given by:
 \begin{equation}
-\frac{d(J^2)}{d\mathbf{u}^{FAC}}=C \frac{dP}{d\mathbf{u}^{FAC}}
+\frac{d(J^2)}{d\mathbf{u}^{FAC}}=C \frac{dP}{d\mathbf{u}^{FAC}},
 \end{equation}
-Where $\frac{dP}{d\mathbf{u}^{FAC}}$ can be solved using general minimized residual calculation of
+where $\frac{dP}{d\mathbf{u}^{FAC}}$ can be solved using general minimized residual calculation of
 \begin{equation}
-A \frac{dP}{d\mathbf{u}^{FAC}}=B \mathbf{u}^{FAC}
+A \frac{dP}{d\mathbf{u}^{FAC}}=B \mathbf{u}^{FAC}.
 \end{equation}
-to find $\frac{dP}{d\mathbf{u}^{FAC}}$.
-
-%Optimization of $\mathbf{u}^{PAC}$ was done by restricting $\frac{d(J^2)}{d\mathbf{u}^{FAC}}$ in ambient space to search a low dimensional subspace with a linear lift operator D.
-Then the subspace search is given by.
-%\begin{equation}
-%\frac{d(J^2)}{d\mathbf{u}^{PAC}}=C \frac{dP}{d\mathbf{u}^{PAC}} = C D \frac{dP}{d\mathbf{u}^{FAC}}  
-%\end{equation}
-
-%Root mean squared error was calculated by taking the square root of the score.
-%\begin{equation}
-%RMSE=\sqrt{J}
-%\end{equation}
 
@@
 \vspace{-0.1in}
-\caption{Stochastic simulations of the time delayed system show decreasing control performance using either the FAC or PAC.
-(A-C) Joint (left) and marginal distributions (right) of the at different levels of time delay for the FAC.
-(D-F) Joint (left) and marginal distributions(right) at three levels of time delay using the PAC.
-(G) plots of RMSE control performance versus time delay for both FAC and PAC.}
+\caption{{\bf Effects of time delay on control performance.} (A-C) Joint (left, color scale shown at top) and marginal distributions (right) of the controlled system at different levels of time delay using the FAC controller.
+The target state, $\mathbf{T}$ is denoted by a small circles on the left panels and dashed lines on the right panels.
+The steady state RMSE is shown in each case.
+(D-F) Same as (A-C) but for the PAC controller.
+(G) RMSE control performance versus time delay for both FAC (blue) and PAC (green).
+Letters A-F correspond to panels A-F.
+Dashed red line corresponds to optimal performance with no feedback (i.e., constant input).
+Dashed yellow line corresponds to characteristic system time, $\tau_c = 1/\gamma$.}
 \label{Time}
@@
 \end{figure*}
-% XXX - This figure needs larger fonts for all the labels and axes. 
-% MMM- Did the change
-% Update panel letters as written in caption
-% Add tic marks to left side of the A and C heat maps
-% Fix the ref to baumschlager paper.
 
-In general, feedback control can only be effective if one can quickly make measurements, compute adjustments to the control signal, and implement the needed changes within an appropriate amount of time relative to the characteristic times scales of the system.
+In general, feedback control can only be effective if one can quickly make measurements, compute adjustments to the control signal, and implement the needed changes within an appropriate amount of time relative to the characteristic timescale of the system.
 As the time required for any of these steps increases, control performance will be degraded, perhaps even leading to large fluctuations or instability.
-To explore how time delays would affect the noise-enhanced controllers $\mathbf{u}^{\rm FAC}$ and $\mathbf{u}^{\rm PAC}$, we generated large sets of time-delayed stochastic simulations (see Methods) for different lengths of the time delay.
+To explore how time delays affect the noise-enhanced controllers $\mathbf{u}^{\rm FAC}$ and $\mathbf{u}^{\rm PAC}$, we generated large sets of time-delayed stochastic simulations (see Section \ref{sec:Delay}) for different lengths of the time delay.
 Each SSA was sub-sampled for 1000 times over 10000 minutes of simulation time after a burnin period of 10000 minutes.
@@
 \vspace{-0.1in}
-\caption{Stochastic simulations of the FAC driven system and PAC driven system show different failure modes at at different $(\tau,\alpha)$ pairs.
-(A-I)Joint probability distributions at different $[\tau,\alpha]$ pairs.
-$\alpha$ across columns and $\tau$ across rows.
-(J) Control performance measured over a domain of $(\tau,\alpha)$ pairs show good control performance when both $\alpha$ is large and $\tau$ small.
-(K) Trajectories where good control was expected but not observed show oscillations instead.}
+\caption{{\bf Joint effects of granularity and time delay on control performance.} (A-I) Joint probability distributions at different combinations of $\tau$ (rows) and $\alpha$ (columns).
+Overall RMSE shown at top.
+(J) Heat-map of control performance versus $(\tau,\alpha)$.
+Points A-I correspond to panels A-I.
+Dashed yellow line shows characteristic time $\tau_c = \alpha/\gamma/$.
+\brian[Label 'J' is missing from the figure. Replace XXX below.]
+(K) Controlled system trajectory for $\alpha = XXX$ and $\tau = $ min (denoted by red star in panel J).}
 \label{DG}
@@
 \vspace{-0.1in}
-\caption{The FAC and PAC control methods were used to optimize control inputs and probability distributions for static set points in a 2D domain.
-Optimized control inputs and probability distributions were optimized using the FAC control method(A and B) or the PAC control method (C and D).
-(E) Domains of control performance for the FAC method over a region of static target points show regions of good control and symmetry along the dashed line.}
+\caption{{\bf FAC control laws and performance for different target points.} 
+(A) Optimized control input for target $\mathbf{T} = [20,25]$, with target point denoted by star. 
+(B) Corresponding steady state response distribution. Marginal distributions for $x_1$ and $x_2$  on right and below.
+(C,D) Same as (A,B) but for $\mathbf{T} = [10,25]$.
+ (E) FAC control performance (RMSE) versus targets $\mathbf{T} = [T_1,T_2]$.  Stars correspond to target points in panels A-D. Dashed diagonal shows line of symmetry.} 
 \label{CtrlP}
@@
 \vspace{-0.1in}
-\caption{Finite state projections of the slowly driven system demonstrate good control performance under a variety of phase and frequency scenario.
-(A) Schematic of the control problem shows targeting the number of protein in two cells using a dynamically changing reference point by cycling through a set of individual stable points and under synchronous, phase-lagged or frequency changes reference signal using only one control input.
-(B1-BN) shows the phase space of the reference time series, FSP solution, RMSE, reference phase space, and error residuals respectively for the synchronous control method.
-(C1-C5) Shows the same data for the phase-lagged control input.
-(D1-D5) Shows the same analysis for the frequency-changed control.}
+\caption{{\bf Tracking time-varying reference signal.}
+(A) Schematic of SIMO control to drive two cells to follow different trajectories.
+(B1) Reference signal for $x_1$ (red) and $x_2$ (blue).
+(B2) Controlled response. Distributions shown in shading. Median shown in lines. Three periods are shown after decay of transient dynamics.
+(B3) RMSE performance over time.
+(B4) Phase space of reference signal.
+(B5) Time-averaged distribution of tracking error.
+(C1-C5) Same as (B1-B5) but for phase-lagged reference signal.
+(D1-D5) Same as (B1-B5) but for reference signal with two different frequencies and phases.}
 \label{CR}

### 2023-08-21 IOPLaTeXGuidelines.tex
update

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%    INSTITUTE OF PHYSICS PUBLISHING                                   %
%                                                                      %
%   `Preparing an article for publication in an Institute of Physics   %
%    Publishing journal using LaTeX'                                   %
%                                                                      %
%    LaTeX source code `ioplau2e.tex' used to generate `author         %
%    guidelines', the documentation explaining and demonstrating use   %
%    of the Institute of Physics Publishing LaTeX preprint files       %
%    `iopart.cls, iopart12.clo and iopart10.clo'.                      %
%                                                                      %
%    `ioplau2e.tex' itself uses LaTeX with `iopart.cls'                %
%                                                                      %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%x
%
%
% First we have a character check
%
% ! exclamation mark    " double quote  
% # hash                ` opening quote (grave)
% & ampersand           ' closing quote (acute)
% $ dollar              % percent       
% ( open parenthesis    ) close paren.  
% - hyphen              = equals sign
% | vertical bar        ~ tilde         
% @ at sign             _ underscore
% { open curly brace    } close curly   
% [ open square         ] close square bracket
% + plus sign           ; semi-colon    
% * asterisk            : colon
% < open angle bracket  > close angle   
% , comma               . full stop
% ? question mark       / forward slash 
% \ backslash           ^ circumflex
%
% ABCDEFGHIJKLMNOPQRSTUVWXYZ 
% abcdefghijklmnopqrstuvwxyz 
% 1234567890
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\documentclass[12pt]{iopart}
\newcommand{\gguide}{{\it Preparing graphics for IOP Publishing journals}}
%Uncomment next line if AMS fonts required
%\usepackage{iopams}  
\begin{document}

\title[Author guidelines for IOP Publishing journals in  \LaTeXe]{How to prepare and submit an article for 
publication in an IOP Publishing journal using \LaTeXe}

\author{Content \& Services Team}

\address{IOP Publishing, Temple Circus, Temple Way, Bristol BS1 6HG, UK}
\ead{submissions@iop.org}
\vspace{10pt}
\begin{indented}
\item[]August 2017
\end{indented}

\begin{abstract}
This document describes the  preparation of an article using \LaTeXe\ and 
\verb"iopart.cls" (the IOP Publishing \LaTeXe\ preprint class file).
This class file is designed to help 
authors produce preprints in a form suitable for submission to any of the
journals listed in table~\ref{jlab1} on the next page.  You are not obliged to use this class file---we accept
submissions using all common \LaTeX\ class and style files.  The \verb"iopart.cls"
class file is supplied merely as a convenience for those authors who find it useful.
This document gives both general advice that applies whatever class file you use, and specific advice
that applies if you choose to use \verb"iopart.cls".

We also accept submissions in Word format.  See elsewhere on this site for guidelines on Word submissions.

If you have any queries about this document or any aspect of preparing your article for submission please contact us at the e-mail address given above.
\end{abstract}

%
% Uncomment for keywords
%\vspace{2pc}
%\noindent{\it Keywords}: XXXXXX, YYYYYYYY, ZZZZZZZZZ
%
% Uncomment for Submitted to journal title message
%\submitto{\JPA}
%
% Uncomment if a separate title page is required
%\maketitle
% 
% For two-column output uncomment the next line and choose [10pt] rather than [12pt] in the \documentclass declaration
%\ioptwocol
%

\section{Introduction: file preparation and submission}

The \verb"iopart" \LaTeXe\ article class file is provided to help authors prepare articles for submission to IOP Publishing journals.
  This document gives advice on preparing your submission, and specific instructions on how to use \verb"iopart.cls" to follow this advice.  You
  do not have to use \verb"iopart.cls"; articles prepared using any other common class and style files can also be submitted.
    It is not necessary to mimic the appearance of a published article.

The advice
on \LaTeX\ file preparation in this document applies to
the journals listed in table~\ref{jlab1}.  If your journal is not listed please go to the journal website via \verb"http://iopscience.iop.org/journals" for specific
submission instructions.

\begin{table}
\caption{\label{jlab1}Journals to which this document applies, and macros for the abbreviated journal names in {\tt iopart.cls}. Macros for other journal titles are listed in appendix\,A.}
\footnotesize
\begin{tabular}{@{}llll}
\br
Short form of journal title&Macro name&Short form of journal title&Macro name\\
\mr
2D Mater.&\verb"\TDM"&Mater. Res. Express&\verb"\MRE"\\
Biofabrication&\verb"\BF"&Meas. Sci. Technol.$^c$&\verb"\MST"\\
Bioinspir. Biomim.&\verb"\BB"&Methods Appl. Fluoresc.&\verb"\MAF"\\
Biomed. Mater.&\verb"\BMM"&Modelling Simul. Mater. Sci. Eng.&\verb"\MSMSE"\\
Class. Quantum Grav.&\verb"\CQG"&Nucl. Fusion&\verb"\NF"\\
Comput. Sci. Disc.&\verb"\CSD"&New J. Phys.&\verb"\NJP"\\
Environ. Res. Lett.&\verb"\ERL"&Nonlinearity$^{a,b}$&\verb"\NL"\\
Eur. J. Phys.&\verb"\EJP"&Nanotechnology&\verb"\NT"\\
Inverse Problems&\verb"\IP"&Phys. Biol.$^c$&\verb"\PB"\\
J. Breath Res.&\verb"\JBR"&Phys. Educ.$^a$&\verb"\PED"\\
J. Geophys. Eng.$^d$&\verb"\JGE"&Physiol. Meas.$^{c,d,e}$&\verb"\PM"\\
J. Micromech. Microeng.&\verb"\JMM"&Phys. Med. Biol.$^{c,d,e}$&\verb"\PMB"\\
J. Neural Eng.$^c$&\verb"\JNE"&Plasma Phys. Control. Fusion&\verb"\PPCF"\\
J. Opt.&\verb"\JOPT"&Phys. Scr.&\verb"\PS"\\
J. Phys. A: Math. Theor.&\verb"\jpa"&Plasma Sources Sci. Technol.&\verb"\PSST"\\
J. Phys. B: At. Mol. Opt. Phys.&\verb"\jpb"&Rep. Prog. Phys.$^{e}$&\verb"\RPP"\\
J. Phys: Condens. Matter&\verb"\JPCM"&Semicond. Sci. Technol.&\verb"\SST"\\
J. Phys. D: Appl. Phys.&\verb"\JPD"&Smart Mater. Struct.&\verb"\SMS"\\
J. Phys. G: Nucl. Part. Phys.&\verb"\jpg"&Supercond. Sci. Technol.&\verb"\SUST"\\
J. Radiol. Prot.$^a$&\verb"\JRP"&Surf. Topogr.: Metrol. Prop.&\verb"\STMP"\\
Metrologia&\verb"\MET"&Transl. Mater. Res.&\verb"\TMR"\\
\br
\end{tabular}\\
$^{a}$UK spelling is required; $^{b}$MSC classification numbers are required; $^{c}$titles of articles are required in journal references; $^{d}$Harvard-style references must be used (see section \ref{except}); $^{e}$final page numbers of articles are required in journal references.

\end{table}
\normalsize

Any special submission requirements for the journals are indicated with footnotes in table~\ref{jlab1}.
Journals which require references in a particular format will need special care if you are using BibTeX, and you might need to use a \verb".bst" file
that gives slightly non-standard output in order to supply any extra information required.  It is not
necessary to give references in the exact style of references used in published articles, as long as all of
the required information is present.

Also note that there is an incompatibility
between \verb"amsmath.sty" and \verb"iopart.cls" which cannot be completely worked around.  If your article relies
on commands in \verb"amsmath.sty" that are not available in \verb"iopart.cls", you may wish to consider using a different
class file.

Whatever journal you are submitting to, please look at recent published articles (preferably
articles in your subject area) to familiarize yourself with the features of the journal.  We do not demand
that your \LaTeX\ file closely resembles a published article---a generic `preprint' appearance of the sort
commonly seen on \verb"arXiv.org" is fine---but your submission should be presented
in a way that makes it easy for the referees to form an opinion of whether it is suitable for the journal.
The generic advice in this document---on what to include in an abstract, how best to present complicated
mathematical expressions, and so on---applies whatever class file you are using.

\subsection{What you will need to supply}
Submissions to our journals are handled via the ScholarOne web-based submission system.  When you submit
a new article to us you need only submit a PDF of your article.  When you submit a revised version,
we ask you to submit the source files as well.  Upon acceptance for publication we will use the source files to produce a proof of your article in the journal style.

\subsubsection{Text.}When you send us the source files for a revised version of your submission,
you should send us the \LaTeX\ source code of your paper with all figures read in by 
the source code (see section \ref{figinc}).  Articles can be prepared using almost any version of \TeX\ or \LaTeX{},
not just \LaTeX\ with the class file \verb"iopart.cls".  You may split your \LaTeX\ file into several parts, but please show
which is the `master' \LaTeX\ file that reads in all of the other ones by naming it appropriately.  The `master'
\LaTeX\ file must read in all other \LaTeX\ and figure files from the current directory.  {\it Do not read in files from a different directory, e.g. \verb"\includegraphics{/figures/figure1.eps}" or
\verb"\include{../usr/home/smith/myfiles/macros.tex}"---we store submitted files
all together in a single directory with no subdirectories}.
\begin{itemize}
\item {\bf Using \LaTeX\ packages.} Most \LaTeXe\ packages can be used if they are 
available in common distributions of \LaTeXe; however, if it is essential to use 
a non-standard package then any extra files needed to process the article must 
also be supplied.  Try to avoid using any packages that manipulate or change the standard
\LaTeX\ fonts: published articles use fonts in the Times family, but we prefer that you 
use \LaTeX\ default Computer Modern fonts in your submission.  The use of \LaTeX\ 2.09, and of plain
\TeX\ and variants such as AMSTeX is acceptable, but a complete PDF of your submission should be supplied in these cases.
\end{itemize}
\subsubsection{Figures.} Figures should ideally be included in an article as encapsulated PostScript files
(see section \ref{figinc}) or created using standard \LaTeX\ drawing commands. 
 Please name all figure files using the guidelines in section \ref{fname}.
We accept submissions that use pdf\TeX\ to include
PDF or bitmap figures, but please ensure that you send us a PDF that uses PDF version 1.4 or lower
(to avoid problems in the ScholarOne system).
You can do this by putting \verb"\pdfminorversion=4" at the very start of your TeX file.

\label{fig1}All figures should be included within the body of the text 
at an appropriate point or grouped together with their captions at the end of the article. A standard graphics inclusion package such as \verb"graphicx" should be used for figure inclusion, and the package should be declared in the usual
way, for example with \verb"\usepackage{graphicx}", after the \verb"\documentclass" command.
Authors should avoid using special effects generated by including verbatim
PostScript code in the submitted \LaTeX\ file. Wherever possible, please try to use standard \LaTeX\ tools 
and packages.

\subsubsection{References.\label{bibby}}
You can produce your bibliography in the standard \LaTeX\ way using the \verb"\bibitem" command. Alternatively
you can use BibTeX: our preferred  \verb".bst" styles are:

\begin{itemize}
\item For the numerical (Vancouver) reference style we recommend that authors use 
 \verb"unsrt.bst"; this does not quite follow the style of published articles in our
 journals but this is not a problem.  Alternatively \verb"iopart-num.bst" created by Mark A Caprio
 produces a reference style that closely matches that in published articles.  The file is available from
\verb"http://ctan.org/tex-archive/biblio/bibtex/contrib/iopart-num/" .
\item For alphabetical (Harvard) style references we recommend that authors use the \verb"harvard.sty"
in conjunction with the \verb"jphysicsB.bst" BibTeX style file.  These, and accompanying documentation, can be downloaded
from \penalty-10000 \verb"http://www.ctan.org/tex-archive/macros/latex/contrib/harvard/".
Note that the \verb"jphysicsB.bst" bibliography style does not include article titles
in references to journal articles.
To include the titles of journal articles you can use the style \verb"dcu.bst" which is included
in the \verb"harvard.sty" package.  The output differs a little from the final journal reference
style, but all of the necessary information is present and the reference list will be formatted
into journal house style as part of the production process if your article is accepted for publication.
\end{itemize}

\noindent Please make sure that you include your \verb".bib" bibliographic database file(s) and any 
\verb".bst" style file(s) you have used.

\subsection{\label{copyright}Copyrighted material and ethical policy} If you wish to make use of previously published material for which you do not own the copyright then you must seek permission from the copyright holder, usually both the author and the publisher.  It is your responsibility to obtain copyright permissions and this should be done prior to submitting your article. If you have obtained permission, please provide full details of the permission granted---for example, copies of the text of any e-mails or a copy of any letters you may have received. Figure captions must include an acknowledgment of the original source of the material even when permission to reuse has been obtained.  Please read our ethical policy before writing your article.

\subsection{Naming your files}
\subsubsection{General.}
Please name all your files, both figures and text, as follows:
\begin{itemize}
\item Use only characters from the set a to z, A to Z, 0 to 9 and underscore (\_).
\item Do not use spaces or punctuation characters in file names.
\item Do not use any accented characters such as
\'a, \^e, \~n, \"o.
\item Include an extension to indicate the file type (e.g., \verb".tex", \verb".eps", \verb".txt", etc).
\item Use consistent upper and lower case in filenames and in your \LaTeX\ file.
If your \LaTeX\ file contains the line \verb"\includegraphics{fig1.eps}" the figure file must be called
\verb"fig1.eps" and not \verb"Fig1.eps" or \verb"fig1.EPS".  If you are on a Unix system, please ensure that
there are no pairs of figures whose names differ only in capitalization, such as \verb"fig_2a.eps" and \verb"fig_2A.eps",
as Windows systems will be unable to keep the two files in the same directory.
\end{itemize}
When you submit your article files, they are manipulated
and copied many times across multiple databases and file systems. Including non-standard
characters in your filenames will cause problems when processing your article.
\subsubsection{\label{fname}Naming your figure files.} In addition to the above points, please give each figure file a name which indicates the number of the figure it contains; for example, \verb"figure1.eps", \verb"figure2a.eps", etc. If the figure file contains a figure with multiple parts, for example figure 2(a) to 2(e), give it a name such as \verb"figure2a_2e.eps", and so forth.
\subsection{How to send your files}
Please send your submission via the ScholarOne submission system.  Go to the journal home
page, and use the `Submit an article' link on the right-hand side.

\section{Preparing your article}

\subsection{Sample coding for the start of an article}
\label{startsample}
The code for the start of a title page of a typical paper in the \verb"iopart.cls" style might read:
\small\begin{verbatim}
\documentclass[12pt]{iopart}
\begin{document}
\title[The anomalous magnetic moment of the 
neutrino]{The anomalous magnetic moment of the 
neutrino and its relation to the solar neutrino problem}

\author{P J Smith$^1$, T M Collins$^2$, 
R J Jones$^3$\footnote{Present address:
Department of Physics, University of Bristol, Tyndalls Park Road, 
Bristol BS8 1TS, UK.} and Janet Williams$^3$}

\address{$^1$ Mathematics Faculty, Open University, 
Milton Keynes MK7~6AA, UK}
\address{$^2$ Department of Mathematics, 
Imperial College, Prince Consort Road, London SW7~2BZ, UK}
\address{$^3$ Department of Computer Science, 
University College London, Gower Street, London WC1E~6BT, UK}
\ead{williams@ucl.ac.uk}

\begin{abstract}
...
\end{abstract}
\keywords{magnetic moment, solar neutrinos, astrophysics}
\submitto{\jpg}
\maketitle
\end{verbatim}
\normalsize

At the start of the \LaTeX\ source code please include 
commented material to identify the journal, author, and (if you are sending a revised
version or a resubmission) the reference number that the journal
has given to the submission. The first non-commented line should be 
\verb"\documentclass[12pt]{iopart}"  to load the preprint class 
file.  The normal text will be in the Computer Modern 12pt font.
It is possible to specify 10pt font size by passing the option \verb"[10pt]" to the class file.
Although it is possible to choose a font other than Computer Modern by loading external packages, this is not recommended.

The article text begins after \verb"\begin{document}".
Authors of very long articles may find it convenient to separate 
their article into a series of \LaTeX\ files each containing one section, and each of which is called 
in turn by the primary file.  The files for each section should be read in from the current directory;
please name the primary file clearly so that we know to run \LaTeX\ on this file.

Authors may use any common \LaTeX\ \verb".sty" files.
Authors may also define their own macros and definitions either in the main article \LaTeX\ file
or in a separate \verb".tex" or \verb".sty" file that is read in by the
main file, provided they do not overwrite existing definitions.
It is helpful to the production staff if complicated author-defined macros are explained in a \LaTeX\ comment.
The article class \verb"iopart.cls" can be used with other package files such
as those loading the AMS extension fonts 
\verb"msam" and \verb"msbm", which provide the 
blackboard bold alphabet and various extra maths symbols as well as symbols useful in figure 
captions.  An extra style file \verb"iopams.sty" is provided to load these
packages and provide extra definitions for bold Greek letters.

\subsection{\label{dblcol}Double-column layout}
The \verb"iopart.cls" class file produces single-column output by default, but a two-column layout can be obtained by
using \verb"\documentclass[10pt]" at the start of the file and \verb"\ioptwocol" after the \verb"\maketitle" command.  Two-column output will begin
on a new page (unlike in published double-column articles, where the two-column material
starts on the same page as the abstract).

In general we prefer to receive submissions in single-column format even for journals
published in double-column style; however, the \verb"\ioptwocol" option may be useful to test figure sizes
and equation breaks for these journals.  When setting material
in two columns you can use the asterisked versions of \LaTeX\ commands such as \verb"\begin{figure*} ... \end{figure*}"
to set figures and tables across two columns.  If you have any problems or any queries about producing two-column output, please contact us at \verb"submissions@iop.org".

\section{The title and abstract page} 
If you use \verb"iopart.cls", the code for setting the title page information is slightly different from
the normal default in \LaTeX.  If you are using a different class file, you do not need to mimic the appearance of
an \verb"iopart.cls" title page, but please ensure that all of the necessary information is present.

\subsection{Titles and article types}
The title is set using the command
\verb"\title{#1}", where \verb"#1" is the title of the article. The
first letter 
of the title should be capitalized with the rest in lower case. 
The title appears in bold case, but mathematical expressions within the title may be left in light-face type.

If the title is too long to use as a running head at the top of each page (apart from the
first) a short
form can be provided as an optional argument (in square brackets)
before the full title, i.e.\ \verb"\title[Short title]{Full title}".

For article types other than papers, \verb"iopart.cls"
has a generic heading \verb"\article[Short title]{TYPE}{Full title}" 
and some specific definitions given in table~\ref{arttype}. In each case (apart from Letters
to the Editor and Fast Track Communications) an 
optional argument can be used immediately after the control sequence name
to specify the short title; where no short title is given, the full title
will be used as the running head.  Not every article type has its own macro---use \verb"\article" for
any not listed.  A full list of the types of articles published by a journal is given
in the submission information available via the journal home page.
%For Letters use \verb"\letter{Full title}"; no short title is required as 
%the running head is automatically defined to be {\it Letter to the Editor}.
The generic heading could be used for 
articles such as those presented at a conference or workshop, e.g.
\small\begin{verbatim}
\article[Short title]{Workshop on High-Energy Physics}{Title}
\end{verbatim}\normalsize
Footnotes to titles may be given by using \verb"\footnote{Text of footnote.}" immediately after the title.
Acknowledgment of funding should be included in the acknowledgments section rather than in a footnote.

\begin{table}
\caption{\label{arttype}Types of article defined in the {\tt iopart.cls} 
class file.}
\footnotesize\rm
\begin{tabular*}{\textwidth}{@{}l*{15}{@{\extracolsep{0pt plus12pt}}l}}
\br
Command& Article type\\
\mr
\verb"\title{#1}"&Paper (no surtitle on first page)\\
\verb"\ftc{#1}"&Fast Track Communication\\
\verb"\review{#1}"&Review\\
\verb"\topical{#1}"&Topical Review\\
\verb"\comment{#1}"&Comment\\
\verb"\note{#1}"&Note\\
\verb"\paper{#1}"&Paper (no surtitle on first page)\\
\verb"\prelim{#1}"&Preliminary Communication\\
\verb"\rapid{#1}"&Rapid Communication\\
\verb"\letter{#1}"&Letter to the Editor\\
\verb"\article{#1}{#2}"&Other articles\\\ & (use this for any other type of article; surtitle is whatever is entered as {\tt 
\#1})\\
\br
\end{tabular*}
\end{table}

\subsection{Authors' names and addresses}
For the authors' names type \verb"\author{#1}", 
where \verb"#1" is the 
list of all authors' names. Western-style names should be written as initials then
family name, with a comma after all but the last 
two names, which are separated by `and'. Initials should {\it not} be followed by full stops. First (given) names may be used if 
desired.  Names in Chinese, Japanese and Korean styles should be written as you want them to appear in the published article. Authors in all IOP Publishing journals have the option to include their names in Chinese, Japanese or Korean characters in addition to the English name: see appendix B for details.

If the authors are at different addresses a superscripted number, e.g. $^1$, \verb"$^1$", should be used after each 
name to reference the author to his/her address.
If an author has additional information to appear as a footnote, such as 
a permanent address, a normal \LaTeX\ footnote command
should be given after the family name and address marker 
with this extra information.

The authors' affiliations follow the list of authors. 
Each address is set by using
\verb"\address{#1}" with the address as the single parameter in braces. 
If there is more 
than one address then the appropriate superscripted number, followed by a space, should come at the start of
the address.

E-mail addresses are added by inserting the 
command \verb"\ead{#1}" after the postal address(es) where \verb"#1" is the e-mail address.  
See section~\ref{startsample} for sample coding. For more than one e-mail address, please use the command 
\verb"\eads{\mailto{#1}, \mailto{#2}}" with \verb"\mailto" surrounding each e-mail address.  Please ensure
that, at the very least, you state the e-mail address of the corresponding author.

\subsection{The abstract}
The abstract follows the addresses and
should give readers concise information about the content 
of the article and indicate the main results obtained and conclusions 
drawn. It should be self-contained---there should be no references to 
figures, tables, equations, bibliographic references etc.  It should be enclosed between \verb"\begin{abstract}"
and \verb"\end{abstract}" commands.  The abstract should normally be restricted 
to a single paragraph of around 200 words.

\subsection{Subject classification numbers}
We no longer ask authors to supply Physics and Astronomy Classification System (PACS)
classification numbers.  For submissions to {\it Nonlinearity}\/ we ask that you should
supply Mathematics Subject Classification (MSC) codes.  MSC numbers are included after the abstract 
using \verb"\ams{#1}".

The command
\verb"\submitto{#1}" can be inserted, where \verb"#1" is the journal name written in full or the appropriate control sequence as
given in table~\ref{jlab1}. This command is not essential to the running of the file and can be omitted.

\subsection{Keywords}
Keywords are required for all submissions. Authors should supply a minimum of three (maximum seven) keywords appropriate to their article as a new paragraph starting \verb"\noindent{\it Keywords\/}:" after the end of the abstract.

\subsection{Making a separate title page}
To keep the header material on a separate page from the
body of the text insert \verb"\maketitle" (or \verb"\newpage") before the start of the text. 
If \verb"\maketitle" is not included the text of the
article will start immediately after the abstract.

\section{The text}
\subsection{Sections, subsections and subsubsections}
The text of articles may be divided into sections, subsections and, where necessary, 
subsubsections. To start a new section, end the previous paragraph and 
then include \verb"\section" followed by the section heading within braces. 
Numbering of sections is done {\it automatically} in the headings: 
sections will be numbered 1, 2, 3, etc, subsections will be numbered 
2.1, 2.2,  3.1, etc, and subsubsections will be numbered 2.3.1, 2.3.2, 
etc.  Cross references to other sections in the text should, where
possible, be made using 
labels (see section~\ref{xrefs}) but can also
be made manually. See section~\ref{eqnum} for information on the numbering of displayed equations. Subsections and subsubsections are 
similar to sections but 
the commands are \verb"\subsection" and \verb"\subsubsection" respectively. 
Sections have a bold heading, subsections an italic heading and 
subsubsections an italic heading with the text following on directly.
\small\begin{verbatim}
\section{This is the section title}
\subsection{This is the subsection title}
\end{verbatim}\normalsize

The first section is normally an introduction,  which should state clearly 
the object of the work, its scope and the main advances reported, with 
brief references to relevant results by other workers. In long papers it is 
helpful to indicate the way in which the paper is arranged and the results 
presented.

Footnotes should be avoided whenever possible and can often be included in the text as phrases or sentences in parentheses. If required, they should be used only for brief notes that do not fit conveniently into the text. The use of 
displayed mathematics in footnotes should be avoided wherever possible and no equations within a footnote should be numbered. 
The standard \LaTeX\ macro \verb"\footnote" should be used.  Note that in \verb"iopart.cls" the \verb"\footnote" command
produces footnotes indexed by a variety of different symbols,
whereas in published articles we use numbered footnotes.  This
is not a problem: we will convert symbol-indexed footnotes to numbered ones during the production process.

\subsection{Acknowledgments}
Authors wishing to acknowledge assistance or encouragement from 
colleagues, special work by technical staff or financial support from 
organizations should do so in an unnumbered `Acknowledgments' section 
immediately following the last numbered section of the paper. In \verb"iopart.cls" the 
command \verb"\ack" sets the acknowledgments heading as an unnumbered
section.

Please ensure that you include all of the sources of funding and the funding contract reference numbers that you are contractually obliged to acknowledge. We often receive requests to add such information very late in the production process, or even after the article is published, and we cannot always do this. Please collect all of the necessary information from your co-authors and sponsors as early as possible.

\subsection{Appendices}
Technical detail that it is necessary to include, but that interrupts 
the flow of the article, may be consigned to an appendix. 
Any appendices should be included at the end of the main text of the paper, after the acknowledgments section (if any) but before the reference list.
If there are 
two or more appendices they should be called Appendix A, Appendix B, etc. 
Numbered equations will be in the form (A.1), (A.2), etc,
figures will appear as figure A1, figure B1, etc and tables as table A1,
table B1, etc.

The command \verb"\appendix" is used to signify the start of the
appendices. Thereafter \verb"\section", \verb"\subsection", etc, will 
give headings appropriate for an appendix. To obtain a simple heading of 
`Appendix' use the code \verb"\section*{Appendix}". If it contains
numbered equations, figures or tables the command \verb"\appendix" should
precede it and \verb"\setcounter{section}{1}" must follow it.

\subsection{Some matters of style}
It will help the readers if your article is written in a clear,
consistent and concise manner. During the production process
we will try to make sure that your work is presented to its
readers in the best possible way without sacrificing the individuality of
your writing. Some recommended 
points to note, however, are the following.  These apply to all of the journals listed
in table~\ref{jlab1}.
\begin{enumerate}
\item Authors are often inconsistent in the use of `ize' and `ise' endings.
We recommend using `-ize' spellings (diagonalize, 
renormalization, minimization, etc) but there are some common 
exceptions to this, for example: devise, 
promise and advise.

\item The words table and figure should be written 
in full and {\bf not} abbreviaged to tab. and fig. Do not include `eq.', `equation' etc before an equation number or `ref.'\, `reference' etc before a reference number.
\end{enumerate}

Please check your article carefully for accuracy, consistency and clarity before
submission. Remember that your article will probably be read by many
people whose native language is not English and who may not  
be aware of many of the subtle meanings of words or idiomatic phases
present in the English language. It therefore helps if you try to keep
sentences as short and simple as possible.  If you are not a native English speaker,
please ask a native English speaker to read your paper and check its grammar.

\section{Mathematics}
\subsection{Two-line constructions}
The great advantage of \LaTeX\ 
over other text processing systems is its 
ability to handle mathematics of almost any degree of complexity. However, 
in order to produce an article suitable for publication both within a print journal and online, 
authors should exercise some restraint on the constructions used. Some equations using very small characters which are clear in a preprint style article may be difficult read in a smaller format.

For simple fractions in the text the solidus \verb"/", as in 
$\lambda/2\pi$, should be used instead of \verb"\frac" or \verb"\over", 
using parentheses where necessary to avoid ambiguity, 
for example to distinguish between $1/(n-1)$ and $1/n-1$. Exceptions to 
this are the proper fractions $\frac12$, $\frac13$, $\frac34$, 
etc, which are better left in this form. In displayed equations 
horizontal lines are preferable to solidi provided the equation is 
kept within a height of two lines. A two-line solidus should be 
avoided where possible; the construction $(\ldots)^{-1}$ should be 
used instead. For example use:
\begin{equation*}
\frac{1}{M_{\rm a}}\left(\int^\infty_0{\rm d}
\omega\;\frac{|S_o|^2}{N}\right)^{-1}\qquad\mbox{instead of}\qquad
\frac{1}{M_{\rm a}}\biggl/\int^\infty_0{\rm d}
\omega\;\frac{|S_o|^2}{N}.
\end{equation*}

\subsection{Roman and italic in mathematics}
In mathematics mode \LaTeX\ automatically sets variables in an italic 
font. In most cases authors should accept this italicization. However, 
there are some cases where it is preferable to use a Roman font; for 
instance, a Roman d for a differential d, a Roman e 
for an exponential e and a Roman i for the square root of $-1$. To 
accommodate this and to simplify the  typing of equations, \verb"iopart.cls" provides
some extra definitions. \verb"\rmd", \verb"\rme" and \verb"\rmi" 
now give Roman d, e and i respectively for use in equations, 
e.g.\ $\rmi x\rme^{2x}\rmd x/\rmd y$ 
is obtained by typing \verb"$\rmi x\rme^{2x}\rmd x/\rmd y$".

Certain other common mathematical functions, such as cos, sin, det and 
ker, should appear in Roman type. Standard \LaTeX\ provides macros for 
most of these functions 
(in the cases above, \verb"\cos", \verb"\sin", \verb"\det" and \verb"\ker" 
respectively); \verb"iopart.cls" also provides 
additional definitions for $\Tr$, $\tr$ and 
$\Or$ (\verb"\Tr", \verb"\tr" and \verb"\Or", respectively).

Subscripts and superscripts should be in Roman type if they are labels 
rather than variables or characters that take values. For example in the 
equation
\[
\epsilon_m=-g\mu_{\rm n}Bm
\]
$m$, the $z$ component of the nuclear spin, is italic because it can have 
different values whereas n is Roman because it 
is a label meaning nuclear ($\mu_{\rm n}$ 
is the nuclear magneton).

\subsection{Displayed equations in double-column journals}
Authors should bear in mind that all mathematical formulae in double-column journals will need to fit
into the width of a single column.  You may find it helpful to use a two-column layout (such as the two-column
option in \verb"iopart.cls") in your submission so that you can check the width of equations.

\subsection{Special characters for mathematics}
Bold italic characters can be used in our journals to signify vectors (rather
than using an upright bold or an over arrow). To obtain this effect when using \verb"iopart.cls",
use \verb"\bi{#1}" within maths mode, e.g. $\bi{ABCdef}$. Similarly, in \verb"iopart.cls", if upright 
bold characters are required in maths, use \verb"\mathbf{#1}" within maths
mode, e.g. $\mathbf{XYZabc}$. The calligraphic (script) uppercase alphabet
is obtained with \verb"\mathcal{AB}" or \verb"\cal{CD}" 
($\mathcal{AB}\cal{CD}$).

The American Mathematical Society provides a series of extra symbol fonts
to use with \LaTeX\ and packages containing the character definitions to
use these fonts. Authors wishing to use Fraktur 
\ifiopams$\mathfrak{ABC}$ \fi
or Blackboard Bold \ifiopams$\mathbb{XYZ}$ \fi can include the appropriate
AMS package (e.g. \verb"amsgen.sty", \verb"amsfonts.sty", \verb"amsbsy.sty", \verb"amssymb.sty") with a 
\verb"\usepackage" command or add the command \verb"\usepackage{iopams}"
which loads the four AMS packages mentioned above and also provides
definitions for extra bold characters (all Greek letters and some other
additional symbols).

The package \verb"iopams.sty" uses the definition \verb"\boldsymbol" in \verb"amsbsy.sty"
which allows individual non-alphabetical symbols and Greek letters to be 
made bold within equations.
The bold Greek lowercase letters \ifiopams$\balpha \ldots\bomega$,\fi 
are obtained with the commands 
\verb"\balpha" \dots\ \verb"\bomega" (but note that
bold eta\ifiopams, $\bfeta$,\fi\ is \verb"\bfeta" rather than \verb"\beta")
and the capitals\ifiopams, $\bGamma\ldots\bOmega$,\fi\ with commands 
\verb"\bGamma" \dots\
\verb"\bOmega". Bold versions of the following symbols are
predefined in \verb"iopams.sty": 
bold partial\ifiopams, $\bpartial$,\fi\ \verb"\bpartial",
bold `ell'\ifiopams, $\bell$,\fi\  \verb"\bell", 
bold imath\ifiopams, $\bimath$,\fi\  \verb"\bimath", 
bold jmath\ifiopams, $\bjmath$,\fi\  \verb"\bjmath", 
bold infinity\ifiopams, $\binfty$,\fi\ \verb"\binfty", 
bold nabla\ifiopams, $\bnabla$,\fi\ \verb"\bnabla", 
bold centred dot\ifiopams, $\bdot$,\fi\  \verb"\bdot". Other 
characters are made bold using 
\verb"\boldsymbol{\symbolname}".

Please do not use the style file \verb"amsmath.sty" (part of the AMSTeX package) in conjunction with \verb"iopart.cls". This will result in several errors. To make use of the macros defined in \verb"amsmath.sty", \verb"iopart.cls" provides the file \verb"setstack.sty" which reproduces the following useful macros from \verb"amsmath.sty":
\small\begin{verbatim}
\overset \underset \sideset \substack \boxed   \leftroot
\uproot  \dddot    \ddddot  \varrow   \harrow
\end{verbatim}\normalsize

If the mathematical notation
that you need is best handled in \verb"amsmath.sty" you might want to consider using an article class
other than \verb"iopart.cls". We accept submissions using any class or style files.

Table~\ref{math-tab2} lists some other macros for use in 
mathematics with a brief description of their purpose.

\begin{table}
\caption{\label{math-tab2}Other macros defined in {\tt iopart.cls} for use in maths.}
\begin{tabular*}{\textwidth}{@{}l*{15}{@{\extracolsep{0pt plus
12pt}}l}}
\br
Macro&Result&Description\\
\mr
\verb"\fl"&&Start line of equation full left\\
\verb"\case{#1}{#2}"&$\case{\#1}{\#2}$&Text style fraction in display\\
\verb"\Tr"&$\Tr$&Roman Tr (Trace)\\
\verb"\tr"&$\tr$&Roman tr (trace)\\
\verb"\Or"&$\Or$&Roman O (of order of)\\
\verb"\tdot{#1}"&$\tdot{x}$&Triple dot over character\\
\verb"\lshad"&$\lshad$&Text size left shadow bracket\\
\verb"\rshad"&$\rshad$&Text size right shadow bracket\\
\br
\end{tabular*}
\end{table}

\subsection{Alignment of displayed equations}

The normal style for aligning displayed equations in our published journal articles is to align them left rather than centre. The \verb"iopart.cls" class file automatically does this and indents each line of a display.  In \verb"iopart.cls", to make any line start at the left margin of the page, add \verb"\fl" at start of the line (to indicate full left).

Using the \verb"eqnarray" environment equations will naturally be aligned left and indented without the use of any ampersands for alignment, see equations (\ref{eq1}) and (\ref{eq2})
\begin{eqnarray}
\alpha + \beta =\gamma^2, \label{eq1}\\
\alpha^2 + 2\gamma + \cos\theta = \delta. \label{eq2} 
\end{eqnarray}
This is the normal equation style for our journals.

Where some secondary alignment is needed, for instance a second part of an equation on a second line, a single ampersand is added at the point of alignment in each line  (see  (\ref{eq3}) and (\ref{eq4})).
\begin{eqnarray}
\alpha &=2\gamma^2 + \cos\theta + \frac{XY \sin\theta}{X+ Y\cos\theta} \label{eq3}\\
 & = \delta\theta PQ \cos\gamma. \label{eq4} 
\end{eqnarray}

Two points of alignment are possible using two ampersands for alignment (see  (\ref{eq5}) and (\ref{eq6})).  Note in this case extra space \verb"\qquad" is added before the second ampersand in the longest line (the top one) to separate the condition from the equation. 
\begin{eqnarray}
\alpha &=2\gamma^2 + \cos\theta + \frac{XY \sin\theta}{X+ Y\cos\theta}\qquad& \theta > 1 \label{eq5}\\
 & = \delta\theta PQ \cos\gamma &\theta \leq 1.\label{eq6} 
\end{eqnarray}

For a long equation which has to be split over more than one line the first line should start at the left margin, this is achieved by inserting \verb"\fl" (full left) at the start of the line. The use of the alignment parameter \verb"&" is not necessary unless some secondary alignment is needed.
\begin{eqnarray}
\fl \alpha + 2\gamma^2 = \cos\theta + \frac{XY \sin\theta}{X+ Y\cos\theta} +  \frac{XY \sin\theta}{X- Y\cos\theta} +
+ \left(\frac{XY \sin\theta}{X+ Y\cos\theta}\right)^2 \nonumber\\
+  \left(\frac{XY \sin\theta}{X- Y\cos\theta}\right)^2.\label{eq7} 
\end{eqnarray}

The plain \TeX\ command \verb"\eqalign" can be used within an \verb"equation" environment to obtain a multiline equation with a single centred number, for example
\begin{equation}
\eqalign{\alpha + \beta =\gamma^2 \cr
\alpha^2 + 2\gamma + \cos\theta = \delta.} 
\end{equation}

During the production process we will break equations as appropriate for the page layout of the journal. If you are submitting to a double-column journal and wish to review how your equations will break, you may find the double-column layout described in section \ref{dblcol} useful.

\subsection{Miscellaneous points}
The following points on the layout of mathematics apply whichever class file you use.

Exponential expressions, especially those containing subscripts or 
superscripts, are clearer if the notation $\exp(\ldots)$ is used, except for 
simple examples. For instance $\exp[\rmi(kx-\omega t)]$ and $\exp(z^2)$ are 
preferred to $\e^{\rmi(kx-\omega t)}$ and $\e^{z^2}$, but 
$\e^x$ 
is acceptable.

Similarly the square root sign $\sqrt{\phantom{b}}$ should 
only be used with relatively
simple expressions, e.g.\ $\sqrt2$ and $\sqrt{a^2+b^2}$;
in other cases the 
power $1/2$ should be used; for example, $[(x^2+y^2)/xy(x-y)]^{1/2}$.

It is important to distinguish between $\ln = \log_\e$ and $\lg 
=\log_{10}$. Braces, brackets and parentheses should be used in the 
following order: $\{[(\;)]\}$. The same ordering of brackets should be 
used within each size. However, this ordering can be ignored if the
brackets have a 
special meaning (e.g.\ if they denote an average or a function).

Decimal fractions should always be preceded by a zero: for example 0.123 {\bf not} .123.
For long numbers use thin spaces after every third character away from the position of the decimal point, unless 
this leaves a single separated character: e.g.\ $60\,000$, $0.123\,456\,78$ 
but 4321 and 0.7325.

Equations should be followed by a full stop (periods) when at the end
of a sentence.

\subsection{Equation numbering and layout in {\tt iopart.cls}}
\label{eqnum}
\LaTeX\ provides facilities for automatically numbering equations 
and these should be used where possible. Sequential numbering (1), (2), 
etc, is the default numbering system although in \verb"iopart.cls", if the command
\verb"\eqnobysec" is included in the preamble, equation numbering
by section is obtained, e.g.\ 
(2.1), (2.2), etc. Equation numbering by section is used in appendices automatically when the \verb"\appendix" command is used, even if sequential numbering has been used in the rest of the article. 
Refer to equations in the text using the equation number in parentheses. It is not normally necessary to include the word equation before the number; and abbreviations such as eqn or eq should not be used.
In \verb"iopart.cls", there are alternatives to the standard \verb"\ref" command that you might
find useful---see \tref{abrefs}.

Sometimes it is useful to number equations as parts of the same
basic equation. This can be accomplished in \verb"iopart.cls" by inserting the 
commands \verb"\numparts" before the equations concerned and 
\verb"\endnumparts" when reverting to the normal sequential numbering.
For example using \verb"\numparts \begin{eqnarray}" ... \verb"\end{eqnarray} \endnumparts":

\numparts
\begin{eqnarray}
T_{11}&=(1+P_\e)I_{\uparrow\uparrow}-(1-P_\e)
I_{\uparrow\downarrow},\label{second}\\
T_{-1-1}&=(1+P_\e)I_{\downarrow\downarrow}-(1-P_\e)I_{\uparrow\downarrow},\\
S_{11}&=(3+P_\e)I_{\downarrow\uparrow}-(3-P_e)I_{\uparrow\uparrow},\\
S_{-1-1}&=(3+P_\e)I_{\uparrow\downarrow}-(3-P_\e)
I_{\downarrow\downarrow}.
\end{eqnarray}
\endnumparts

Equation labels within the \verb"\eqnarray" environment will be referenced
as subequations, e.g. (\ref{second}).

\subsection{Miscellaneous extra commands for displayed equations}
The \verb"\cases" command has been amended slightly in \verb"iopart.cls" to 
increase the space between the equation and the condition. 
\Eref{cases} 
demonstrates simply the output from the \verb"\cases" command
\begin{equation}
\label{cases}
X=\cases{1&for $x \ge 0$\\
-1&for $x<0$\\}
\end{equation}
%The code used was:
%\small\begin{verbatim}
%\begin{equation}
%\label{cases}
%X=\cases{1&for $x \ge 0$\\
%-1&for $x<0$\\}
%\end{equation}
%\end{verbatim}
%\normalsize

To obtain text style fractions within displayed maths the command 
\verb"\case{#1}{#2}" can be used instead
of the usual \verb"\frac{#1}{#2}" command or \verb"{#1 \over #2}".

When two or more short equations are on the same line they should be 
separated by a `qquad space' (\verb"\qquad"), rather than
\verb"\quad" or any combination of \verb"\,", \verb"\>", \verb"\;" 
and \verb"\ ".

\section{Referencing\label{except}}
Two different styles of referencing are in common use: 
the Harvard alphabetical system and the Vancouver numerical system. 
All journals to which this document applies allow the use of either the Harvard or Vancouver system, 
except for {\it Physics in Medicine and Biology} and {\it Physiological Measurement}
for which authors {\it must\/} use the Harvard referencing style (with the titles of journal
articles given, and final page numbers given).

\subsection{Harvard (alphabetical) system}
In the Harvard system the name of the author appears in the text together 
with the year of publication. As appropriate, either the date or the name 
and date are included within parentheses. Where there are only two authors 
both names should be given in the text; if there are more than two 
authors only the first name should appear followed by `{\it et al}' 
(which can be obtained in \verb"iopart.cls" by typing \verb"\etal"). When two or 
more references to work by one author or group of authors occur for the 
same year they should be identified by including a, b, etc after the date 
(e.g.\ 2012a). If several references to different pages of the same article 
occur the appropriate page number may be given in the text, e.g.\ Kitchen 
(2011, p 39).

The reference list at the end of an article consists of an 
unnumbered `References' section containing an
alphabetical listing by authors' names. References with the same author list are ordered by date, oldest first.
The reference list in the 
preprint style is started in \verb"iopart.cls" by including the command \verb"\section*{References}" and then
\verb"\begin{harvard}".
Individual references start with \verb"\item[]" and the reference list is completed with \verb"\end{harvard}".
There is also a shortened form of the coding: \verb"\section*{References}"
and \verb"\begin{harvard}" can be replaced by the single command
\verb"\References", and \verb"\end{harvard}" can be shortened to
\verb"\endrefs".

\subsection{Vancouver (numerical) system}
In the Vancouver system references are numbered sequentially 
throughout the text. The numbers occur within square brackets and one 
number can be used to designate several references. A numerical 
reference list in the \verb"iopart" style is started by including the 
command \verb"\section*{References}" and then
\verb"\begin{thebibliography}{<num>}", where \verb"<num>" is the largest
number in the reference list (or any other number with the same number
of digits).  The 
reference list gives the references in 
numerical order, individual references start with \verb"\bibitem{label}". The list is completed by
\verb"\end{thebibliography}". Short forms of the commands are again
available: \verb"\Bibliography{<num>}" can be used at the start of the
references section and \verb"\endbib" at the end.

A variant of this system is to use labels instead of numbers within 
square brackets, in this case references in the list should start with \verb"\bibitem[label-text]". This method is allowed for all journals that accept numerical references.

\subsection{BibTeX\label{bibtex}}
If you are using BibTeX, see the earlier section \ref{bibby} for information on what \verb".bst" file to use.
The output that you get will differ slightly from that specified in the rest of this section,
but this is not a problem as long as all the relevant information is present.

\subsection{References, general}
A complete reference should provide the reader with enough information to 
locate the item concerned. Up to ten authors may be given in a particular reference; where 
there are more than ten only the first should be given followed by 
`{\it et al}'.  If you are using BibTeX
and the \verb".bst" file that you are using includes more than 10 authors, do not worry about this:
we can correct this during the production process.  Abbreviate a journal name only in accordance with the journal's
own recommendations for abbreviation---if in doubt, leave it unabbreviated.

The terms {\it loc.\ cit.}\ and {\it ibid}.\ should not be used. 
Unpublished conferences and reports should generally not be included 
in the reference list if a published version of the work exists. Articles in the course of publication should 
include the article title and the journal of publication, if known. 
A reference to a thesis submitted for a higher degree may be included 
if it has not been superseded by a published 
paper---please state the institution where the work was submitted.

The basic structure of a reference in the reference list is the same in both the alphabetical and numerical systems, the only difference being the code at the start of the reference. Alphabetical references are preceded by \verb"\item[]", numerical by \verb"\bibitem{label}" or just \verb"\item" to generate a number or \verb"\nonum" where a reference is not the first in a group of references under the same number.

Note that footnotes to the text should not be 
included in the reference list, but should appear at the bottom of the relevant page by using the \verb"\footnote" command.

\subsection{References to journal articles}
The following guidance applies if you are producing your reference list `by hand'; that is,
without the help of BibTeX.  See section~\ref{bibby} for BibTeX help.

Article references in published articles in our journals contain three changes of 
font:
the authors and date appear in Roman type, the journal title in 
italic, the volume number in bold and the page numbers in Roman again. 
A typical journal entry would be:

\smallskip
\begin{harvard}
\item[] Spicer P E, Nijhoff F W and van der Kamp P H 2011 {\it Nonlinearity} {\bf 24} 2229
\end{harvard}
\smallskip

\noindent which would be obtained by typing, within the references
environment 
\small\begin{verbatim}
\item[] Spicer P E, Nijhoff F W and van der Kamp P H 2011 {\it Nonlinearity} 
{\bf 24} 2229
\end{verbatim}\normalsize

\noindent Features to note are the following.

\begin{enumerate}
\item The authors should be in the form of surname (with only the first 
letter capitalized) followed by the initials with no 
periods after the initials. Authors should be separated by a comma 
except for the last two which should be separated by `and' with no 
comma preceding it.

\item The year of publication follows the authors and is not in parentheses.

\item Titles of journal articles can also be included (in Roman (upright) text after the year). Article titles are required in reference lists for {\it Inverse Problems, Journal of Neural Engineering, Measurement Science and Technology, Physical Biology, Physics in Medicine and Biology\/} and {\it Physiological Measurement}.

\item The journal is in italic and is abbreviated. If a journal has several parts denoted by 
different letters the part letter should be inserted after the journal in Roman type (e.g.\ 
{\it Phys.\ Rev.\ \rm A}). \verb"iopart.cls" includes macros for abbreviated titles of all journals handled by IOP Publishing (see table~\ref{jlab2}) and some other common titles (table \ref{jlab3}).

\item The volume number is bold; the page number is Roman.
 Both the initial and final page numbers should be given where possible---note that for {\it Reports on Progress in Physics\/}, {\it Physiological Measurement} and {\it Physics in Medicine and Biology}
 the final page number is {\it required}. The final page number should be in 
the shortest possible form and separated from the initial page number by an en rule (\verb"--"), e.g.\ 1203--14.

\item Where there are two or more references with identical authors, 
the authors' names should be repeated for the second and subsequent references. Each individual publication should be presented as a separate reference, although in the numerical system one number can be used for several references. This facilitates linking in the online journal. 
\end{enumerate}

\subsubsection{Article numbering.}
Many journals now use article-numbering systems that do not fit the conventional {\it year-journal-volume-page numbers} pattern. Some examples are:

\numrefs{1}
\item Carlip S and Vera R 1998 {\it Phys. Rev.} D {\bf 58} 011345 
\item Davies K and Brown G 1997 {\it J. High Energy Phys.} JHEP12(1997)002
\item Hannestad S 2005 {\it J. Cosmol. Astropart. Phys.} JCAP02(2005)011
\item Hilhorst H J 2005 {\it J. Stat. Mech.} L02003
\item Gundlach C 1999 {\it Liv. Rev. Rel.} 1994-4
\endnumrefs

\noindent The website of the journal you are citing should state the correct format for citations.

\subsection{Preprint references}
Preprints may be referenced but if the article concerned has been published in a peer-reviewed journal, that reference should take precedence. If only a preprint reference can be given, it is helpful to include the article title. Examples are:
\vskip6pt
\numrefs{1}
\item Neilson D and Choptuik M 2000 {\it Class. Quantum Grav.} {\bf 17} 761 (arXiv:gr-qc/9812053)
\item Sundu H, Azizi K, S\"ung\"u J Y and Yinelek N 2013 Properties of $D_{s2}^*(2573)$ charmed-strange tensor meson arXiv:1307.6058
\endnumrefs

\noindent For preprints added to arXiv.org after April 2007 it is not necessary to include the subject area, however this information can be included in square brackets after the number if desired, e.g.
\numrefs{1}
\item Sundu H, Azizi K, S\"ung\"u J Y and Yinelek N 2013 Properties of $D_{s2}^*(2573)$ charmed-strange tensor meson arXiv:1307.6058 [hep-ph]
\endnumrefs

\subsection{References to books, conference proceedings and reports}

References to books, proceedings and reports are similar, but have only two
changes of font. The authors and date of publication are in Roman, the 
title of the book is in italic, and the editors, publisher, 
town of publication 
and page number are in Roman. A typical reference to a book and a
conference paper might be

\smallskip
\begin{harvard}
\item[] Dorman L I 1975 {\it Variations of Galactic Cosmic Rays} 
(Moscow: Moscow State University Press) p~103
\item[] Caplar R and Kulisic P 1973 {\it Proc.\
Int.\ Conf.\ on Nuclear Physics (Munich)} vol~1 (Amsterdam:  
North-Holland/American Elsevier) p~517
\end{harvard}
\smallskip

\noindent which would be obtained by with the code
\small\begin{verbatim}
\item[] Dorman L I 1975 {\it Variations of Galactic Cosmic Rays} 
(Moscow: Moscow State University Press) p~103
\item[] Caplar R and Kulisic P 1973 {\it Proc. Int. Conf. on Nuclear 
Physics (Munich)} vol~1 (Amsterdam: North-Holland/American 
Elsevier) p~517
\end{verbatim}\normalsize
\noindent

\noindent Features to note are the following.
\begin{enumerate}
\item Book titles are in italic and should be spelt out in full with 
initial capital letters for all except minor words. Words such as 
Proceedings, Symposium, International, Conference, Second, etc should 
be abbreviated to Proc., Symp., Int., Conf., 2nd, 
respectively, but the rest of the title should be given in full, 
followed by the date of the conference and the 
town or city where the conference was held. For 
laboratory reports the laboratory should be spelt out wherever 
possible, e.g.\ {\it Argonne National Laboratory Report}.

\item The volume number, for example, vol~2, should be followed by 
the editors, if any, in the form ed~A~J~Smith and P~R~Jones. Use 
\etal if there are more than two editors. Next comes the town of 
publication and publisher, within brackets and separated by a colon, 
and finally the page numbers preceded by p if only one number is given 
or pp if both the initial and final numbers are given.

\item If a book is part of a series (for examples, {\it Springer Tracts in Modern Physics\/}), the series title and volume number is given in parentheses after the book title. Whereas for an individual volume in a multivolume set, the set title is given first, then the volume title. 
\end{enumerate}
\smallskip
\begin{harvard}
\item[]Morse M 1996 Supersonic beam sources {\it Atomic Molecular and Optical Physics\/} ({\it Experimental Methods in the Physical Sciences\/} vol 29) ed F B Dunning and R Hulet (San Diego, CA: Academic)  
\item[]Fulco C E, Liverman C T and Sox H C (eds) 2000 {\it Gulf War and Health\/} vol 1 {\it Depleted Uranium, Pyridostigmine Bromide, Sarin, and Vaccines\/} (Washington, DC: The National Academies Press)
\end{harvard}

\section{Cross-referencing\label{xrefs}}
The facility to cross reference items in the text is very useful when 
composing articles as the precise form of the article may be uncertain at the start 
and  revisions and amendments may subsequently be made. 
\LaTeX\ provides excellent facilities for doing cross-referencing
and these can be very useful in preparing articles.

\subsection{References}
\label{refs}
Cross referencing is useful for numeric reference lists because, if it 
is used, adding 
another reference to the list does not then involve renumbering all 
subsequent references. It is not necessary for referencing 
in the Harvard system where the final reference list is alphabetical 
and normally no other changes are necessary when a reference is added or
deleted.
When using \LaTeX , two passes (under certain circumstances, three passes)
are necessary initially to get the cross references right 
but once they are correct a single run is usually sufficient provided an 
\verb".aux" file is available and the file 
is run to the end each time.
If the 
reference list contains an entry \verb"\bibitem{label}", 
this command 
will produce the correct number in the reference list and 
\verb"\cite{label}" will produce the number within square brackets in the 
text. \verb"label" may contain letters, numbers 
or punctuation characters but must not contain spaces or commas. It is also
recommended that the underscore character \_{} is not used in cross
referencing. 
Thus labels of the form 
\verb"eq:partial", \verb"fig:run1", \verb"eq:dy'", 
etc, may be used. When several 
references occur together in the text \verb"\cite" may be used with 
multiple labels with commas but no spaces separating them; 
the output will be the 
numbers within a single pair of square brackets with a comma and a 
thin space separating the numbers. Thus \verb"\cite{label1,label2,label4}"
would give [1,\,2,\,4]. Note that no attempt is made by the style file to sort the 
labels and no shortening of groups of consecutive numbers is done.
Authors should therefore either try to use multiple labels in the correct 
order, or use a package such as \verb"cite.sty" that reorders labels
correctly.

The numbers for the cross referencing are generated in the order the 
references appear in the reference list, so that if the entries in the 
list are not in the order in which the references appear in the text 
then the 
numbering within the text will not be sequential. To correct this 
change the ordering of the entries in the reference list and then 
rerun the \LaTeX\ file {\it twice}.  Please ensure that all references resolve correctly: check the \verb".log" file
for undefined or multiply-defined citations, and check that the output does not contain question
marks that indicate unresolved references.

\subsection{Equation numbers, sections, subsections, figures and 
tables}
Labels for equation numbers, sections, subsections, figures and tables 
are all defined with the \verb"\label{label}" command and cross references 
to them are made with the \verb"\ref{label}" command.

Any section, subsection, subsubsection, appendix or subappendix 
command defines a section type label, e.g. 1, 2.2, A2, A1.2 depending 
on context. A typical article might have in the code of its introduction 
`The results are discussed in section\verb"~\ref{disc}".' and
the heading for the discussion section would be:
\small\begin{verbatim}
\section{Results}\label{disc}
\end{verbatim}\normalsize
Labels to sections, etc, may occur anywhere within that section except
within another numbered environment. 
Within a maths environment labels can be used to tag equations which are 
referred to within the text.

In addition to the standard \verb"\ref{<label>}", in \verb"iopart.cls" the abbreviated
forms given in \tref{abrefs}
are available for reference to standard parts of the text.

\Table{\label{abrefs}Alternatives to the normal references command {\tt $\backslash$ref} available in {\tt iopart.cls},
and the text generated by
them. Note it is not normally necessary to include the word equation
before an equation number except where the number starts a sentence. The
versions producing an initial capital should only be used at the start of
sentences.} 
\br
Reference&Text produced\\
\mr
\verb"\eref{<label>}"&(\verb"<num>")\\
\verb"\Eref{<label>}"&Equation (\verb"<num>")\\
\verb"\fref{<label>}"&figure \verb"<num>"\\
\verb"\Fref{<label>}"&Figure \verb"<num>"\\
\verb"\sref{<label>}"&section \verb"<num>"\\
\verb"\Sref{<label>}"&Section \verb"<num>"\\
\verb"\tref{<label>}"&table \verb"<num>"\\
\verb"\Tref{<label>}"&Table \verb"<num>"\\
\br
\endTable

\section{Tables and table captions}
Tables are numbered serially and referred to in the text 
by number (table 1, etc, {\bf not} tab. 1). Each table should have an 
explanatory caption which should be as concise as possible. If a table 
is divided into parts these should be labelled \pt(a), \pt(b), 
\pt(c), etc but there should be only one caption for the whole 
table, not separate ones for each part.

In the preprint style the tables may be included in the text 
or listed separately after the reference list 
starting on a new page.

\subsection{The basic table format}
The standard form for a table in \verb"iopart.cls" is:
\small\begin{verbatim}
\begin{table}
\caption{\label{label}Table caption.}
\begin{indented}
\item[]\begin{tabular}{@{}llll}
\br
Head 1&Head 2&Head 3&Head 4\\
\mr
1.1&1.2&1.3&1.4\\
2.1&2.2&2.3&2.4\\
\br
\end{tabular}
\end{indented}
\end{table}
\end{verbatim}\normalsize

\noindent Points to note are:
\begin{enumerate}
\item The caption comes before the table. It should have a period at
the end.

\item Tables are normally set in a smaller type than the text.
The normal style is for tables to be indented. This is accomplished
by using \verb"\begin{indented}" \dots\ \verb"\end{indented}"
and putting \verb"\item[]" before the start of the tabular environment.
Omit these
commands for any tables which will not fit on the page when indented.

\item The default is for columns to be aligned left and 
adding \verb"@{}" omits the extra space before the first column.

\item Tables have only horizontal rules and no vertical ones. The rules at
the top and bottom are thicker than internal rules and are set with
\verb"\br" (bold rule). 
The rule separating the headings from the entries is set with
\verb"\mr" (medium rule).  These are special \verb"iopart.cls" commands.

\item Numbers in columns should be aligned on the decimal point;
to help do this a control sequence \verb"\lineup" has been defined
in \verb"iopart.cls"
which sets \verb"\0" equal to a space the size of a digit, \verb"\m"
to be a space the width of a minus sign, and \verb"\-" to be a left
overlapping minus sign. \verb"\-" is for use in text mode while the other
two commands may be used in maths or text.
(\verb"\lineup" should only be used within a table
environment after the caption so that \verb"\-" has its normal meaning
elsewhere.) See table~\ref{tabone} for an example of a table where
\verb"\lineup" has been used.
\end{enumerate}

\begin{table}
\caption{\label{tabone}A simple example produced using the standard table commands 
and $\backslash${\tt lineup} to assist in aligning columns on the 
decimal point. The width of the 
table and rules is set automatically by the 
preamble.}

\begin{indented}
\lineup
\item[]\begin{tabular}{@{}*{7}{l}}
\br                              
$\0\0A$&$B$&$C$&\m$D$&\m$E$&$F$&$\0G$\cr 
\mr
\0\023.5&60  &0.53&$-20.2$&$-0.22$ &\01.7&\014.5\cr
\0\039.7&\-60&0.74&$-51.9$&$-0.208$&47.2 &146\cr 
\0123.7 &\00 &0.75&$-57.2$&\m---   &---  &---\cr 
3241.56 &60  &0.60&$-48.1$&$-0.29$ &41   &\015\cr 
\br
\end{tabular}
\end{indented}
\end{table}

\subsection{Simplified coding and extra features for tables}
The basic coding format can be simplified using extra commands provided in
the \verb"iopart" class file. The commands up to and including 
the start of the tabular environment
can be replaced by
\small\begin{verbatim}
\Table{\label{label}Table caption}
\end{verbatim}\normalsize
and this also activates the definitions within \verb"\lineup".
The final three lines can also be reduced to \verb"\endTable" or
\verb"\endtab". Similarly for a table which does not fit on the page when indented
\verb"\fulltable{\label{label}caption}" \dots\ \verb"\endfulltable"
can be used. \LaTeX\ optional positional parameters can, if desired, be added after 
\verb"\Table{\label{label}caption}" and \verb"\fulltable{\label{label}caption}".

\verb"\centre{#1}{#2}" can be used to centre a heading 
\verb"#2" over \verb"#1" 
columns and \verb"\crule{#1}" puts a rule across 
\verb"#1" columns. A negative 
space \verb"\ns" is usually useful to reduce the space between a centred 
heading and a centred rule. \verb"\ns" should occur immediately after the 
\verb"\\" of the row containing the centred heading (see code for
\tref{tabl3}). A small space can be 
inserted between rows of the table 
with \verb"\ms" and a half line space with \verb"\bs" 
(both must follow a \verb"\\" but should not have a 
\verb"\\" following them).

\Table{\label{tabl3}A table with headings spanning two columns and containing notes. 
To improve the 
visual effect a negative skip ($\backslash${\tt ns})
has been put in between the lines of the 
headings. Commands set-up by $\backslash${\tt lineup} are used to aid 
alignment in columns. $\backslash${\tt lineup} is defined within
the $\backslash${\tt Table} definition.}
\br
&&&\centre{2}{Separation energies}\\
\ns
&Thickness&&\crule{2}\\
Nucleus&(mg\,cm$^{-2}$)&Composition&$\gamma$, n (MeV)&$\gamma$, 2n (MeV)\\
\mr
$^{181}$Ta&$19.3\0\pm 0.1^{\rm a}$&Natural&7.6&14.2\\
$^{208}$Pb&$\03.8\0\pm 0.8^{\rm b}$&99\%\ enriched&7.4&14.1\\
$^{209}$Bi&$\02.86\pm 0.01^{\rm b}$&Natural&7.5&14.4\\
\br
\end{tabular}
\item[] $^{\rm a}$ Self-supporting.
\item[] $^{\rm b}$ Deposited over Al backing.
\end{indented}
\end{table}

Units should not normally be given within the body of a table but 
given in brackets following the column heading; however, they can be 
included in the caption for long column headings or complicated units. 
Where possible tables should not be broken over pages. 
If a table has related notes these should appear directly below the table
rather than at the bottom of the page. Notes can be designated with
footnote symbols (preferable when there are only a few notes) or
superscripted small roman letters. The notes are set to the same width as
the table and in normal tables follow after \verb"\end{tabular}", each
note preceded by \verb"\item[]". For a full width table \verb"\noindent"
should precede the note rather than \verb"\item[]". To simplify the coding 
\verb"\tabnotes" can, if desired, replace \verb"\end{tabular}" and 
\verb"\endtabnotes" replaces
\verb"\end{indented}\end{table}".

If all the tables are grouped at the end of a document
the command \verb"\Tables" is used to start a new page and 
set a heading `Tables and table captions'. If the tables follow an appendix then add the command \verb"\noappendix" to revert to normal style numbering.

\section{Figures and figure captions}

Figures (with their captions) can be incorporated into the text at the appropriate position or grouped together
at the end of the article. If the figures are at the end of the article and follow an appendix then in \verb"iopart.cls" you can add the command \verb"\noappendix" to revert to normal style numbering.  We remind you that you must seek permission
to reuse any previously-published figures, and acknowledge their use correctly---see section \ref{copyright}.

\subsection{Inclusion of graphics files\label{figinc}}
Using the \verb"graphicx" package graphics files can 
be included within figure and center environments at an 
appropriate point within the text using code such as:
\small\begin{verbatim}
\includegraphics{file.eps}
\end{verbatim}\normalsize
The \verb"graphicx" package supports various optional arguments
to control the appearance of the figure. Other similar 
packages can also be used (e.g. \verb"graphics", \verb"epsf").   Whatever package you use,
you must include it explicitly after the \verb"\documentclass" declaration using (say)
\verb"\usepackage{graphicx}".

For more detail about graphics inclusion see the documentation 
of the \verb"graphicx" package, refer to one of the books on \LaTeX , e.g. Goosens M, Rahtz S and Mittelbach F 1997 {\it The }\LaTeX\ {\it Graphics Companion\/} 
(Reading, MA: Addison-Wesley),
or download some of the excellent free documentation available via the Comprehensive
TeX Archive Network (CTAN) \verb"http://www.ctan.org"---{in particular see Reckdahl K 2006 {\it Using Imported Graphics in }\LaTeX\ {\it and }pdf\LaTeX\ \verb"http://www.ctan.org/tex-archive/info/epslatex".

IOP Publishing's graphics guidelines provide further information on preparing \verb".eps" files.

We prefer you to use \verb".eps" files for your graphics, but we realise that converting other
formats of graphics to \verb".eps" format can be troublesome.  If you use PDF or bitmap-format graphics
such as JPG or PNG that need to be included using the pdf\TeX\ package, this is OK, but please bear
in mind that the PDF you submit should use PDF standard 1.4 or lower (use \verb"\pdfminorversion=4" at the
start of the file).

The main \LaTeX\ file must read in graphics files and subsidiary \LaTeX\ files from the current directory,
{\it not} from a subdirectory.  Your submission files are stored on our systems in a single location and we will not be able to process your
TeX file automatically if it relies on organization of the files into subdirectories.

\subsection{Captions}
Below each figure should be a brief caption describing it and, if 
necessary, interpreting the various lines and symbols on the figure. 
As much lettering as possible should be removed from the figure itself 
and included in the caption. If a figure has parts, these should be 
labelled ($a$), ($b$), ($c$), etc and all parts should be described 
within a single caption. \Tref{blobs} gives the definitions for describing 
symbols and lines often used within figure captions (more symbols are 
available when using the optional packages loading the AMS extension fonts).

\subsection{Supplementary Data}
All of our journals encourage authors to submit supplementary data attachments to 
enhance the online versions of published research articles. Supplementary data 
enhancements typically consist of video clips, animations or
data files, tables of extra information or extra figures. They can 
add to the reader's understanding and present results in attractive ways that go 
beyond what can be presented in the PDF version of the article. 
See our supplementary data guidelines for further details.

Software, in the form of input scripts for mathematical packages (such as Mathematica notebook files), or
source code that can be interpreted or compiled (such as Python scripts or Fortran or C programs), or executable
files, can sometimes be accepted as supplementary data, but the journal may ask you for assurances about
the software and distribute them from the article web page only subject to a disclaimer.  Contact the journal
in the first instance if you want to submit software.

\begin{table}[t]
\caption{\label{blobs}Control sequences to describe lines and symbols in figure 
captions.}
\begin{indented}
\item[]\begin{tabular}{@{}lllll}
\br
Control sequence&Output&&Control sequence&Output\\
\mr
\verb"\dotted"&\dotted        &&\verb"\opencircle"&\opencircle\\
\verb"\dashed"&\dashed        &&\verb"\opentriangle"&\opentriangle\\
\verb"\broken"&\broken&&\verb"\opentriangledown"&\opentriangledown\\
\verb"\longbroken"&\longbroken&&\verb"\fullsquare"&\fullsquare\\
\verb"\chain"&\chain          &&\verb"\opensquare"&\opensquare\\
\verb"\dashddot"&\dashddot    &&\verb"\fullcircle"&\fullcircle\\
\verb"\full"&\full            &&\verb"\opendiamond"&\opendiamond\\
\br
\end{tabular}
\end{indented}
\end{table}

\clearpage

\appendix

\section{List of macros for formatting text, figures and tables}

\begin{table}[hb]
\caption{Macros available for use in text in {\tt iopart.cls}. Parameters in square brackets are optional.}
\footnotesize\rm
\begin{tabular}{@{}*{7}{l}}
\br
Macro name&Purpose\\
\mr
\verb"\title[#1]{#2}"&Title of article and short title (optional)\\
\verb"\paper[#1]{#2}"&Title of paper and short title (optional)\\
\verb"\letter{#1}"&Title of Letter to the Editor\\
\verb"\ftc{#1}"&Title of Fast Track Communication\\
\verb"\rapid[#1]{#2}"&Title of Rapid Communication and short title (optional)\\
\verb"\comment[#1]{#2}"&Title of Comment and short title (optional)\\
\verb"\topical[#1]{#2}"&Title of Topical Review and short title 
(optional)\\
\verb"\review[#1]{#2}"&Title of review article and short title (optional)\\
\verb"\note[#1]{#2}"&Title of Note and short title (optional)\\
\verb"\prelim[#1]{#2}"&Title of Preliminary Communication \& short title\\
\verb"\author{#1}"&List of all authors\\
\verb"\article[#1]{#2}{#3}"&Type and title of other articles and 
short title (optional)\\
\verb"\address{#1}"&Address of author\\
\verb"\ams{#1}"&Mathematics Classification Scheme\\
\verb"\submitto{#1}"&`Submitted to' message\\
\verb"\maketitle"&Creates title page\\
\verb"\begin{abstract}"&Start of abstract\\
\verb"\end{abstract}"&End of abstract\\
\verb"\nosections"&Inserts space before text when no sections\\
\verb"\section{#1}"&Section heading\\
\verb"\subsection{#1}"&Subsection heading\\
\verb"\subsubsection{#1}"&Subsubsection heading\\
\verb"\appendix"&Start of appendixes\\
\verb"\ack"&Acknowledgments heading\\
\verb"\References"&Heading for reference list\\
\verb"\begin{harvard}"&Start of alphabetic reference list\\
\verb"\end{harvard}"&End of alphabetic reference list\\
\verb"\begin{thebibliography}{#1}"&Start of numeric reference list\\
\verb"\end{thebibliography}"&End of numeric reference list\\
\verb"\etal"&\etal for text and reference lists\\
\verb"\nonum"&Unnumbered entry in numerical reference list\\
\br
\end{tabular}
\end{table}

\clearpage

\begin{table}
\caption{Macros defined within {\tt iopart.cls}
for use with figures and tables.}
\begin{indented}
\item[]\begin{tabular}{@{}l*{15}{l}}
\br
Macro name&Purpose\\
\mr
\verb"\Figures"&Heading for list of figure captions\\
\verb"\Figure{#1}"&Figure caption\\
\verb"\Tables"&Heading for tables and table captions\\
\verb"\Table{#1}"&Table caption\\
\verb"\fulltable{#1}"&Table caption for full width table\\
\verb"\endTable"&End of table created with \verb"\Table"\\
\verb"\endfulltab"&End of table created with \verb"\fulltable"\\
\verb"\endtab"&End of table\\
\verb"\br"&Bold rule for tables\\
\verb"\mr"&Medium rule for tables\\
\verb"\ns"&Small negative space for use in table\\
\verb"\centre{#1}{#2}"&Centre heading over columns\\
\verb"\crule{#1}"&Centre rule over columns\\
\verb"\lineup"&Set macros for alignment in columns\\
\verb"\m"&Space equal to width of minus sign\\
\verb"\-"&Left overhanging minus sign\\
\verb"\0"&Space equal to width of a digit\\
\br
\end{tabular}
\end{indented}
\end{table}

\clearpage

\begin{table}[hb]
\caption{\label{jlab2}Abbreviations in {\tt iopart.cls} for journals handled by IOP Publishing.}
\footnotesize
\begin{tabular}{@{}llll}
\br
{\rm Short form of journal title} & Macro & {\rm Short form of journal title} & Macro \\
\mr
2D Mater.&\verb"\TDM"&J. Radiol. Prot.&\verb"\JRP"\\
AJ&\verb"\AJ"&J. Semicond.&\verb"\JOS"\\
ApJ&\verb"\APJ"&J. Stat. Mech.&\verb"\JSTAT"\\
ApJL&\verb"\APJL"&Laser Phys.&\verb"\LP"\\
ApJS&\verb"\APJS"&Laser Phys. Lett.&\verb"\LPL"\\
Adv. Nat. Sci: Nanosci. Nanotechnol.&\verb"\ANSN"&Metrologia&\verb"\MET"\\
Appl. Phys. Express&\verb"\APEX"&Mater. Res. Express&\verb"\MRE"\\
Biofabrication&\verb"\BF"&Meas. Sci. Technol.&\verb"\MST"\\
Bioinspir. Biomim.&\verb"\BB"&Methods Appl. Fluoresc.&\verb"\MAF"\\
Biomed. Mater.&\verb"\BMM"&Modelling Simul. Mater. Sci. Eng.&\verb"\MSMSE"\\
Chin. J. Chem. Phys.&\verb"\CJCP"&Nucl. Fusion&\verb"\NF"\\
Chinese Phys. B&\verb"\CPB"&New J. Phys.&\verb"\NJP"\\
Chinese Phys. C&\verb"\CPC"&Nonlinearity&\verb"\NL"\\
Chinese Phys. Lett.&\verb"\CPL"&Nanotechnology&\verb"\NT"\\
Class. Quantum Grav.&\verb"\CQG"&Phys. Biol.&\verb"\PB"\\
Commun. Theor. Phys.&\verb"\CTP"&Phys. Educ.&\verb"\PED"\\
Comput. Sci. Disc.&\verb"\CSD"&Phys.-Usp.&\verb"\PHU"\\
Environ. Res. Lett.&\verb"\ERL"&Physiol. Meas.&\verb"\PM"\\
EPL&\verb"\EPL"&Phys. Med. Biol.&\verb"\PMB"\\
Eur. J. Phys.&\verb"\EJP"&Phys. Scr.&\verb"\PS"\\
Fluid Dyn. Res.&\verb"\FDR"&Plasma Phys. Control. Fusion&\verb"\PPCF"\\
Inverse Problems&\verb"\IP"&Plasma Sci. Technol.&\verb"\PST"\\
Izv. Math.&\verb"\IZV"&Plasma Sources Sci. Technol.&\verb"\PSST"\\
Jpn. J. Appl. Phys.&\verb"\JJAP"&Quantum Electron.&\verb"\QEL"\\
J. Breath Res.&\verb"\JBR"&Rep. Prog. Phys.&\verb"\RPP"\\
JCAP&\verb"\JCAP"&Res. Astron. Astrophys.&\verb"\RAA"\\
J. Geophys. Eng.&\verb"\JGE"&Russ. Chem. Rev.&\verb"\RCR"\\
JINST&\verb"\JINST"&Russ. Math. Surv.&\verb"\RMS"\\
J. Micromech. Microeng.&\verb"\JMM"&Sb. Math.&\verb"\MSB"\\
J. Neural Eng.&\verb"\JNE"&Science Foundation in China&\verb"\SFC"\\
J. Opt.&\verb"\JOPT"&Sci. Technol. Adv. Mater.&\verb"\STAM"\\
J. Phys. A: Math. Theor.&\verb"\jpa"&Semicond. Sci. Technol.&\verb"\SST"\\
J. Phys. B: At. Mol. Opt. Phys.&\verb"\jpb"&Smart Mater. Struct.&\verb"\SMS"\\
J. Phys: Condens. Matter&\verb"\JPCM"&Supercond. Sci. Technol.&\verb"\SUST"\\
J. Phys. D: Appl. Phys.&\verb"\JPD"&Surf. Topogr.: Metrol. Prop.&\verb"\STMP"\\
J. Phys. G: Nucl. Part. Phys.&\verb"\jpg"&Transl. Mater. Res.&\verb"\TMR"\\
\mr
{\it IOP Conference Series journals}\\
\mr
J. Phys.: Conf. Ser.&\verb"\JPCS"\\
IOP Conf. Ser.: Earth Environ. Sci.&\verb"\EES"\\
IOP Conf. Ser.: Mater. Sci. Eng.&\verb"\MSE"\\

\br
\end{tabular}

\end{table}

\clearpage

\begin{table}

\caption{\label{jlab2b}Abbreviations for IOP Publishing journals that are no longer published.}
\begin{indented}
\item[]\begin{tabular}{@{}lll}
\br
{\rm Short form of journal title} & Macro name & Years relevant\\
\mr
J. Phys. A: Math. Gen.&\verb"\JPA"&1975--2006\\
J. Phys. B: At. Mol. Phys.&\verb"\JPB"&1968--1987\\
J. Phys. C: Solid State Phys.&\verb"\JPC"&1968--1988\\
J. Phys. E: Sci. Instrum.&\verb"\JPE"&1968--1989\\
J. Phys. F: Met. Phys.&\verb"\JPF"&1971--1988\\
J. Phys. G: Nucl. Phys.&\verb"\JPG"&1975--1988\\
Pure Appl. Opt.&\verb"\PAO"&1992--1998\\
Quantum Opt.&\verb"\QO"&1989--1994\\
Quantum Semiclass. Opt.&\verb"\QSO"&1995--1998\\
J. Opt. A: Pure Appl. Opt.&\verb"\JOA"&1999--2009\\
J. Opt. B: Quantum Semiclass. Opt.&\verb"\JOB"&1999--2005\\
\br
\end{tabular}
\end{indented}
\end{table}

\begin{table}[h]

\caption{\label{jlab3}Abbreviations in {\tt iopart.cls} for some 
common journals not handled by IOP Publishing.}
\begin{indented}
\item[]\begin{tabular}{@{}llll}
\br
Short form of journal & Macro & Short form of Journal & Macro\\
\mr
Acta Crystallogr.&\verb"\AC"&J. Quant. Spectrosc. Radiat. Transfer&\verb"\JQSRT"\\
Acta Metall.&\verb"\AM"&Nuovo Cimento&\verb"\NC"\\
Ann. Phys., Lpz&\verb"\AP"&Nucl. Instrum. Methods&\verb"\NIM"\\
Ann. Phys., NY&\verb"\APNY"&Nucl. Phys.&\verb"\NP"\\
Ann. Phys., Paris&\verb"\APP"&Phys. Fluids&\verb"\PF"\\
Can. J. Phys.&\verb"\CJP"&Phys. Lett.&\verb"\PL"\\
Gen. Rel. Grav.&\verb"\GRG"&Phys. Rev.&\verb"\PR"\\
J. Appl. Phys.&\verb"\JAP"&Phys. Rev. Lett.&\verb"\PRL"\\
J. Chem. Phys.&\verb"\JCP"&Proc. R. Soc.&\verb"\PRS"\\
J. High Energy Phys.&\verb"\JHEP"&Phys. Status Solidi&\verb"\PSS"\\
J. Magn. Magn. Mater.&\verb"\JMMM"&Phil. Trans. R. Soc.&\verb"\PTRS"\\
J. Math. Phys.&\verb"\JMP"&Rev. Mod. Phys.&\verb"\RMP"\\
J. Opt. Soc. Am.&\verb"\JOSA"&Rev. Sci. Instrum.&\verb"\RSI"\\
J. Physique&\verb"\JP"&Solid State Commun.&\verb"\SSC"\\
J. Phys. Chem.&\verb"\JPhCh"&Sov. Phys.--JETP&\verb"\SPJ"\\
J. Phys. Soc. Jpn.&\verb"\JPSJ"&Z. Phys.&\verb"\ZP"\\

\br
\end{tabular}
\end{indented}
\end{table}

\clearpage

\section{Including author names using Chinese, Japanese and Korean characters in submissions to IOP Publishing journals}
Authors in all IOP Publishing journals have the option to include names in Chinese, Japanese or Korean (CJK) characters in addition to the English name. The names will be displayed in the print issue and the online PDF, abstract and table of contents, in parentheses after the English name.

It is the decision of the individual authors whether or not to include a CJK version of their names; for a single article it is not necessary for all authors to include a CJK name if only one author wishes to do so. It is the responsibility of the authors to check the accuracy and formatting of the names in the final proofs that they receive prior to publication.

To include names in CJK characters, authors should use the \verb"cjk.sty" package, available from \verb"http://www.ctan.org/tex-archive/language/chinese/CJK/". Users should be aware that this is a very large and complicated package which relies on a large number of fonts.  We recommend using a TeX package that includes this package and all of the fonts by default, so that manual configuration is not required (e.g. the TeXLive distribution, which is available on all platforms (Macintosh, Windows and Linux)).

The documentation for the \verb"cjk.sty" package gives information on how CJK characters can be included in TeX files.  Most authors will find it convenient to include the characters in one of the standard encodings such as UTF-8, GB or JIS, if they have access to a text editor that supports such encodings.

Example TeX coding might be:
\begin{verbatim}
\documentclass[12pt]{iopart}
\usepackage{CJK}
.
.
.
\begin{document}
\begin{CJK*}{GBK}{ }

\title[]{Title of article}
\author{Author Name (CJK characters)}
\address{Department, University, City, Country}
.
.
.
\end{CJK*}
\end{verbatim}

To avoid potential problems in handling the CJK characters in submissions, authors should always include a PDF of the full version of their papers (including all figure files, tables, references etc) with the CJK characters in it.

\end{document}

### 2023-08-21 paper.tex
update

@@
 
-Uncertain fluctuations, or `noise,' is a common theme throughout many fields of engineering, and is a frequent concern when attempting to control any human-made system such as vehicles \brian[\cite{XXX}], production machinery \brian[\cite{XXX}], or chemical processes \brian[\cite{XXX}].
+Uncertain fluctuations, or `noise,' is a common theme throughout many fields of engineering, and robust control is a frequent concern when attempting to control any human-made system such as vehicles \michael[\cite{Liang2021}], chemical processes \michael[\cite{Lucia2013}], or biology \michael[\cite{Baetica2020}].
 \brian[Please look for some relevant references or a review of control under noise. I highlight several missing refs in red.]
@@
 Rather, statistical models are usually employed where key mechanisms or dynamics are subject to stochastic variations or inputs that act as a proxy for random or un-modeled fluctuations elsewhere in the system.
-In this context, much effort has been placed on enhancing the robustness of controlled processes, where uncertain or unpredictable variations in internal or external parameters are modeled as intrinsic or extrinsic noise \brian[\cite{XXX}].
-For many macroscopic, human-engineered systems, the resulting system dynamics can be modeled effectively simply by combining a deterministic (i.e., noise-free) model with additive noise (usually assumed to be Gaussian under arguments based on the central limit theorem) \brian[\cite{XXX}], and most current approaches seek to control the system to minimize variations that result from the noisy inputs.
+In this context, much effort has been placed on enhancing the robustness of controlled processes, where uncertain or unpredictable variations in internal or external parameters are modeled as intrinsic or extrinsic noise \brian[cite XXX].
+For many macroscopic, human-engineered systems, the resulting system dynamics can be modeled effectively simply by combining a deterministic (i.e., noise-free) model with additive noise (usually assumed to be Gaussian under arguments based on the central limit theorem) \brian[\cite{Feng2022}], and most current approaches seek to control the system to minimize variations that result from the noisy inputs.
 
-Analyses of noise and control theory are equally relevant to understand the basic biological processes of gene transcription regulation \brian[\cite{XXX}] or mRNA translation regulation \brian[\cite{XXX}]or to modify these processes for practical use \brian[\cite{XXX}].
+Analyses of noise and control theory are equally relevant to understand the basic biological processes of gene transcription regulation \michael[\cite{Sun2020}] or mRNA translation regulation \brian[\cite{Cialek2022}]or to modify these processes for practical use \brian[\cite{Dionisi2022,Baumschlager2017}].
 However, at this mesoscopic scale of cellular biology, where transcription factors compete to activate or deactivate individual genes or where mRNA or protein molecules are present at just a few copies (or none at all) per cell, the additive noise model is much less realistic.
@@
 These digital-synthetic actuators enable fine-tuned, computer-modulated control of cellular systems, previously unattainable, with faster response times compared to chemical diffusion\cite{Rullan2018, Baumschlager2017} .
-Classical and modern control methods like PID control and model predictive control have been implemented in such systems \brian[\cite{XXX}] to control synthetic systems to different stable points.
+Classical and modern control methods like PID control and model predictive control have been implemented in such systems \brian[\cite{Filo2022}] to control synthetic systems to different stable points.
 
-While most previous control efforts focussed on deterministic ordinary differential equation (ODE) formulations of the control problem, a few pioneering theoretical works have extended these analysis to include noise and have used stochastic simulations to show how systems can be guided towards unstable points by oscillating around those points \cite{Guarino2020}.
 It has recently been shown that new control techniques that leverage the complete probability distribution information of the system could actually harness the noise of single-cell gene regulation to achieve more complicated control objectives.
@@
 In May et al.~\cite{May2021}, we identified a simplified stochastic model to reproduce data measured in Baumschlager, et al.
-\brian[\cite{XXX}] for the expression from a transcription promoter under the optogenetic control of a UV-activated T7 polymerase (see model in Figure \ref{cartoons}A, top promoter).
+\brian[\cite{Baumschlager2017}] for the expression from a transcription promoter under the optogenetic control of a UV-activated T7 polymerase (see model in Figure \ref{cartoons}A, top promoter).
 We then proposed the addition of a positive auto-regulation (Figure \ref{cartoons}A, bottom promoter) to help maintain an elevated expression phenotype in the presence of UV excitation, and we demonstrated how a Single-Input-Multiple-Output (SIMO) multicellular controller could control multiple cells to arbitrary phenotypes using only a single input.
@@
 The second, much simpler, single-species model was developed by assuming quasi-steady equilibrium for the first five species. 
-Both models were independently parameterized using the same experimental data from Baumschlager et al. \brian[\cite{XXX}]. 
+Both models were independently parameterized using the same experimental data from Baumschlager et al. \brian[\cite{Baumschlager2017}]. 
 Furthermore, an extension was made to each model to incorporate a secondary self-activated promoter-gene construct, where the expression rate was determined by a Hill function \ref{prodRate}.
@@
 In realistic applications, models are never exact, but are often chosen as simplifications of known processes. 
-For example, when analyzing discrete stochastic chemical kinetics, it is common to project the CME onto lower-dimensional spaces using finite state projection \brian[\cite{XXX}], time scale separations \brian[\cite{Peles2006}], Krylov subspaces \brian[\cite{Macnemara&Sidje?}], principle orthogonal decompositions \brian[\cite{Vo2019}], or other coarse meshes \brian[\cite{Munsky2008IEEE,Tapia2012CDC}].
+For example, when analyzing discrete stochastic chemical kinetics, it is common to project the CME onto lower-dimensional spaces using finite state projection \brian[\cite{Munsky2006}], time scale separations \brian[\cite{Peles2006}], Krylov subspaces \brian[\cite{Macnemara2008}], principle orthogonal decompositions \brian[\cite{Vo2019}], or other coarse meshes \brian[\cite{Munsky2008IEEE,Tapia2012CDC}].
 Similarly, measurements are also always inexact and in many cases may only provide information about relative changes -- for example, although fluorescent proteins usually cannot be counted exactly, one may reasonably assume that a cell's total fluorescence intensity varies linearly with the fluorescent protein concentration.
@@
 \url{https://journals.plos.org/ploscompbiol/article?id=10.1371/journal.pcbi.1004923}].
-
+\michael[added Voliotis2016 to the section after this comment]
+The Extrande method \cite{Voliotis2016} was used to update the control input at an average frequency of 50 updates per minute, far exceeding the dynamics of the system.
 \subsection{Tracking Time-Varying Trajectories}\label{sec:Dynamic}
@@
 In other words, the process becomes more predictable and therefore more controllable.
-At the extreme as the system size increases, the dynamics converge towards a deterministic process, except for certain exceptional initial conditions lying on manifolds that would separate different steady state behaviors \brian[\cite{XXX}].
+At the extreme as the system size increases, the dynamics converge towards a deterministic process, except for certain exceptional initial conditions lying on manifolds that would separate different steady state behaviors \brian[\cite{Strasser2012}].
 
@@
 Dashed yellow line shows characteristic time $\tau_c = \alpha/\gamma/$.
-\brian[Label 'J' is missing from the figure. Replace XXX below.]
-(K) Controlled system trajectory for $\alpha = XXX$ and $\tau = $ min (denoted by red star in panel J).}
+\brian[Label 'J' is missing from the figure. Replace XXX below.] \michael[added J. Replace XXX with 4.0]
+(K) Controlled system trajectory for $\alpha = 4.0$ and $\tau = $ min (denoted by red star in panel J).}
 \label{DG}
@@
 
-\begin{thebibliography}{10}
-\providecommand{\url}[1]{#1}
-\csname url@samestyle\endcsname
-\providecommand{\newblock}{\relax}
-\providecommand{\bibinfo}[2]{#2}
-\providecommand{\BIBentrySTDinterwordspacing}{\spaceskip=0pt\relax}
-\providecommand{\BIBentryALTinterwordstretchfactor}{4}
-\providecommand{\BIBentryALTinterwordspacing}{\spaceskip=\fontdimen2\font plus
-\BIBentryALTinterwordstretchfactor\fontdimen3\font minus
-  \fontdimen4\font\relax}
-\providecommand{\BIBforeignlanguage}[2]{{%
-\expandafter\ifx\csname l@#1\endcsname\relax
-\typeout{** WARNING: IEEEtran.bst: No hyphenation pattern has been}%
-\typeout{** loaded for the language `#1'. Using the pattern for}%
-\typeout{** the default language instead.}%
-\else
-\language=\csname l@#1\endcsname
-\fi
-#2}}
-\providecommand{\BIBdecl}{\relax}
-\BIBdecl
-
-\bibitem{May2021}
-M.~May and B.~Munsky, ``{Exploiting noise, nonlinearity, and feedback to
-  differentially control multiple synthetic cells with a single optogenetic
-  input},'' pp. 1--28, 2021.
-eps
-\bibitem{Ng2019}
-A.~H. Ng, T.~H. Nguyen, M.~G{\'{o}}mez-Schiavon, G.~Dods, R.~A. Langan, S.~E.
-  Boyken, J.~A. Samson, L.~M. Waldburger, J.~E. Dueber, D.~Baker, and
-  H.~El-Samad, ``{Modular and tunable biological feedback control using a de
-  novo protein switch},'' pp. 265--269, 2019.
-
-\bibitem{Liu2018}
-C.~C. Liu, M.~C. Jewett, J.~W. Chin, and C.~A. Voigt, ``{Toward an orthogonal
-  central dogma},'' \emph{Nature Chemical Biology}, vol.~14, no.~2, pp.
-  103--106, 2018.
-
-\bibitem{Sheets2020}
-M.~B. Sheets, W.~W. Wong, and M.~J. Dunlop, ``{Light-Inducible Recombinases for
-  Bacterial Optogenetics},'' \emph{ACS Synthetic Biology}, vol.~9, no.~2, pp.
-  227--235, 2020.
-
-\bibitem{Groseclose2020}
-\BIBentryALTinterwordspacing
-T.~M. Groseclose, R.~E. Rondon, Z.~D. Herde, C.~A. Aldrete, and C.~J. Wilson,
-  ``{Engineered systems of inducible anti-repressors for the next generation of
-  biological programming},'' \emph{Nature Communications}, vol.~11, no.~1, pp.
-  1--15, 2020. [Online]. Available:
-  \url{http://dx.doi.org/10.1038/s41467-020-18302-1}
-\BIBentrySTDinterwordspacing
-
-\bibitem{Shin2020}
-J.~Shin, S.~Zhang, B.~S. Der, A.~A. Nielsen, and C.~A. Voigt, ``{ Programming
-  Escherichia coli to function as a digital display },'' \emph{Molecular
-  Systems Biology}, vol.~16, no.~3, pp. 1--12, 2020.
-
-\bibitem{Baumschlager2017}
-A.~Baumschlager, S.~K. Aoki, and M.~Khammash, ``{Dynamic blue light-inducible
-  T7 RNA polymerases (Opto-T7RNAPs) for precise spatiotemporal gene expression
-  control},'' \emph{ACS synthetic biology}, vol.~6, no.~11, pp. 2157--2167,
-  2017.
-
-\bibitem{Chen2020}
-\BIBentryALTinterwordspacing
-S.~Y. Chen, L.~C. Osimiri, M.~Chevalier, L.~J. Bugaj, T.~H. Nguyen, R.~A.
-  Greenstein, A.~H. Ng, J.~Stewart-Ornstein, L.~T. Neves, and H.~El-Samad,
-  ``{Optogenetic Control Reveals Differential Promoter Interpretation of
-  Transcription Factor Nuclear Translocation Dynamics},'' \emph{Cell Systems},
-  vol.~11, no.~4, pp. 336----353.e24, 2020. [Online]. Available:
-  \url{http://dx.doi.org/10.1016/j.cels.2020.08.009}
-\BIBentrySTDinterwordspacing
-
-\bibitem{Lillacci2018}
-G.~Lillacci, Y.~Benenson, and M.~Khammash, ``{Synthetic control systems for
-  high performance gene expression in mammalian cells},'' \emph{Nucleic acids
-  research}, vol.~46, no.~18, pp. 9855--9863, 2018.
-
-\bibitem{Fox2021}
-\BIBentryALTinterwordspacing
-Z.~R. Fox, S.~Fletcher, A.~Fraisse, C.~Aditya, and S.~Sosa, ``{MicroMator: Open
-  and Flexible Software for Reactive Microscopy},'' \emph{bioRxiv}, pp. 1--9,
-  2021. [Online]. Available:
-  \url{http://biorxiv.org/cgi/content/short/2021.03.12.435206v1}
-\BIBentrySTDinterwordspacing
-
-\bibitem{Baumschlager2021}
-\BIBentryALTinterwordspacing
-A.~Baumschlager and M.~Khammash, ``{Synthetic Biological Approaches for
-  Optogenetics and Tools for Transcriptional Light-Control in Bacteria},''
-  \emph{Advanced Biology}, vol.~5, no.~5, p. 2000256, 2021. [Online].
-  Available:
-  \url{https://onlinelibrary.wiley.com/doi/abs/10.1002/adbi.202000256}
-\BIBentrySTDinterwordspacing
-
-\bibitem{Lugagne2017}
-\BIBentryALTinterwordspacing
-J.~B. Lugagne, S.~{Sosa Carrillo}, M.~Kirch, A.~K{\"{o}}hler, G.~Batt, and
-  P.~Hersen, ``{Balancing a genetic toggle switch by real-time feedback control
-  and periodic forcing},'' \emph{Nature Communications}, vol.~8, no.~1, pp.
-  1--7, 2017. [Online]. Available:
-  \url{http://dx.doi.org/10.1038/s41467-017-01498-0}
-\BIBentrySTDinterwordspacing
-
-\bibitem{Rullan2018}
-\BIBentryALTinterwordspacing
-M.~Rullan, D.~Benzinger, G.~W. Schmidt, A.~Milias-Argeitis, and M.~Khammash,
-  ``{An Optogenetic Platform for Real-Time, Single-Cell Interrogation of
-  Stochastic Transcriptional Regulation},'' \emph{Molecular Cell}, vol.~70,
-  no.~4, pp. 745--756.e6, 2018. [Online]. Available:
-  \url{https://doi.org/10.1016/j.molcel.2018.04.012}
-\BIBentrySTDinterwordspacing
-
-\bibitem{Guarino2020}
-A.~Guarino, D.~Fiore, D.~Salzano, and M.~{Di Bernardo}, ``{Balancing Cell
-  Populations Endowed with a Synthetic Toggle Switch via Adaptive Pulsatile
-  Feedback Control},'' \emph{ACS Synthetic Biology}, vol.~9, no.~4, pp.
-  793--803, 2020.
-
-\bibitem{Munsky2012}
-B.~Munsky, G.~Neuert, and A.~Van~Oudenaarden, ``{Using gene expression noise to
-  understand gene regulation},'' \emph{Science}, vol. 336, no. 6078, pp.
-  183--187, 2012.
-
-\bibitem{Szymanska2015}
-P.~Szyma{\'{n}}ska, N.~Gritti, J.~M. Keegstra, M.~Soltani, and B.~Munsky,
-  ``{Using noise to control heterogeneity of isogenic populations in homogenous
-  environments},'' \emph{Physical Biology}, vol.~12, no.~4, 2015.
-
-\bibitem{Gillespie1992}
-D.~T. Gillespie, ``{A rigorous derivation of the chemical master equation},''
-  \emph{Physica A: Statistical Mechanics and its Applications}, vol. 188, no.
-  1-3, pp. 404--425, 1992.
-
-\bibitem{Gillespie1977}
-D.~T. Gillespie, ``{Exact stochastic simulation of coupled chemical reactions},''
-  \emph{The journal of physical chemistry}, vol.~81, no.~25, pp. 2340--2361,
-  1977.
-
-\end{thebibliography}
-
+\section*{References}
+\bibliographystyle{plain}
+\bibliography{refs}
 \end{document}

### 2023-09-14 paper.tex
paper update

@@
 
-Uncertain fluctuations, or `noise,' is a common theme throughout many fields of engineering, and robust control is a frequent concern when attempting to control any human-made system such as vehicles \michael[\cite{Liang2021}], chemical processes \michael[\cite{Lucia2013}], or biology \michael[\cite{Baetica2020}].
+Uncertain fluctuations, or `noise,' is a common theme throughout many fields of engineering, and robust control is a frequent concern when attempting to control any human-made system such as vehicles \cite{Liang2021}, chemical processes \cite{Lucia2013}, or biology \cite{Baetica2020}.
 \brian[Please look for some relevant references or a review of control under noise. I highlight several missing refs in red.]
@@
 Rather, statistical models are usually employed where key mechanisms or dynamics are subject to stochastic variations or inputs that act as a proxy for random or un-modeled fluctuations elsewhere in the system.
-In this context, much effort has been placed on enhancing the robustness of controlled processes, where uncertain or unpredictable variations in internal or external parameters are modeled as intrinsic or extrinsic noise \brian[cite XXX].
-For many macroscopic, human-engineered systems, the resulting system dynamics can be modeled effectively simply by combining a deterministic (i.e., noise-free) model with additive noise (usually assumed to be Gaussian under arguments based on the central limit theorem) \brian[\cite{Feng2022}], and most current approaches seek to control the system to minimize variations that result from the noisy inputs.
+In this context, much effort has been placed on enhancing the robustness of controlled processes, where uncertain or unpredictable variations in internal or external parameters are modeled as intrinsic or extrinsic noise \cite{Bhattacharyya2017}.
+For many macroscopic, human-engineered systems, the resulting system dynamics can be modeled effectively simply by combining a deterministic (i.e., noise-free) model with additive noise (usually assumed to be Gaussian under arguments based on the central limit theorem) \cite{Feng2022}, and most current approaches seek to control the system to minimize variations that result from the noisy inputs.
 
-Analyses of noise and control theory are equally relevant to understand the basic biological processes of gene transcription regulation \michael[\cite{Sun2020}] or mRNA translation regulation \brian[\cite{Cialek2022}]or to modify these processes for practical use \brian[\cite{Dionisi2022,Baumschlager2017}].
+Analyses of noise and control theory are equally relevant to understand the basic biological processes of gene transcription regulation \cite{Sun2020}or mRNA translation regulation \cite{Cialek2022}or to modify these processes for practical use \cite{Dionisi2022,Baumschlager2017}.
 However, at this mesoscopic scale of cellular biology, where transcription factors compete to activate or deactivate individual genes or where mRNA or protein molecules are present at just a few copies (or none at all) per cell, the additive noise model is much less realistic.
@@
 These digital-synthetic actuators enable fine-tuned, computer-modulated control of cellular systems, previously unattainable, with faster response times compared to chemical diffusion\cite{Rullan2018, Baumschlager2017} .
-Classical and modern control methods like PID control and model predictive control have been implemented in such systems \brian[\cite{Filo2022}] to control synthetic systems to different stable points.
+Classical and modern control methods like PID control and model predictive control have been implemented in such systems \cite{Filo2022}to control synthetic systems to different stable points.
 
 It has recently been shown that new control techniques that leverage the complete probability distribution information of the system could actually harness the noise of single-cell gene regulation to achieve more complicated control objectives.
-For example, inspired by the genetic toggle switch from Kobayashi et al.\cite{Kobayashi:2004}, Szymanska et al.~\cite{Szymanska2015} showed that noise could be exploited to achieve independent control of multiple cells using a single input, even despite uncertain parameters or time delays due to maturation of fluorescent proteins or limited observation of the regulatory proteins.
+For example, inspired by the genetic toggle switch from Kobayashi et al.\cite{Kobayashi2004}, Szymanska et al.~\cite{Szymanska2015} showed that noise could be exploited to achieve independent control of multiple cells using a single input, even despite uncertain parameters or time delays due to maturation of fluorescent proteins or limited observation of the regulatory proteins.
 In May et al.~\cite{May2021}, we identified a simplified stochastic model to reproduce data measured in Baumschlager, et al.
-\brian[\cite{Baumschlager2017}] for the expression from a transcription promoter under the optogenetic control of a UV-activated T7 polymerase (see model in Figure \ref{cartoons}A, top promoter).
+\cite{Baumschlager2017} for the expression from a transcription promoter under the optogenetic control of a UV-activated T7 polymerase (see model in Figure \ref{cartoons}A, top promoter).
 We then proposed the addition of a positive auto-regulation (Figure \ref{cartoons}A, bottom promoter) to help maintain an elevated expression phenotype in the presence of UV excitation, and we demonstrated how a Single-Input-Multiple-Output (SIMO) multicellular controller could control multiple cells to arbitrary phenotypes using only a single input.
@@
 The second, much simpler, single-species model was developed by assuming quasi-steady equilibrium for the first five species. 
-Both models were independently parameterized using the same experimental data from Baumschlager et al. \brian[\cite{Baumschlager2017}]. 
+Both models were independently parameterized using the same experimental data from Baumschlager et al. \cite{Baumschlager2017}. 
 Furthermore, an extension was made to each model to incorporate a secondary self-activated promoter-gene construct, where the expression rate was determined by a Hill function \ref{prodRate}.
@@
 To address this issue, we first truncate the system at a finite number for each species and then apply a reflecting boundary condition, resulting in a finite dimensional master equation.
-To assess the time interval over which this truncation is valid, we also solve the Finite State Projection \brian[\cite{Munsky:2006}] for the same truncation, which allows us to compute an upper bound on the truncation error as a function of time.
+To assess the time interval over which this truncation is valid, we also solve the Finite State Projection \cite{Munsky2006} for the same truncation, which allows us to compute an upper bound on the truncation error as a function of time.
 
@@
 \subsection{Scaling for system granularity}\label{sec:Scaling}
-%Previous studies have addressed the computational challenge of analyzing discrete stochastic chemical kinetics by projecting the FSP onto lower-dimensional spaces using techniques such as time scale separations \cite{Peles2006}, Krylov subspaces \cite{Macnemara&Sidje?}, coarse meshes \cite{Munsky2008IEEE,Tapia2012CDC}, or principle orthogonal decompositions \cite{Vo2019}.
-%
-%However, using CME analyses is often hindered by an experimental limitations.
-Gene expression quantification based on fluorescence intensities, such as using fluorescent protein reporters, typically captures relative changes in expression over time rather than providing direct molecular counts.
-%% XXX - would be good to look for a few other examples outside of our own papers where the CME dimension is reduced by linear projections.
-%
-%It would be convenient to create a parameter that could increase or decrease the intrinsic noise of a system without changing the overall dynamics of the system.
-This would enable the analysis of models with more or less noise without the need of new unique models each time.
-A rescaling of the system could decrease the intrinsic noise while maintaining the same general behavior.
-%To accomplish this rescaling, let $M$ denote the size of the origional unscaled system and let $M'$ denote the size of the newly scaled system.
 
 In realistic applications, models are never exact, but are often chosen as simplifications of known processes. 
-For example, when analyzing discrete stochastic chemical kinetics, it is common to project the CME onto lower-dimensional spaces using finite state projection \brian[\cite{Munsky2006}], time scale separations \brian[\cite{Peles2006}], Krylov subspaces \brian[\cite{Macnemara2008}], principle orthogonal decompositions \brian[\cite{Vo2019}], or other coarse meshes \brian[\cite{Munsky2008IEEE,Tapia2012CDC}].
+For example, when analyzing discrete stochastic chemical kinetics, it is common to project the CME onto lower-dimensional spaces using finite state projection \cite{Munsky2006}, time scale separations \cite{Peles2006}, Krylov subspaces \cite{Macnamara2008}, principle orthogonal decompositions \cite{Vo2019}, or other coarse meshes \cite{Munsky2008IEEE,Tapia2012}.
 Similarly, measurements are also always inexact and in many cases may only provide information about relative changes -- for example, although fluorescent proteins usually cannot be counted exactly, one may reasonably assume that a cell's total fluorescence intensity varies linearly with the fluorescent protein concentration.
@@
 However, at large granularity, we find that the opposite is true - the control performance actually improves when applied to an incorrect model.
-The reason for this surprising result is that relative amplitude of stochastic fluctuations (i.e., the standard deviation divided by the mean) in a chemical process decreases with the inverse square root of the process scale \brian[\cite{VanKampen}].
+The reason for this surprising result is that relative amplitude of stochastic fluctuations (i.e., the standard deviation divided by the mean) in a chemical process decreases with the inverse square root of the process scale \cite{VanKampen1992}.
 In other words, the process becomes more predictable and therefore more controllable.
-At the extreme as the system size increases, the dynamics converge towards a deterministic process, except for certain exceptional initial conditions lying on manifolds that would separate different steady state behaviors \brian[\cite{Strasser2012}].
+At the extreme as the system size increases, the dynamics converge towards a deterministic process, except for certain exceptional initial conditions lying on manifolds that would separate different steady state behaviors \cite{Strasser2012}.
 
@@
 \section*{References}
-\bibliographystyle{plain}
+\bibliographystyle{unsrt}
 \bibliography{refs}

### 2023-10-03 paper.tex
Changes in my office.

@@
 \end{abstract}
-Keywords: Stochastic Control, Gene Regulation, Optogenetics
+Keywords: Stochastic Gene Regulation, Optogenetic Feedback Control, Chemical Master Equation
 
-%
-% Uncomment for keywords
-%\vspace{2pc}
-%\noindent{\it Keywords}: Stochastic control, optogenetic, synthetic biology
-%
-% Uncomment for Submitted to journal title message
-%\submitto{\JPA}
-%
-% Uncomment if a separate title page is required
 \maketitle
-% 
-% For two-column output uncomment the next line and choose [10pt] rather than [12pt] in the \documentclass declaration
-%\ioptwocol
 
@@
 Uncertain fluctuations, or `noise,' is a common theme throughout many fields of engineering, and robust control is a frequent concern when attempting to control any human-made system such as vehicles \cite{Liang2021}, chemical processes \cite{Lucia2013}, or biology \cite{Baetica2020}.
-\brian[Please look for some relevant references or a review of control under noise. I highlight several missing refs in red.]
 Noise arrises from many sources, occasionally from quantum events and thermally-induced fluctuations, but more commonly from unknown or uncharacterized physical processes, and as such, these fluctuations usually cannot be efficiently reproduced or predicted using purely physical or mechanistic models.
@@
 To address this issue, we first truncate the system at a finite number for each species and then apply a reflecting boundary condition, resulting in a finite dimensional master equation.
-To assess the time interval over which this truncation is valid, we also solve the Finite State Projection \cite{Munsky2006} for the same truncation, which allows us to compute an upper bound on the truncation error as a function of time.
-
-\brian[We should use the FSP to compute the error of the truncation for a couple cases.  We can talk about how to do this when we meet.]
-\michael[ok!]
+%To assess the time interval over which this truncation is valid, we also solve the Finite State Projection \cite{Munsky2006} for the same truncation, which allows us to compute an upper bound on the truncation error as a function of time.
 
@@
 We note that the time delay stochastic process was only simulated using the SSA because to our knowledge an appropriate direct FSP/CME integration procedure has not yet been developed.
-\brian[Still missing reference to the extrande method at this website
-\url{https://journals.plos.org/ploscompbiol/article?id=10.1371/journal.pcbi.1004923}].
-\michael[added Voliotis2016 to the section after this comment]
 The Extrande method \cite{Voliotis2016} was used to update the control input at an average frequency of 50 updates per minute, far exceeding the dynamics of the system.
+
 \subsection{Tracking Time-Varying Trajectories}\label{sec:Dynamic}
@@
 and assign that controller.
-%
-%controllers for each target point and cycling through the controllers over time to create the final controller $\mathbf{R}(\mathbf{x},t)$.
-%Let $\mathbf{p}(t)$ be a continuous and cyclicly moving point in state space
-%\begin{equation}
-%\mathbf{p(t)}=[x_1(t),x_2(t)]
-%\end{equation}
-%where $x_1(t)$, and $x_2(t)$ are both cyclic functions of time $\in \mathbf{R}$.
-%Then, $\mathbf{q}$ is a collection of points $q_i$ which evaluate the cyclic function over its natural period (T) in n steps. 
-%\begin{equation}
-%\mathbf{q}=[q_1,\ldots ,q_n]=[\mathbf{p}(t_1),\ldots,\mathbf{p}(t_n)]
-%\end{equation}
-%where each $t_i$ comes from linearly spaced points between $0$ and $T$, in $n$ steps 
-%\begin{equation}
-%t_i = [t_1,\ldots,t_n]=[0,\frac{1}{n} T,\ldots,\frac{n}{n}T]
-%\end{equation}
-%Then, we optimize a set of controllers which target each $q_i$ in the discretized path.
-%\begin{equation}
-%\mathbf{U}_i(\mathbf{x})=[\mathbf{u}_1(\mathbf{x}),...,\mathbf{u}_n(\mathbf{x})]
-%\end{equation}
-%Next, we cycle through each controller at a fixed frequency by using ($\phi(t)$). 
-%\begin{equation}
-%\phi(t)=\lceil \frac{n}{T} mod(t,T)\rceil = \lceil n \ f \ mod(t,\frac{1}{f})\rceil
-%\end{equation}
-%where $f$ the cycling frequency given by $1/T$. 
-%Then the final control equation is given by 
-% \begin{equation}
-%\mathbf{R}(\mathbf{x},t)=\bf{U}_{\phi(t)}(x)
-%\end{equation}
-%Where $\mathbf{R}(\mathbf{x},t)$ returns the control input scalar as a function of state and time. 
 Under this piecewise constant controller law, the full time-varying CME becomes
@@
 \end{equation}
+
+\subsection{Computational details}
+\brian[Add section here describing the computer resources used and how long computations took for each type of analysis: What Matlab version and toolboxes, what cluster used (number/type of processors), OS,  ]
+
+\subsection{Code and data availability}
+All codes and data needed to reproduce figures is provided at \url{https://github.com/MunskyGroup/May_2023_Exploiting_Intrinsic_Noise}.
+A frozen version of this repository will be provided via Zenodo upon acceptance of the manuscript.
+
 \section{Results}
 
-%\brian[{\bf We still need this.} Add a paragraph or two to explain the big picture here.
-Use figure 1 to remind the reader of (1) the control goals, (2) the control law, (3) the resulting distributions, (4) the objective function and (5) the optimization of the control law to minimize the error.
-Refer to the methods for each such calculation.
-This section is critical to make sure the reader understands what is coming next.
-Then provide an outline of the remaining subsections of the results section so that the reader knows what to expect.]
-
-As developed in the Methods section, we focus on a SIMO controller where {\em all cells receive the same input} at every instant of time.
-For this, the control signal depends upon the observed state, e.g., $u(t) = \mathbf{u}(\mathbf{x})$ as developed above and illustrated in Fig.
+Using the analyses described in the Methods section, we focus on a single-input-multiple-output (SIMO) controller where {\em all cells receive the same input} at every instant of time.
+For this, the control signal depends upon the observed state, e.g., $u(t) = \mathbf{u}(\mathbf{x})$ as illustrated in Fig.
 \ref{cartoons}(I,III).
-These controllers have been optimized to break symmetry so that multiple cells can be controlled to different targets as illustrated in Fig.
+These controllers have been optimized to break symmetry so that multiple cells can be controlled to different target pstates as illustrated in Fig.
 \ref{cartoons}(II,IV), and we quantify performance according to the RMSE error introduced in Eq.
 \ref{Euclid}.
-To explore how such controllers may perform in realistic settings where the models are approximate and parameter are unknown or extrinsically variable, we now fix the parameters of those controllers and explore performance robustness to different types of model uncertainties, including parameter errors or variations (Section \ref{sec:ResPars}), incorrect assumptions on system scales (Section \ref{sec:ResGran}) and time delays (Section \ref{sec:ResTime}).
+To explore how such controllers may perform in realistic settings (e.g., where the models are approximate or the parameter are unknown or extrinsically variable), we now fix the parameters of those controllers and explore performance robustness to different types of model uncertainties, including parameter errors or variations (Section \ref{sec:ResPars}), incorrect assumptions on system scales (Section \ref{sec:ResGran}) and time delays (Section \ref{sec:ResTime}).
 Finally, in Section \ref{sec:ResTraj}, we extend the control analysis to consider the performance of the controller for tracking variable reference signals with different frequencies and phases.
 
- \subsection{Stochastic SIMO Optogenetic Control can Remain Effective Despite Small Parameter Errors or Extrinsic Uncertainties}\label{sec:ResPars}
+ \subsection{Stochastic SIMO Optogenetic control can remain effective despite small parameter errors or extrinsic uncertainties}\label{sec:ResPars}
 \begin{figure*}
@@
 Figure \ref{Volume} (A - F) shows the joint probability distributions (left plots) and marginal probability distributions (right plots) of the system at a low granularity ($\alpha=0.2$), the original granularity ($\alpha=1.0$), and at an increased granularity ($\alpha=2.0$).
-%As the system increases, the scale the size of the dynamics of the system becomes stretched over a larger region in state space. 
-%This effect can be observed in the smoothness of C with respect to A despite using the same system and the same controller. 
-%\brian[the previous two sentences do not make sense to me.]
 Figure \ref{Volume}G shows the trend of the performance versus alpha for both the FAC (solid cyan line) and the PAC (solid magenta line) controller.
@@
 
-\begin{figure*}[t!]
+\begin{figure}[t!]
 \begin{center}
@@
 \vspace{-0.2in}
-\end{figure*}
-
- % XXX - update this to explain what you see in this new figure.
-
-%These data suggest that increasing granularity increases steady state control performance even if the controller itself depends on noise for its implementation.
-Also, designing a controller to work for a system of one volume size can result in a controller that works even for another much larger volume.
-This has practical significance, because it is relatively easy to search over a control law defined on a small $M\times M$ grid of states (e.g., when $M$=50 as in \cite{May2021}), but to implement the controller for a larger system which may be computationally prohibitive.
-This result implies that it may be possible to optimize controller using coarse-grained FSP analyses for small granularity problems and adapt these via interpolation for use with larger, more realistic systems, that exceed the computational limit of standard FSP computations.
+\end{figure}
 
@@
 From the figures, it is clear that performance is rapidly degraded as the delay approaches and then exceeds the characteristic time ($\tau_c = 1 / \gamma = 49$ min) set by the degradation rate of the process (yellow dashed line).
-At very small time delays ($\tau< 3.4$ min), the FAC outperforms the PAC  but at moderate time delays ($\tau > 3.4$ min) the PAC outperforms the FAC.
-%
-%These data show that time delays greater than $\tau = 0.1 / \gamma = 4.9$ min begin to degrade FAC control performance and plateau at $\tau_c = 1.0 / \gamma = 49$ min.
-\brian[Check these times.] \michael[rewrote the sentence]The choice of best controller depends on the level of time delay in the system.
-At high time delays, both controllers lose their asymmetry, resulting in significantly worse performances ($RMSE = 32$ and $23$) and even perform worse than under a simple constant control input without feedback (depicted by a horizontal red line in Fig.\ \ref{Time}G).
-% XXX - This figure needs much larger fonts for all the labels and axes.
-% MMM - Did the change
-% There are errors in the parameter names.  What is 'alpha'?
-% remove the A,B,C,D,E.
-% replace I, II, III with A, B, C
-% what is parameter $\mu$? Is that an error?  Please use $\mu$ for reaction index as in the methods section.
+At very small time delays ($\tau< 3.4$ min), the FAC outperforms the PAC, but at moderate time delays ($\tau > 3.4$ min), the PAC outperforms the FAC.
+At high time delays, both controllers lose their asymmetry, resulting in substantially worse performances ($RMSE = 32$ and $23$) and even perform worse than under a simple constant control input without feedback (depicted by a horizontal red line in Fig.\ \ref{Time}G).
 
@@
 Dashed yellow line shows characteristic time $\tau_c = \alpha/\gamma/$.
-\brian[Label 'J' is missing from the figure. Replace XXX below.] \michael[added J. Replace XXX with 4.0]
 (K) Controlled system trajectory for $\alpha = 4.0$ and $\tau = $ min (denoted by red star in panel J).}
@@
 Control performance errors measured by RMSE improved as $\alpha$ increased when the delays were small ($\tau=1 \ $ min and $\tau=10 \ $ min) but not when $\tau=100 \ $ min.
-%
-%\brian[What are the units of tau in Figure5?
-This section is confusing in that it is not clear to me how tau and gamma are changing relative to one another when alpha is changed.
-From Eq 21 it seems that gamma changes with 1/alpha.
-If this is correct, does that mean that the critical time changes with $\tau_c = 1/gamma = alpha/gamma ~= 5*alpha$?
-If you plot that as a thick dashed dark line in Figure5J,  would it match to the contours?
-] \michael[$tc=1/(0.7*alpha)$ matches the counters well]
 
 Recalling that under the system granularity rescaling (Eq.\ \ref{eq:rescaling}), as $\alpha$ changes, the effective degradation rate scales according to $\gamma' = \gamma/\alpha$.
-Therefore, the critical limit for the delays should also change with the system scale according to $\tau_c = 1\alpha/\gamma$.
+Therefore, the critical limit for the delays should also change with the system scale according to $\tau_c = \alpha/\gamma$.
 Figure\ref{DG}(J) depicts this characteristic line and shows that as the time delay approaches and then exceeds this level, the steady state performance becomes dramatically worse.
@@
 
-\begin{figure*}
+\begin{figure}
 \begin{center}
@@
 \vspace{-0.2in}
-\end{figure*}
+\end{figure}
 
@@
 
-%We next analyzed the control performance over a domain of target points to find if some states are easier to achieve than others.
-For a two-cell system, stability analysis of the auto-regulated system using a moderate light input can create up to four stable points and one unstable point in the state space.
-These points give rise to basins which may attract probability more easily than other points.
 We next optimized the FAC controller and calculated its performance for the two-cell system over a two dimensional domain of discrete target points $\mathbf{T}_i = [T_{i1},T_{i2}]$ between between five and forty-five.
@@
 Figure \ref{CtrlP} (C and D) show the the same thing but for a different target state of $\mathbf{T}=[10, 25]$.
-%\brian[Let's remove the PAC from this section.
-Show only the FAC and and replace the Figure6C/D with the FAC at a different target point.
-Put a new star of a different color to match to the new 6C/D.
-Remove Figure6F.
-Make sure to change all the figure labels and titles and captions.] \michael[ok, I also moved the settpoint score heatmap to the right]
 Figure \ref{CtrlP}(E) shows the overall steady state FAC control performance over the entire domain of static set points, and illustrates that some regions are easier to attain than others.
-%For example, in the space contained in the dashed-green box which also hold the four stable points of the system, the control performance is quite strong, whereas near the unstable point (dashed-blue box), the FAC results in relatively worse control.
-%For example, the top left and bottom right edge regions were the most difficult to control, since the control input does not push probability in these directions.
-\brian[I don't understand this statement.] \michael[editied it better]
+For example, the top-left and bottom-right edge regions (i.e., where one species is well above its high equilibrium point and the other is well below its low equilibrium point) are the most difficult to control, while other target points near the two equilibrium points are more easily obtained.
 
@@
 $$
-For each reference signal, the driving frequency was defined as $f=10^{-4}$ cycles per minute.
+For each reference signal, a slow driving frequency was applied at $f=10^{-4}$ cycles per minute.
 
@@
 The characteristic frequency (given by the predicted $\tau_c^{-1}$ over a range of $\alpha$) is plotted as the yellow dashed line.
-%RMSE control performance of 128 simulations were simulated over a range of frequency between $10^-3$ and $10^-1$ Figure\ref{CRG}(D) for 32 cycles to obtain a better measurement of the true RMSE Figure\ref{CRG}(D) at $\alpha=2.0$, $\alpha=4.0$, and $\alpha=8.0$.
 In particular, the heat map of control performance over $(f,\alpha)$ shows that the worst performance occurs at moderate $f$ near $f=10^{-2}$ cycles/min and that control performance is best when granularity is high and frequency is low \ref{CRG}(E).
